### 2023-09-10 MOB_paper.Rnw
First commit
First version

\documentclass[doc,floatsintext,natbib]{apa7}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{doi}
\usepackage{booktabs}
%\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{setspace} % for adjusting line spread
\usepackage{hyperref}

\renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
\renewcommand{\thefootnote}{\fnsymbol{footnote}} % uses symbols instead of numbers for footnotes
\usepackage[colorinlistoftodos, textsize=footnotesize]{todonotes}
%%\todo[inline, color=green!40]{This is a green inline comment.}

\linespread{1.3}

\title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
\shorttitle{shorttitle}
\authorsnames{Marjolein Fokkema$^1$ \& Carolin Strobl$^2$}
\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}

\abstract{Model-based recursive partitioning \cite{ZeilyHoth08} is a flexible framework for detecting subgroups with differential effects in a wide range of parametric models. It provides a versatile tool for detecting and explaining heterogeneity of intervention effects. In this tutorial paper, we provide an introduction to the general MOB framework. In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item-response theory models. In the first case study, we show how GLMM trees \cite{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models. We apply GLMM trees to a dataset from a study of \cite{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program. Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start. In a second case study, we show how Rasch trees \cite{StroyKopf15} can be used to detect subgroups with different parameters in IRT models. We show how a recently developed stopping criterion \cite{HennyDeba23} can be used to guide subgroup detection based on effect sizes instead of statistical significance.\\}

\begin{document}
\SweaveOpts{concordance=TRUE}
\renewenvironment{Schunk}{\small}{}

<<echo=FALSE>>=
## suppress printing of > and + before code, it is annoying for copy-pasting
options(prompt=" ", continue = " ") 
@

\maketitle

\todo[inline]{Target journal: \textit{Journal of School Psychology}. This Special Issue is on "Conceptual and Methodological Advances for Understanding Contextual, Identity, and Cultural Effects in Intervention Research". Submission deadline: October 31, 2023}

\newpage
\section{Introduction}
\label{sec:Introduction}

\todo[inline]{Introduce and explain MOB.}

\todo[inline]{\textbf{Previous Papers on IRT and Rasch Models in Journal of School Psychology}

Rasch and IRT models are often mentioned in this journal's papers as being used for scoring school assessments (e.g., Virginia Standards of Learning Assessments;  Oregon Assessment of Knowledge and Skills; Peabody individual achievement test, \cite{Luth92}).

A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}. There are also papers describing extensions to mixure IRT models \cite{AyalySant17} and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.

\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).}

\newpage
\section{Tutorial: Subgroup Detection in Mixed-Effects Models}
\label{sec:TutorialMixed}

<<data_prep_mixed, echo=FALSE>>=
## Load data
data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))

## Keep only eligible children in HS or no preschool
data <- data[data$Elig2_90 == 1, ]
data <- data[data$None2_90 == 1 | data$HS2_90 == 1, ]
data <- data[data$MotherID %in% names(table(data$MotherID) >= 2), ]
data$program <- factor(ifelse(data$None2_90 == 1, "None", 
                              ifelse(data$Pre2_90 == 1, "Pre", "HS")))

## Keep only families in which children differ in HS participation
selected_family <- rowSums(table(data$MotherID, data$program) > 0) == 2
selected_family <- names(selected_family)[selected_family]
data <- data[data$MotherID %in% selected_family, ]

## Node-specific predictor
data$program <- factor(ifelse(data$None2_90 == 1, "None", 
                              ifelse(data$Pre2_90 == 1, "Pre", "HS")),
                       levels = c("None", "HS"))

## Partitioning variables

## ageAFTQ: age-adjusted AFQT
## Race_child: Hispanic=Race_Child==1; Black=Race_Child==2; White=Race_Child==3
data$Race_Child <- factor(ifelse(data$Race_Child == 1, "Hispanic", 
                          ifelse(data$Race_Child == 2, "Black", "White")))
## family income: PermInc
## Mom's years of education: MothED
part_vars <- c("AgeAFQT", "Race_Child", "PermInc", "Height_Mom81", "MothED")
id_vars <- c("ChildID", "MotherID")

## Outcome: Peabody Picture Vocabulary Test (PPVT)
testscore_vars <- c(paste0("PPVT_Raw", 2*(43:52)), paste0("AgeTest_Yr", 2*(43:52)))

## Create long dataset
pred_vars <- "program"
HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
                    varying = list(paste0("PPVT_Raw", 2*(43:52)),
                                   paste0("AgeTest_Yr", 2*(43:52))),
                    direction = "long", 
                    v.names = c("PPVT_Raw", "AgeTest_Yr"))
HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
                     "ChildID", "MotherID", "Program", "PPVT", "Age")
HS_dat <- HS_dat[complete.cases(HS_dat), ]
HS_dat$Age <- sqrt(HS_dat$Age)
rownames(HS_dat) <- NULL
saveRDS(HS_dat, file = "HS_dat.Rda")
@

\subsection{Dataset}

To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start. Head Start is a federally funded nationwide pre-school program for children from low-income families. \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).

The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program. Data from children in those families who participated in another preschool program were excluded. As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start. As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time. As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.

Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height. Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables. The dataset comprises data from families and children for whom complete data was available.

We load the data and inspect the first rows:

<<>>=
HS_dat <- readRDS("HS_dat.Rda")
head(HS_dat, 3)
@

We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red). To show the effect of age and Head Start participation, we fitted a mixed-effects model comprising their main and interaction effects. To account for the correlation between repeated assessments on the same child, we specified a random intercept. The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference persists over time. This result agrees with the findings of \cite{Demi09}.

\begin{figure}%
\caption{}
\begin{subfigure}{.7\textwidth}
<<echo=FALSE, fig=TRUE>>=
library("lme4")
lmm <- lmer(PPVT ~ Program*Age + (1|ChildID), data = HS_dat)
beta <- fixef(lmm)
plot(jitter(HS_dat$Age), HS_dat$PPVT, col = HS_dat$Program,
     cex = .5, #cex.lab = .7, cex.axis=.7, cex.main = .7,
     xlab = "Age (years)", ylab = "PPVT score", xaxt = "n",
     main = paste0("Global Model (", length(unique(HS_dat$ChildID)), " children)"))
abline(a = beta["(Intercept)"], b = beta["Age"], col = "black")
abline(a = beta["(Intercept)"]+beta["ProgramHS"], 
       b = beta["Age"]+beta["ProgramHS:Age"], col = "red")
axis(1, at=sqrt(c(3, 5, 8, 12, 17)), labels=c(3, 5, 8, 12, 17), cex.axis = .8)
@
\end{subfigure}
\label{fig:global_lmm}
\end{figure}%

\FloatBarrier
\subsection{Linear mixed effects model tree}

We now test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:

<<>>=
library("glmertree")
HS_tree <- lmertree(PPVT ~ Program*Age | (1|ChildID) | AFTQ + Race + 
                      Income + Mom_edu_yrs + Mom_height, 
                    data = HS_dat, cluster = ChildID, minsize = 250)
@

With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance. The middle part (\texttt{1|ChildID}) specified the random effects. The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.

With the second argument, we specified the dataset which contain the variables. Because we are dealing with repeated measurements on the same children, we additionally specified that the parameter stability tests should be performed on the child level using the \verb|cluster| argument. Using the default observation-level parameter stability tests may artificially inflate power. Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.

Next, we plot the tree. With multiple fixed-effects predictors of interest, the default plots may become too crowded or difficult to interpret. We therefore specify \verb|type = "simple"| to facilitate interpretation, and using the \verb|nodesize_level| argument, we specified that the sample size printed above every terminal node should count the number of children, not the number of individual observations:

<<eval=FALSE>>=
plot(HS_tree, type = "simple", nodesize_level = 2)
@

\begin{figure}%
\caption{}
<<fig=TRUE, echo=FALSE, fig.width=9, fig.height=4>>=
plot(HS_tree, type = "simple", which = "tree", gp = gpar(cex = .5), nodesize_level = 2)
@

\vspace*{-3cm}

<<fig=TRUE, echo=FALSE, width=7.6, height=2.5>>=
nodes <- predict(HS_tree, type = "node")
beta <- coef(HS_tree)
par(mfrow = c(1, 4))
for (node in sort(unique(nodes))) {
  plot(HS_dat$Age[nodes == node], HS_dat$PPVT[nodes == node],
       col = HS_dat$Program[nodes == node],
       cex = .5, cex.lab = .7, cex.axis=.7, cex.main = .7,
       xlab = "Age", ylab = "PPVT", xlim = c(sqrt(3), sqrt(20)), ylim = c(0, 150))
  abline(a = beta[as.character(node), "(Intercept)"], b = beta[as.character(node), "Age"], col = "black")
  abline(a = beta[as.character(node), "(Intercept)"] + beta[as.character(node), "ProgramHS"], 
         b = beta[as.character(node), "Age"] + beta[as.character(node), "ProgramHS:Age"], col = "red")
}
@
\label{fig:lmm_tree}
\end{figure}%

The resulting tree is presented in Figure~\ref{fig:lmm_tree}. Below each terminal node, we plotted the observations and the two regression curves given by the coefficients in that node. The first split was made based on the AFTQ variable, which represents the mother's score on the Armed Forced Qualification Test, adjusted for the age at which they completed the test. The group with higher mother's AFTQ scores is further split based on race. The Black and Hispanic group is further split based on income.

<<echo=FALSE, results=tex>>=
coefs <- round(coef(HS_tree), digits = 2)
design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
                        Program = factor(rep(c("None", "HS"))))
design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(6)), 
                                     re.form = ~0)
design_df$`PPVT at age 18` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(18)), 
                                     re.form = ~0)
names(design_df)[1] <- "Node"
knitr::kable(design_df, format = "latex", digits = 2, 
             caption = "Predicted PPVT scores at different ages.", 
             label = "predictions", centering = FALSE,
             linesep = "", # linesep command suppressess addlinesep every 5 rows
             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)
@

To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18. All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT. This benefit remains the same over time for the group with lower mother's AFTQ scores. The benefit increases over time for White children with higher mother's AFTQ scores. Strikingly, the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores. These results correspond to the conclusions of \cite{Demi09}.

To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation. We separated the \Sexpr{length(unique(HS_dat$MotherID))} families in the dataset into ten equally-sized folds. We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree (i.e., an LMM with subgroups) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child). We evaluated performance on the remaining folds, by computing and evaluating accuracy of the predictions. \todo{explain MSE and $R^2$} We repeated this procedure ten times, so that all folds were used as a test set once. The results are presented in Table~\ref{tab:performance}, which shows that LMM trees generalize well: They provide better predictive accuracy compared to LMMs, while implementing only few splits.

<<cv_lmm_tree, echo=FALSE, results=tex>>=
results <- data.frame(MSE_tree = rep(NA, times = 10), 
                      MSE_lmm = rep(NA, times = 10),
                      tree_size = rep(NA, times = 10))

set.seed(42)
folds <- sample(1:10, replace = TRUE, size = length(unique(HS_dat$MotherID)))
HS_dat$fold <- NA
for (mom in unique(HS_dat$MotherID)) {
  HS_dat$fold[HS_dat$MotherID == mom] <- folds[unique(HS_dat$MotherID) == mom]
}
for (i in 1:10) {
  ## Fit models on training data
  lmmt <- lmertree(PPVT ~ Age*Program | ChildID | AFTQ + Race + Income + 
                     Mom_height + Mom_edu_yrs, 
                   cluster = ChildID, data = HS_dat[HS_dat$fold != i, ], 
                   minsize = 225)
  lmm <- lmer(PPVT ~ Age*Program + (1|ChildID), 
              data = HS_dat[HS_dat$fold != i, ])

  ## Evaluate performance on test data
  results$tree_size[i] <- (length(lmmt$tree)-1)/2
  preds <- predict(lmmt, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
  results$MSE_tree[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
  preds <- predict(lmm, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
  results$MSE_lmm[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
}
res_df <- data.frame(method = c("LMM tree", "LMM"),
                     MSE = colMeans(results[1:2]),
                     SD = sapply(results, sd)[1:2],
                     `number of splits` = c(colMeans(results)[3], NA),
                     R2 = colMeans(1 - results[ , 1:2] / var(HS_dat$PPVT)))
knitr::kable(res_df, format = "latex", digits = 3, 
             caption = "Cross-validated performance of LMMs and LMM trees.", 
             label = "performance", centering = FALSE,
             linesep = "", # linesep command suppressess addlinesep every 5 rows
             row.names = FALSE, escape=TRUE, align = c("ccccc"), booktabs = TRUE)
@

\todo[inline]{I am not sure this dataset is the best GLMM tree illustration, because visualized differences between subgroups are very small and perhaps the fixed-effect part is already too complex. Alternative: Do not focus on effect of Head Start, but use as partitioning variable. Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.}

\newpage
\FloatBarrier
\section{Tutorial: Subgroup Detection in Rasch Models}
\label{sec:TutorialRasch}

Using Rasch trees, we assess DIF for items on math and reading \todo{decide on subscale(s) to include} from the Self-Description Questionnaire \cite{Boyl94}. We use a dataset from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}. Data were collected from 1,018 schools across the USA. Assessments took place from kindergarten through 8th grade. Here we focus on the 8th grade assessment, in which four SDQ-II items were administered to assess perceived interest and competence in reading, and another four to assess perceived interest and competence in math. Items are presented in Appendix~\ref{sec:AppendixA}. The items were rated by the children on a 4-point scale (“not at all true” through “very true”). We coded responses 1 and 2 as 0, and responses 3 and 4 as 1, to allow for fitting a Rasch model, which assumes binary responses.

<<echo=FALSE>>=
SDQ_dat <- readRDS("ECLSK_SDQ.Rda")
@

We used three covariates as possible partitioning variables: Gender (1=Male; 2=Female), race (8 categories: 1=White, non-Hispanic; 2=Black or African-American, non-Hispanic; 3=Hispanic, race specified; 4=Hispanic, race not specified; 5=Asian; 6=Native Hawaiian or other Pasific Islander; 7= American Indian or Alaska native; 7=More than one race) and socio-economic status (range $-$5 to 3). \todo{Add more partitioning variables? This does increase number of splits. But could add one that should not have an effect, e.g., month of birth.} We analyze observations of children with complete data only, yielding a total sample size of \Sexpr{formatC(nrow(SDQ_dat), big.mark=",")}.

We load the data and inspect the first three rows as follows:

<<>>=
SDQ_dat <- readRDS("ECLSK_SDQ.Rda")
head(SDQ_dat, 3)
@

To fit Rasch trees, we use function \verb|raschtree| which is available from R package \textbf{psychotree}. However, the stopping criterion based on the Mantel–Haenszel effect size measure is implemented in R package \verb|raschtreeMH|, which we therefore use here. \todo{might be nice to implement the MH effect-size criterion in the psychotree package for this paper, currently it seems to interfere with plotting the tree}. It can be installed and loaded as follows:

<<warning=FALSE, message=FALSE, eval=FALSE>>=
devtools::install_github("mirka-henninger/raschtreeMH")
library("raschtreeMH")
@

<<warning=FALSE, message=FALSE, echo=FALSE>>=
library("raschtreeMH")
@

To fit a Rasch tree, we first create a \verb|data.frame| that contains the item responses and possible partitioning variables:

<<>>=
math_items <- c("C7MTHBST", "C7MTHGD", "C7LIKMTH", "C7ENJMTH")
part_vars <- c("GENDER", "RACE", "WKSESL")
mydata <- SDQ_dat[ , part_vars]
mydata$resp <- sapply(SDQ_dat[ , math_items], function(x) ifelse(x > 2, 1, 0)) 
@

Next, we apply function \verb|raschtree| and plot the result:

<<>>=
stop_fun <- stopfun_mantelhaenszel(purification = "iterative", stopcrit = "C")
math_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
@

<<eval=FALSE>>=
plot(read_tree, gp = gpar(cex = .7))
@

The resulting tree is presented in Figure~\ref{fig:math_tree}. The splitting nodes show the $p$-value of the corresponding parameter stability tests. The terminal depict the node-specific item difficulties. The difficulties of items 1 (C7MTHBST: Math is one of my best subjects) and 3 (C7LIKMTH: I like math) appear relatively stable and show average difficulty in every subgroup. The difficulties of items 2 (C7MTHGD: I get good grades in math) and 4 (C7ENJMTH: I enjoy doing work in math) are always lower and higher, respectively, and also seem to vary more strongly.

\begin{figure}%
\caption{Rasch tree for the four math items.}
\begin{subfigure}{1.25\textwidth}
<<fig=TRUE, echo=FALSE, width= 9, height=6>>=
library("psychotree") ## needed for plotting, but interferes with stopfun_mantelhaenszel
plot(math_tree, gp = gpar(cex = .7))
detach("package:psychotree", unload = TRUE)
@
\end{subfigure}
\label{fig:math_tree}
\end{figure}%

We repeat the procedure for the reading items:

<<>>=
read_items <-  c("C7ENGBST", "C7GRDENG", "C7LIKRD", "C7ENJRD")
mydata$resp <- sapply(SDQ_dat[ , read_items], function(x) ifelse(x > 2, 1, 0))
read_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
@

<<eval=FALSE>>=
plot(read_tree, gp = gpar(cex = .7))
@

The resulting tree is presented in Figure~\ref{fig:read_tree}, which reveals a pattern of difficulties and splits quite similar to Figure~\ref{fig:math_tree}.

\begin{figure}%
\caption{Rasch tree for the four reading items.}
\begin{subfigure}{1.25\textwidth}
<<fig=TRUE, width=9, height=6, echo=FALSE>>=
library("psychotree") ## needed for plotting, interferes with stopfun_mantelhaenszel
plot(read_tree, gp = gpar(cex = .7))
detach("package:psychotree", unload = TRUE)
@
\end{subfigure}
\label{fig:read_tree}
\end{figure}

\FloatBarrier
\section{Discussion}

\todo[inline]{Write summary of findings.}

\todo[inline]{Mention shortcomings.}

\todo[inline]{Write about future work.}

\bibliography{MOB_paper}

\todo{Add DOIs.}

\newpage
\appendix

\section{Appendix A: SDQ-II items administered in the ECLS}
\label{sec:AppendixA}

\todo{Fix formatting.}

How true is each of these about you? (1="not at all true"; 2="a little bit true"; 3="mostly true" or 4="very true")

\begin{itemize}

  \item C7MTHBST (math): Math is one of my best subjects.

  \item C7ANGRY (internalizing): I feel angry when I have trouble learning.

  \item C7LIKRD (reading): I like reading.

  \item C7WRYTST (internalizing): I worry about taking tests.

  \item C7MTHGD (math): I get good grades in math.

  \item C7LONLY (internalizing): I often feel lonely.

  \item C7ENGBST (reading): English is one of my best subjects.

  \item C7SAD (internalizing): I feel sad a lot of the time.

  \item C7LIKMTH (math): I like math.

  \item C7WRYWEL (internalizing): I worry about doing well in school.

  \item C7ENJRD (reading): I enjoy doing work in reading.

  \item C7WRYFIN (internalizing): I worry about finishing my work.

  \item C7ENJMTH (math): I enjoy doing work in math.

  \item C7WRYHNG (internalizing): I worry about having someone to hang out with at school.

  \item C7GRDENG (reading): I get good grades in English.

  \item C7ASHAME (internalizing): I feel ashamed when I make mistakes at school.

\end{itemize}

\end{document}

### 2023-09-10 _Rasch trees.Rmd
First commit
First version

---
title: "DIF detection with Rasch Trees in SDQ items from ECLSK data"
author: "Marjolein Fokkema"
date: "2023-08-15"
output: pdf_document
---

# Load data

```{r, eval=FALSE}
library("data.table")
ICPSR <- as.data.frame(fread( "28023-0001-Data.tsv" ))
```

We could analyse data from grade 8 students from the ECLSK (Early Childhood Longitudinal Study - Kindergarten class of ...).

Student quesionnaires are included in most waves, including adaptations of the Self-Description Questionnaire II (SDQII; Marsh, 1990).

In the ECLSK data, a variable starting with:

* 'T' indicates teacher rating, 
* 'C' indicates child self-report, 
* 'P' indicates parent report.

The number afterwards indicates the assessment wave.

# Get SDQ-II item responses and partitioning variables

### Original SDQ-II has 102 items, comprising 11 subscales:

Subscale | Number of items
-------------------------
Physical Abilities | 8 
Physical Appearance | 8
Opposite-Sex Relationships | 10 
Same-Sex Relationships | 8 
Parent Relationships | 8 
Honesty–Trustworthiness | 10 
Emotional Stability | 10 
Self-Esteem | 10
Math | 10 
Verbal | 10 
General School | 10
------------------------

### ECLSK grade 3 and 5 assessments

42 items were administered from the SDQ-I forming shorter subscales:

* Perceived Interest/ Competence in Reading (8 items)

* Perceived Interest/ Competence in Math (8 items)

* Perceived Interest/ Competence in All School Subjects (6 items)

* Perceived Interest/ Competence in Peer Relations (6 items)

* Externalizing Problem Behaviors (6 items)

* Internalizing Problem Behaviors (8 items)

### ECLSK grade 8 (C7)

The self-description questionnaire consists of 16 statements. Children rated whether each item was “not at all true,” “a little bit true,” “mostly true,” or “very true.” Three subscales were produced from the self-description questionnaire items. The scale scores on all self-description questionnaire scales represent the mean rating of the items included in the scale.

* The SDQ Perceived Interest/Competence–Reading subscale includes four items on grades in English and the child’s interest in and enjoyment of reading.

* The SDQ Perceived Interest/Competence–Math subscale includes four items on mathematics grades and the child’s interest in and enjoyment of mathematics.

* The SDQ Internalizing Behavior subscale includes eight items on internalizing problem behaviors such as feeling “sad a lot of the time,” feeling lonely, feeling ashamed of mistakes, feeling frustrated, and worrying about school and friendships.

```{r}
sdq <- c("C7MTHBST", "C7ANGRY", "C7LIKRD", "C7WRYTST", "C7MTHGD", "C7LONLY",
  "C7ENGBST", "C7SAD", "C7LIKMTH", "C7WRYWEL", "C7ENJRD", "C7WRYFIN",
  "C7ENJMTH", "C7WRYHNG", "C7GRDENG", "C7ASHAME")
math <- c("C7MTHBST", "C7MTHGD", "C7LIKMTH", "C7ENJMTH")
intern <- c("C7ANGRY", "C7WRYTST", "C7LONLY", "C7SAD", "C7WRYWEL", "C7WRYFIN", 
            "C7WRYHNG", "C7ASHAME")
read <- c("C7LIKRD", "C7ENGBST", "C7ENJRD", "C7GRDENG")
part_vars <- c("GENDER", "RACE", "WKSESL")#, "T6INTERN", "T6EXTERN", "T6INTERP", 
               #"T6CONTRO") ## some omitted to avoid large numbers of splits
```

```{r, eval=FALSE}
dat <- ICPSR[ , sdq]
sapply(dat, table, useNA = "ifany")
dat <- cbind(dat, ICPSR[ , part_vars])
dat <- data.frame(sapply(dat, function(x) ifelse(x == -9, NA, x)))
dat <- dat[complete.cases(dat), ]
dat$GENDER <- factor(dat$GENDER)
dat$RACE <- factor(dat$RACE)
saveRDS(dat, file = "ECLSK_SDQ.Rda")
```

Note that some partitioning variables are teacher assessments from the previous assessment ('6'). These were not available for the '7' assessment. Loads of other potential partitioning variables are available.

# Fit Rasch trees

```{r, warning=FALSE, message=FALSE}
dat <- readRDS("ECLSK_SDQ.Rda")
##devtools::install_github("mirka-henninger/raschtreeMH")
table(complete.cases(dat))
library("raschtreeMH")
library("psychotree") ## needed for plotting, but interferes with stopfun_mantelhaenszel
```

### Math

* Q20. How true is each of these about you? Q20a. Math is one of my best subjects.
Variable Name : C7MTHBST (math)

* Q20. How true is each of these about you? Q20e. I get good grades in math.
Variable Name : C7MTHGD (math)

* Q20. How true is each of these about you? Q20i. I like math.
Variable Name : C7LIKMTH (math)

* Q20. How true is each of these about you? Q20m. I enjoy doing work in math.
Variable Name : C7ENJMTH (math)

```{r, warning=FALSE, message=FALSE}
## Construct data.frame for analysis
mydata <- dat[ , part_vars] 
mydata$resp <- sapply(dat[ , math], function(x) ifelse(x > 2, 1, 0)) 
rasch_tree <- raschtree(resp ~ ., data = mydata)
length(rasch_tree)
detach("package:psychotree", unload = TRUE)
rasch_tree_es <- raschtree(resp ~ ., data = mydata,
                    stopfun = stopfun_mantelhaenszel(purification = "iterative", 
                                                     stopcrit = "C"))
length(rasch_tree_es)
library("psychotree") 
plot(rasch_tree_es, gp = gpar(cex = .5))
```

### Reading

* Q20. How true is each of these about you? Q20c. I like reading.
Variable Name : C7LIKRD (reading)

* Q20. How true is each of these about you? Q20g. English is one of my best subjects.
Variable Name : C7ENGBST (reading)

* Q20. How true is each of these about you? Q20k. I enjoy doing work in reading.
Variable Name : C7ENJRD (reading)

* Q20. How true is each of these about you? Q20o. I get good grades in English.
Variable Name : C7GRDENG (reading)

```{r, warning=FALSE, message=FALSE}
mydata <- dat[ , part_vars] 
read ## rearrange so as to have same order as with math
mydata$resp <- sapply(dat[ , read[c(2,4,1,3)]], function(x) ifelse(x > 2, 1, 0)) 
rasch_tree <- raschtree(resp ~ ., data = mydata)
length(rasch_tree)
detach("package:psychotree", unload = TRUE)
rasch_tree_es <- raschtree(resp ~ ., data = mydata,
                    stopfun = stopfun_mantelhaenszel(purification = "iterative",
                                                     stopcrit = "C"))
length(rasch_tree_es)
library("psychotree") 
plot(rasch_tree_es, gp = gpar(cex = .5))
```

### Internalizing

* Q20. How true is each of these about you? Q20b. I feel angry when I have trouble learning.
Variable Name : C7ANGRY (internalizing)

* Q20. How true is each of these about you? Q20d. I worry about taking tests.
Variable Name : C7WRYTST (internalizing)

* Q20. How true is each of these about you? Q20f. I often feel lonely.
Variable Name : C7LONLY (internalizing)

* Q20. How true is each of these about you? Q20h. I feel sad a lot of the time.
Variable Name : C7SAD (internalizing)

* Q20. How true is each of these about you? Q20j. I worry about doing well in school.
Variable Name : C7WRYWEL (internalizing)

* Q20. How true is each of these about you? Q20l. I worry about finishing my work.
Variable Name : C7WRYFIN (internalizing)

* Q20. How true is each of these about you? Q20n. I worry about having someone to hang out with at school.
Variable Name : C7WRYHNG (internalizing)

* Q20. How true is each of these about you? Q20p. I feel ashamed when I make mistakes at school.
Variable Name : C7ASHAME (internalizing)

```{r,  warning=FALSE, message=FALSE}
mydata <- dat[ , part_vars] 
mydata$resp <- sapply(dat[ , intern], function(x) ifelse(x > 2, 1, 0)) 
rasch_tree <- raschtree(resp ~ ., data = mydata)
length(rasch_tree)
detach("package:psychotree", unload = TRUE)
rasch_tree_es <- raschtree(resp ~ ., data = mydata,
                    stopfun = stopfun_mantelhaenszel(purification = "iterative", 
                                                     stopcrit = "C"))
length(rasch_tree_es)
library("psychotree") 
plot(rasch_tree_es, gp = gpar(cex = .5))
```

\newpage
# Other possible questionnaire in ECLSK

* Q21. How do you feel about each of the following statements? Q21a. I feel good about myself.
Variable Name : C7FLGOOD

* Q21. How do you feel about each of the following statements? Q21b. I don’t have enough control over the direction my life is taking.
Variable Name : C7NOCNTR

* Q21. How do you feel about each of the following statements? Q21c. In my life, good luck is more important than hard work for success.
Variable Name : C7LUCK

* Q21. How do you feel about each of the following statements? Q21d. I feel I am a person of worth, the equal of other people.
Variable Name : C7WORTH

* Q21. How do you feel about each of the following statements? Q21e. I am able to do things as well as most other people.
Variable Name : C7ABLE

* Q21. How do you feel about each of the following statements? Q21f. Every time I try to get ahead, something or somebody stops me.
Variable Name : C7STOPS

* Q21. How do you feel about each of the following statements? Q21g. My plans hardly ever work out, so planning only makes me unhappy.
Variable Name : C7NOPLAN

* Q21. How do you feel about each of the following statements? Q21h. On the whole, I am satisfied with myself.
Variable Name : C7SATISF

* Q21. How do you feel about each of the following statements? Q21i. I certainly feel useless at times.
Variable Name : C7USELES

* Q21. How do you feel about each of the following statements? Q21j. At times I think I am no good at all.
Variable Name : C7NOGOOD

* Q21. How do you feel about each of the following statements? Q21k. When I make plans, I am almost certain I can make them work.
Variable Name : C7PLANS

Q21. How do you feel about each of the following statements? Q21l. I fell I do not have much to be proud of.

Variable Name : C7NOPRD

Q21. How do you feel about each of the following statements? Q21m. Chance and luck are very important for
what happens in my life.
Variable Name : C7CHANCE

### 2023-09-10 Deming data files/Deming script2.R

@@ -1,0 +1,286 @@
+## Load dataset
+
+## This is the openly available dataset on which I ran the STATA script 
+data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
+names(data)
+
+## This is the openly available dataset
+data2 <- haven::read_dta("data_Deming_2008_0217.dta")
+names(data2)
+
+names(data)[!names(data) %in% names(data2)]
+
+
+#############################
+##
+## Sample Eligibility
+##
+
+## First create rules that establish sample eligibility
+## Rule is 5 years old by 1990, so that they will be 19 by 2004
+## Next restrict the sample to families with at least 2 age-eligible children
+## Finally restrict to families where at least one (but not all) children were in Head Start
+
+############################################################################################
+# I create 3 sets of variables that define preschool participation. The first simply uses   
+# the age variable and excludes those with any missing data. The second substitutes the PPVT
+# age (ie age at test) when the original variable is unavailable, and also substitutes past 
+# or future age variables from other survey years (plus or minus 24 months). The third is   
+# the most restrictive definition. It codes inconsistent or missing responses across years  
+# as zeros, as well as for children for whom mothers report "3 months or less" participation
+# in the program. The variables are coded HS1, HS2, HS3 and Pre1, Pre2, Pre3. In general    
+# different rules have very little impact on the estimates. I use #2 in the paper.          
+############################################################################################
+
+# Load the dataset (assuming it's in Stata format)
+data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
+
+## Keep only eligible children in HS or no preschool
+data <- data[data$Elig2_90 == 1, ]
[…]

### 2023-09-10 _MOB_paper.tex
changed file names

--- 2023-09-10 MOB_paper.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{MOB_paper-concordance}
 \renewenvironment{Schunk}{\small}{}
-
-<<echo=FALSE>>=
-## suppress printing of > and + before code, it is annoying for copy-pasting
-options(prompt=" ", continue = " ") 
-@
 
@@
 
-<<data_prep_mixed, echo=FALSE>>=
-## Load data
-data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
-
-## Keep only eligible children in HS or no preschool
-data <- data[data$Elig2_90 == 1, ]
-data <- data[data$None2_90 == 1 | data$HS2_90 == 1, ]
-data <- data[data$MotherID %in% names(table(data$MotherID) >= 2), ]
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")))
-
-## Keep only families in which children differ in HS participation
-selected_family <- rowSums(table(data$MotherID, data$program) > 0) == 2
-selected_family <- names(selected_family)[selected_family]
-data <- data[data$MotherID %in% selected_family, ]
-
-## Node-specific predictor
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")),
-                       levels = c("None", "HS"))
-
-## Partitioning variables
-
-## ageAFTQ: age-adjusted AFQT
-## Race_child: Hispanic=Race_Child==1; Black=Race_Child==2; White=Race_Child==3
-data$Race_Child <- factor(ifelse(data$Race_Child == 1, "Hispanic", 
-                          ifelse(data$Race_Child == 2, "Black", "White")))
-## family income: PermInc
-## Mom's years of education: MothED
-part_vars <- c("AgeAFQT", "Race_Child", "PermInc", "Height_Mom81", "MothED")
-id_vars <- c("ChildID", "MotherID")
-
-## Outcome: Peabody Picture Vocabulary Test (PPVT)
-testscore_vars <- c(paste0("PPVT_Raw", 2*(43:52)), paste0("AgeTest_Yr", 2*(43:52)))
-
-## Create long dataset
-pred_vars <- "program"
-HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
-                    varying = list(paste0("PPVT_Raw", 2*(43:52)),
-                                   paste0("AgeTest_Yr", 2*(43:52))),
-                    direction = "long", 
-                    v.names = c("PPVT_Raw", "AgeTest_Yr"))
-HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
-names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
-                     "ChildID", "MotherID", "Program", "PPVT", "Age")
-HS_dat <- HS_dat[complete.cases(HS_dat), ]
-HS_dat$Age <- sqrt(HS_dat$Age)
-rownames(HS_dat) <- NULL
-saveRDS(HS_dat, file = "HS_dat.Rda")
-@
-
 \subsection{Dataset}
@@
 
-The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
+The sample consists of 273 families with at least two children where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
 Data from children in those families who participated in another preschool program were excluded.
@@
 
-<<>>=
+\begin{Schunk}
+\begin{Sinput}
  HS_dat <- readRDS("HS_dat.Rda")
  head(HS_dat, 3)
-@
+\end{Sinput}
+\begin{Soutput}
+       AFTQ     Race   Income Mom_height Mom_edu_yrs ChildID MotherID Program
+1  3.478122 Hispanic 37731.07        502          12   20502      205      HS
+2  3.478122 Hispanic 37731.07        502          12   20501      205    None
+3 15.964368    Black 16119.13        504          10   22403      224    None
+  PPVT      Age
+1   18 2.000000
+2   48 2.645751
+3   69 2.645751
+\end{Soutput}
+\end{Schunk}
 
@@
 \begin{subfigure}{.7\textwidth}
-<<echo=FALSE, fig=TRUE>>=
-library("lme4")
-lmm <- lmer(PPVT ~ Program*Age + (1|ChildID), data = HS_dat)
-beta <- fixef(lmm)
-plot(jitter(HS_dat$Age), HS_dat$PPVT, col = HS_dat$Program,
-     cex = .5, #cex.lab = .7, cex.axis=.7, cex.main = .7,
-     xlab = "Age (years)", ylab = "PPVT score", xaxt = "n",
-     main = paste0("Global Model (", length(unique(HS_dat$ChildID)), " children)"))
-abline(a = beta["(Intercept)"], b = beta["Age"], col = "black")
-abline(a = beta["(Intercept)"]+beta["ProgramHS"], 
-       b = beta["Age"]+beta["ProgramHS:Age"], col = "red")
-axis(1, at=sqrt(c(3, 5, 8, 12, 17)), labels=c(3, 5, 8, 12, 17), cex.axis = .8)
-@
+\includegraphics{MOB_paper-004}
 \end{subfigure}
@@
 
-<<>>=
+\begin{Schunk}
+\begin{Sinput}
  library("glmertree")
@@
                      data = HS_dat, cluster = ChildID, minsize = 250)
-@
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<eval=FALSE>>=
+\begin{Schunk}
+\begin{Sinput}
  plot(HS_tree, type = "simple", nodesize_level = 2)
-@
+\end{Sinput}
+\end{Schunk}
 
@@
 \caption{}
-<<fig=TRUE, echo=FALSE, fig.width=9, fig.height=4>>=
-plot(HS_tree, type = "simple", which = "tree", gp = gpar(cex = .5), nodesize_level = 2)
-@
+\includegraphics{MOB_paper-007}
 
@@
 
-<<fig=TRUE, echo=FALSE, width=7.6, height=2.5>>=
-nodes <- predict(HS_tree, type = "node")
-beta <- coef(HS_tree)
-par(mfrow = c(1, 4))
-for (node in sort(unique(nodes))) {
-  plot(HS_dat$Age[nodes == node], HS_dat$PPVT[nodes == node],
-       col = HS_dat$Program[nodes == node],
-       cex = .5, cex.lab = .7, cex.axis=.7, cex.main = .7,
-       xlab = "Age", ylab = "PPVT", xlim = c(sqrt(3), sqrt(20)), ylim = c(0, 150))
-  abline(a = beta[as.character(node), "(Intercept)"], b = beta[as.character(node), "Age"], col = "black")
-  abline(a = beta[as.character(node), "(Intercept)"] + beta[as.character(node), "ProgramHS"], 
-         b = beta[as.character(node), "Age"] + beta[as.character(node), "ProgramHS:Age"], col = "red")
-}
-@
+\includegraphics{MOB_paper-008}
 \label{fig:lmm_tree}
@@
 
-<<echo=FALSE, results=tex>>=
-coefs <- round(coef(HS_tree), digits = 2)
-design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
-                        Program = factor(rep(c("None", "HS"))))
-design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(6)), 
-                                     re.form = ~0)
-design_df$`PPVT at age 18` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(18)), 
-                                     re.form = ~0)
-names(design_df)[1] <- "Node"
-knitr::kable(design_df, format = "latex", digits = 2, 
-             caption = "Predicted PPVT scores at different ages.", 
-             label = "predictions", centering = FALSE,
-             linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)
-@
+\begin{table}
 
+\caption{\label{tab:predictions}Predicted PPVT scores at different ages.}
+\begin{tabular}[t]{cccc}
+\toprule
+Node & Program & PPVT at age 6 & PPVT at age 18\\
+\midrule
+2 & None & 46.63 & 131.23\\
+2 & HS & 49.39 & 134.48\\
+5 & None & 50.55 & 144.09\\
+5 & HS & 55.15 & 146.43\\
+6 & None & 57.04 & 151.51\\
+6 & HS & 60.09 & 151.63\\
+7 & None & 62.10 & 155.74\\
+7 & HS & 66.36 & 163.32\\
+\bottomrule
+\end{tabular}
+\end{table}
 To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
@@
 To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
-We separated the \Sexpr{length(unique(HS_dat$MotherID))} families in the dataset into ten equally-sized folds.
+We separated the 258 families in the dataset into ten equally-sized folds.
 We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree (i.e., an LMM with subgroups) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child).
@@
 
-<<cv_lmm_tree, echo=FALSE, results=tex>>=
-results <- data.frame(MSE_tree = rep(NA, times = 10), 
-                      MSE_lmm = rep(NA, times = 10),
-                      tree_size = rep(NA, times = 10))
+\begin{table}
 
-set.seed(42)
-folds <- sample(1:10, replace = TRUE, size = length(unique(HS_dat$MotherID)))
-HS_dat$fold <- NA
-for (mom in unique(HS_dat$MotherID)) {
-  HS_dat$fold[HS_dat$MotherID == mom] <- folds[unique(HS_dat$MotherID) == mom]
-}
-for (i in 1:10) {
-  ## Fit models on training data
-  lmmt <- lmertree(PPVT ~ Age*Program | ChildID | AFTQ + Race + Income + 
-                     Mom_height + Mom_edu_yrs, 
-                   cluster = ChildID, data = HS_dat[HS_dat$fold != i, ], 
-                   minsize = 225)
-  lmm <- lmer(PPVT ~ Age*Program + (1|ChildID), 
-              data = HS_dat[HS_dat$fold != i, ])
-
-  ## Evaluate performance on test data
-  results$tree_size[i] <- (length(lmmt$tree)-1)/2
-  preds <- predict(lmmt, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
-  results$MSE_tree[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
-  preds <- predict(lmm, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
-  results$MSE_lmm[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
-}
-res_df <- data.frame(method = c("LMM tree", "LMM"),
-                     MSE = colMeans(results[1:2]),
-                     SD = sapply(results, sd)[1:2],
-                     `number of splits` = c(colMeans(results)[3], NA),
-                     R2 = colMeans(1 - results[ , 1:2] / var(HS_dat$PPVT)))
-knitr::kable(res_df, format = "latex", digits = 3, 
-             caption = "Cross-validated performance of LMMs and LMM trees.", 
-             label = "performance", centering = FALSE,
-             linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("ccccc"), booktabs = TRUE)
-@
+\caption{\label{tab:performance}Cross-validated performance of LMMs and LMM trees.}
+\begin{tabular}[t]{ccccc}
+\toprule
+method & MSE & SD & number.of.splits & R2\\
+\midrule
+LMM tree & 221.022 & 55.727 & 2.7 & 0.813\\
+LMM & 258.233 & 71.451 & NA & 0.781\\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-<<echo=FALSE>>=
-SDQ_dat <- readRDS("ECLSK_SDQ.Rda")
-@
-
 We used three covariates as possible partitioning variables: Gender (1=Male; 2=Female), race (8 categories: 1=White, non-Hispanic; 2=Black or African-American, non-Hispanic; 3=Hispanic, race specified; 4=Hispanic, race not specified; 5=Asian; 6=Native Hawaiian or other Pasific Islander; 7= American Indian or Alaska native; 7=More than one race) and socio-economic status (range $-$5 to 3).
@@
 This does increase number of splits.
-But could add one that should not have an effect, e.g., month of birth.} We analyze observations of children with complete data only, yielding a total sample size of \Sexpr{formatC(nrow(SDQ_dat), big.mark=",")}.
+But could add one that should not have an effect, e.g., month of birth.} We analyze observations of children with complete data only, yielding a total sample size of 7,417.
 
@@
 
-<<>>=
+\begin{Schunk}
+\begin{Sinput}
  SDQ_dat <- readRDS("ECLSK_SDQ.Rda")
  head(SDQ_dat, 3)
-@
+\end{Sinput}
+\begin{Soutput}
+   C7MTHBST C7ANGRY C7LIKRD C7WRYTST C7MTHGD C7LONLY C7ENGBST C7SAD C7LIKMTH
+2         4       4       4        2       4       2        4     3        4
+10        4       1       4        2       4       1        4     1        4
+16        3       3       1        2       3       1        4     1        4
+   C7WRYWEL C7ENJRD C7WRYFIN C7ENJMTH C7WRYHNG C7GRDENG C7ASHAME GENDER RACE
+2         4       4        4        3        2        3        4      2    1
+10        1       3        1        4        1        4        1      2    1
+16        2       1        2        4        1        4        1      1    1
+   WKSESL T6INTERN T6EXTERN T6INTERP T6CONTRO
+2    1.56     1.50     1.00      4.0     3.75
+10   1.41     1.25     1.50      4.0     4.00
+16  -2.93     1.50     2.33      2.8     3.00
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<warning=FALSE, message=FALSE, eval=FALSE>>=
+\begin{Schunk}
+\begin{Sinput}
  devtools::install_github("mirka-henninger/raschtreeMH")
  library("raschtreeMH")
-@
-
-<<warning=FALSE, message=FALSE, echo=FALSE>>=
-library("raschtreeMH")
-@
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<>>=
+\begin{Schunk}
+\begin{Sinput}
  math_items <- c("C7MTHBST", "C7MTHGD", "C7LIKMTH", "C7ENJMTH")
@@
  mydata$resp <- sapply(SDQ_dat[ , math_items], function(x) ifelse(x > 2, 1, 0)) 
-@
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<>>=
+\begin{Schunk}
+\begin{Sinput}
  stop_fun <- stopfun_mantelhaenszel(purification = "iterative", stopcrit = "C")
  math_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
-@
+\end{Sinput}
+\end{Schunk}
 
-<<eval=FALSE>>=
+\begin{Schunk}
+\begin{Sinput}
  plot(read_tree, gp = gpar(cex = .7))
-@
+\end{Sinput}
+\end{Schunk}
 
@@
 \begin{subfigure}{1.25\textwidth}
-<<fig=TRUE, echo=FALSE, width= 9, height=6>>=
-library("psychotree") ## needed for plotting, but interferes with stopfun_mantelhaenszel
-plot(math_tree, gp = gpar(cex = .7))
-detach("package:psychotree", unload = TRUE)
-@
+\includegraphics{MOB_paper-018}
 \end{subfigure}
@@
 
-<<>>=
+\begin{Schunk}
+\begin{Sinput}
  read_items <-  c("C7ENGBST", "C7GRDENG", "C7LIKRD", "C7ENJRD")
@@
  read_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
-@
+\end{Sinput}
+\end{Schunk}
 
-<<eval=FALSE>>=
+\begin{Schunk}
+\begin{Sinput}
  plot(read_tree, gp = gpar(cex = .7))
-@
+\end{Sinput}
+\end{Schunk}
 
@@
 \begin{subfigure}{1.25\textwidth}
-<<fig=TRUE, width=9, height=6, echo=FALSE>>=
-library("psychotree") ## needed for plotting, interferes with stopfun_mantelhaenszel
-plot(read_tree, gp = gpar(cex = .7))
-detach("package:psychotree", unload = TRUE)
-@
+\includegraphics{MOB_paper-021}
 \end{subfigure}

### 2023-09-22 _MOB_paper.Rnw
started Intro

--- 2023-09-10 MOB_paper.Rnw
@@
 
+%text from Marjolein edited by Caro
+\newcommand{\edc}[1]{\textcolor{blue}{#1}}
+
 \linespread{1.3}
 
-\title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
+\title{One Model May Not Fit All: Subgroup Detection Using MOB}
 \shorttitle{shorttitle}
@@
 
-\abstract{Model-based recursive partitioning \cite{ZeilyHoth08} is a flexible framework for detecting subgroups with differential effects in a wide range of parametric models.
+\abstract{MOB \citep[MOB]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models.
 It provides a versatile tool for detecting and explaining heterogeneity of intervention effects.
@@
 In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item-response theory models.
-In the first case study, we show how GLMM trees \cite{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
-We apply GLMM trees to a dataset from a study of \cite{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
+In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
+We apply GLMM trees to a dataset from a study of \citep{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
 Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start.
-In a second case study, we show how Rasch trees \cite{StroyKopf15} can be used to detect subgroups with different parameters in IRT models.
-We show how a recently developed stopping criterion \cite{HennyDeba23} can be used to guide subgroup detection based on effect sizes instead of statistical significance.\\}
+In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
+DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on effect sizes instead of statistical significance.\\}
 
@@
 
-\todo[inline]{Introduce and explain MOB.}
+MOB \citep[MOB]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical
+model between groups of persons. This method is a generalization of the principle of recursive partitioning, that is also used in classification and regression trees \citet{Breetal:1984}. 
+In classification and regression trees, the aim is to detect groups of persons, specified by (combinations of) covariates, that differ in a response variable.
+An example could be that children with ????
+and ????
+show higher average exam grades than others.
+MOB generalizes this idea by identifying groups of persons, specified by (combinations of) covariates, that differ in a the parameters of a statistical model.
+An example could be that children with ????
+and ????
+show higher slopes in a regression model relating time spent studying and exam grades than others.
+
+The framework of MOB is very flexible and can be applied in combination with a variety of statistical models.
+In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models and for models from Rasch measurement models.
+Mixed-effects models become relevant whenever data are collected in a repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are again nested in schools) participate in a study.
+In application example 1 we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
+
+\todo[inline]{should we call a psychological test a test/an instrument/??? to distinguish it from the verb and noun for statistical test?}
+
+While for this first application example we assume that the administered ???
+items/test has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample.
+Before using the test in our research, we need to make sure that test results are comparable, for example between children of different genders.
+If certain items pose an advantage or disadvantage to either group, they are said to violate measurement invariance or exhibit differential item functioning (DIF).
+In order to test for DIF in the framework of Item Response Theory or Rasch modelling \cite{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, groups of persons are compared with respect to the item parameters for their specific group.
+This can be done in a way that allows to detect DIF on top of any true group differences in ability.
+While many standard approaches for testing DIF are limited to comparing typically two pre-specified groups at a time, we will show in application example 2 that MOB more flexibly allows to detect groups of persons with different item parameters in a data-driven way.
+
+In the following section, we will give a short introduction into the algorithm and statistical concepts behind MOB.
+Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}.
+An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
+Mixture modelling aims at identifying latent classes of persons with different properties.
+It can also be combined with covariates.
+A comparison of both approches is given by \citet{FriStrZei:2014}.
+
+\todo[inline]{Caro: I mostly added our own citations so far, need to add more from other authors}
+
+\todo[inline]{Caro: move this to next section}
+One important aspect that we would like to point out, however, is that some algorithms for classification and regression trees and related methods suffer from a problem called variable selection bias.
+This means that, due to their construction principles, some algorithms systematically prefer variables of certain types.
+
+\section{The algorithm behind MOB}
+
+What you see displayed in Figures ???
+and ???, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure ???) and for a binary Rasch model (Figure ???).
+Both/???all figures share the tree-structure at the top.
+The starting point at the very top of each tree is called the root node (so it is actually an upside down tree).
+The root node contains the entire sample of persons in the data set.
+From there, the persons are devided into subgroups.
+The subgroups are defined by covariates and cutpoints in those covariates (for example, in the semi-continuous covariate ???
+\todo{adapt to final examples} the cutpoint ???
+was selected; in a binary covariate like ???, on the other hand, there is only one possible cutpoint, namely between the two categories).
+Which splitting variable and cutpoint is selected in each step of the recursive partitiong algorithms is further explained below.
+The parametric model to which the MOB algorithm is applied is visualized in the end nodes at the bottom of each tree.
+While in Figure ???
+each end node contains a linear mixed-effecs model with a group-specific effect for the ???
+intervention, in Figure ???
+each end node contains the group-specific item parameters of a Rasch model for the ???
+items.
+
+The fact that Figures ??? display more than one end node means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
+
+Figure ???
+also shows that, out of the ???
+covariates that were presented to the algorithm, only ???
+were actually selected.
+Variables are selected based on ????.
+Variables that are no associated with a difference in the model parameters are not included in the tree, i.e., the tree performs automatic variable selection.
+
+still no type I error inflation because closed testing procedure ???cite wie in Raschtree paper
+
+\todo[inline]{Caro: following text copy pasted from Raschtree paper, ueberarbeiten}
+
+What is important to note here is that the entire structure identified by the trees was not pre-specified and provided to the algorithm, but was learned from the data in an exploratory way.
+This is a key feature of the MOB approach that makes it very flexible and distinguishes it from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
+
+Technically, the following consecutive steps are used to infer the structure of a MOB
+
+%
+\begin{enumerate}
+  \item Estimate the model parameters jointly for all subjects in the current sample, starting with the full sample.
+  \item Assess the stability of the parameters with respect
+        to each available covariate.
+  \item If there is significant instability, split the sample
+        along the covariate with the strongest instability and in
+        the cutpoint leading to the highest improvement of the model fit.
+  \item Repeat steps 1--3 recursively in the resulting subsamples
+        until there are no more significant instabilities (or the
+    subsample becomes too small).
+\end{enumerate}
+%
 
@@
 
-A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}.
-There are also papers describing extensions to mixure IRT models \cite{AyalySant17} and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.
+A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}.%included
+There are also papers describing extensions to mixure IRT models \cite{AyalySant17}%included
+and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.
 
-\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).}
+\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).%included
+}
 
 \newpage
-\section{Tutorial: Subgroup Detection in Mixed-Effects Models}
+\section{\edc{Application Example 1}: Subgroup Detection in Mixed-Effects Models}
 \label{sec:TutorialMixed}
@@
 To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start.
-Head Start is a federally funded nationwide pre-school program for children from low-income families.
+Head Start is a federally funded nationwide \todo{what nation?} pre-school program for children from low-income families.
 \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).
 
-The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
+The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children, where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
 Data from children in those families who participated in another preschool program were excluded.
@@
 As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time.
-As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.
+As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.\todo{this would scare me a little as a reader: how would I know what to do/that I should do this?}
 
@@
 To account for the correlation between repeated assessments on the same child, we specified a random intercept.
-The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference persists over time.
+The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference \edc{is slightly diminished but} persists over time.
 This result agrees with the findings of \cite{Demi09}.
@@
 \end{figure}%
+
+\fixme{code for figure not displayed on purpose? why only those particular x axis ticks?}
 
@@
 
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
+To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 \todo{x axis for age starts at 2?} and 18.
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
@@
 The benefit increases over time for White children with higher mother's AFTQ scores.
-Strikingly, the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
+Strikingly,\todo{possibly less extreme word like: Interestingly,} the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
 These results correspond to the conclusions of \cite{Demi09}.
+\todo[inline]{Generally the effects are not large but the example is very clear and I like that it is an intervention effect.
+I suggest to state early before the figure that the effects are not large in this example but we are interested in the group differences; also try to make figure bigger and wider to make slopes more visible}
 
 To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
-We separated the \Sexpr{length(unique(HS_dat$MotherID))} families in the dataset into ten equally-sized folds.
-We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree (i.e., an LMM with subgroups) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child).
-We evaluated performance on the remaining folds, by computing and evaluating accuracy of the predictions.
+We \edc{randomly} separated the \Sexpr{length(unique(HS_dat$MotherID))} families in the dataset into ten equally-sized folds.
+We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree \edc{like the one shown above} (i.e., an LMM with subgroups \todo{necessary?
+confused me}) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child).
+We evaluated performance on the remaining folds by computing and evaluating accuracy of the predictions.
 \todo{explain MSE and $R^2$} We repeated this procedure ten times, so that all folds were used as a test set once.
@@
 Alternative: Do not focus on effect of Head Start, but use as partitioning variable.
-Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.}
+Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.
+-- response Caro: I like it, see comment above: try to prime that effects are not large}
 
@@
 \FloatBarrier
-\section{Tutorial: Subgroup Detection in Rasch Models}
+\section{\edc{Application Example 2}: Subgroup Detection in Rasch Models}
 \label{sec:TutorialRasch}
@@
 
-<<warning=FALSE, message=FALSE, echo=FALSE>>=
+<<warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE>>=
 library("raschtreeMH")
@@
 
-<<>>=
+<<eval=FALSE>>=
 math_items <- c("C7MTHBST", "C7MTHGD", "C7LIKMTH", "C7ENJMTH")
@@
 
-<<>>=
+<<eval=FALSE>>=
 stop_fun <- stopfun_mantelhaenszel(purification = "iterative", stopcrit = "C")
@@
 \begin{subfigure}{1.25\textwidth}
-<<fig=TRUE, echo=FALSE, width= 9, height=6>>=
+<<fig=TRUE, echo=FALSE, width= 9, height=6, eval=FALSE>>=
 library("psychotree") ## needed for plotting, but interferes with stopfun_mantelhaenszel
@@
 
-<<>>=
+<<eval=FALSE>>=
 read_items <-  c("C7ENGBST", "C7GRDENG", "C7LIKRD", "C7ENJRD")
@@
 \begin{subfigure}{1.25\textwidth}
-<<fig=TRUE, width=9, height=6, echo=FALSE>>=
+<<fig=TRUE, width=9, height=6, echo=FALSE, eval=FALSE>>=
 library("psychotree") ## needed for plotting, interferes with stopfun_mantelhaenszel
@@
 
-\bibliography{MOB_paper}
+\bibliography{_MOB_paper}
 

### 2023-09-22 _MOB_paper.tex
not sure why those files are shown in status, tex and pdf should be ignored

@@
 \begin{document}
-\input{MOB_paper-concordance}
+\input{_MOB_paper-concordance}
 \renewenvironment{Schunk}{\small}{}
@@
 To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start.
-Head Start is a federally funded nationwide pre-school program for children from low-income families.
+Head Start is a federally funded nationwide \todo{what nation?} pre-school program for children from low-income families.
 \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).
 
-The sample consists of 273 families with at least two children where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
+The sample consists of 273 families with at least two children, where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
 Data from children in those families who participated in another preschool program were excluded.
@@
 \begin{subfigure}{.7\textwidth}
-\includegraphics{MOB_paper-004}
+\includegraphics{_MOB_paper-004}
 \end{subfigure}
@@
 \caption{}
-\includegraphics{MOB_paper-007}
+\includegraphics{_MOB_paper-007}
 
@@
 
-\includegraphics{MOB_paper-008}
+\includegraphics{_MOB_paper-008}
 \label{fig:lmm_tree}
@@
 \end{table}
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
+To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 \todo{x axis for age starts at 2?} and 18.
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
@@
 The benefit increases over time for White children with higher mother's AFTQ scores.
-Strikingly, the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
+Strikingly,\todo{possibly less extreme word like: Interestingly,} the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
 These results correspond to the conclusions of \cite{Demi09}.
+\todo[inline]{Generally the effects are not large but the example is very clear and I like that it is an intervention effect.
+I suggest to state early before the figure that the effects are not large in this example but we are interested in the group differences; also try to make figure bigger and wider to make slopes more visible}
 
 To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
-We separated the 258 families in the dataset into ten equally-sized folds.
-We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree (i.e., an LMM with subgroups) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child).
-We evaluated performance on the remaining folds, by computing and evaluating accuracy of the predictions.
+We \edc{randomly} separated the 258 families in the dataset into ten equally-sized folds.
+We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree \edc{like the one shown above} (i.e., an LMM with subgroups \todo{necessary?
+confused me}) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child).
+We evaluated performance on the remaining folds by computing and evaluating accuracy of the predictions.
 \todo{explain MSE and $R^2$} We repeated this procedure ten times, so that all folds were used as a test set once.
@@
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{MOB_paper-018}
 \end{subfigure}
@@
 \begin{subfigure}{1.25\textwidth}
-\includegraphics{MOB_paper-021}
 \end{subfigure}

### 2023-09-25 _MOB_paper.Rnw
intro

@@
 
-\abstract{MOB \citep[MOB]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models.
+\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models.
 It provides a versatile tool for detecting and explaining heterogeneity of intervention effects.
@@
 In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
-DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on effect sizes instead of statistical significance.\\}
+DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}
 
@@
 
-MOB \citep[MOB]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical
+Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical
 model between groups of persons. This method is a generalization of the principle of recursive partitioning, that is also used in classification and regression trees \citet{Breetal:1984}. 
 In classification and regression trees, the aim is to detect groups of persons, specified by (combinations of) covariates, that differ in a response variable.
-An example could be that children with ????
-and ????
-show higher average exam grades than others.
+An example could be that children who are highly motivated and have good reading skills show higher average math exam grades.
 MOB generalizes this idea by identifying groups of persons, specified by (combinations of) covariates, that differ in a the parameters of a statistical model.
-An example could be that children with ????
-and ????
-show higher slopes in a regression model relating time spent studying and exam grades than others.
+An example could be that children who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.
 
-The framework of MOB is very flexible and can be applied in combination with a variety of statistical models.
-In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models and for models from Rasch measurement models.
-Mixed-effects models become relevant whenever data are collected in a repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are again nested in schools) participate in a study.
+The framework of MOB is very flexible and can be applied in combination with a variety of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS}.
+In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \cite{FokkySmit18} and for models from Rasch measurement and Item Response Theory \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
+Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are again nested in schools) participate in a study.
 In application example 1 we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
 
-\todo[inline]{should we call a psychological test a test/an instrument/??? to distinguish it from the verb and noun for statistical test?}
-
-While for this first application example we assume that the administered ???
-items/test has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample.
+While for this first application example we assume that the administered test \todo{should we call a psychological test a test/an instrument/???
+to distinguish it from the verb and noun for statistical test?} has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample.
 Before using the test in our research, we need to make sure that test results are comparable, for example between children of different genders.
@@
 This can be done in a way that allows to detect DIF on top of any true group differences in ability.
-While many standard approaches for testing DIF are limited to comparing typically two pre-specified groups at a time, we will show in application example 2 that MOB more flexibly allows to detect groups of persons with different item parameters in a data-driven way.
+While many standard approaches for testing DIF are limited to comparing typically two pre-specified groups at a time, we will show in application example 2 that MOB more flexibly allows to detect groups of persons with different item parameters in a data-driven
+way.
 
@@
 
-\todo[inline]{Caro: I mostly added our own citations so far, need to add more from other authors}
-
-\todo[inline]{Caro: move this to next section}
-One important aspect that we would like to point out, however, is that some algorithms for classification and regression trees and related methods suffer from a problem called variable selection bias.
-This means that, due to their construction principles, some algorithms systematically prefer variables of certain types.
-
 \section{The algorithm behind MOB}
 
-What you see displayed in Figures ???
-and ???, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure ???) and for a binary Rasch model (Figure ???).
-Both/???all figures share the tree-structure at the top.
-The starting point at the very top of each tree is called the root node (so it is actually an upside down tree).
-The root node contains the entire sample of persons in the data set.
-From there, the persons are devided into subgroups.
-The subgroups are defined by covariates and cutpoints in those covariates (for example, in the semi-continuous covariate ???
-\todo{adapt to final examples} the cutpoint ???
-was selected; in a binary covariate like ???, on the other hand, there is only one possible cutpoint, namely between the two categories).
-Which splitting variable and cutpoint is selected in each step of the recursive partitiong algorithms is further explained below.
+What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
+These figures share the tree-structure in the upper part, but differ in their lower part.
+The starting point of the tree structure is at the very top of each tree.
+The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the
+bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.
+
+The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
+For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \income{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups.
+The algorithm has identified this cutpoint as being the location of the strongest parameter difference between the two resulting groups.
+The categorical covariate \texttt{Race}, which was coded in three categories in this example, has been divided into two groups of categories: Black and Hispanic vs.~White.
+For a binary covariate, \todo{like ...
+-- add name if is there one in the other example?}
+on the other hand, there is only one possible cutpoint, namely between the two categories.
+While it is also possible to create more than two groups in each split \citep{KimLoh:2001,Qui:1993}, binary splitting algorithms are typically preferred because they lead to more concise trees.
+
 The parametric model to which the MOB algorithm is applied is visualized in the end nodes at the bottom of each tree.
-While in Figure ???
-each end node contains a linear mixed-effecs model with a group-specific effect for the ???
-intervention, in Figure ???
-each end node contains the group-specific item parameters of a Rasch model for the ???
-items.
+While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items.
+The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
+Figure~\ref{fig:lmm_tree} also shows that, out of the  covariates that were presented to the algorithm (\texttt{AFTQ}, \texttt{Race}, \texttt{Income},\texttt{Mom\_edu\_yrs}, and \texttt{Mom\_height}), only \texttt{AFTQ}, \texttt{Race}, and \texttt{Income} were actually selected for splitting.
+Variables are selected based on a statistical test for parameter differences, also termed a test for structural change.
+Structural change is present if, for example, the model parameters systematically differ for children from lower vs.~higher income families.
+At each node, the algorithm will select the covariate showing the strongest structural change as the next splitting variable.
+Within this variables, the optimal cutpoint is chosen in a separate step.
 
-The fact that Figures ??? display more than one end node means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
+The way the variable selection and cutpoint selection steps are separated in modern classification and regression tree and MOB algorithms, including the ones used here, distinguish them from the traditional classification and regression tree algorithms of \citet{Breetal:1984} and \citet{Qui:1993}.
+The traditional algorithms performed variable and cutpoint selection in one step.
+However, this leads to an undesirable behavior called variable selection bias.
+It means that the traditional algorithms prefer variables offering more cutpoints in the selection process -- regardless of their true information content.
+An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}.
+Modern algorithms for classification and regression trees that have solved this problem are the unbiased approaches \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, available in R in the \texttt{party} and \texttt{partykit} packages \citep{partykit:pkg}.
+The latter forms the basis for all MOB approaches presented in this paper.
 
-Figure ???
-also shows that, out of the ???
-covariates that were presented to the algorithm, only ???
-were actually selected.
-Variables are selected based on ????.
-Variables that are no associated with a difference in the model parameters are not included in the tree, i.e., the tree performs automatic variable selection.
+Once there are no more covariates that show a significant structural change in any node, the MOB algorithm stops splitting.
+In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection.
+Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected.
+This is the second difference between the modern classification and regression tree and MOB algorithms used here, compared to traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}: While the traditional algorithms grew very large trees and then cut them back by means of so called pruning, the framework employed here is based on statistical significance tests and party also on effect size measures (see application example 2) to stop the trees already in the growing phase when no more significant change can be detected.
+Other stopping criteria are based on the number of persons in the end nodes.
+These crieria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}
 
-still no type I error inflation because closed testing procedure ???cite wie in Raschtree paper
+In summary, the rationale of a MOB algorithm consists of the following steps:
 
-\todo[inline]{Caro: following text copy pasted from Raschtree paper, ueberarbeiten}
+\begin{enumerate}
+  \item The model parameters are first estimated jointly for all persons in the current node, starting with the full sample.
+  \item Structural change in the model parameters is assessed with respect to each available covariate.
+  \item If there is significant structural change, groups are split along the covariate with the strongest change and using the optimal cutpoint.
+  \item Steps 1--3 are repeated recursively in the resulting groups
+        until there is no more significant structural change (or the
+   groups becomes too small).
+\end{enumerate}
 
-What is important to note here is that the entire structure identified by the trees was not pre-specified and provided to the algorithm, but was learned from the data in an exploratory way.
+For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.
+
+An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory way, but is learned from the data in an exploratory way.
 This is a key feature of the MOB approach that makes it very flexible and distinguishes it from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
-
-Technically, the following consecutive steps are used to infer the structure of a MOB
-
-%
-\begin{enumerate}
-  \item Estimate the model parameters jointly for all subjects in the current sample, starting with the full sample.
-  \item Assess the stability of the parameters with respect
-        to each available covariate.
-  \item If there is significant instability, split the sample
-        along the covariate with the strongest instability and in
-        the cutpoint leading to the highest improvement of the model fit.
-  \item Repeat steps 1--3 recursively in the resulting subsamples
-        until there are no more significant instabilities (or the
-    subsample becomes too small).
-\end{enumerate}
-%
+While there are phases in psychological and educational research where it is very important to specify hypothesis a priori and test them in a confirmatory way, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers.
 
@@
 
-A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}.%included
-There are also papers describing extensions to mixure IRT models \cite{AyalySant17}%included
+A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}. \edc{included}
+There are also papers describing extensions to mixure IRT models \cite{AyalySant17} \edc{included}
 and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.
 
-\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).%included
+\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).
+\edc{included}
 }

### 2023-09-25 _MOB_paper.tex
intro

@@
 
-\todo[inline]{Caro: I mostly added our own citations so far, need to add more from other authors}
-
-\todo[inline]{Caro: move this to next section}
-One important aspect that we would like to point out, however, is that some algorithms for classification and regression trees and related methods suffer from a problem called variable selection bias.
-This means that, due to their construction principles, some algorithms systematically prefer variables of certain types.
-
 \section{The algorithm behind MOB}
 
-What you see displayed in Figures ???
-and ???, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure ???) and for a binary Rasch model (Figure ???).
-Both/???all figures share the tree-structure at the top.
-The starting point at the very top of each tree is called the root node (so it is actually an upside down tree).
-The root node contains the entire sample of persons in the data set.
-From there, the persons are devided into subgroups.
-The subgroups are defined by covariates and cutpoints in those covariates (for example, in the semi-continuous covariate ???
-\todo{adapt to final examples} the cutpoint ???
-was selected; in a binary covariate like ???, on the other hand, there is only one possible cutpoint, namely between the two categories).
-Which splitting variable and cutpoint is selected in each step of the recursive partitiong algorithms is further explained below.
+What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
+These figures share the tree-structure in the upper part, but differ in their lower part.
+The starting point of the tree structure is at the very top of each tree.
+The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the
+bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.
+
+The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
+For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \income{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups.
+The algorithm has identified this cutpoint as being the location of the strongest parameter difference between the two resulting groups.
+The categorical covariate \texttt{Race}, which was coded in three categories in this example, has been divided into two groups of categories: Black and Hispanic vs.~White.
+For a binary covariate, \todo{like ...
+-- add name if is there one in the other example?}
+on the other hand, there is only one possible cutpoint, namely between the two categories.
+While it is also possible to create more than two groups in each split \citep{KimLoh:2001,Qui:1993}, binary splitting algorithms are typically preferred because they lead to more concise trees.
+
 The parametric model to which the MOB algorithm is applied is visualized in the end nodes at the bottom of each tree.
-While in Figure ???
-each end node contains a linear mixed-effecs model with a group-specific effect for the ???
-intervention, in Figure ???
-each end node contains the group-specific item parameters of a Rasch model for the ???
-items.
+While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items.
+The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
+Figure~\ref{fig:lmm_tree} also shows that, out of the  covariates that were presented to the algorithm (\texttt{AFTQ}, \texttt{Race}, \texttt{Income},\texttt{Mom\_edu\_yrs}, and \texttt{Mom\_height}), only \texttt{AFTQ}, \texttt{Race}, and \texttt{Income} were actually selected for splitting.
+Variables are selected based on a statistical test for parameter differences, also termed a test for structural change.
+Structural change is present if, for example, the model parameters systematically differ for children from lower vs.~higher income families.
+At each node, the algorithm will select the covariate showing the strongest structural change as the next splitting variable.
+Within this variables, the optimal cutpoint is chosen in a separate step.
 
-The fact that Figures ??? display more than one end node means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
+The way the variable selection and cutpoint selection steps are separated in modern classification and regression tree and MOB algorithms, including the ones used here, distinguish them from the traditional classification and regression tree algorithms of \citet{Breetal:1984} and \citet{Qui:1993}.
+The traditional algorithms performed variable and cutpoint selection in one step.
+However, this leads to an undesirable behavior called variable selection bias.
+It means that the traditional algorithms prefer variables offering more cutpoints in the selection process -- regardless of their true information content.
+An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}.
+Modern algorithms for classification and regression trees that have solved this problem are the unbiased approaches \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, available in R in the \texttt{party} and \texttt{partykit} packages \citep{partykit:pkg}.
+The latter forms the basis for all MOB approaches presented in this paper.
 
-Figure ???
-also shows that, out of the ???
-covariates that were presented to the algorithm, only ???
-were actually selected.
-Variables are selected based on ????.
-Variables that are no associated with a difference in the model parameters are not included in the tree, i.e., the tree performs automatic variable selection.
+Once there are no more covariates that show a significant structural change in any node, the MOB algorithm stops splitting.
+In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection.
+Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected.
+This is the second difference between the modern classification and regression tree and MOB algorithms used here, compared to traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}: While the traditional algorithms grew very large trees and then cut them back by means of so called pruning, the framework employed here is based on statistical significance tests and party also on effect size measures (see application example 2) to stop the trees already in the growing phase when no more significant change can be detected.
+Other stopping criteria are based on the number of persons in the end nodes.
+These crieria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}
 
-still no type I error inflation because closed testing procedure ???cite wie in Raschtree paper
+In summary, the rationale of a MOB algorithm consists of the following steps:
 
-\todo[inline]{Caro: following text copy pasted from Raschtree paper, ueberarbeiten}
+\begin{enumerate}
+  \item The model parameters are first estimated jointly for all persons in the current node, starting with the full sample.
+  \item Structural change in the model parameters is assessed with respect to each available covariate.
+  \item If there is significant structural change, groups are split along the covariate with the strongest change and using the optimal cutpoint.
+  \item Steps 1--3 are repeated recursively in the resulting groups
+        until there is no more significant structural change (or the
+   groups becomes too small).
+\end{enumerate}
 
-What is important to note here is that the entire structure identified by the trees was not pre-specified and provided to the algorithm, but was learned from the data in an exploratory way.
+For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StroyKopf15,StrMalTut:2009:PM}.
+
+An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory way, but is learned from the data in an exploratory way.
 This is a key feature of the MOB approach that makes it very flexible and distinguishes it from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
-
-Technically, the following consecutive steps are used to infer the structure of a MOB
-
-%
-\begin{enumerate}
-  \item Estimate the model parameters jointly for all subjects in the current sample, starting with the full sample.
-  \item Assess the stability of the parameters with respect
-        to each available covariate.
-  \item If there is significant instability, split the sample
-        along the covariate with the strongest instability and in
-        the cutpoint leading to the highest improvement of the model fit.
-  \item Repeat steps 1--3 recursively in the resulting subsamples
-        until there are no more significant instabilities (or the
-    subsample becomes too small).
-\end{enumerate}
-%
+While there are phases in psychological and educational research where it is very important to specify hypothesis a priori and test them in a confirmatory way, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers.
 

### 2023-10-02 _MOB_paper.Rnw
intro ctd

@@
 \shorttitle{shorttitle}
-\authorsnames{Marjolein Fokkema$^1$ \& Carolin Strobl$^2$}
-\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}
+\author{Marjolein Fokkema, Mirka Henninger \& Carolin Strobl}
+
+%\authorsnames{Marjolein Fokkema$^1$ \& Carolin Strobl$^2$}
+%\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}
 
@@
 \maketitle
+\todo[inline]{Marjolein: what document style for authorsnames? commented out}
 
@@
 
-\section{The algorithm behind MOB}
+\section{The Algorithm Behind MOB}
 
@@
 The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
-For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \income{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups.
+For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \texttt{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups.
 The algorithm has identified this cutpoint as being the location of the strongest parameter difference between the two resulting groups.
@@
 
+\todo{why are the spaces so big? I didn't change the enumerate command}
 \begin{enumerate}
@@
 }
+
+The general framework of MOB introduced above will now be applied to two types of statistical models that are particularly relevant in school psychology research: Mixed-effects models for repeated measures or nested data structures and measurement models for validating psychological and educational tests.
+
+\subsection{Using MOB for Subgroup Detection in Mixed-Effects Models}
+
+In mixed-effects models contain two types of effects: fixed effects and random effects.
+Fixed effects are typically used, for example, to describe the effect of an intervention on the average student performance.
+Random effects are used to describe the variation in the intercept and slope for, e.g., students in different classes.
+
+When mixed-effects models are combined with MOB, ...
+
+\todo[inline]{Marjolein please edit and finish}
+
+\subsection{Using MOB for Detecting DIF in Measurement Models}
+
+Before being able to use a psychological or educational test for comparing different groups of persons in a fair way, different assumptions of the measurement model have to be checked.
+In the framework of IRT and Rasch measurement, test items are typically investigated with respect item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}.
+Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance.
+Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
+
+MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model.
+For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustration in application example 2.
+We will see that, just like for the mixed-effects model trees \todo{use same names for methods everywhere}, we need to specify which variables go into the parametric mode.
+In the case of the Rasch model, this will be the test items.
+Moreover, we need to specify, which variables are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.
+
+If one joint Rasch model holds for the entire sample, i.e., if there is no DIF a Rasch tree should show no splits.
+However, in certain settings in educational research, such as large scale assessements, very large sample sizes are available for testing for DIF.
+Large sample sizes are good for detecting even small effects or model violations with a high statistical power.
+The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected with large samples.
+However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
+As we will illustrate in the first part of application example 2, for a large sample of university students \todo{are only students in the sample?} who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
+\citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
+In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
+\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.\todo{Mirka: is it A vs.
+B,C or A,B vs.
+C by default?}
+
+As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
+Together with a purification step \citet[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
+For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words used in the different languages are equally frequently used, etc.
+In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test.
+Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.
+
+Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
+This would be the case in test inteded to measure math aptitude containing pure algebra problems as well as story problems.
+Students whose native language is not the same as the test language can have disadvantage in answering the story problems, for example when they contain seldomly used words.
+These items will then show DIF between native and non-native speakers.
+When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, e.g., by using more frequently used words, or whether the test should be considered two-dimensional rather than one-dimensional \citep[see also][for a discussion of the connection between DIF and multidimensionality]{Ack:1992,Stretal:2021:APM}.
+
+We will now illustrate how to use the R packages \texttt{glmertree} and \texttt{psychotree} to conduct the MOB analyses.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\fixme{code for figure not displayed on purpose? why only those particular x axis ticks?}
+\todo{code for figure not displayed on purpose? why only those particular x axis ticks?}
 
@@
 
+In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.
+The main advantage of MOB is that it can detect groups of persons with different model parameters in a data driven way.
+This makes it more flexible for detecting differences that were not hypothesized by the researchers.
+For example in DIF analysis, it is often the case that any obvious sources of DIF have already been avoided by the content expernts in the item creation phase.
+Any remaining DIF is unexpected.
+Therefore, any DIF detection approach that relies on the researchers correctly specifying the exact groups of persons that exhibit DIF can miss DIF if it is associated with other (combinations of) covariates or other cutpoints than the ones investigated.
+In this sense, a more exploratory approach like a Rasch tree has a higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
+The same holds for parameter differences in mixed-effects models, where the substantial hypothesis are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
+\todo{Marjolein add if this makes any sense here}
+
 \todo[inline]{Write summary of findings.}
+
+For our two application examples, we could show that MOB can detect ..., performs automated variable selection \todo[inline]{or did you, Marjolein, mean repeat substantive findings?
+not very exciting for SPISA}
 
@@
 
+not possible to detect certain patterns: XOR problem, known for all trees
+
+stability of trees, small changes in data can lead to quite different looking tree.
+The tree structure should always be interpreted jointly for all variables: rather than considering the first splitting variable to be the most important one, it is actually the combination of all splitting variables that creates the groups.\todo{in interpretation of trees in examples: do this, point out that it is interactions -- Caro also add in intro?}.
+The \texttt{stabletree} R package also provides descriptive statistics and graphics that help judge the stability of tree and MOB results \citep{PhiRusHor:2018:JoCaGS,PhiZeiStr:2016}.
+
 \todo[inline]{Write about future work.}
+
+sample size dependence, similar ideas like MH trees also possible for other MOB?
+
+The Mantel-Haenszel effect size measure for DIF will be integrated in the \texttt{psychotools} package.
 

### 2023-10-02 _MOB_paper.tex
intro ctd

@@
 
-For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StroyKopf15,StrMalTut:2009:PM}.
+For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.
 

### 2023-10-04 _MOB_paper.Rnw
Small changes in Intro + Comment in glmer example

@@
 Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical
-model between groups of persons. This method is a generalization of the principle of recursive partitioning, that is also used in classification and regression trees \citet{Breetal:1984}. 
+model between groups of persons. This method is a generalization of the principle of recursive partitioning, that is also used in classification and regression trees \citep{Breetal:1984}. 
 In classification and regression trees, the aim is to detect groups of persons, specified by (combinations of) covariates, that differ in a response variable.
@@
 The framework of MOB is very flexible and can be applied in combination with a variety of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS}.
-In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \cite{FokkySmit18} and for models from Rasch measurement and Item Response Theory \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
+In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and for models from Rasch measurement and Item Response Theory \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
 Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are again nested in schools) participate in a study.
@@
 While for this first application example we assume that the administered test \todo{should we call a psychological test a test/an instrument/???
-to distinguish it from the verb and noun for statistical test?} has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample.
+to distinguish it from the verb and noun for statistical test?
+(Mirka: I like the term instrument to distinguish it from a statistical test)} has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample.
 Before using the test in our research, we need to make sure that test results are comparable, for example between children of different genders.
 If certain items pose an advantage or disadvantage to either group, they are said to violate measurement invariance or exhibit differential item functioning (DIF).
-In order to test for DIF in the framework of Item Response Theory or Rasch modelling \cite{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, groups of persons are compared with respect to the item parameters for their specific group.
+In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, groups of persons are compared with respect to the item parameters for their specific group.
 This can be done in a way that allows to detect DIF on top of any true group differences in ability.
@@
 
-What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
+What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?)}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
 These figures share the tree-structure in the upper part, but differ in their lower part.
@@
 Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected.
-This is the second difference between the modern classification and regression tree and MOB algorithms used here, compared to traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}: While the traditional algorithms grew very large trees and then cut them back by means of so called pruning, the framework employed here is based on statistical significance tests and party also on effect size measures (see application example 2) to stop the trees already in the growing phase when no more significant change can be detected.
+This is the second difference between the modern classification and regression tree and MOB algorithms used here, compared to traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}: While the traditional algorithms grew very large trees and then cut them back by means of so called pruning, the framework employed here is based on statistical significance tests and partly also on effect size measures (see application example 2) to stop the trees already in the growing phase when no more significant change can be detected.
 Other stopping criteria are based on the number of persons in the end nodes.
@@
   \item If there is significant structural change, groups are split along the covariate with the strongest change and using the optimal cutpoint.
-  \item Steps 1--3 are repeated recursively in the resulting groups
-        until there is no more significant structural change (or the
-   groups becomes too small).
+  \item Steps 1--3 are repeated recursively in the resulting groups until there is no more significant structural change (or the groups becomes too small).
 \end{enumerate}
@@
 For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustration in application example 2.
-We will see that, just like for the mixed-effects model trees \todo{use same names for methods everywhere}, we need to specify which variables go into the parametric mode.
+We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables go into the parametric mode.
 In the case of the Rasch model, this will be the test items.
@@
 However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
-As we will illustrate in the first part of application example 2, for a large sample of university students \todo{are only students in the sample?} who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+As we will illustrate in the first part of application example 2, for a large sample of university students \todo{are only students in the sample?
+(Mirka: I can take a subsample in which there are only students)} who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
 Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
@@
 B,C or A,B vs.
-C by default?}
+C by default?; Response Mirka: A vs.
+BC is the default, but it can be changed}
 
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citet[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+Together with a purification step \citep[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
@@
 Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
-This would be the case in test inteded to measure math aptitude containing pure algebra problems as well as story problems.
+This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
 Students whose native language is not the same as the test language can have disadvantage in answering the story problems, for example when they contain seldomly used words.
@@
 As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time.
-As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.\todo{this would scare me a little as a reader: how would I know what to do/that I should do this?}
+As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.\todo{this would scare me a little as a reader: how would I know what to do/that I should do this?
+(Mirka: Maybe we can add that the sqrt or the log is a typical way to model trajectories that first increase rapidly, but then later increase more slowly?)}
 
@@
 
+\todo[inline]{I (Mirka) am confused: the nested structure in this example is timepoints (age) within children?
+Therefore, ChildID is the cluster-variable.
+But then, there should be residual errors between the two children from the same family?
+In the text it says that this is accounted for by the random intercept, but I don't think this is the case (but maybe I am wrong?).
+Is it maybe accounted for by the cluster argument that is also said to ChildID?
+To me, the model specification is not yet clear from the text.
+Plus: If I am right, and the cluster is the child and the timepoints are on Level-1, I think all the covariates are on Level-2 (child level), and then maybe we should comment on the Type-I error problem?}
 We now test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:

### 2023-10-04 _MOB_paper.Rnw
Comment on glmertree example

@@
 With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance.
-The middle part (\texttt{1|ChildID}) specified the random effects.
+The middle part (\texttt{1|ChildID}) specified the random effects\todo{(Mirka) Maybe add information that this is for each child, average DV across timepoints}.
 The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.
 
+\todo[inline]{I (Mirka) am not sure what exactly the cluster argument does.
+Maybe Marjolein you have more information from your work with Achim.
+But my knowledge/intuition is that it has something to do with clustered covariances, so maybe it accounts for the sibling structure, but I don't think it accounts for the level of the predictor variables.}
 With the second argument, we specified the dataset which contain the variables.
 Because we are dealing with repeated measurements on the same children, we additionally specified that the parameter stability tests should be performed on the child level using the \verb|cluster| argument.
-Using the default observation-level parameter stability tests may artificially inflate power.
+Using the default observation-level parameter stability tests may artificially inflate power\todo{Mirka: I'd rather say it has an inflated type-I error rate.}.
 Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.
@@
 \todo[inline]{Generally the effects are not large but the example is very clear and I like that it is an intervention effect.
-I suggest to state early before the figure that the effects are not large in this example but we are interested in the group differences; also try to make figure bigger and wider to make slopes more visible}
+I suggest to state early before the figure that the effects are not large in this example but we are interested in the group differences; also try to make figure bigger and wider to make slopes more visible; (Mirka): I think the main effect of the intervention and the interaction effect is not too small, but it is hard to see because the age effect is so pronounced in this example.
+Would it be an option to only use the intervention effect in the end nodes?
+}
 
@@
 Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.
--- response Caro: I like it, see comment above: try to prime that effects are not large}
+-- response Caro: I like it, see comment above: try to prime that effects are not large -- response Mirka: I think that's okay.
+Only if you had data at hand that is not longitudinal (rather students in classes/schools), with covariates on Level-1, I would opt for that because it doesn't have the Level-2 splitting issue.
+}
 

### 2023-10-04 _MOB_paper.Rnw
New example for Rasch tree and backup of old example

@@
 
-Using Rasch trees, we assess DIF for items on math and reading \todo{decide on subscale(s) to include} from the Self-Description Questionnaire \cite{Boyl94}.
-We use a dataset from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}.
-Data were collected from 1,018 schools across the USA.
-Assessments took place from kindergarten through 8th grade.
-Here we focus on the 8th grade assessment, in which four SDQ-II items were administered to assess perceived interest and competence in reading, and another four to assess perceived interest and competence in math.
-Items are presented in Appendix~\ref{sec:AppendixA}.
-The items were rated by the children on a 4-point scale (“not at all true” through “very true”).
-We coded responses 1 and 2 as 0, and responses 3 and 4 as 1, to allow for fitting a Rasch model, which assumes binary responses.
+\begin{itemize}
+\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet)
+\item take the subsample of students (only students), and a random subsample of 5000 respondents (seed 04102023)
+\item items could probably appear in Appendix??
+\begin{itemize}
+\item Which painter created this painting? – Andy Warhol.
+\item What do these four buildings have in common? – All four were designed by the same architects.
+\item Roman numbers: What is the meaning of CLVI? – 156.
+\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
+\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
+\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
+\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
+\item In which city is this building located? – Paris.
+\item Which one of the following operas is not by Mozart? – Aida.
+\end{itemize}
+\end{itemize}
 
-<<echo=FALSE>>=
-SDQ_dat <- readRDS("ECLSK_SDQ.Rda")
+\todo[inline]{at Caro: Ich habe die Daten (glaube ich) mal von dir bekommen (\texttt{spisa\_ges.RData}) und die Item-Formulierungen habe ich aktuell aus dem psychotree Paket.
+Stimmt das so?
+Ich glaube, der Originaltest hatte mehrere parallele Skalen, zumindest sind in dem PDF/BUCH von Trepte sehr viel mehr, und sehr ähnliche Items enthalten.
+Das könnten wir vielleicht nochmal nachgucken.}
+
+We load the data and inspect the first rows.
+there are several covariates: age (continuous), gender (male, female, missing), area (Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, Sports) and 9 items from the culture scale (true/false 0/1).
+
+@Caro: I excluded the covariates student and occupation, because I took the student subsample;
+
+<<echo = FALSE>>=
+load("SPISA_example/dat_SPISA.rda")
+covar_data$Student <- NULL
+covar_data$Occupation <- NULL
 @
 
-We used three covariates as possible partitioning variables: Gender (1=Male; 2=Female), race (8 categories: 1=White, non-Hispanic; 2=Black or African-American, non-Hispanic; 3=Hispanic, race specified; 4=Hispanic, race not specified; 5=Asian; 6=Native Hawaiian or other Pasific Islander; 7= American Indian or Alaska native; 7=More than one race) and socio-economic status (range $-$5 to 3).
-\todo{Add more partitioning variables?
-This does increase number of splits.
-But could add one that should not have an effect, e.g., month of birth.} We analyze observations of children with complete data only, yielding a total sample size of \Sexpr{formatC(nrow(SDQ_dat), big.mark=",")}.
+<<>>=
+head(covar_data)
+head(culture_scale)
+@
 
-We load the data and inspect the first three rows as follows:
+We create a dataset \texttt{dat\_SPISA} that contains the data of the covariate and the item responses from the culture scale.
+By assigning the whole scale to the dataset using \texttt{\$}, we can later access item responses from all items using \texttt{\$culture}.
+<<>>=
+dat_SPISA <- covar_data
+dat_SPISA$culture <- culture_scale
+@
+
+The Rasch tree can be fitted with the function \texttt{raschtree} from the R package \texttt{psychotree}.
+We use the typical formula syntax with the item responses of the culture scale on the left hand side of $\sim$, and the covariates on the right hand side.
+
+<<echo = FALSE>>=
+if(!file.exists("SPISA_example/Raschtree.rda")){
+  library(psychotree)
+  Raschtree_culture <- raschtree(culture ~  Gender + Age + Area, 
+                                 data = dat_SPISA)
+  save(Raschtree_culture, file = "SPISA_example/Raschtree.rda")
+}
+load("SPISA_example/Raschtree.rda")
+@
+
+<<eval = FALSE>>=
+library(psychotree)
+Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
+                               data = dat_SPISA)
+@
+
+We can plot the Raschtree using the plot function.
+But it is very big, lots of splits, we don't know whether the splits are due to substantial differences in item difficulty parameters, because the item difficulty profiles in the end nodes are hard to interpret/compare.
+
+<<fig = TRUE, height = 8, width = 15>>=
+plot(Raschtree_culture)
+@
+
+Mantel-Haenszel trees use the Mantel-Haenszel effect size measure for DIF.
+it has three categories (A: negligible, B: moderate, C: large).
+If none of the items has DIF in category B or C (default, but can also be changed so that DIF has to be C), the tree is stopped from growing.
+
+The software must be installed from github, but will (probably) be implemented in the psychotree package some time in the future
 
 <<>>=
-SDQ_dat <- readRDS("ECLSK_SDQ.Rda")
-head(SDQ_dat, 3)
+devtools::install_github("mirka-henninger/raschtreeMH")
+library(raschtreeMH)
 @
 
-To fit Rasch trees, we use function \verb|raschtree| which is available from R package \textbf{psychotree}.
-However, the stopping criterion based on the Mantel–Haenszel effect size measure is implemented in R package \verb|raschtreeMH|, which we therefore use here.
-\todo{might be nice to implement the MH effect-size criterion in the psychotree package for this paper, currently it seems to interfere with plotting the tree}.
-It can be installed and loaded as follows:
+Syntax is very similar to the classical Rasch tree, but an additional argument \texttt{stopfun}, where the mantelhaneszel stopping function can be selected (but also own stopping functions can be provided; see Henninger et al., 2023).
+The user also has to indicate what kind of purification strategy should be used (none, 2step, iterative).
+Iterative is recommended.
 
-<<warning=FALSE, message=FALSE, eval=FALSE>>=
-devtools::install_github("mirka-henninger/raschtreeMH")
-library("raschtreeMH")
+<<echo = FALSE>>=
+if(!file.exists("SPISA_example/Raschtree.rda")){
+  library(psychotree)
+  Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area, 
+                                  data = dat_SPISA, 
+                                  stopfun= stopfun_mantelhaenszel(purification = "iterative"))
+  Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, purification = "iterative")
+  save(Raschtree_MH_culture, file = "SPISA_example/Raschtree_MH_culture.rda")
+}
+load("SPISA_example/Raschtree_MH_culture.rda")
 @
 
-<<warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE>>=
-library("raschtreeMH")
+<<eval = FALSE>>=
+Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area, 
+                                  data = dat_SPISA, 
+                                  stopfun= stopfun_mantelhaenszel(
+                                    purification = "iterative"))
 @
 
-To fit a Rasch tree, we first create a \verb|data.frame| that contains the item responses and possible partitioning variables:
+For technical reasons, the information about the effect size and ETS classification are not saved in the tree object itself but have to be added afterwards using the \texttt{add\_mantelhaenszel} function.
+The information can then be accessed from the object using \texttt{\$info\$mantelhaenszel}
 
-<<eval=FALSE>>=
-math_items <- c("C7MTHBST", "C7MTHGD", "C7LIKMTH", "C7ENJMTH")
-part_vars <- c("GENDER", "RACE", "WKSESL")
-mydata <- SDQ_dat[ , part_vars]
-mydata$resp <- sapply(SDQ_dat[ , math_items], function(x) ifelse(x > 2, 1, 0)) 
+<<eval = FALSE>>=
+Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, 
+                                           purification = "iterative")
+Raschtree_MH_culture$info$mantelhaenszel
 @
 
-Next, we apply function \verb|raschtree| and plot the result:
+If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.
+In addition, we see the number of items classified as A, B, or C in each node.
+For instance, we see that one item is classified as B in Node 1, and no item as C.
 
-<<eval=FALSE>>=
-stop_fun <- stopfun_mantelhaenszel(purification = "iterative", stopcrit = "C")
-math_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
+<<fig = TRUE, width = 12, height = 7>>=
+plot(Raschtree_MH_culture)
 @
 
-<<eval=FALSE>>=
-plot(read_tree, gp = gpar(cex = .7))
+We can also color the items in the end node profiles according to a split in the tree.
+Here, we see that for Node 1 (split on the covariate Area), item 2 shows DIF in category B.
+Item 2 seems to be easier (y-axis shows item easiness) in certain areas compared to Language, Culture, Medicine, Health, Arts.
+
+<<fig = TRUE, width = 12, height = 7>>=
+plot(Raschtree_MH_culture, color_by_node = 1)
 @
 
-The resulting tree is presented in Figure~\ref{fig:math_tree}.
-The splitting nodes show the $p$-value of the corresponding parameter stability tests.
-The terminal depict the node-specific item difficulties.
-The difficulties of items 1 (C7MTHBST: Math is one of my best subjects) and 3 (C7LIKMTH: I like math) appear relatively stable and show average difficulty in every subgroup.
-The difficulties of items 2 (C7MTHGD: I get good grades in math) and 4 (C7ENJMTH: I enjoy doing work in math) are always lower and higher, respectively, and also seem to vary more strongly.
+We can also color by Node 3 (also split on covariate Area). Also here, Item 2 shows DIF. Similarly, item 2 seems to be more difficult (less easy) for respondents who study Arts.
 
-\begin{figure}%
-\caption{Rasch tree for the four math items.}
-\begin{subfigure}{1.25\textwidth}
-<<fig=TRUE, echo=FALSE, width= 9, height=6, eval=FALSE>>=
-library("psychotree") ## needed for plotting, but interferes with stopfun_mantelhaenszel
-plot(math_tree, gp = gpar(cex = .7))
-detach("package:psychotree", unload = TRUE)
-@
-\end{subfigure}
-\label{fig:math_tree}
-\end{figure}%
-
-We repeat the procedure for the reading items:
-
-<<eval=FALSE>>=
-read_items <-  c("C7ENGBST", "C7GRDENG", "C7LIKRD", "C7ENJRD")
-mydata$resp <- sapply(SDQ_dat[ , read_items], function(x) ifelse(x > 2, 1, 0))
-read_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
+<<fig = TRUE, width = 12, height = 7>>=
+plot(Raschtree_MH_culture, color_by_node = 3)
 @
 
-<<eval=FALSE>>=
-plot(read_tree, gp = gpar(cex = .7))
-@
+This is interesting, because it is the culture scale.
+Why should people who study Arts have more problems to solve the item?
+When we look at the item, it probably makes sense: the items asks what four buildings have in common (Allianz-Arena München, Tate Modern NY, Olympia-Stadion Peking, Elbphilharmonie Hamburg).
+Two of these buildings are sports stadiums.
+So probably respondents who are interested in sports, and know these buildings well have an advantage on the item (independent of the level of knowledge about culture).
 
-The resulting tree is presented in Figure~\ref{fig:read_tree}, which reveals a pattern of difficulties and splits quite similar to Figure~\ref{fig:math_tree}.
-
-\begin{figure}%
-\caption{Rasch tree for the four reading items.}
-\begin{subfigure}{1.25\textwidth}
-<<fig=TRUE, width=9, height=6, echo=FALSE, eval=FALSE>>=
-library("psychotree") ## needed for plotting, interferes with stopfun_mantelhaenszel
-plot(read_tree, gp = gpar(cex = .7))
-detach("package:psychotree", unload = TRUE)
-@
-\end{subfigure}
-\label{fig:read_tree}
-\end{figure}
+\todo[inline]{Maybe a small note on anchoring?
+Item difficulty profiles are anchoring on the items that do not show DIF.
+This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
 

### 2023-10-04 example2_backup.Rnw
New example for Rasch tree and backup of old example

--- 2023-09-10 MOB_paper.Rnw
@@
-\documentclass[doc,floatsintext,natbib]{apa7}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{doi}
-\usepackage{booktabs}
-%\usepackage[dvipsnames]{xcolor}
-\usepackage{placeins}
-\usepackage{enumitem}
-\usepackage{xcolor}
-\usepackage{setspace} % for adjusting line spread
-\usepackage{hyperref}
-
-\renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
-\renewcommand{\thefootnote}{\fnsymbol{footnote}} % uses symbols instead of numbers for footnotes
-\usepackage[colorinlistoftodos, textsize=footnotesize]{todonotes}
-%%\todo[inline, color=green!40]{This is a green inline comment.}
-
-\linespread{1.3}
-
-\title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
-\shorttitle{shorttitle}
-\authorsnames{Marjolein Fokkema$^1$ \& Carolin Strobl$^2$}
-\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}
-
-\abstract{Model-based recursive partitioning \cite{ZeilyHoth08} is a flexible framework for detecting subgroups with differential effects in a wide range of parametric models.
-It provides a versatile tool for detecting and explaining heterogeneity of intervention effects.
-In this tutorial paper, we provide an introduction to the general MOB framework.
-In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item-response theory models.
-In the first case study, we show how GLMM trees \cite{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
-We apply GLMM trees to a dataset from a study of \cite{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
-Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start.
-In a second case study, we show how Rasch trees \cite{StroyKopf15} can be used to detect subgroups with different parameters in IRT models.
-We show how a recently developed stopping criterion \cite{HennyDeba23} can be used to guide subgroup detection based on effect sizes instead of statistical significance.\\}
-
-\begin{document}
-\SweaveOpts{concordance=TRUE}
-\renewenvironment{Schunk}{\small}{}
-
-<<echo=FALSE>>=
-## suppress printing of > and + before code, it is annoying for copy-pasting
-options(prompt=" ", continue = " ") 
-@
-
-\maketitle
-
-\todo[inline]{Target journal: \textit{Journal of School Psychology}.
-This Special Issue is on "Conceptual and Methodological Advances for Understanding Contextual, Identity, and Cultural Effects in Intervention Research".
-Submission deadline: October 31, 2023}
-
-\newpage
-\section{Introduction}
-\label{sec:Introduction}
-
-\todo[inline]{Introduce and explain MOB.}
-
-\todo[inline]{\textbf{Previous Papers on IRT and Rasch Models in Journal of School Psychology}
-
-Rasch and IRT models are often mentioned in this journal's papers as being used for scoring school assessments (e.g., Virginia Standards of Learning Assessments;  Oregon Assessment of Knowledge and Skills; Peabody individual achievement test, \cite{Luth92}).
-
-A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}.
-There are also papers describing extensions to mixure IRT models \cite{AyalySant17} and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.
-
-\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).}
-
-\newpage
-\section{Tutorial: Subgroup Detection in Mixed-Effects Models}
-\label{sec:TutorialMixed}
-
-<<data_prep_mixed, echo=FALSE>>=
-## Load data
-data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
-
-## Keep only eligible children in HS or no preschool
-data <- data[data$Elig2_90 == 1, ]
-data <- data[data$None2_90 == 1 | data$HS2_90 == 1, ]
-data <- data[data$MotherID %in% names(table(data$MotherID) >= 2), ]
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")))
-
-## Keep only families in which children differ in HS participation
-selected_family <- rowSums(table(data$MotherID, data$program) > 0) == 2
-selected_family <- names(selected_family)[selected_family]
-data <- data[data$MotherID %in% selected_family, ]
-
-## Node-specific predictor
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")),
-                       levels = c("None", "HS"))
-
-## Partitioning variables
-
-## ageAFTQ: age-adjusted AFQT
-## Race_child: Hispanic=Race_Child==1; Black=Race_Child==2; White=Race_Child==3
-data$Race_Child <- factor(ifelse(data$Race_Child == 1, "Hispanic", 
-                          ifelse(data$Race_Child == 2, "Black", "White")))
-## family income: PermInc
-## Mom's years of education: MothED
-part_vars <- c("AgeAFQT", "Race_Child", "PermInc", "Height_Mom81", "MothED")
-id_vars <- c("ChildID", "MotherID")
-
-## Outcome: Peabody Picture Vocabulary Test (PPVT)
-testscore_vars <- c(paste0("PPVT_Raw", 2*(43:52)), paste0("AgeTest_Yr", 2*(43:52)))
-
-## Create long dataset
-pred_vars <- "program"
-HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
-                    varying = list(paste0("PPVT_Raw", 2*(43:52)),
-                                   paste0("AgeTest_Yr", 2*(43:52))),
-                    direction = "long", 
-                    v.names = c("PPVT_Raw", "AgeTest_Yr"))
-HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
-names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
-                     "ChildID", "MotherID", "Program", "PPVT", "Age")
-HS_dat <- HS_dat[complete.cases(HS_dat), ]
-HS_dat$Age <- sqrt(HS_dat$Age)
-rownames(HS_dat) <- NULL
-saveRDS(HS_dat, file = "HS_dat.Rda")
-@
-
-\subsection{Dataset}
-
-To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start.
-Head Start is a federally funded nationwide pre-school program for children from low-income families.
-\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).
-
-The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
-Data from children in those families who participated in another preschool program were excluded.
-As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start.
-As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time.
-As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.
-
-Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height.
-Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables.
-The dataset comprises data from families and children for whom complete data was available.
-
-We load the data and inspect the first rows:
-
-<<>>=
-HS_dat <- readRDS("HS_dat.Rda")
-head(HS_dat, 3)
-@
-
-We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red).
-To show the effect of age and Head Start participation, we fitted a mixed-effects model comprising their main and interaction effects.
-To account for the correlation between repeated assessments on the same child, we specified a random intercept.
-The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference persists over time.
-This result agrees with the findings of \cite{Demi09}.
-
-\begin{figure}%
-\caption{}
-\begin{subfigure}{.7\textwidth}
-<<echo=FALSE, fig=TRUE>>=
-library("lme4")
-lmm <- lmer(PPVT ~ Program*Age + (1|ChildID), data = HS_dat)
-beta <- fixef(lmm)
-plot(jitter(HS_dat$Age), HS_dat$PPVT, col = HS_dat$Program,
-     cex = .5, #cex.lab = .7, cex.axis=.7, cex.main = .7,
-     xlab = "Age (years)", ylab = "PPVT score", xaxt = "n",
-     main = paste0("Global Model (", length(unique(HS_dat$ChildID)), " children)"))
-abline(a = beta["(Intercept)"], b = beta["Age"], col = "black")
-abline(a = beta["(Intercept)"]+beta["ProgramHS"], 
-       b = beta["Age"]+beta["ProgramHS:Age"], col = "red")
-axis(1, at=sqrt(c(3, 5, 8, 12, 17)), labels=c(3, 5, 8, 12, 17), cex.axis = .8)
-@
-\end{subfigure}
-\label{fig:global_lmm}
-\end{figure}%
-
-\FloatBarrier
-\subsection{Linear mixed effects model tree}
-
-We now test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:
-
-<<>>=
-library("glmertree")
-HS_tree <- lmertree(PPVT ~ Program*Age | (1|ChildID) | AFTQ + Race + 
-                      Income + Mom_edu_yrs + Mom_height, 
-                    data = HS_dat, cluster = ChildID, minsize = 250)
-@
-
-With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance.
-The middle part (\texttt{1|ChildID}) specified the random effects.
-The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.
-
-With the second argument, we specified the dataset which contain the variables.
-Because we are dealing with repeated measurements on the same children, we additionally specified that the parameter stability tests should be performed on the child level using the \verb|cluster| argument.
-Using the default observation-level parameter stability tests may artificially inflate power.
-Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.
-
-Next, we plot the tree.
-With multiple fixed-effects predictors of interest, the default plots may become too crowded or difficult to interpret.
-We therefore specify \verb|type = "simple"| to facilitate interpretation, and using the \verb|nodesize_level| argument, we specified that the sample size printed above every terminal node should count the number of children, not the number of individual observations:
-
-<<eval=FALSE>>=
-plot(HS_tree, type = "simple", nodesize_level = 2)
-@
-
-\begin{figure}%
-\caption{}
-<<fig=TRUE, echo=FALSE, fig.width=9, fig.height=4>>=
-plot(HS_tree, type = "simple", which = "tree", gp = gpar(cex = .5), nodesize_level = 2)
-@
-
-\vspace*{-3cm}
-
-<<fig=TRUE, echo=FALSE, width=7.6, height=2.5>>=
-nodes <- predict(HS_tree, type = "node")
-beta <- coef(HS_tree)
-par(mfrow = c(1, 4))
-for (node in sort(unique(nodes))) {
-  plot(HS_dat$Age[nodes == node], HS_dat$PPVT[nodes == node],
-       col = HS_dat$Program[nodes == node],
-       cex = .5, cex.lab = .7, cex.axis=.7, cex.main = .7,
-       xlab = "Age", ylab = "PPVT", xlim = c(sqrt(3), sqrt(20)), ylim = c(0, 150))
-  abline(a = beta[as.character(node), "(Intercept)"], b = beta[as.character(node), "Age"], col = "black")
-  abline(a = beta[as.character(node), "(Intercept)"] + beta[as.character(node), "ProgramHS"], 
-         b = beta[as.character(node), "Age"] + beta[as.character(node), "ProgramHS:Age"], col = "red")
-}
-@
-\label{fig:lmm_tree}
-\end{figure}%
-
-The resulting tree is presented in Figure~\ref{fig:lmm_tree}.
-Below each terminal node, we plotted the observations and the two regression curves given by the coefficients in that node.
-The first split was made based on the AFTQ variable, which represents the mother's score on the Armed Forced Qualification Test, adjusted for the age at which they completed the test.
-The group with higher mother's AFTQ scores is further split based on race.
-The Black and Hispanic group is further split based on income.
-
-<<echo=FALSE, results=tex>>=
-coefs <- round(coef(HS_tree), digits = 2)
-design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
-                        Program = factor(rep(c("None", "HS"))))
-design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(6)), 
-                                     re.form = ~0)
-design_df$`PPVT at age 18` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(18)), 
-                                     re.form = ~0)
-names(design_df)[1] <- "Node"
-knitr::kable(design_df, format = "latex", digits = 2, 
-             caption = "Predicted PPVT scores at different ages.", 
-             label = "predictions", centering = FALSE,
-             linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)
-@
-
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
-All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
-This benefit remains the same over time for the group with lower mother's AFTQ scores.
-The benefit increases over time for White children with higher mother's AFTQ scores.
-Strikingly, the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
-These results correspond to the conclusions of \cite{Demi09}.
-
-To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
-We separated the \Sexpr{length(unique(HS_dat$MotherID))} families in the dataset into ten equally-sized folds.
-We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree (i.e., an LMM with subgroups) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child).
-We evaluated performance on the remaining folds, by computing and evaluating accuracy of the predictions.
-\todo{explain MSE and $R^2$} We repeated this procedure ten times, so that all folds were used as a test set once.
-The results are presented in Table~\ref{tab:performance}, which shows that LMM trees generalize well: They provide better predictive accuracy compared to LMMs, while implementing only few splits.
-
-<<cv_lmm_tree, echo=FALSE, results=tex>>=
-results <- data.frame(MSE_tree = rep(NA, times = 10), 
-                      MSE_lmm = rep(NA, times = 10),
-                      tree_size = rep(NA, times = 10))
-
-set.seed(42)
-folds <- sample(1:10, replace = TRUE, size = length(unique(HS_dat$MotherID)))
-HS_dat$fold <- NA
-for (mom in unique(HS_dat$MotherID)) {
-  HS_dat$fold[HS_dat$MotherID == mom] <- folds[unique(HS_dat$MotherID) == mom]
-}
-for (i in 1:10) {
-  ## Fit models on training data
-  lmmt <- lmertree(PPVT ~ Age*Program | ChildID | AFTQ + Race + Income + 
-                     Mom_height + Mom_edu_yrs, 
-                   cluster = ChildID, data = HS_dat[HS_dat$fold != i, ], 
-                   minsize = 225)
-  lmm <- lmer(PPVT ~ Age*Program + (1|ChildID), 
-              data = HS_dat[HS_dat$fold != i, ])
-
-  ## Evaluate performance on test data
-  results$tree_size[i] <- (length(lmmt$tree)-1)/2
-  preds <- predict(lmmt, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
-  results$MSE_tree[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
-  preds <- predict(lmm, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
-  results$MSE_lmm[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
-}
-res_df <- data.frame(method = c("LMM tree", "LMM"),
-                     MSE = colMeans(results[1:2]),
-                     SD = sapply(results, sd)[1:2],
-                     `number of splits` = c(colMeans(results)[3], NA),
-                     R2 = colMeans(1 - results[ , 1:2] / var(HS_dat$PPVT)))
-knitr::kable(res_df, format = "latex", digits = 3, 
-             caption = "Cross-validated performance of LMMs and LMM trees.", 
-             label = "performance", centering = FALSE,
-             linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("ccccc"), booktabs = TRUE)
-@
-
-\todo[inline]{I am not sure this dataset is the best GLMM tree illustration, because visualized differences between subgroups are very small and perhaps the fixed-effect part is already too complex.
-Alternative: Do not focus on effect of Head Start, but use as partitioning variable.
-Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.}
-
-\newpage
-\FloatBarrier
-\section{Tutorial: Subgroup Detection in Rasch Models}
-\label{sec:TutorialRasch}
-
 Using Rasch trees, we assess DIF for items on math and reading \todo{decide on subscale(s) to include} from the Self-Description Questionnaire \cite{Boyl94}.
@@
 \end{figure}
-
-\FloatBarrier
-\section{Discussion}
-
-\todo[inline]{Write summary of findings.}
-
-\todo[inline]{Mention shortcomings.}
-
-\todo[inline]{Write about future work.}
-
-\bibliography{MOB_paper}
-
-\todo{Add DOIs.}
-
-\newpage
-\appendix
-
-\section{Appendix A: SDQ-II items administered in the ECLS}
-\label{sec:AppendixA}
-
-\todo{Fix formatting.}
-
-How true is each of these about you? (1="not at all true"; 2="a little bit true"; 3="mostly true" or 4="very true")
-
-\begin{itemize}
-
-  \item C7MTHBST (math): Math is one of my best subjects.
-
-  \item C7ANGRY (internalizing): I feel angry when I have trouble learning.
-
-  \item C7LIKRD (reading): I like reading.
-
-  \item C7WRYTST (internalizing): I worry about taking tests.
-
-  \item C7MTHGD (math): I get good grades in math.
-
-  \item C7LONLY (internalizing): I often feel lonely.
-
-  \item C7ENGBST (reading): English is one of my best subjects.
-
-  \item C7SAD (internalizing): I feel sad a lot of the time.
-
-  \item C7LIKMTH (math): I like math.
-
-  \item C7WRYWEL (internalizing): I worry about doing well in school.
-
-  \item C7ENJRD (reading): I enjoy doing work in reading.
-
-  \item C7WRYFIN (internalizing): I worry about finishing my work.
-
-  \item C7ENJMTH (math): I enjoy doing work in math.
-
-  \item C7WRYHNG (internalizing): I worry about having someone to hang out with at school.
-
-  \item C7GRDENG (reading): I get good grades in English.
-
-  \item C7ASHAME (internalizing): I feel ashamed when I make mistakes at school.
-
-\end{itemize}
-
-\end{document}

### 2023-10-10 _MOB_paper.Rnw
misc

@@
 \shorttitle{shorttitle}
-\author{Marjolein Fokkema, Mirka Henninger \& Carolin Strobl}
 
-%\authorsnames{Marjolein Fokkema$^1$ \& Carolin Strobl$^2$}
-%\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}
+\authorsnames{Marjolein Fokkema$^1$, Mirka Henninger$^2$ \& Carolin Strobl$^2$}
+\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}
 
@@
 In this tutorial paper, we provide an introduction to the general MOB framework.
-In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item-response theory models.
+In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models.
 In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
@@
 \maketitle
-\todo[inline]{Marjolein: what document style for authorsnames? commented out}
 
@@
 MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model.
-For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustration in application example 2.
-We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables go into the parametric mode.
+For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustrated in application example 2.
+We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model.
 In the case of the Rasch model, this will be the test items.
@@
 
-If one joint Rasch model holds for the entire sample, i.e., if there is no DIF a Rasch tree should show no splits.
+If one joint Rasch model holds for the entire sample, i.e., if there is no DIF, a Rasch tree should show no splits.
 However, in certain settings in educational research, such as large scale assessements, very large sample sizes are available for testing for DIF.
@@
 However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
-As we will illustrate in the first part of application example 2, for a large sample of university students \todo{are only students in the sample?
-(Mirka: I can take a subsample in which there are only students)} who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
 Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
@@
 In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
-\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.\todo{Mirka: is it A vs.
-B,C or A,B vs.
-C by default?; Response Mirka: A vs.
-BC is the default, but it can be changed}
+\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
 
@@
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
-For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words used in the different languages are equally frequently used, etc.
+\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
+For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc.
 In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test.
@@
 This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
-Students whose native language is not the same as the test language can have disadvantage in answering the story problems, for example when they contain seldomly used words.
+Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
 These items will then show DIF between native and non-native speakers.
@@
 
+\todo[inline]{not clear what this information is necessary/useful for? since it is not displayed, leave this out?}
+
 If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.

### 2023-10-10 _MOB_paper.tex
misc

@@
 \item If there is significant structural change, groups are split along the covariate with the strongest change and using the optimal cutpoint.
-  \item Steps 1--3 are repeated recursively in the resulting groups
-        until there is no more significant structural change (or the
-   groups becomes too small).
+\item Steps 1--3 are repeated recursively in the resulting groups until there is no more significant structural change (or the groups becomes too small).
 \end{enumerate}
@@
 MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model.
-For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustration in application example 2.
-We will see that, just like for the mixed-effects model trees \todo{use same names for methods everywhere}, we need to specify which variables go into the parametric mode.
+For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustrated in application example 2.
+We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model.
 In the case of the Rasch model, this will be the test items.
@@
 However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
-As we will illustrate in the first part of application example 2, for a large sample of university students \todo{are only students in the sample?} who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
 Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
@@
 In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
-\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.\todo{Mirka: is it A vs.
-B,C or A,B vs.
-C by default?}
+\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
 
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citet[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+Together with a purification step \citep[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
-For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words used in the different languages are equally frequently used, etc.
+\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
+For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc.
 In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test.
@@
 Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
-This would be the case in test inteded to measure math aptitude containing pure algebra problems as well as story problems.
-Students whose native language is not the same as the test language can have disadvantage in answering the story problems, for example when they contain seldomly used words.
+This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
+Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
 These items will then show DIF between native and non-native speakers.
@@
 
-Using Rasch trees, we assess DIF for items on math and reading \todo{decide on subscale(s) to include} from the Self-Description Questionnaire \cite{Boyl94}.
-We use a dataset from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}.
-Data were collected from 1,018 schools across the USA.
-Assessments took place from kindergarten through 8th grade.
-Here we focus on the 8th grade assessment, in which four SDQ-II items were administered to assess perceived interest and competence in reading, and another four to assess perceived interest and competence in math.
-Items are presented in Appendix~\ref{sec:AppendixA}.
-The items were rated by the children on a 4-point scale (“not at all true” through “very true”).
-We coded responses 1 and 2 as 0, and responses 3 and 4 as 1, to allow for fitting a Rasch model, which assumes binary responses.
+\begin{itemize}
+\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet)
+\item take the subsample of students (only students), and a random subsample of 5000 respondents (seed 04102023)
+\item items could probably appear in Appendix??
+\begin{itemize}
+\item Which painter created this painting? – Andy Warhol.
+\item What do these four buildings have in common? – All four were designed by the same architects.
+\item Roman numbers: What is the meaning of CLVI? – 156.
+\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
+\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
+\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
+\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
+\item In which city is this building located? – Paris.
+\item Which one of the following operas is not by Mozart? – Aida.
+\end{itemize}
+\end{itemize}
 
-We used three covariates as possible partitioning variables: Gender (1=Male; 2=Female), race (8 categories: 1=White, non-Hispanic; 2=Black or African-American, non-Hispanic; 3=Hispanic, race specified; 4=Hispanic, race not specified; 5=Asian; 6=Native Hawaiian or other Pasific Islander; 7= American Indian or Alaska native; 7=More than one race) and socio-economic status (range $-$5 to 3).
-\todo{Add more partitioning variables?
-This does increase number of splits.
-But could add one that should not have an effect, e.g., month of birth.} We analyze observations of children with complete data only, yielding a total sample size of 7,417.
+\todo[inline]{at Caro: Ich habe die Daten (glaube ich) mal von dir bekommen (\texttt{spisa\_ges.RData}) und die Item-Formulierungen habe ich aktuell aus dem psychotree Paket.
+Stimmt das so?
+Ich glaube, der Originaltest hatte mehrere parallele Skalen, zumindest sind in dem PDF/BUCH von Trepte sehr viel mehr, und sehr ähnliche Items enthalten.
+Das könnten wir vielleicht nochmal nachgucken.}
 
-We load the data and inspect the first three rows as follows:
+We load the data and inspect the first rows.
+there are several covariates: age (continuous), gender (male, female, missing), area (Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, Sports) and 9 items from the culture scale (true/false 0/1).
 
@@
 \begin{Sinput}
- SDQ_dat <- readRDS("ECLSK_SDQ.Rda")
- head(SDQ_dat, 3)
+ head(covar_data)
 \end{Sinput}
 \begin{Soutput}
-   C7MTHBST C7ANGRY C7LIKRD C7WRYTST C7MTHGD C7LONLY C7ENGBST C7SAD C7LIKMTH
-2         4       4       4        2       4       2        4     3        4
-10        4       1       4        2       4       1        4     1        4
-16        3       3       1        2       3       1        4     1        4
-   C7WRYWEL C7ENJRD C7WRYFIN C7ENJMTH C7WRYHNG C7GRDENG C7ASHAME GENDER RACE
-2         4       4        4        3        2        3        4      2    1
-10        1       3        1        4        1        4        1      2    1
-16        2       1        2        4        1        4        1      1    1
-   WKSESL T6INTERN T6EXTERN T6INTERP T6CONTRO
-2    1.56     1.50     1.00      4.0     3.75
-10   1.41     1.25     1.50      4.0     4.00
-16  -2.93     1.50     2.33      2.8     3.00
+       Age Gender                Area
+445472  28   Male  Language & Culture
+167408  26 Female         Engineering
+422161  25   Male  Language & Culture
+350734  27   Male         Engineering
+570206  23   Male \nMedicine & Health
+265570  21 Female                Arts
+\end{Soutput}
+\begin{Sinput}
+ head(culture_scale)
+\end{Sinput}
+\begin{Soutput}
+     i27 i28 i29 i30 i31 i32 i33 i34 i35
+[1,]   1   1   0   0   1   1   1   1   1
+[2,]   1   0   1   0   0   0   1   1   1
+[3,]   1   1   1   0   1   1   1   1   1
+[4,]   0   1   0   1   0   1   1   0   1
+[5,]   1   1   0   1   1   1   0   1   0
+[6,]   1   1   0   1   1   0   1   1   1
 \end{Soutput}
@@
 
-To fit Rasch trees, we use function \verb|raschtree| which is available from R package \textbf{psychotree}.
-However, the stopping criterion based on the Mantel–Haenszel effect size measure is implemented in R package \verb|raschtreeMH|, which we therefore use here.
-\todo{might be nice to implement the MH effect-size criterion in the psychotree package for this paper, currently it seems to interfere with plotting the tree}.
-It can be installed and loaded as follows:
+We create a dataset \texttt{dat\_SPISA} that contains the data of the covariate and the item responses from the culture scale.
+By assigning the whole scale to the dataset using \texttt{\$}, we can later access item responses from all items using \texttt{\$culture}.
+\begin{Schunk}
+\begin{Sinput}
+ dat_SPISA <- covar_data
+ dat_SPISA$culture <- culture_scale
+\end{Sinput}
+\end{Schunk}
+
+The Rasch tree can be fitted with the function \texttt{raschtree} from the R package \texttt{psychotree}.
+We use the typical formula syntax with the item responses of the culture scale on the left hand side of $\sim$, and the covariates on the right hand side.
+
+\begin{Schunk}
+\begin{Sinput}
+ library(psychotree)
+ Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
+                                data = dat_SPISA)
+\end{Sinput}
+\end{Schunk}
+
+We can plot the Raschtree using the plot function.
+But it is very big, lots of splits, we don't know whether the splits are due to substantial differences in item difficulty parameters, because the item difficulty profiles in the end nodes are hard to interpret/compare.
+
+\begin{Schunk}
+\begin{Sinput}
+ plot(Raschtree_culture)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{_MOB_paper-016}
+
+Mantel-Haenszel trees use the Mantel-Haenszel effect size measure for DIF.
+it has three categories (A: negligible, B: moderate, C: large).
+If none of the items has DIF in category B or C (default, but can also be changed so that DIF has to be C), the tree is stopped from growing.
+
+The software must be installed from github, but will (probably) be implemented in the psychotree package some time in the future
 
@@
  devtools::install_github("mirka-henninger/raschtreeMH")
- library("raschtreeMH")
+ library(raschtreeMH)
 \end{Sinput}
@@
 
-To fit a Rasch tree, we first create a \verb|data.frame| that contains the item responses and possible partitioning variables:
+Syntax is very similar to the classical Rasch tree, but an additional argument \texttt{stopfun}, where the mantelhaneszel stopping function can be selected (but also own stopping functions can be provided; see Henninger et al., 2023).
+The user also has to indicate what kind of purification strategy should be used (none, 2step, iterative).
+Iterative is recommended.
 
@@
 \begin{Sinput}
- math_items <- c("C7MTHBST", "C7MTHGD", "C7LIKMTH", "C7ENJMTH")
- part_vars <- c("GENDER", "RACE", "WKSESL")
- mydata <- SDQ_dat[ , part_vars]
- mydata$resp <- sapply(SDQ_dat[ , math_items], function(x) ifelse(x > 2, 1, 0)) 
+ Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area, 
+                                   data = dat_SPISA, 
+                                   stopfun= stopfun_mantelhaenszel(
+                                     purification = "iterative"))
 \end{Sinput}
@@
 
-Next, we apply function \verb|raschtree| and plot the result:
+For technical reasons, the information about the effect size and ETS classification are not saved in the tree object itself but have to be added afterwards using the \texttt{add\_mantelhaenszel} function.
+The information can then be accessed from the object using \texttt{\$info\$mantelhaenszel}
 
@@
 \begin{Sinput}
- stop_fun <- stopfun_mantelhaenszel(purification = "iterative", stopcrit = "C")
- math_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
+ Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, 
+                                            purification = "iterative")
+ Raschtree_MH_culture$info$mantelhaenszel
 \end{Sinput}
@@
 
-\begin{Schunk}
-\begin{Sinput}
- plot(read_tree, gp = gpar(cex = .7))
-\end{Sinput}
-\end{Schunk}
+\todo[inline]{not clear what this information is necessary/useful for? since it is not displayed, leave this out?}
 
-The resulting tree is presented in Figure~\ref{fig:math_tree}.
-The splitting nodes show the $p$-value of the corresponding parameter stability tests.
-The terminal depict the node-specific item difficulties.
-The difficulties of items 1 (C7MTHBST: Math is one of my best subjects) and 3 (C7LIKMTH: I like math) appear relatively stable and show average difficulty in every subgroup.
-The difficulties of items 2 (C7MTHGD: I get good grades in math) and 4 (C7ENJMTH: I enjoy doing work in math) are always lower and higher, respectively, and also seem to vary more strongly.
-
-\begin{figure}%
-\caption{Rasch tree for the four math items.}
-\begin{subfigure}{1.25\textwidth}
-\end{subfigure}
-\label{fig:math_tree}
-\end{figure}%
-
-We repeat the procedure for the reading items:
+If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.
+In addition, we see the number of items classified as A, B, or C in each node.
+For instance, we see that one item is classified as B in Node 1, and no item as C.
 
@@
 \begin{Sinput}
- read_items <-  c("C7ENGBST", "C7GRDENG", "C7LIKRD", "C7ENJRD")
- mydata$resp <- sapply(SDQ_dat[ , read_items], function(x) ifelse(x > 2, 1, 0))
- read_tree <- raschtree(resp ~ ., data = mydata, stopfun = stop_fun)
+ plot(Raschtree_MH_culture)
 \end{Sinput}
 \end{Schunk}
+\includegraphics{_MOB_paper-021}
+
+We can also color the items in the end node profiles according to a split in the tree.
+Here, we see that for Node 1 (split on the covariate Area), item 2 shows DIF in category B.
+Item 2 seems to be easier (y-axis shows item easiness) in certain areas compared to Language, Culture, Medicine, Health, Arts.
 
@@
 \begin{Sinput}
- plot(read_tree, gp = gpar(cex = .7))
+ plot(Raschtree_MH_culture, color_by_node = 1)
 \end{Sinput}
 \end{Schunk}
+\includegraphics{_MOB_paper-022}
 
-The resulting tree is presented in Figure~\ref{fig:read_tree}, which reveals a pattern of difficulties and splits quite similar to Figure~\ref{fig:math_tree}.
+We can also color by Node 3 (also split on covariate Area). Also here, Item 2 shows DIF. Similarly, item 2 seems to be more difficult (less easy) for respondents who study Arts.
 
-\begin{figure}%
-\caption{Rasch tree for the four reading items.}
-\begin{subfigure}{1.25\textwidth}
-\end{subfigure}
-\label{fig:read_tree}
-\end{figure}
+\begin{Schunk}
+\begin{Sinput}
+ plot(Raschtree_MH_culture, color_by_node = 3)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{_MOB_paper-023}
+
+This is interesting, because it is the culture scale.
+Why should people who study Arts have more problems to solve the item?
+When we look at the item, it probably makes sense: the items asks what four buildings have in common (Allianz-Arena München, Tate Modern NY, Olympia-Stadion Peking, Elbphilharmonie Hamburg).
+Two of these buildings are sports stadiums.
+So probably respondents who are interested in sports, and know these buildings well have an advantage on the item (independent of the level of knowledge about culture).
+
+\todo[inline]{Maybe a small note on anchoring?
+Item difficulty profiles are anchoring on the items that do not show DIF.
+This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
 

### 2023-10-10 _MOB_paper.Rnw
todos from meetin on okt 10

@@
 In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
-We apply GLMM trees to a dataset from a study of \citep{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
+We apply GLMM trees to a dataset from a study of \citep{Demi09}\todo{take that out?}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
 Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start.
@@
 
-What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?)}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
+What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?); Marjolein: please read and decide whether we do need a picture of a regression tree and/or mob tree here, if yes, simulate but with abstract x1 x2, no story}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
 These figures share the tree-structure in the upper part, but differ in their lower part.
@@
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citep[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+Together with a purification step \citep[see][for details]{HennyDeba23} \todo{Caro can add more text on achoring, incl.
+word anchoring, here to prepare example 2 - only shortly mention here and explain in more detail later in the example (will become more clear in example)}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
@@
 
+\todo[inline]{Marjolein add here comment that typically affect of age is typically strongest; take out next part, but comment on it in discussion: CV would be a technique to see how stable results are, cite something about CV in general}
+
 To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
@@
 
-\todo[inline]{Write summary of findings.}
+\todo[inline]{Write summary of findings. -- now think: rather not discuss substantive findings, only properties of the methods}
 
@@
 \todo[inline]{Mention shortcomings.}
-
-not possible to detect certain patterns: XOR problem, known for all trees
 
@@
 sample size dependence, similar ideas like MH trees also possible for other MOB?
+\todo[inline]{possibly Mirka add something about visual interpretation?
+because effect sizes in mixed effects models are not very clear}
 

### 2023-10-11 Raschtree_MH.Rnw
Take Rasch Chapter out to work on it

--- 2023-10-10 _MOB_paper.Rnw
@@
-\documentclass[doc,floatsintext,natbib]{apa7}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{doi}
-\usepackage{booktabs}
-%\usepackage[dvipsnames]{xcolor}
-\usepackage{placeins}
-\usepackage{enumitem}
-\usepackage{xcolor}
-\usepackage{setspace} % for adjusting line spread
-\usepackage{hyperref}
-
-\renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
-\renewcommand{\thefootnote}{\fnsymbol{footnote}} % uses symbols instead of numbers for footnotes
-\usepackage[colorinlistoftodos, textsize=footnotesize]{todonotes}
-%%\todo[inline, color=green!40]{This is a green inline comment.}
-
-%text from Marjolein edited by Caro
-\newcommand{\edc}[1]{\textcolor{blue}{#1}}
-
-\linespread{1.3}
-
-\title{One Model May Not Fit All: Subgroup Detection Using MOB}
-\shorttitle{shorttitle}
-
-\authorsnames{Marjolein Fokkema$^1$, Mirka Henninger$^2$ \& Carolin Strobl$^2$}
-\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}
-
-\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models.
-It provides a versatile tool for detecting and explaining heterogeneity of intervention effects.
-In this tutorial paper, we provide an introduction to the general MOB framework.
-In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models.
-In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
-We apply GLMM trees to a dataset from a study of \citep{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
-Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start.
-In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
-DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}
-
-\begin{document}
-\SweaveOpts{concordance=TRUE}
-\renewenvironment{Schunk}{\small}{}
-
-<<echo=FALSE>>=
-## suppress printing of > and + before code, it is annoying for copy-pasting
-options(prompt=" ", continue = " ") 
-@
-
-\maketitle
-
-\todo[inline]{Target journal: \textit{Journal of School Psychology}.
-This Special Issue is on "Conceptual and Methodological Advances for Understanding Contextual, Identity, and Cultural Effects in Intervention Research".
-Submission deadline: October 31, 2023}
-
-\newpage
-\section{Introduction}
-\label{sec:Introduction}
-
-Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical
-model between groups of persons. This method is a generalization of the principle of recursive partitioning, that is also used in classification and regression trees \citep{Breetal:1984}. 
-In classification and regression trees, the aim is to detect groups of persons, specified by (combinations of) covariates, that differ in a response variable.
-An example could be that children who are highly motivated and have good reading skills show higher average math exam grades.
-MOB generalizes this idea by identifying groups of persons, specified by (combinations of) covariates, that differ in a the parameters of a statistical model.
-An example could be that children who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.
-
-The framework of MOB is very flexible and can be applied in combination with a variety of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS}.
-In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and for models from Rasch measurement and Item Response Theory \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
-Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are again nested in schools) participate in a study.
-In application example 1 we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
-
-While for this first application example we assume that the administered test \todo{should we call a psychological test a test/an instrument/???
-to distinguish it from the verb and noun for statistical test?
-(Mirka: I like the term instrument to distinguish it from a statistical test)} has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample.
-Before using the test in our research, we need to make sure that test results are comparable, for example between children of different genders.
-If certain items pose an advantage or disadvantage to either group, they are said to violate measurement invariance or exhibit differential item functioning (DIF).
-In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, groups of persons are compared with respect to the item parameters for their specific group.
-This can be done in a way that allows to detect DIF on top of any true group differences in ability.
-While many standard approaches for testing DIF are limited to comparing typically two pre-specified groups at a time, we will show in application example 2 that MOB more flexibly allows to detect groups of persons with different item parameters in a data-driven
-way.
-
-In the following section, we will give a short introduction into the algorithm and statistical concepts behind MOB.
-Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}.
-An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
-Mixture modelling aims at identifying latent classes of persons with different properties.
-It can also be combined with covariates.
-A comparison of both approches is given by \citet{FriStrZei:2014}.
-
-\section{The Algorithm Behind MOB}
-
-What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?)}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
-These figures share the tree-structure in the upper part, but differ in their lower part.
-The starting point of the tree structure is at the very top of each tree.
-The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the
-bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.
-
-The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
-For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \texttt{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups.
-The algorithm has identified this cutpoint as being the location of the strongest parameter difference between the two resulting groups.
-The categorical covariate \texttt{Race}, which was coded in three categories in this example, has been divided into two groups of categories: Black and Hispanic vs.~White.
-For a binary covariate, \todo{like ...
--- add name if is there one in the other example?}
-on the other hand, there is only one possible cutpoint, namely between the two categories.
-While it is also possible to create more than two groups in each split \citep{KimLoh:2001,Qui:1993}, binary splitting algorithms are typically preferred because they lead to more concise trees.
-
-The parametric model to which the MOB algorithm is applied is visualized in the end nodes at the bottom of each tree.
-While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items.
-The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
-Figure~\ref{fig:lmm_tree} also shows that, out of the  covariates that were presented to the algorithm (\texttt{AFTQ}, \texttt{Race}, \texttt{Income},\texttt{Mom\_edu\_yrs}, and \texttt{Mom\_height}), only \texttt{AFTQ}, \texttt{Race}, and \texttt{Income} were actually selected for splitting.
-Variables are selected based on a statistical test for parameter differences, also termed a test for structural change.
-Structural change is present if, for example, the model parameters systematically differ for children from lower vs.~higher income families.
-At each node, the algorithm will select the covariate showing the strongest structural change as the next splitting variable.
-Within this variables, the optimal cutpoint is chosen in a separate step.
-
-The way the variable selection and cutpoint selection steps are separated in modern classification and regression tree and MOB algorithms, including the ones used here, distinguish them from the traditional classification and regression tree algorithms of \citet{Breetal:1984} and \citet{Qui:1993}.
-The traditional algorithms performed variable and cutpoint selection in one step.
-However, this leads to an undesirable behavior called variable selection bias.
-It means that the traditional algorithms prefer variables offering more cutpoints in the selection process -- regardless of their true information content.
-An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}.
-Modern algorithms for classification and regression trees that have solved this problem are the unbiased approaches \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, available in R in the \texttt{party} and \texttt{partykit} packages \citep{partykit:pkg}.
-The latter forms the basis for all MOB approaches presented in this paper.
-
-Once there are no more covariates that show a significant structural change in any node, the MOB algorithm stops splitting.
-In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection.
-Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected.
-This is the second difference between the modern classification and regression tree and MOB algorithms used here, compared to traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}: While the traditional algorithms grew very large trees and then cut them back by means of so called pruning, the framework employed here is based on statistical significance tests and partly also on effect size measures (see application example 2) to stop the trees already in the growing phase when no more significant change can be detected.
-Other stopping criteria are based on the number of persons in the end nodes.
-These crieria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}
-
-In summary, the rationale of a MOB algorithm consists of the following steps:
-
-\todo{why are the spaces so big? I didn't change the enumerate command}
-\begin{enumerate}
-\item The model parameters are first estimated jointly for all persons in the current node, starting with the full sample.
-\item Structural change in the model parameters is assessed with respect to each available covariate.
-\item If there is significant structural change, groups are split along the covariate with the strongest change and using the optimal cutpoint.
-\item Steps 1--3 are repeated recursively in the resulting groups until there is no more significant structural change (or the groups becomes too small).
-\end{enumerate}
-
-For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.
-
-An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory way, but is learned from the data in an exploratory way.
-This is a key feature of the MOB approach that makes it very flexible and distinguishes it from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
-While there are phases in psychological and educational research where it is very important to specify hypothesis a priori and test them in a confirmatory way, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers.
-
-\todo[inline]{\textbf{Previous Papers on IRT and Rasch Models in Journal of School Psychology}
-
-Rasch and IRT models are often mentioned in this journal's papers as being used for scoring school assessments (e.g., Virginia Standards of Learning Assessments;  Oregon Assessment of Knowledge and Skills; Peabody individual achievement test, \cite{Luth92}).
-
-A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}. \edc{included}
-There are also papers describing extensions to mixure IRT models \cite{AyalySant17} \edc{included}
-and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.
-
-\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).
-\edc{included}
-}
-
-The general framework of MOB introduced above will now be applied to two types of statistical models that are particularly relevant in school psychology research: Mixed-effects models for repeated measures or nested data structures and measurement models for validating psychological and educational tests.
-
-\subsection{Using MOB for Subgroup Detection in Mixed-Effects Models}
-
-In mixed-effects models contain two types of effects: fixed effects and random effects.
-Fixed effects are typically used, for example, to describe the effect of an intervention on the average student performance.
-Random effects are used to describe the variation in the intercept and slope for, e.g., students in different classes.
-
-When mixed-effects models are combined with MOB, ...
-
-\todo[inline]{Marjolein please edit and finish}
-
-\subsection{Using MOB for Detecting DIF in Measurement Models}
-
-Before being able to use a psychological or educational test for comparing different groups of persons in a fair way, different assumptions of the measurement model have to be checked.
-In the framework of IRT and Rasch measurement, test items are typically investigated with respect item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}.
-Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance.
-Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
-
-MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model.
-For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustrated in application example 2.
-We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model.
-In the case of the Rasch model, this will be the test items.
-Moreover, we need to specify, which variables are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.
-
-If one joint Rasch model holds for the entire sample, i.e., if there is no DIF, a Rasch tree should show no splits.
-However, in certain settings in educational research, such as large scale assessements, very large sample sizes are available for testing for DIF.
-Large sample sizes are good for detecting even small effects or model violations with a high statistical power.
-The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected with large samples.
-However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
-As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
-Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
-\citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
-In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
-\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
-
-As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citep[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
-This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
-\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
-For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc.
-In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test.
-Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.
-
-Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
-This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
-Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
-These items will then show DIF between native and non-native speakers.
-When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, e.g., by using more frequently used words, or whether the test should be considered two-dimensional rather than one-dimensional \citep[see also][for a discussion of the connection between DIF and multidimensionality]{Ack:1992,Stretal:2021:APM}.
-
-We will now illustrate how to use the R packages \texttt{glmertree} and \texttt{psychotree} to conduct the MOB analyses.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
-\section{\edc{Application Example 1}: Subgroup Detection in Mixed-Effects Models}
-\label{sec:TutorialMixed}
-
-<<data_prep_mixed, echo=FALSE>>=
-## Load data
-data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
-
-## Keep only eligible children in HS or no preschool
-data <- data[data$Elig2_90 == 1, ]
-data <- data[data$None2_90 == 1 | data$HS2_90 == 1, ]
-data <- data[data$MotherID %in% names(table(data$MotherID) >= 2), ]
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")))
-
-## Keep only families in which children differ in HS participation
-selected_family <- rowSums(table(data$MotherID, data$program) > 0) == 2
-selected_family <- names(selected_family)[selected_family]
-data <- data[data$MotherID %in% selected_family, ]
-
-## Node-specific predictor
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")),
-                       levels = c("None", "HS"))
-
-## Partitioning variables
-
-## ageAFTQ: age-adjusted AFQT
-## Race_child: Hispanic=Race_Child==1; Black=Race_Child==2; White=Race_Child==3
-data$Race_Child <- factor(ifelse(data$Race_Child == 1, "Hispanic", 
-                                 ifelse(data$Race_Child == 2, "Black", "White")))
-## family income: PermInc
-## Mom's years of education: MothED
-part_vars <- c("AgeAFQT", "Race_Child", "PermInc", "Height_Mom81", "MothED")
-id_vars <- c("ChildID", "MotherID")
-
-## Outcome: Peabody Picture Vocabulary Test (PPVT)
-testscore_vars <- c(paste0("PPVT_Raw", 2*(43:52)), paste0("AgeTest_Yr", 2*(43:52)))
-
-## Create long dataset
-pred_vars <- "program"
-HS_dat <- reshape(data[ , c(part_vars, id_vars, testscore_vars, pred_vars)], 
-                  varying = list(paste0("PPVT_Raw", 2*(43:52)),
-                                 paste0("AgeTest_Yr", 2*(43:52))),
-                  direction = "long", 
-                  v.names = c("PPVT_Raw", "AgeTest_Yr"))
-HS_dat <- HS_dat[ , -which(names(HS_dat) %in% c("time", "id"))]
-names(HS_dat) <- c("AFTQ", "Race", "Income", "Mom_height", "Mom_edu_yrs", 
-                   "ChildID", "MotherID", "Program", "PPVT", "Age")
-HS_dat <- HS_dat[complete.cases(HS_dat), ]
-HS_dat$Age <- sqrt(HS_dat$Age)
-rownames(HS_dat) <- NULL
-saveRDS(HS_dat, file = "HS_dat.Rda")
-@
-
-\subsection{Dataset}
-
-To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start.
-Head Start is a federally funded nationwide \todo{what nation?} pre-school program for children from low-income families.
-\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).
-
-The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children, where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
-Data from children in those families who participated in another preschool program were excluded.
-As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start.
-As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time.
-As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.\todo{this would scare me a little as a reader: how would I know what to do/that I should do this?
-(Mirka: Maybe we can add that the sqrt or the log is a typical way to model trajectories that first increase rapidly, but then later increase more slowly?)}
-
-Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height.
-Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables.
-The dataset comprises data from families and children for whom complete data was available.
-
-We load the data and inspect the first rows:
-
-<<>>=
-HS_dat <- readRDS("HS_dat.Rda")
-head(HS_dat, 3)
-@
-
-We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red).
-To show the effect of age and Head Start participation, we fitted a mixed-effects model comprising their main and interaction effects.
-To account for the correlation between repeated assessments on the same child, we specified a random intercept.
-The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference \edc{is slightly diminished but} persists over time.
-This result agrees with the findings of \cite{Demi09}.
-
-\begin{figure}%
-\caption{}
-\begin{subfigure}{.7\textwidth}
-<<echo=FALSE, fig=TRUE>>=
-library("lme4")
-lmm <- lmer(PPVT ~ Program*Age + (1|ChildID), data = HS_dat)
-beta <- fixef(lmm)
-plot(jitter(HS_dat$Age), HS_dat$PPVT, col = HS_dat$Program,
-     cex = .5, #cex.lab = .7, cex.axis=.7, cex.main = .7,
-     xlab = "Age (years)", ylab = "PPVT score", xaxt = "n",
-     main = paste0("Global Model (", length(unique(HS_dat$ChildID)), " children)"))
-abline(a = beta["(Intercept)"], b = beta["Age"], col = "black")
-abline(a = beta["(Intercept)"]+beta["ProgramHS"], 
-       b = beta["Age"]+beta["ProgramHS:Age"], col = "red")
-axis(1, at=sqrt(c(3, 5, 8, 12, 17)), labels=c(3, 5, 8, 12, 17), cex.axis = .8)
-@
-\end{subfigure}
-\label{fig:global_lmm}
-\end{figure}%
-
-\todo{code for figure not displayed on purpose? why only those particular x axis ticks?}
-
-\FloatBarrier
-\subsection{Linear mixed effects model tree}
-
-\todo[inline]{I (Mirka) am confused: the nested structure in this example is timepoints (age) within children?
-Therefore, ChildID is the cluster-variable.
-But then, there should be residual errors between the two children from the same family?
-In the text it says that this is accounted for by the random intercept, but I don't think this is the case (but maybe I am wrong?).
-Is it maybe accounted for by the cluster argument that is also said to ChildID?
-To me, the model specification is not yet clear from the text.
-Plus: If I am right, and the cluster is the child and the timepoints are on Level-1, I think all the covariates are on Level-2 (child level), and then maybe we should comment on the Type-I error problem?}
-We now test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:
-
-<<>>=
-library("glmertree")
-HS_tree <- lmertree(PPVT ~ Program*Age | (1|ChildID) | AFTQ + Race + 
-                      Income + Mom_edu_yrs + Mom_height, 
-                    data = HS_dat, cluster = ChildID, minsize = 250)
-@
-
-With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance.
-The middle part (\texttt{1|ChildID}) specified the random effects\todo{(Mirka) Maybe add information that this is for each child, average DV across timepoints}.
-The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.
-
-\todo[inline]{I (Mirka) am not sure what exactly the cluster argument does.
-Maybe Marjolein you have more information from your work with Achim.
-But my knowledge/intuition is that it has something to do with clustered covariances, so maybe it accounts for the sibling structure, but I don't think it accounts for the level of the predictor variables.}
-With the second argument, we specified the dataset which contain the variables.
-Because we are dealing with repeated measurements on the same children, we additionally specified that the parameter stability tests should be performed on the child level using the \verb|cluster| argument.
-Using the default observation-level parameter stability tests may artificially inflate power\todo{Mirka: I'd rather say it has an inflated type-I error rate.}.
-Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.
-
-Next, we plot the tree.
-With multiple fixed-effects predictors of interest, the default plots may become too crowded or difficult to interpret.
-We therefore specify \verb|type = "simple"| to facilitate interpretation, and using the \verb|nodesize_level| argument, we specified that the sample size printed above every terminal node should count the number of children, not the number of individual observations:
-
-<<eval=FALSE>>=
-plot(HS_tree, type = "simple", nodesize_level = 2)
-@
-
-\begin{figure}%
-\caption{}
-<<fig=TRUE, echo=FALSE, fig.width=9, fig.height=4>>=
-plot(HS_tree, type = "simple", which = "tree", gp = gpar(cex = .5), nodesize_level = 2)
-@
-
-\vspace*{-3cm}
-
-<<fig=TRUE, echo=FALSE, width=7.6, height=2.5>>=
-nodes <- predict(HS_tree, type = "node")
-beta <- coef(HS_tree)
-par(mfrow = c(1, 4))
-for (node in sort(unique(nodes))) {
-  plot(HS_dat$Age[nodes == node], HS_dat$PPVT[nodes == node],
-       col = HS_dat$Program[nodes == node],
-       cex = .5, cex.lab = .7, cex.axis=.7, cex.main = .7,
-       xlab = "Age", ylab = "PPVT", xlim = c(sqrt(3), sqrt(20)), ylim = c(0, 150))
-  abline(a = beta[as.character(node), "(Intercept)"], b = beta[as.character(node), "Age"], col = "black")
-  abline(a = beta[as.character(node), "(Intercept)"] + beta[as.character(node), "ProgramHS"], 
-         b = beta[as.character(node), "Age"] + beta[as.character(node), "ProgramHS:Age"], col = "red")
-}
-@
-\label{fig:lmm_tree}
-\end{figure}%
-
-The resulting tree is presented in Figure~\ref{fig:lmm_tree}.
-Below each terminal node, we plotted the observations and the two regression curves given by the coefficients in that node.
-The first split was made based on the AFTQ variable, which represents the mother's score on the Armed Forced Qualification Test, adjusted for the age at which they completed the test.
-The group with higher mother's AFTQ scores is further split based on race.
-The Black and Hispanic group is further split based on income.
-
-<<echo=FALSE, results=tex>>=
-coefs <- round(coef(HS_tree), digits = 2)
-design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
-                        Program = factor(rep(c("None", "HS"))))
-design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(6)), 
-                                     re.form = ~0)
-design_df$`PPVT at age 18` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(18)), 
-                                      re.form = ~0)
-names(design_df)[1] <- "Node"
-knitr::kable(design_df, format = "latex", digits = 2, 
-             caption = "Predicted PPVT scores at different ages.", 
-             label = "predictions", centering = FALSE,
-             linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)
-@
-
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 \todo{x axis for age starts at 2?} and 18.
-All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
-This benefit remains the same over time for the group with lower mother's AFTQ scores.
-The benefit increases over time for White children with higher mother's AFTQ scores.
-Strikingly,\todo{possibly less extreme word like: Interestingly,} the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
-These results correspond to the conclusions of \cite{Demi09}.
-\todo[inline]{Generally the effects are not large but the example is very clear and I like that it is an intervention effect.
-I suggest to state early before the figure that the effects are not large in this example but we are interested in the group differences; also try to make figure bigger and wider to make slopes more visible; (Mirka): I think the main effect of the intervention and the interaction effect is not too small, but it is hard to see because the age effect is so pronounced in this example.
-Would it be an option to only use the intervention effect in the end nodes?
-}
-
-To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
-We \edc{randomly} separated the \Sexpr{length(unique(HS_dat$MotherID))} families in the dataset into ten equally-sized folds.
-We took nine of the ten folds as a training dataset on which we fitted two models: an LMM tree \edc{like the one shown above} (i.e., an LMM with subgroups \todo{necessary?
-confused me}) and an LMM without subgroups (comprising main and interaction effects of age and Head Start participation, and a random intercept with respect to child).
-We evaluated performance on the remaining folds by computing and evaluating accuracy of the predictions.
-\todo{explain MSE and $R^2$} We repeated this procedure ten times, so that all folds were used as a test set once.
-The results are presented in Table~\ref{tab:performance}, which shows that LMM trees generalize well: They provide better predictive accuracy compared to LMMs, while implementing only few splits.
-
-<<cv_lmm_tree, echo=FALSE, results=tex>>=
-results <- data.frame(MSE_tree = rep(NA, times = 10), 
-                      MSE_lmm = rep(NA, times = 10),
-                      tree_size = rep(NA, times = 10))
-
-set.seed(42)
-folds <- sample(1:10, replace = TRUE, size = length(unique(HS_dat$MotherID)))
-HS_dat$fold <- NA
-for (mom in unique(HS_dat$MotherID)) {
-  HS_dat$fold[HS_dat$MotherID == mom] <- folds[unique(HS_dat$MotherID) == mom]
-}
-for (i in 1:10) {
-  ## Fit models on training data
-  lmmt <- lmertree(PPVT ~ Age*Program | ChildID | AFTQ + Race + Income + 
-                     Mom_height + Mom_edu_yrs, 
-                   cluster = ChildID, data = HS_dat[HS_dat$fold != i, ], 
-                   minsize = 225)
-  lmm <- lmer(PPVT ~ Age*Program + (1|ChildID), 
-              data = HS_dat[HS_dat$fold != i, ])
-
-  ## Evaluate performance on test data
-  results$tree_size[i] <- (length(lmmt$tree)-1)/2
-  preds <- predict(lmmt, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
-  results$MSE_tree[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
-  preds <- predict(lmm, newdata = HS_dat[HS_dat$fold == i, ], re.form = ~0)
-  results$MSE_lmm[i] <- mean((preds - HS_dat[HS_dat$fold == i, "PPVT"])^2)
-}
-res_df <- data.frame(method = c("LMM tree", "LMM"),
-                     MSE = colMeans(results[1:2]),
-                     SD = sapply(results, sd)[1:2],
-                     `number of splits` = c(colMeans(results)[3], NA),
-                     R2 = colMeans(1 - results[ , 1:2] / var(HS_dat$PPVT)))
-knitr::kable(res_df, format = "latex", digits = 3, 
-             caption = "Cross-validated performance of LMMs and LMM trees.", 
-             label = "performance", centering = FALSE,
-             linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("ccccc"), booktabs = TRUE)
-@
-
-\todo[inline]{I am not sure this dataset is the best GLMM tree illustration, because visualized differences between subgroups are very small and perhaps the fixed-effect part is already too complex.
-Alternative: Do not focus on effect of Head Start, but use as partitioning variable.
-Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.
--- response Caro: I like it, see comment above: try to prime that effects are not large -- response Mirka: I think that's okay.
-Only if you had data at hand that is not longitudinal (rather students in classes/schools), with covariates on Level-1, I would opt for that because it doesn't have the Level-2 splitting issue.
-}
-
-\newpage
-\FloatBarrier
-\section{\edc{Application Example 2}: Subgroup Detection in Rasch Models}
-\label{sec:TutorialRasch}
-
 \begin{itemize}
@@
 This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
-
-\FloatBarrier
-\section{Discussion}
-
-In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.
-The main advantage of MOB is that it can detect groups of persons with different model parameters in a data driven way.
-This makes it more flexible for detecting differences that were not hypothesized by the researchers.
-For example in DIF analysis, it is often the case that any obvious sources of DIF have already been avoided by the content expernts in the item creation phase.
-Any remaining DIF is unexpected.
-Therefore, any DIF detection approach that relies on the researchers correctly specifying the exact groups of persons that exhibit DIF can miss DIF if it is associated with other (combinations of) covariates or other cutpoints than the ones investigated.
-In this sense, a more exploratory approach like a Rasch tree has a higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
-The same holds for parameter differences in mixed-effects models, where the substantial hypothesis are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
-\todo{Marjolein add if this makes any sense here}
-
-\todo[inline]{Write summary of findings.}
-
-For our two application examples, we could show that MOB can detect ..., performs automated variable selection \todo[inline]{or did you, Marjolein, mean repeat substantive findings?
-not very exciting for SPISA}
-
-\todo[inline]{Mention shortcomings.}
-
-not possible to detect certain patterns: XOR problem, known for all trees
-
-stability of trees, small changes in data can lead to quite different looking tree.
-The tree structure should always be interpreted jointly for all variables: rather than considering the first splitting variable to be the most important one, it is actually the combination of all splitting variables that creates the groups.\todo{in interpretation of trees in examples: do this, point out that it is interactions -- Caro also add in intro?}.
-The \texttt{stabletree} R package also provides descriptive statistics and graphics that help judge the stability of tree and MOB results \citep{PhiRusHor:2018:JoCaGS,PhiZeiStr:2016}.
-
-\todo[inline]{Write about future work.}
-
-sample size dependence, similar ideas like MH trees also possible for other MOB?
-
-The Mantel-Haenszel effect size measure for DIF will be integrated in the \texttt{psychotools} package.
-
-\bibliography{_MOB_paper}
-
-\todo{Add DOIs.}
-
-\newpage
-\appendix
-
-\section{Appendix A: SDQ-II items administered in the ECLS}
-\label{sec:AppendixA}
-
-\todo{Fix formatting.}
-
-How true is each of these about you? (1="not at all true"; 2="a little bit true"; 3="mostly true" or 4="very true")
-
-\begin{itemize}
-
-\item C7MTHBST (math): Math is one of my best subjects.
-
-\item C7ANGRY (internalizing): I feel angry when I have trouble learning.
-
-\item C7LIKRD (reading): I like reading.
-
-\item C7WRYTST (internalizing): I worry about taking tests.
-
-\item C7MTHGD (math): I get good grades in math.
-
-\item C7LONLY (internalizing): I often feel lonely.
-
-\item C7ENGBST (reading): English is one of my best subjects.
-
-\item C7SAD (internalizing): I feel sad a lot of the time.
-
-\item C7LIKMTH (math): I like math.
-
-\item C7WRYWEL (internalizing): I worry about doing well in school.
-
-\item C7ENJRD (reading): I enjoy doing work in reading.
-
-\item C7WRYFIN (internalizing): I worry about finishing my work.
-
-\item C7ENJMTH (math): I enjoy doing work in math.
-
-\item C7WRYHNG (internalizing): I worry about having someone to hang out with at school.
-
-\item C7GRDENG (reading): I get good grades in English.
-
-\item C7ASHAME (internalizing): I feel ashamed when I make mistakes at school.
-
-\end{itemize}
-
-\end{document}

### 2023-10-11 _MOB_paper.Rnw
Take Rasch Chapter out to work on it

@@
 
-\begin{itemize}
-\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet)
-\item take the subsample of students (only students), and a random subsample of 5000 respondents (seed 04102023)
-\item items could probably appear in Appendix??
-\begin{itemize}
-\item Which painter created this painting? – Andy Warhol.
-\item What do these four buildings have in common? – All four were designed by the same architects.
-\item Roman numbers: What is the meaning of CLVI? – 156.
-\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
-\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
-\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
-\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
-\item In which city is this building located? – Paris.
-\item Which one of the following operas is not by Mozart? – Aida.
-\end{itemize}
-\end{itemize}
-
-\todo[inline]{at Caro: Ich habe die Daten (glaube ich) mal von dir bekommen (\texttt{spisa\_ges.RData}) und die Item-Formulierungen habe ich aktuell aus dem psychotree Paket.
-Stimmt das so?
-Ich glaube, der Originaltest hatte mehrere parallele Skalen, zumindest sind in dem PDF/BUCH von Trepte sehr viel mehr, und sehr ähnliche Items enthalten.
-Das könnten wir vielleicht nochmal nachgucken.}
-
-We load the data and inspect the first rows.
-there are several covariates: age (continuous), gender (male, female, missing), area (Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, Sports) and 9 items from the culture scale (true/false 0/1).
-
-@Caro: I excluded the covariates student and occupation, because I took the student subsample;
-
-<<echo = FALSE>>=
-load("SPISA_example/dat_SPISA.rda")
-covar_data$Student <- NULL
-covar_data$Occupation <- NULL
-@
-
-<<>>=
-head(covar_data)
-head(culture_scale)
-@
-
-We create a dataset \texttt{dat\_SPISA} that contains the data of the covariate and the item responses from the culture scale.
-By assigning the whole scale to the dataset using \texttt{\$}, we can later access item responses from all items using \texttt{\$culture}.
-<<>>=
-dat_SPISA <- covar_data
-dat_SPISA$culture <- culture_scale
-@
-
-The Rasch tree can be fitted with the function \texttt{raschtree} from the R package \texttt{psychotree}.
-We use the typical formula syntax with the item responses of the culture scale on the left hand side of $\sim$, and the covariates on the right hand side.
-
-<<echo = FALSE>>=
-if(!file.exists("SPISA_example/Raschtree.rda")){
-  library(psychotree)
-  Raschtree_culture <- raschtree(culture ~  Gender + Age + Area, 
-                                 data = dat_SPISA)
-  save(Raschtree_culture, file = "SPISA_example/Raschtree.rda")
-}
-load("SPISA_example/Raschtree.rda")
-@
-
-<<eval = FALSE>>=
-library(psychotree)
-Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
-                               data = dat_SPISA)
-@
-
-We can plot the Raschtree using the plot function.
-But it is very big, lots of splits, we don't know whether the splits are due to substantial differences in item difficulty parameters, because the item difficulty profiles in the end nodes are hard to interpret/compare.
-
-<<fig = TRUE, height = 8, width = 15>>=
-plot(Raschtree_culture)
-@
-
-Mantel-Haenszel trees use the Mantel-Haenszel effect size measure for DIF.
-it has three categories (A: negligible, B: moderate, C: large).
-If none of the items has DIF in category B or C (default, but can also be changed so that DIF has to be C), the tree is stopped from growing.
-
-The software must be installed from github, but will (probably) be implemented in the psychotree package some time in the future
-
-<<>>=
-devtools::install_github("mirka-henninger/raschtreeMH")
-library(raschtreeMH)
-@
-
-Syntax is very similar to the classical Rasch tree, but an additional argument \texttt{stopfun}, where the mantelhaneszel stopping function can be selected (but also own stopping functions can be provided; see Henninger et al., 2023).
-The user also has to indicate what kind of purification strategy should be used (none, 2step, iterative).
-Iterative is recommended.
-
-<<echo = FALSE>>=
-if(!file.exists("SPISA_example/Raschtree.rda")){
-  library(psychotree)
-  Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area, 
-                                  data = dat_SPISA, 
-                                  stopfun= stopfun_mantelhaenszel(purification = "iterative"))
-  Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, purification = "iterative")
-  save(Raschtree_MH_culture, file = "SPISA_example/Raschtree_MH_culture.rda")
-}
-load("SPISA_example/Raschtree_MH_culture.rda")
-@
-
-<<eval = FALSE>>=
-Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area, 
-                                  data = dat_SPISA, 
-                                  stopfun= stopfun_mantelhaenszel(
-                                    purification = "iterative"))
-@
-
-For technical reasons, the information about the effect size and ETS classification are not saved in the tree object itself but have to be added afterwards using the \texttt{add\_mantelhaenszel} function.
-The information can then be accessed from the object using \texttt{\$info\$mantelhaenszel}
-
-<<eval = FALSE>>=
-Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, 
-                                           purification = "iterative")
-Raschtree_MH_culture$info$mantelhaenszel
-@
-
-\todo[inline]{not clear what this information is necessary/useful for? since it is not displayed, leave this out?}
-
-If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.
-In addition, we see the number of items classified as A, B, or C in each node.
-For instance, we see that one item is classified as B in Node 1, and no item as C.
-
-<<fig = TRUE, width = 12, height = 7>>=
-plot(Raschtree_MH_culture)
-@
-
-We can also color the items in the end node profiles according to a split in the tree.
-Here, we see that for Node 1 (split on the covariate Area), item 2 shows DIF in category B.
-Item 2 seems to be easier (y-axis shows item easiness) in certain areas compared to Language, Culture, Medicine, Health, Arts.
-
-<<fig = TRUE, width = 12, height = 7>>=
-plot(Raschtree_MH_culture, color_by_node = 1)
-@
-
-We can also color by Node 3 (also split on covariate Area). Also here, Item 2 shows DIF. Similarly, item 2 seems to be more difficult (less easy) for respondents who study Arts.
-
-<<fig = TRUE, width = 12, height = 7>>=
-plot(Raschtree_MH_culture, color_by_node = 3)
-@
-
-This is interesting, because it is the culture scale.
-Why should people who study Arts have more problems to solve the item?
-When we look at the item, it probably makes sense: the items asks what four buildings have in common (Allianz-Arena München, Tate Modern NY, Olympia-Stadion Peking, Elbphilharmonie Hamburg).
-Two of these buildings are sports stadiums.
-So probably respondents who are interested in sports, and know these buildings well have an advantage on the item (independent of the level of knowledge about culture).
-
-\todo[inline]{Maybe a small note on anchoring?
-Item difficulty profiles are anchoring on the items that do not show DIF.
-This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
+\SweaveInput{Raschtree_MH.Rnw}
 

### 2023-10-11 Raschtree_MH.Rnw
New Rasch tree example

@@
 \begin{itemize}
-\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet)
-\item take the subsample of students (only students), and a random subsample of 5000 respondents (seed 04102023)
+\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet; Bogen 20 (as in psychotree, but from the whole dataset, not only a subsample))
+\item take the subsample of students (only students); complete cases; $N = 9769$
 \item items could probably appear in Appendix??
-\begin{itemize}
+\begin{enumerate}
 \item Which painter created this painting? – Andy Warhol.
@@
 \item Which one of the following operas is not by Mozart? – Aida.
+\end{enumerate}
 \end{itemize}
-\end{itemize}
-
-\todo[inline]{at Caro: Ich habe die Daten (glaube ich) mal von dir bekommen (\texttt{spisa\_ges.RData}) und die Item-Formulierungen habe ich aktuell aus dem psychotree Paket.
-Stimmt das so?
-Ich glaube, der Originaltest hatte mehrere parallele Skalen, zumindest sind in dem PDF/BUCH von Trepte sehr viel mehr, und sehr ähnliche Items enthalten.
-Das könnten wir vielleicht nochmal nachgucken.}
 
@@
 
-<<fig = TRUE, height = 8, width = 15>>=
+<<fig = TRUE, width = 15, height = 8>>=
 plot(Raschtree_culture)
@@
 Iterative is recommended.
+Optionally, the user can specify the stopping criterion.
+In this case, tree growing is stopped when all items fall into category ``A''.
 
 <<echo = FALSE>>=
-if(!file.exists("SPISA_example/Raschtree.rda")){
-  library(psychotree)
+if(!file.exists("SPISA_example/Raschtree_MH_culture.rda")){
   Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area, 
                                     data = dat_SPISA, 
-                                    stopfun= stopfun_mantelhaenszel(purification = "iterative"))
-  Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, purification = "iterative")
+                                    stopfun= stopfun_mantelhaenszel(purification = "iterative",
+                                                                    stopcrit = "A"))
   save(Raschtree_MH_culture, file = "SPISA_example/Raschtree_MH_culture.rda")
@@
                                   stopfun= stopfun_mantelhaenszel(
-                                    purification = "iterative"))
+                                    purification = "iterative",
+                                    stopcrit = "A"))
 @
@@
 For technical reasons, the information about the effect size and ETS classification are not saved in the tree object itself but have to be added afterwards using the \texttt{add\_mantelhaenszel} function.
-The information can then be accessed from the object using \texttt{\$info\$mantelhaenszel}
 
-<<eval = FALSE>>=
+<<>>=
 Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, 
                                            purification = "iterative")
-Raschtree_MH_culture$info$mantelhaenszel
 @
 
-\todo[inline]{not clear what this information is necessary/useful for? since it is not displayed, leave this out?}
-
+After we have added the information about the Mantel-Haenszel effect size to the Rasch tree, we can now visualize the Rasch tree together with the effect size measure.
 If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.
 In addition, we see the number of items classified as A, B, or C in each node.
-For instance, we see that one item is classified as B in Node 1, and no item as C.
+For instance, we see that one item is classified in category B in Node 1, and 4 items as C.
 
-<<fig = TRUE, width = 12, height = 7>>=
+<<fig = TRUE, width = 9, height = 7>>=
 plot(Raschtree_MH_culture)
@@
 
-We can also color the items in the end node profiles according to a split in the tree.
-Here, we see that for Node 1 (split on the covariate Area), item 2 shows DIF in category B.
-Item 2 seems to be easier (y-axis shows item easiness) in certain areas compared to Language, Culture, Medicine, Health, Arts.
+We can also color the items in the end node profiles according to a split in the tree. Here, we see that for Node 1 (split on the covariate Gender), items 1, 6, 7, 8, 9 show DIF in categories B or C.
 
-<<fig = TRUE, width = 12, height = 7>>=
+\begin{itemize}
+\item 1 painting by Andy Warhol
+\item 6: novel by Daniel Kehlmann: Die Vermessung der Welt
+\item 7: city of the Buddenbrooks: Lübeck
+\item 8: city with building: Paris
+\item 9: opera not by Mozart: Aida
+\end{itemize}
+
+<<fig = TRUE, width = 9, height = 7>>=
 plot(Raschtree_MH_culture, color_by_node = 1)
@@
 
-We can also color by Node 3 (also split on covariate Area). Also here, Item 2 shows DIF. Similarly, item 2 seems to be more difficult (less easy) for respondents who study Arts.
+We can also color by Node 2 (split on covariate Age).
+Also here, Item 1 shows DIF in category B.
+It seems to be easier for younger respondents ($\leq 24$) compared to older respondents (painting by Andy Warhol).
 
-<<fig = TRUE, width = 12, height = 7>>=
-plot(Raschtree_MH_culture, color_by_node = 3)
+<<fig = TRUE, width = 9, height = 7>>=
+plot(Raschtree_MH_culture, color_by_node = 2)
 @
 
-This is interesting, because it is the culture scale.
-Why should people who study Arts have more problems to solve the item?
-When we look at the item, it probably makes sense: the items asks what four buildings have in common (Allianz-Arena München, Tate Modern NY, Olympia-Stadion Peking, Elbphilharmonie Hamburg).
-Two of these buildings are sports stadiums.
-So probably respondents who are interested in sports, and know these buildings well have an advantage on the item (independent of the level of knowledge about culture).
+For Node 4 (also split on covariate Age), Items 1 and 9 show DIF, such that Item 1 and 9 are easier for younger respondents (Age $\leq 29$).
+(Item 1: Painting by Andy Warhol; Item 9: Opera not by Mozart: Aida)
+
+<<fig = TRUE, width = 9, height = 7>>=
+plot(Raschtree_MH_culture, color_by_node = 4)
+@
 
@@
 This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
+
+The information about the Mantel-Haenszel effect size measure for each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
+it contains the information on classification of each item in e ach node, the computed Mantel-Haenszel effect size measure, the type of purification, and if purification was iterative how many iterations were realized.
+
+<<>>=
+Raschtree_MH_culture$info$mantelhaenszel
+@
+
+@Caro: for instance, from the signs we can see that items are less easy (i.e.
+more difficult) for females compared to men; it is probably easier to see that from the sign compared to the item difficulty profiles.

### 2023-10-11 SPISA_example/SPISA.R

@@ -1,0 +1,85 @@
+library(psychotree)
+library(raschtreeMH)
+load("/home/mhenninger/Arbeit/Forschung/Data/SPISA/spisa_ges.RData")
+
+spisa <- spisa[spisa$Bogen == 20, ]
+
+# translate covariates into english
+covar_dat <- spisa[, c("Alter", "Geschlecht", "Student", "Tat", "Richtung")]
+colnames(covar_dat) <- c("Age", "Gender", "Student", "Occupation", "Area")
+levels(covar_dat$Gender) <- c("Female", "Male", "Missing")
+levels(covar_dat$Student) <- c("No", "Yes", "Missing")
+
+# combine factor levels of occupation
+covar_dat$Occupation <- forcats::fct_collapse(covar_dat$Occupation, 
+                                              Arbeitnehmer = c("Vollzeit", "Teilzeit"),
+                                              nicht_berufstaetig = c("arbeitlos", "nicht berufstaetig"),
+                                              Ruhestand = c("Ruhestand", "Ruhe+Tat")
+)
+levels(covar_dat$Occupation) <- c("\n\n\nPupil", "\n\n\nStudent", "\nPhD",
+                                  "\nin Education", "\nEmployed",
+                                  "\nUnemployed", "\nRetired",
+                                  "\nOthers")
+covar_dat[!(covar_dat$Gender %in% c("Female", "Male")),]$Gender <- "Missing"
+levels(covar_dat$Area) <- c("Language & Culture", "\nLaw and Economics", "\nMedicine & Health", "\nEngineering",
+                            "\nSciences, Pharmacy, Geography", "\nAgriculture & Nutrition", "\nSports", "\nArts",
+                            "\nno Information")
+
+culture_scale <- apply(spisa[,28:36],2,as.numeric)
+
+dat <- covar_dat
+dat$culture_scale <- culture_scale
+
+# only students
+dat <- dat[dat$Student == "Yes", ]
+dat <- dat[complete.cases(dat),]
+# dat <- dat[dat$Area != "Missing", ]
+
+# set.seed(13081990)
+# dat_SPISA <- dat[sample(nrow(dat), 5000),]
+dat_SPISA <- dat
+
+culture_scale <- dat_SPISA$culture_scale
+covar_data <- dat_SPISA[,1:5]
+save(culture_scale, covar_data, file = "SPISA_example/dat_SPISA.rda")
+
+
[…]

### 2023-10-11 _MOB_paper.Rnw
Add Interpretation Techniques to DiscussioN

@@
 In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
-DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}
+DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}
 
@@
 As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
-Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
+Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
 \citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
 In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
-\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
+\citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
 
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citep[see][for details]{HennyDeba23} \todo{Caro can add more text on achoring, incl.
+Together with a purification step \citep[see][for details]{HenDebStr:2023:EPM} \todo{Caro can add more text on achoring, incl.
 word anchoring, here to prepare example 2 - only shortly mention here and explain in more detail later in the example (will become more clear in example)}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
@@
 
-stability of trees, small changes in data can lead to quite different looking tree.
-The tree structure should always be interpreted jointly for all variables: rather than considering the first splitting variable to be the most important one, it is actually the combination of all splitting variables that creates the groups.\todo{in interpretation of trees in examples: do this, point out that it is interactions -- Caro also add in intro?}.
-The \texttt{stabletree} R package also provides descriptive statistics and graphics that help judge the stability of tree and MOB results \citep{PhiRusHor:2018:JoCaGS,PhiZeiStr:2016}.
-
 \todo[inline]{Write about future work.}
@@
 because effect sizes in mixed effects models are not very clear}
+
+The Rasch tree method has been extended to include an effect size measure that can stop the tree from growing if the effect size is non-substantial, but also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
+At the same time, effect sizes are less straightforward to calculate and interpret in linear mixed models, and therewith in the linear mixed effects model tree.
+A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
+These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s).
+A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.
 

### 2023-10-11 Raschtree_MH.Rnw
Comments about changes

@@
+\todo[inline]{Comment on new example: I now use data from Bogen 20 of the SPISA dataset.
+I use the student subset, and complete cases on all variables.
+The story is less clear, but the tree is still stopped early, and we can display which items have dif.
+}
+
 \begin{itemize}
@@
 covar_data$Occupation <- NULL
+covar_data$Area <- sub("\n", "", covar_data$Area)
 @
@@
 @
+
+\todo[inline]{The are covariate split does not look pretty yet. But I will try to make it more pretty later on.. }
 

### 2023-10-20 _MOB_paper.Rnw
Comments about changes
Paper updates
Revised manuscript and analyses. Did not touch application to Rasch trees. Still quite some work to do, but this is as far as I could get.

@@
 
-\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups \edc{of persons showing different} effects in a wide range of parametric models.
+\abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models.
 It provides a versatile tool for detecting and explaining heterogeneity of intervention effects.
@@
 In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
-We apply GLMM trees to a dataset from a study of \citep{Demi09}\todo{take that out?}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
-Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start.
-In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
-DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}
+We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance.
+In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF).
+%DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.
+% Test fairness is more related to use of test scores in practice than group-level comparisons I would think?
+We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}
 
@@
 
-Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical
-model between groups of persons. This method is a generalization of the principle of recursive partitioning, that is also used in classification and regression trees \citep{Breetal:1984}. 
-In classification and regression trees, the aim is to detect groups of persons, specified by (combinations of) covariates, that differ in a response variable.
-An example could be that children who are highly motivated and have good reading skills show higher average math exam grades.
-MOB generalizes this idea by identifying groups of persons, specified by (combinations of) covariates, that differ in a the parameters of a statistical model.
+Model-based recursive partitioning \citep[MOB;][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical model between groups of persons.
+It generalizes the principle of recursive partitioning, that is also used in classification and regression trees \citep[CART; ][]{Breetal:1984}.
+In CART, the aim is to detect groups of persons, defined by (combinations of) covariate values, that differ in their mean on a response variable.
+For example, if we want to predict math exam grades using student characteristics assessed in a previous year, we could find that the subgroup of children who are highly motivated and have good reading skills show higher average math exam grades.
+MOB generalizes this idea: It also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean.
 An example could be that children who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.
 
-The framework of MOB is very flexible and can be applied in combination with a variety of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS}.
-In this article, we will highlight two such combinations that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and for models from Rasch measurement and Item Response Theory \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
-Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are again nested in schools) participate in a study.
-In application example 1 we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
+The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}.
+In this article, we will highlight two specific MOB methods that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
+Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are in turn nested in schools) participate in a study.
+In application example 1 \ref{sec:TutorialMixed} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
+In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model.
+We thus assume that the psychological test score(s) of interest \todo{(Mirka: I like the term instrument to distinguish it from a statistical test) (Marjolein: I hope 'psychological test score' clears up possible confusion)} have already been validated.
 
-While for this first application example we assume that the administered test \todo{should we call a psychological test a test/an instrument/???
-to distinguish it from the verb and noun for statistical test?
-(Mirka: I like the term instrument to distinguish it from a statistical test)} has already been thoroughly validated, for application example 2 we go back in the research process to the point where a new test is administered to a validation sample.
-Before using the test in our research, we need to make sure that test results are comparable, for example between children of different genders.
-If certain items pose an advantage or disadvantage to either group, they are said to violate measurement invariance or exhibit differential item functioning (DIF).
-In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, groups of persons are compared with respect to the item parameters for their specific group.
-This can be done in a way that allows to detect DIF on top of any true group differences in ability.
-While many standard approaches for testing DIF are limited to comparing typically two pre-specified groups at a time, we will show in application example 2 that MOB more flexibly allows to detect groups of persons with different item parameters in a data-driven
-way.
+For application example 2\todo{add link to section?
+not sure how to do}, we go back in the research process to the point where a new test has been administered to a validation sample.
+Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model.
+Validity assessment requires that we make sure that test results are comparable, for example between children of different genders.
+If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage, and these items are said to violate measurement invariance or to exhibit differential item functioning (DIF).
+In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, the item parameters are compared between groups of persons.
+This can be done in a way that allows to detect DIF, while accounting for possible true group differences in ability.
+Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF.
+In application example 2 we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.
 
-In the following section, we will give a short introduction into the algorithm and statistical concepts behind MOB.
+In the following section, we first give a short introduction into the algorithm and statistical concepts behind MOB.
 Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}.
@@
 It can also be combined with covariates.
-A comparison of both approches is given by \citet{FriStrZei:2014}.
+A comparison of both approaches is given by \citet{FriStrZei:2014}.
+\todo{Last 5 sentences seem distracting here.
+Omit or move to discussion?}
 
-\section{The Algorithm Behind MOB}
+\section{The MOB Algorithm}
 
-What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?); Marjolein: please read and decide whether we do need a picture of a regression tree and/or mob tree here, if yes, simulate but with abstract x1 x2, no story}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
-These figures share the tree-structure in the upper part, but differ in their lower part.
-The starting point of the tree structure is at the very top of each tree.
-The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the
-bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.
+\begin{figure}
+\caption{Univariate distribution of the response variable, with global mean indicated by triangle.}
+\begin{subfigure}{.5\textwidth}
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
+library("colorspace")
+set.seed(3)
+x1 <- runif(250, min = 5, max = 11)
+x2 <- runif(250, min = 2, max = 8)
+x3 <- runif(250, min = 4, max = 10)
+error <- rnorm(250, sd = 1.5)
+y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
+toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
+boxplot(toy_data$y, cex = .7)
+points(x = 1, y = mean(toy_data$y), col = rainbow_hcl(1), pch = 17)
+@
+\end{subfigure}
+\label{fig:global}
+\end{figure}
 
-The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
+The main rationale of MOB is that one global model may not fit all observations in a dataset equally well.
+In many studies, additional covariates may be available.
+It may then be possible to uncover subgroups defined by these covariates, and obtain better-fitting models in each of those subgroups \citep{ZeilyHoth08}.
+
+We illustrate the idea using a very simple toy dataset, comprising \Sexpr{nrow(toy_data)} observations, and four variables: A continuous response variable $y$, and three possible partitioning (or predictor) variables $x_1$, $x_2$ and $x_3$.
+To keep the example simple, we apply MOB to a very simply global model, which comprises only an intercept.
+Figure~\ref{fig:global} shows the distribution of $y$ in a boxplot, with the global intercept depicted as a triangle.
+Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
+
+To detect possible subgroups with different values for the parameters, MOB cycles iteratively through the following steps:
+
+\begin{enumerate}
+\setlength\itemsep{0.25em}
+\item The model parameters are first estimated jointly for all persons in the current node, starting with the root node containing the full sample.
+\item Structural change in the model parameters is assessed with respect to each available covariate.
+\item If there is significant structural change, the observations in the current node are split using the covariate associated with the strongest change.
+\item Steps 1--3 are repeated recursively in each resulting node until there is no more significant structural change (or the groups becomes too small).
+\end{enumerate}
+
+We applied MOB to the observations in Figure~\ref{fig:global}, specifying $x_1$, $x_2$ and $x_3$ as the additional covariates.
+The resulting tree is shown in Figure~\ref{fig:tree}.
+In the root node, which contains all observations, structural change tests were performed for each of the three covariates (step 2).
+The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected for splitting (step 3).
+The $p$ value resulting from the structural change test for $x_1$ is depicted in this root node.
+Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible between the groups.
+In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with instability in this subgroup.
+In the right daughter node, there was significant instability associated with $x_2$, a categorical covariate.
+Again, the cutpoint was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible.
+The subgroup-specific model parameters are presented in end nodes at the bottom of the tree.
+
+\begin{figure}
+\caption{Decision tree with group-specific parameters in the terminal nodes.}
+\begin{subfigure}{.5\textwidth}
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
+library("partykit")
+toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
+tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
+plot(tree, gp = gpar(cex = .7))
+@
+~
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
+plot(tree, gp = gpar(cex = .7), type = "simple")
+@
+\end{subfigure}
+\label{fig:tree}
+\end{figure}
+
+%The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
 For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \texttt{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups.
@@
 -- add name if is there one in the other example?}
-on the other hand, there is only one possible cutpoint, namely between the two categories.
-While it is also possible to create more than two groups in each split \citep{KimLoh:2001,Qui:1993}, binary splitting algorithms are typically preferred because they lead to more concise trees.
+%on the other hand, there is only one possible cutpoint, namely between the two categories.
 
-The parametric model to which the MOB algorithm is applied is visualized in the end nodes at the bottom of each tree.
-While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items.
-The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
-Figure~\ref{fig:lmm_tree} also shows that, out of the  covariates that were presented to the algorithm (\texttt{AFTQ}, \texttt{Race}, \texttt{Income},\texttt{Mom\_edu\_yrs}, and \texttt{Mom\_height}), only \texttt{AFTQ}, \texttt{Race}, and \texttt{Income} were actually selected for splitting.
+%While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items.
+%The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
+
+%Figure~\ref{fig:lmm_tree} also shows that, out of the  covariates that were presented to the algorithm (\texttt{AFTQ}, \texttt{Race}, \texttt{Income},\texttt{Mom\_edu\_yrs}, and \texttt{Mom\_height}), only \texttt{AFTQ}, \texttt{Race}, and \texttt{Income} were actually selected for splitting.
 Variables are selected based on a statistical test for parameter differences, also termed a test for structural change.
@@
 
-The way the variable selection and cutpoint selection steps are separated in modern classification and regression tree and MOB algorithms, including the ones used here, distinguish them from the traditional classification and regression tree algorithms of \citet{Breetal:1984} and \citet{Qui:1993}.
-The traditional algorithms performed variable and cutpoint selection in one step.
-However, this leads to an undesirable behavior called variable selection bias.
-It means that the traditional algorithms prefer variables offering more cutpoints in the selection process -- regardless of their true information content.
+%While it is also possible to create more than two groups in each split \citep{KimLoh:2001,Qui:1993}, binary splitting algorithms are typically preferred because they lead to more concise trees.
+% Think we can leave this sentence out, I'm not convinced readers will consider that other than binary splits may be implemented.
+
+There are two main differences between MOB and and traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}.
+The first difference relates to the separation of variable and cutpoint selection.
+The traditional algorithms performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias.
+That is, the traditional algorithms prefer variables offering more possible cutpoints in the selection process -- regardless of their true information content.
 An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}.
-Modern algorithms for classification and regression trees that have solved this problem are the unbiased approaches \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, available in R in the \texttt{party} and \texttt{partykit} packages \citep{partykit:pkg}.
+More recent algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}.
 The latter forms the basis for all MOB approaches presented in this paper.
+For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.
 
-Once there are no more covariates that show a significant structural change in any node, the MOB algorithm stops splitting.
+The second difference relates to the way MOB stops splitting: It uses a criterion of statistical significance.
+Once there are no more covariates that show a significant structural change in any node, splitting is halted.
 In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection.
 Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected.
-This is the second difference between the modern classification and regression tree and MOB algorithms used here, compared to traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}: While the traditional algorithms grew very large trees and then cut them back by means of so called pruning, the framework employed here is based on statistical significance tests and partly also on effect size measures (see application example 2) to stop the trees already in the growing phase when no more significant change can be detected.
+While the traditional algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (i.e., \textit{pruning}), the MOB framework employs significance tests, but it can also (partly) use effect size measures, as we will see in application example 2 \todo{include link}.
+This allows for stopping tree growing as soon as no significant differences can be detected anymore.
 Other stopping criteria are based on the number of persons in the end nodes.
-These crieria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}
+These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}
 
-In summary, the rationale of a MOB algorithm consists of the following steps:
-
-\todo{why are the spaces so big? I didn't change the enumerate command}
-\begin{enumerate}
-\item The model parameters are first estimated jointly for all persons in the current node, starting with the full sample.
-\item Structural change in the model parameters is assessed with respect to each available covariate.
-\item If there is significant structural change, groups are split along the covariate with the strongest change and using the optimal cutpoint.
-\item Steps 1--3 are repeated recursively in the resulting groups until there is no more significant structural change (or the groups becomes too small).
-\end{enumerate}
-
-For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.
-
-An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory way, but is learned from the data in an exploratory way.
-This is a key feature of the MOB approach that makes it very flexible and distinguishes it from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
-While there are phases in psychological and educational research where it is very important to specify hypothesis a priori and test them in a confirmatory way, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers.
-
-\todo[inline]{\textbf{Previous Papers on IRT and Rasch Models in Journal of School Psychology}
-
-Rasch and IRT models are often mentioned in this journal's papers as being used for scoring school assessments (e.g., Virginia Standards of Learning Assessments;  Oregon Assessment of Knowledge and Skills; Peabody individual achievement test, \cite{Luth92}).
-
-A general introduction to IRT in the journal is provided by \cite{AnthyDiPe16}. \edc{included}
-There are also papers describing extensions to mixure IRT models \cite{AyalySant17} \edc{included}
-and many-facet Rasch measurement (MFRM) approaches that allow controlling for rater effects \cite{StycyAntr21}.
-
-\cite{Mall97} assessed DIF by comparing Rasch model item difficulties of WISC-III subtests between 110 severely and profoundly deaf children, 110 matched nonreferred hearing children, and the WISC-III standardization sample (N = 2,200).
-\edc{included}
-}
+An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner.
+This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
+While there are phases in psychological and educational research where it is very important to specify hypotheses a-priori and test them in a confirmatory manner, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers.
+Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of covariates or possible partitioning variables and to choose the settings of the MOB algorithm.
+The current paper aims to provide some guidance.
 
@@
 
-In mixed-effects models contain two types of effects: fixed effects and random effects.
-Fixed effects are typically used, for example, to describe the effect of an intervention on the average student performance.
-Random effects are used to describe the variation in the intercept and slope for, e.g., students in different classes.
+Mixed-effects models contain two types of effects: Fixed and random effects.
+Fixed effects are typically used to capture population-averaged effects, while random effects are used to capture inter-individual variation deviating from these fixed effects.
+In many studies, researchers are specifically interested in testing hypotheses relating to the population-averaged effects, while the random effects are included in the model to properly account for inter-individual variation, and correlations between observations within the same unit.
 
-When mixed-effects models are combined with MOB, ...
-
-\todo[inline]{Marjolein please edit and finish}
+GLMM trees combine MOB and mixed-effects models and were introduced by \cite{FokkySmit18}.
+Because researchers' interests commonly focus on the fixed effects (of time or treatment, for example), the GLMM tree algorithm only targets structural change in the fixed-effects parameters.
+The random-effects parameters can be specified as usual and are assumed constant; they are estimated using all observations in the dataset.
+The MOB algorithm is thus applied to the fixed-effects parameters only and the detected subgroups will only obtain different estimates for the fixed-effects coefficients.
+This allows for detecting subgroups with different means on the response variable, but also with differential treatment effects or differential growth over time, to name but a few examples.
+While the 'standard' MOB trees \cite{ZeilyHoth08} allow for partitioning fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for partitioning mixed-effects regression models.
+Further mathematical and computational details about the GLMM tree algorithm are described in \cite{FokkySmit18} and \citep{FokkyZeil23}.
 
@@
 
-Before being able to use a psychological or educational test for comparing different groups of persons in a fair way, different assumptions of the measurement model have to be checked.
-In the framework of IRT and Rasch measurement, test items are typically investigated with respect item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}.
-Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance.
+Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups.
+In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}.
+\todo{Omit last sentence or move to discussion?
+I think better to focus on DIF only here.} Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance.
 Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
 
-MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model.
-For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?} The usage of this function will be illustrated in application example 2.
+MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model\todo{possibly confusing: covariates are supplied by user, not searched for.
+Suggestion: by means of testing whether the parameters of the measurement model exhibit significant instability with respect to (sets of) covariates?}.
+For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?
+(Marjolein: I think software should only be mentioned in the applications.)} The usage of this function will be illustrated in application example 2.
 We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model.
@@
 If one joint Rasch model holds for the entire sample, i.e., if there is no DIF, a Rasch tree should show no splits.
-However, in certain settings in educational research, such as large scale assessements, very large sample sizes are available for testing for DIF.
+However, in certain settings in educational research, such as large scale assessments, very large sample sizes are available for testing for DIF.
 Large sample sizes are good for detecting even small effects or model violations with a high statistical power.
@@
 \newpage
-\section{\edc{Application Example 1}: Subgroup Detection in Mixed-Effects Models}
+\section{Application Example 1: Subgroup Detection in Mixed-Effects Models}
 \label{sec:TutorialMixed}
@@
 HS_dat <- HS_dat[complete.cases(HS_dat), ]
-HS_dat$Age <- sqrt(HS_dat$Age)
+HS_dat$Age_orig <- HS_dat$Age
+#HS_dat$Age <- sqrt(HS_dat$Age)
 rownames(HS_dat) <- NULL
@@
 To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start.
-Head Start is a federally funded nationwide \todo{what nation?} pre-school program for children from low-income families.
+Head Start is a federally funded nationwide pre-school program for children from low-income families in the United States.
+Participation in Head Start takes place from ages 3 through 5.
 \cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).
@@
 As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start.
-As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time.
-As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.\todo{this would scare me a little as a reader: how would I know what to do/that I should do this?
-(Mirka: Maybe we can add that the sqrt or the log is a typical way to model trajectories that first increase rapidly, but then later increase more slowly?)}
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF), we will this model PPVT trajectories over time.
+There were an average number of \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(1-mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there was only a single PPVT score available.
+
+%Developmental trajectories rarely if ever show a linear pattern over time; they typically show strong increases early on, which level off over time.
+We therefore took the square root of the child's age in years, to obtain a pattern of approximately linear increases over time.
+Appendix A shows PPVT scores as a function of time, both before and after the transformation.
 
@@
 We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red).
-To show the effect of age and Head Start participation, we fitted a mixed-effects model comprising their main and interaction effects.
-To account for the correlation between repeated assessments on the same child, we specified a random intercept.
-The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference \edc{is slightly diminished but} persists over time.
+To show the effect of age and Head Start participation, we first fit a mixed-effects model comprising their main and interaction effects, using package \texttt{lme4}.
+To account for the correlation between repeated assessments on the same child, and between siblings with the same mother, we specify a random intercept for children, nested within mothers:
+
+<<>>=
+library("lme4")
+lmm <- lmer(PPVT ~ Program*Age + (1|MotherID/ChildID), data = HS_dat)
+@
+
+The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference is slightly diminished but persistent over time.
 This result agrees with the findings of \cite{Demi09}.
+The code for the figure is omitted here for space considerations, because we want to focus on fitting and interpreting GLMM trees.
+Code for exact replication of the results presented here is provided in the Supplementary Materials.
 
@@
 <<echo=FALSE, fig=TRUE>>=
-library("lme4")
-lmm <- lmer(PPVT ~ Program*Age + (1|ChildID), data = HS_dat)
 beta <- fixef(lmm)
@@
        b = beta["Age"]+beta["ProgramHS:Age"], col = "red")
-axis(1, at=sqrt(c(3, 5, 8, 12, 17)), labels=c(3, 5, 8, 12, 17), cex.axis = .8)
+axis(1, #at=sqrt(c(3, 5, 8, 12, 17)), labels=c(3, 5, 8, 12, 17), 
+     cex.axis = .8)
 @
@@
 
-\todo[inline]{I (Mirka) am confused: the nested structure in this example is timepoints (age) within children?
-Therefore, ChildID is the cluster-variable.
-But then, there should be residual errors between the two children from the same family?
-In the text it says that this is accounted for by the random intercept, but I don't think this is the case (but maybe I am wrong?).
-Is it maybe accounted for by the cluster argument that is also said to ChildID?
-To me, the model specification is not yet clear from the text.
-Plus: If I am right, and the cluster is the child and the timepoints are on Level-1, I think all the covariates are on Level-2 (child level), and then maybe we should comment on the Type-I error problem?}
-We now test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:
+Next, we test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:
 
@@
 library("glmertree")
-HS_tree <- lmertree(PPVT ~ Program*Age | (1|ChildID) | AFTQ + Race + 
+HS_tree <- lmertree(PPVT ~ Program*Age | (1|MotherID/ChildID) | AFTQ + Race + 
                       Income + Mom_edu_yrs + Mom_height, 
-                    data = HS_dat, cluster = ChildID, minsize = 250)
+                    data = HS_dat, cluster = MotherID, minsize = 250)
 @
@@
 With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance.
-The middle part (\texttt{1|ChildID}) specified the random effects\todo{(Mirka) Maybe add information that this is for each child, average DV across timepoints}.
+The middle part (\texttt{1|MotherID/ChildID}) specified the random effect: Repeated PPVT assessments are nested within children, which are nested within mothers.
 The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.
 
-\todo[inline]{I (Mirka) am not sure what exactly the cluster argument does.
-Maybe Marjolein you have more information from your work with Achim.
-But my knowledge/intuition is that it has something to do with clustered covariances, so maybe it accounts for the sibling structure, but I don't think it accounts for the level of the predictor variables.}
 With the second argument, we specified the dataset which contain the variables.
-Because we are dealing with repeated measurements on the same children, we additionally specified that the parameter stability tests should be performed on the child level using the \verb|cluster| argument.
-Using the default observation-level parameter stability tests may artificially inflate power\todo{Mirka: I'd rather say it has an inflated type-I error rate.}.
+With the \texttt{cluster} argument, we specified that the partitioning variables are measured at the level of the mothers.
+As a result, parameter stability tests will be performed on the appropriate level \citep{FokkyZeil23}.
+Not specifying the \texttt{cluster} argument would result in use of the default observation-level parameter stability tests, which would yield inflated type-I error rates.
+That is, it could result in detection of spurious subgroups.
 Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.
@@
        cex = .5, cex.lab = .7, cex.axis=.7, cex.main = .7,
-       xlab = "Age", ylab = "PPVT", xlim = c(sqrt(3), sqrt(20)), ylim = c(0, 150))
+       xlab = "Age", ylab = "PPVT", #xlim = c(sqrt(3), sqrt(20)), 
+       ylim = c(0, 150))
   abline(a = beta[as.character(node), "(Intercept)"], b = beta[as.character(node), "Age"], col = "black")
@@
 
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 \todo{x axis for age starts at 2?} and 18.
+To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
@@
 These results correspond to the conclusions of \cite{Demi09}.
-\todo[inline]{Generally the effects are not large but the example is very clear and I like that it is an intervention effect.
-I suggest to state early before the figure that the effects are not large in this example but we are interested in the group differences; also try to make figure bigger and wider to make slopes more visible; (Mirka): I think the main effect of the intervention and the interaction effect is not too small, but it is hard to see because the age effect is so pronounced in this example.
-Would it be an option to only use the intervention effect in the end nodes?
-}
+
+\todo[inline]{With different specification of the GLMM tree (no square root, parameter stability tests at mom level, random intercepts of children nested within mothers) the results have changed, still need to adjust description}.
 
@@
 
-To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
+%To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
 We \edc{randomly} separated the \Sexpr{length(unique(HS_dat$MotherID))} families in the dataset into ten equally-sized folds.
@@
 
-\todo[inline]{I am not sure this dataset is the best GLMM tree illustration, because visualized differences between subgroups are very small and perhaps the fixed-effect part is already too complex.
-Alternative: Do not focus on effect of Head Start, but use as partitioning variable.
-Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.
--- response Caro: I like it, see comment above: try to prime that effects are not large -- response Mirka: I think that's okay.
-Only if you had data at hand that is not longitudinal (rather students in classes/schools), with covariates on Level-1, I would opt for that because it doesn't have the Level-2 splitting issue.
-}
-
 \newpage
@@
 In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.
-The main advantage of MOB is that it can detect groups of persons with different model parameters in a data driven way.
-This makes it more flexible for detecting differences that were not hypothesized by the researchers.
-For example in DIF analysis, it is often the case that any obvious sources of DIF have already been avoided by the content expernts in the item creation phase.
+The main advantage of MOB is that it can detect groups of persons with different model parameters in a data-driven manner.
+This makes it more flexible for detecting differences that were not hypothesized by the researcher.
+For example in DIF analysis, it is often the case that obvious sources of DIF have already been avoided by the content experts during item creation.
 Any remaining DIF is unexpected.
 Therefore, any DIF detection approach that relies on the researchers correctly specifying the exact groups of persons that exhibit DIF can miss DIF if it is associated with other (combinations of) covariates or other cutpoints than the ones investigated.
-In this sense, a more exploratory approach like a Rasch tree has a higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
-The same holds for parameter differences in mixed-effects models, where the substantial hypothesis are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
+In this sense, the more exploratory approach of Rasch trees has higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
+
+The same holds for parameter differences in mixed-effects models, where the substantial hypotheses are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
 \todo{Marjolein add if this makes any sense here}
 
-\todo[inline]{Write summary of findings. -- now think: rather not discuss substantive findings, only properties of the methods}
+For our two application examples, we could show that MOB can detect ..., performs automated variable selection ...
 
-For our two application examples, we could show that MOB can detect ..., performs automated variable selection \todo[inline]{or did you, Marjolein, mean repeat substantive findings?
-not very exciting for SPISA}
+\subsection{Limitations}
+
+Due to the exploratory nature of MOB, it does not allow for hypothesis tests, as there is no valid way to account for the exploratory searching of the subgroups.
+Thus, even if detected subgroups are substantively meaningful or differences are large, if researchers want to ascertain statistical significance of the subgroup differences, this should be done on new data using confirmatory techniques.
 
@@
 A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.
+\todo[inline]{This is my (Mirka) suggestion for the interpretation techniques. Please feel free to add, edit, delete, however you prefer!}
 
 The Mantel-Haenszel effect size measure for DIF will be integrated in the \texttt{psychotools} package.
+
+\todo{Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
+In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}
+
+\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
+In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}
 
@@
 
-\section{Appendix A: SDQ-II items administered in the ECLS}
+\section{Appendix A: Association between Age and PPVT scores before and after transformation}
 \label{sec:AppendixA}
 
-\todo{Fix formatting.}
-
-How true is each of these about you? (1="not at all true"; 2="a little bit true"; 3="mostly true" or 4="very true")
-
-\begin{itemize}
-
-\item C7MTHBST (math): Math is one of my best subjects.
-
-\item C7ANGRY (internalizing): I feel angry when I have trouble learning.
-
-\item C7LIKRD (reading): I like reading.
-
-\item C7WRYTST (internalizing): I worry about taking tests.
-
-\item C7MTHGD (math): I get good grades in math.
-
-\item C7LONLY (internalizing): I often feel lonely.
-
-\item C7ENGBST (reading): English is one of my best subjects.
-
-\item C7SAD (internalizing): I feel sad a lot of the time.
-
-\item C7LIKMTH (math): I like math.
-
-\item C7WRYWEL (internalizing): I worry about doing well in school.
-
-\item C7ENJRD (reading): I enjoy doing work in reading.
-
-\item C7WRYFIN (internalizing): I worry about finishing my work.
-
-\item C7ENJMTH (math): I enjoy doing work in math.
-
-\item C7WRYHNG (internalizing): I worry about having someone to hang out with at school.
-
-\item C7GRDENG (reading): I get good grades in English.
-
-\item C7ASHAME (internalizing): I feel ashamed when I make mistakes at school.
-
-\end{itemize}
+\begin{figure}
+\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
+<<echo=FALSE, fig=TRUE, width=6, height=4>>=
+par(mfrow = c(1, 2))
+plot(HS_dat$Age, HS_dat$PPVT, cex = .7, xlab = "Age", ylab = "PPVT",
+     cex.lab = .7, cex.axis = .7)
+plot(sqrt(HS_dat$Age), HS_dat$PPVT, ylab = "PPVT", xlab = "Age (square root)", 
+     cex = .7, cex.lab = .7, cex.axis = .7)
+@
+\end{figure}
 

### 2023-10-20 _MOB_paper.tex
Paper updates
Revised manuscript and analyses. Did not touch application to Rasch trees. Still quite some work to do, but this is as far as I could get.

@@
 In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
-We apply GLMM trees to a dataset from a study of \citep{Demi09}, who compared longitudinal performance of siblings who did, and siblings who did not participate in the Head Start program.
-Using GLMM trees, we identify subgroups of families in which children show comparatively larger or smaller gains in performance following participation in Head Start.
-In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, \edc{i.e.~differential item functioning (DIF).
-DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.} We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}
+We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance.
+In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF).
+%DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.
+% Test fairness is more related to use of test scores in practice than group-level comparisons I would think?
+We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}
 
@@
 It can also be combined with covariates.
-A comparison of both approches is given by \citet{FriStrZei:2014}.
+A comparison of both approaches is given by \citet{FriStrZei:2014}.
+\todo{Last 5 sentences seem distracting here.
+Omit or move to discussion?}
 
-\section{The Algorithm Behind MOB}
+\section{The MOB Algorithm}
 
-What you see displayed in Figures~\ref{fig:lmm_tree}, \ref{fig:math_tree}, and \ref{fig:read_tree}\todo{adjust if raschtree example changes (Mirka: maybe it would be helpful to have a simplified figure of a GLMERtree and a Raschtree early in the manuscript already, so we can refer to it here?)}, which will be explained in detail in the application example sections, is the result of the MOB algorithm for a linear mixed-effects model (Figure~\ref{fig:lmm_tree}) and for a binary Rasch model (Figures~\ref{fig:math_tree} and \ref{fig:read_tree}).
-These figures share the tree-structure in the upper part, but differ in their lower part.
-The starting point of the tree structure is at the very top of each tree.
-The very top of each tree is called the root node, so what you see are actually ``upside down'' trees with the root at the top and the branches at the
-bottom. The root node at the very top contains the entire sample of persons in the data set. From there, the persons are divided into subgroups.
+\begin{figure}
+\caption{Univariate distribution of the response variable, with global mean indicated by triangle.}
+\begin{subfigure}{.5\textwidth}
+\includegraphics{_MOB_paper-002}
+\end{subfigure}
+\label{fig:global}
+\end{figure}
 
-The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
+The main rationale of MOB is that one global model may not fit all observations in a dataset equally well.
+In many studies, additional covariates may be available.
+It may then be possible to uncover subgroups defined by these covariates, and obtain better-fitting models in each of those subgroups \citep{ZeilyHoth08}.
+
+We illustrate the idea using a very simple toy dataset, comprising 250 observations, and four variables: A continuous response variable $y$, and three possible partitioning (or predictor) variables $x_1$, $x_2$ and $x_3$.
+To keep the example simple, we apply MOB to a very simply global model, which comprises only an intercept.
+Figure~\ref{fig:global} shows the distribution of $y$ in a boxplot, with the global intercept depicted as a triangle.
+Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
+
+To detect possible subgroups with different values for the parameters, MOB cycles iteratively through the following steps:
+
+\begin{enumerate}
+\setlength\itemsep{0.25em}
+\item The model parameters are first estimated jointly for all persons in the current node, starting with the root node containing the full sample.
+\item Structural change in the model parameters is assessed with respect to each available covariate.
+\item If there is significant structural change, the observations in the current node are split using the covariate associated with the strongest change.
+\item Steps 1--3 are repeated recursively in each resulting node until there is no more significant structural change (or the groups becomes too small).
+\end{enumerate}
+
+We applied MOB to the observations in Figure~\ref{fig:global}, specifying $x_1$, $x_2$ and $x_3$ as the additional covariates.
+The resulting tree is shown in Figure~\ref{fig:tree}.
+In the root node, which contains all observations, structural change tests were performed for each of the three covariates (step 2).
+The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected for splitting (step 3).
+The $p$ value resulting from the structural change test for $x_1$ is depicted in this root node.
+Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible between the groups.
+In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with instability in this subgroup.
+In the right daughter node, there was significant instability associated with $x_2$, a categorical covariate.
+Again, the cutpoint was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible.
+The subgroup-specific model parameters are presented in end nodes at the bottom of the tree.
+
+\begin{figure}
+\caption{Decision tree with group-specific parameters in the terminal nodes.}
+\begin{subfigure}{.5\textwidth}
+\includegraphics{_MOB_paper-003}
+~
+\includegraphics{_MOB_paper-004}
+\end{subfigure}
+\label{fig:tree}
+\end{figure}
+
+%The subgroups are defined by the covariates that are used for splitting and together with cutpoints in those covariates.
 For example, in Figure~\ref{fig:lmm_tree} for the metric covariates \texttt{AFTQ} and \texttt{Income} a numeric cutpoint has been selected by the MOB algorithm to separate the groups.
@@
 As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
-Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
+Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
 \citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
 In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
-\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
+\citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
 
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citep[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+Together with a purification step \citep[see][for details]{HenDebStr:2023:EPM} \todo{Caro can add more text on achoring, incl.
+word anchoring, here to prepare example 2 - only shortly mention here and explain in more detail later in the example (will become more clear in example)}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
@@
 As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start.
-As the outcome variable, we take the Peabody Picture Vocabulary Test (PPVT; REF) and model PPVT trajectories over time.
-As the timing metric, we took the square root of the child's age in years, which yielded a pattern of approximately linear increases over time.\todo{this would scare me a little as a reader: how would I know what to do/that I should do this?
-(Mirka: Maybe we can add that the sqrt or the log is a typical way to model trajectories that first increase rapidly, but then later increase more slowly?)}
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF), we will this model PPVT trajectories over time.
+There were an average number of 2.14 PPVT scores per child, for 29\% of the children there was only a single PPVT score available.
+
+%Developmental trajectories rarely if ever show a linear pattern over time; they typically show strong increases early on, which level off over time.
+We therefore took the square root of the child's age in years, to obtain a pattern of approximately linear increases over time.
+Appendix A shows PPVT scores as a function of time, both before and after the transformation.
 
@@
 3 15.964368    Black 16119.13        504          10   22403      224    None
-  PPVT      Age
-1   18 2.000000
-2   48 2.645751
-3   69 2.645751
+  PPVT Age Age_orig
+1   18   4        4
+2   48   7        7
+3   69   7        7
 \end{Soutput}
@@
 We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red).
-To show the effect of age and Head Start participation, we fitted a mixed-effects model comprising their main and interaction effects.
-To account for the correlation between repeated assessments on the same child, we specified a random intercept.
-The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference \edc{is slightly diminished but} persists over time.
+To show the effect of age and Head Start participation, we first fit a mixed-effects model comprising their main and interaction effects, using package \texttt{lme4}.
+To account for the correlation between repeated assessments on the same child, and between siblings with the same mother, we specify a random intercept for children, nested within mothers:
+
+\begin{Schunk}
+\begin{Sinput}
+ library("lme4")
+ lmm <- lmer(PPVT ~ Program*Age + (1|MotherID/ChildID), data = HS_dat)
+\end{Sinput}
+\end{Schunk}
+
+The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference is slightly diminished but persistent over time.
 This result agrees with the findings of \cite{Demi09}.
+The code for the figure is omitted here for space considerations, because we want to focus on fitting and interpreting GLMM trees.
+Code for exact replication of the results presented here is provided in the Supplementary Materials.
 
@@
 \begin{subfigure}{.7\textwidth}
-\includegraphics{_MOB_paper-004}
+\includegraphics{_MOB_paper-008}
 \end{subfigure}
@@
  library("glmertree")
- HS_tree <- lmertree(PPVT ~ Program*Age | (1|ChildID) | AFTQ + Race + 
+ HS_tree <- lmertree(PPVT ~ Program*Age | (1|MotherID/ChildID) | AFTQ + Race + 
                        Income + Mom_edu_yrs + Mom_height, 
-                     data = HS_dat, cluster = ChildID, minsize = 250)
+                     data = HS_dat, cluster = MotherID, minsize = 250)
 \end{Sinput}
@@
 \caption{}
-\includegraphics{_MOB_paper-007}
+\includegraphics{_MOB_paper-011}
 
@@
 
-\includegraphics{_MOB_paper-008}
+\includegraphics{_MOB_paper-012}
 \label{fig:lmm_tree}
@@
 \midrule
-2 & None & 46.63 & 131.23\\
-2 & HS & 49.39 & 134.48\\
-5 & None & 50.55 & 144.09\\
-5 & HS & 55.15 & 146.43\\
-6 & None & 57.04 & 151.51\\
-6 & HS & 60.09 & 151.63\\
-7 & None & 62.10 & 155.74\\
-7 & HS & 66.36 & 163.32\\
+2 & None & 16.16 & 31.08\\
+2 & HS & 20.75 & 35.26\\
+5 & None & 16.58 & 33.64\\
+5 & HS & 23.42 & 38.74\\
+6 & None & 21.08 & 37.36\\
+6 & HS & 27.02 & 43.67\\
+7 & None & 25.61 & 42.80\\
+7 & HS & 27.75 & 45.84\\
 \bottomrule
@@
 \end{table}
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 \todo{x axis for age starts at 2?} and 18.
+To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
@@
 These results correspond to the conclusions of \cite{Demi09}.
-\todo[inline]{Generally the effects are not large but the example is very clear and I like that it is an intervention effect.
-I suggest to state early before the figure that the effects are not large in this example but we are interested in the group differences; also try to make figure bigger and wider to make slopes more visible; (Mirka): I think the main effect of the intervention and the interaction effect is not too small, but it is hard to see because the age effect is so pronounced in this example.
-Would it be an option to only use the intervention effect in the end nodes?
-}
 
-To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
+\todo[inline]{With different specification of the GLMM tree (no square root, parameter stability tests at mom level, random intercepts of children nested within mothers) the results have changed, still need to adjust description}.
+
+\todo[inline]{Marjolein add here comment that typically affect of age is typically strongest; take out next part, but comment on it in discussion: CV would be a technique to see how stable results are, cite something about CV in general}
+
+%To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
 We \edc{randomly} separated the 258 families in the dataset into ten equally-sized folds.
@@
 \midrule
-LMM tree & 221.022 & 55.727 & 2.7 & 0.813\\
-LMM & 258.233 & 71.451 & NA & 0.781\\
+LMM tree & 259.595 & 58.786 & 2.5 & 0.780\\
+LMM & 297.926 & 77.296 & NA & 0.747\\
 \bottomrule
@@
 \end{table}
-
-\todo[inline]{I am not sure this dataset is the best GLMM tree illustration, because visualized differences between subgroups are very small and perhaps the fixed-effect part is already too complex.
-Alternative: Do not focus on effect of Head Start, but use as partitioning variable.
-Then can also use data from more families (HS or none not needed), partition using child-level characteristics, and  model specification becomes simpler.
--- response Caro: I like it, see comment above: try to prime that effects are not large -- response Mirka: I think that's okay.
-Only if you had data at hand that is not longitudinal (rather students in classes/schools), with covariates on Level-1, I would opt for that because it doesn't have the Level-2 splitting issue.
-}
 
@@
 
+\todo[inline]{Comment on new example: I now use data from Bogen 20 of the SPISA dataset.
+I use the student subset, and complete cases on all variables.
+The story is less clear, but the tree is still stopped early, and we can display which items have dif.
+}
+
 \begin{itemize}
-\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet)
-\item take the subsample of students (only students), and a random subsample of 5000 respondents (seed 04102023)
+\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet; Bogen 20 (as in psychotree, but from the whole dataset, not only a subsample))
+\item take the subsample of students (only students); complete cases; $N = 9769$
 \item items could probably appear in Appendix??
-\begin{itemize}
+\begin{enumerate}
 \item Which painter created this painting? – Andy Warhol.
@@
     Age Gender                    Area
-445472  28   Male  Language & Culture
-167408  26 Female         Engineering
-422161  25   Male  Language & Culture
-350734  27   Male         Engineering
-570206  23   Male \nMedicine & Health
-265570  21 Female                Arts
+90   21   Male       Law and Economics
+163  25   Male          no Information
+402  22   Male Agriculture & Nutrition
+449  20   Male          no Information
+518  21 Female          no Information
+649  23   Male          no Information
 \end{Soutput}
@@
 \begin{Soutput}
-     i27 i28 i29 i30 i31 i32 i33 i34 i35
-[1,]   1   1   0   0   1   1   1   1   1
-[2,]   1   0   1   0   0   0   1   1   1
-[3,]   1   1   1   0   1   1   1   1   1
-[4,]   0   1   0   1   0   1   1   0   1
-[5,]   1   1   0   1   1   1   0   1   0
-[6,]   1   1   0   1   1   0   1   1   1
+     i28 i29 i30 i31 i32 i33 i34 i35 i36
+[1,]   1   1   0   1   0   1   1   1   1
+[2,]   1   1   1   1   0   1   1   0   1
+[3,]   0   1   0   1   1   0   1   1   0
+[4,]   0   1   1   1   1   1   1   1   1
+[5,]   1   0   1   0   0   1   1   1   1
+[6,]   1   1   1   1   0   1   1   1   1
 \end{Soutput}
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-016}
+\includegraphics{_MOB_paper-020}
+
+\todo[inline]{The are covariate split does not look pretty yet. But I will try to make it more pretty later on.. }
 
@@
 Iterative is recommended.
+Optionally, the user can specify the stopping criterion.
+In this case, tree growing is stopped when all items fall into category ``A''.
 
@@
                                    stopfun= stopfun_mantelhaenszel(
-                                     purification = "iterative"))
+                                     purification = "iterative",
+                                     stopcrit = "A"))
 \end{Sinput}
@@
 For technical reasons, the information about the effect size and ETS classification are not saved in the tree object itself but have to be added afterwards using the \texttt{add\_mantelhaenszel} function.
-The information can then be accessed from the object using \texttt{\$info\$mantelhaenszel}
 
@@
                                             purification = "iterative")
- Raschtree_MH_culture$info$mantelhaenszel
 \end{Sinput}
@@
 
-\todo[inline]{not clear what this information is necessary/useful for? since it is not displayed, leave this out?}
-
+After we have added the information about the Mantel-Haenszel effect size to the Rasch tree, we can now visualize the Rasch tree together with the effect size measure.
 If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.
 In addition, we see the number of items classified as A, B, or C in each node.
-For instance, we see that one item is classified as B in Node 1, and no item as C.
+For instance, we see that one item is classified in category B in Node 1, and 4 items as C.
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-021}
+\includegraphics{_MOB_paper-025}
 
-We can also color the items in the end node profiles according to a split in the tree.
-Here, we see that for Node 1 (split on the covariate Area), item 2 shows DIF in category B.
-Item 2 seems to be easier (y-axis shows item easiness) in certain areas compared to Language, Culture, Medicine, Health, Arts.
+We can also color the items in the end node profiles according to a split in the tree. Here, we see that for Node 1 (split on the covariate Gender), items 1, 6, 7, 8, 9 show DIF in categories B or C.
+
+\begin{itemize}
+\item 1 painting by Andy Warhol
+\item 6: novel by Daniel Kehlmann: Die Vermessung der Welt
+\item 7: city of the Buddenbrooks: Lübeck
+\item 8: city with building: Paris
+\item 9: opera not by Mozart: Aida
+\end{itemize}
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-022}
+\includegraphics{_MOB_paper-026}
 
-We can also color by Node 3 (also split on covariate Area). Also here, Item 2 shows DIF. Similarly, item 2 seems to be more difficult (less easy) for respondents who study Arts.
+We can also color by Node 2 (split on covariate Age).
+Also here, Item 1 shows DIF in category B.
+It seems to be easier for younger respondents ($\leq 24$) compared to older respondents (painting by Andy Warhol).
 
@@
 \begin{Sinput}
- plot(Raschtree_MH_culture, color_by_node = 3)
+ plot(Raschtree_MH_culture, color_by_node = 2)
 \end{Sinput}
 \end{Schunk}
-\includegraphics{_MOB_paper-023}
+\includegraphics{_MOB_paper-027}
 
-This is interesting, because it is the culture scale.
-Why should people who study Arts have more problems to solve the item?
-When we look at the item, it probably makes sense: the items asks what four buildings have in common (Allianz-Arena München, Tate Modern NY, Olympia-Stadion Peking, Elbphilharmonie Hamburg).
-Two of these buildings are sports stadiums.
-So probably respondents who are interested in sports, and know these buildings well have an advantage on the item (independent of the level of knowledge about culture).
+For Node 4 (also split on covariate Age), Items 1 and 9 show DIF, such that Item 1 and 9 are easier for younger respondents (Age $\leq 29$).
+(Item 1: Painting by Andy Warhol; Item 9: Opera not by Mozart: Aida)
+
+\begin{Schunk}
+\begin{Sinput}
+ plot(Raschtree_MH_culture, color_by_node = 4)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{_MOB_paper-028}
 
@@
 
+The information about the Mantel-Haenszel effect size measure for each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
+it contains the information on classification of each item in e ach node, the computed Mantel-Haenszel effect size measure, the type of purification, and if purification was iterative how many iterations were realized.
+
+\begin{Schunk}
+\begin{Sinput}
+ Raschtree_MH_culture$info$mantelhaenszel
+\end{Sinput}
+\begin{Soutput}
+$classification
+      node4 node2 node1
+item1 "B"   "B"   "C"  
+item2 "A"   "A"   "A"  
+item3 "A"   "A"   "A"  
+item4 "A"   "A"   "A"  
+item5 "A"   "A"   "A"  
+item6 "A"   "A"   "C"  
+item7 "A"   "A"   "C"  
+item8 "A"   "A"   "B"  
+item9 "B"   "A"   "C"
+
+$mantelhaenszel
+            node4       node2        node1
+item1 -1.53377668 -1.03559992 -1.802977840
+item2  0.72945873 -0.14242027  0.377084860
+item3 -0.35893502  0.51100000  0.127575191
+item4 -0.05897255  0.40879442 -0.007385896
+item5  0.41615400 -0.08996266  0.128463797
+item6  0.42418247 -0.16529252 -1.875191790
+item7 -0.87987154 -0.57949560 -1.620265527
+item8  0.69240041  0.54735808 -1.083093749
+item9 -1.00725807 -0.39502640 -1.798520587
+
+$purification
+      node4       node2       node1 
+"iterative" "iterative" "iterative"
+
+$purification_counter
+node4 node2 node1 
+    1     1     2 
+\end{Soutput}
+\end{Schunk}
+
 \FloatBarrier
@@
 In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.
-The main advantage of MOB is that it can detect groups of persons with different model parameters in a data driven way.
-This makes it more flexible for detecting differences that were not hypothesized by the researchers.
-For example in DIF analysis, it is often the case that any obvious sources of DIF have already been avoided by the content expernts in the item creation phase.
+The main advantage of MOB is that it can detect groups of persons with different model parameters in a data-driven manner.
+This makes it more flexible for detecting differences that were not hypothesized by the researcher.
+For example in DIF analysis, it is often the case that obvious sources of DIF have already been avoided by the content experts during item creation.
 Any remaining DIF is unexpected.
 Therefore, any DIF detection approach that relies on the researchers correctly specifying the exact groups of persons that exhibit DIF can miss DIF if it is associated with other (combinations of) covariates or other cutpoints than the ones investigated.
-In this sense, a more exploratory approach like a Rasch tree has a higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
-The same holds for parameter differences in mixed-effects models, where the substantial hypothesis are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
+In this sense, the more exploratory approach of Rasch trees has higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
+
+The same holds for parameter differences in mixed-effects models, where the substantial hypotheses are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
 \todo{Marjolein add if this makes any sense here}
 
-\todo[inline]{Write summary of findings.}
+For our two application examples, we could show that MOB can detect ..., performs automated variable selection ...
 
-For our two application examples, we could show that MOB can detect ..., performs automated variable selection \todo[inline]{or did you, Marjolein, mean repeat substantive findings?
-not very exciting for SPISA}
+\subsection{Limitations}
+
+Due to the exploratory nature of MOB, it does not allow for hypothesis tests, as there is no valid way to account for the exploratory searching of the subgroups.
+Thus, even if detected subgroups are substantively meaningful or differences are large, if researchers want to ascertain statistical significance of the subgroup differences, this should be done on new data using confirmatory techniques.
 
 \todo[inline]{Mention shortcomings.}
-
-not possible to detect certain patterns: XOR problem, known for all trees
-
-stability of trees, small changes in data can lead to quite different looking tree.
-The tree structure should always be interpreted jointly for all variables: rather than considering the first splitting variable to be the most important one, it is actually the combination of all splitting variables that creates the groups.\todo{in interpretation of trees in examples: do this, point out that it is interactions -- Caro also add in intro?}.
-The \texttt{stabletree} R package also provides descriptive statistics and graphics that help judge the stability of tree and MOB results \citep{PhiRusHor:2018:JoCaGS,PhiZeiStr:2016}.
 
@@
 sample size dependence, similar ideas like MH trees also possible for other MOB?
+\todo[inline]{possibly Mirka add something about visual interpretation?
+because effect sizes in mixed effects models are not very clear}
+
+The Rasch tree method has been extended to include an effect size measure that can stop the tree from growing if the effect size is non-substantial, but also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
+At the same time, effect sizes are less straightforward to calculate and interpret in linear mixed models, and therewith in the linear mixed effects model tree.
+A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
+These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s).
+A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.
+\todo[inline]{This is my (Mirka) suggestion for the interpretation techniques. Please feel free to add, edit, delete, however you prefer!}
 
 The Mantel-Haenszel effect size measure for DIF will be integrated in the \texttt{psychotools} package.
+
+\todo{Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
+In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}
+
+\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
+In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}
 
@@
 
-\section{Appendix A: SDQ-II items administered in the ECLS}
+\section{Appendix A: Association between Age and PPVT scores before and after transformation}
 \label{sec:AppendixA}
 
-\todo{Fix formatting.}
-
-How true is each of these about you? (1="not at all true"; 2="a little bit true"; 3="mostly true" or 4="very true")
-
-\begin{itemize}
-
-\item C7MTHBST (math): Math is one of my best subjects.
-
-\item C7ANGRY (internalizing): I feel angry when I have trouble learning.
-
-\item C7LIKRD (reading): I like reading.
-
-\item C7WRYTST (internalizing): I worry about taking tests.
-
-\item C7MTHGD (math): I get good grades in math.
-
-\item C7LONLY (internalizing): I often feel lonely.
-
-\item C7ENGBST (reading): English is one of my best subjects.
-
-\item C7SAD (internalizing): I feel sad a lot of the time.
-
-\item C7LIKMTH (math): I like math.
-
-\item C7WRYWEL (internalizing): I worry about doing well in school.
-
-\item C7ENJRD (reading): I enjoy doing work in reading.
-
-\item C7WRYFIN (internalizing): I worry about finishing my work.
-
-\item C7ENJMTH (math): I enjoy doing work in math.
-
-\item C7WRYHNG (internalizing): I worry about having someone to hang out with at school.
-
-\item C7GRDENG (reading): I get good grades in English.
-
-\item C7ASHAME (internalizing): I feel ashamed when I make mistakes at school.
-
-\end{itemize}
+\begin{figure}
+\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
+\includegraphics{_MOB_paper-030}
+\end{figure}
 

### 2023-10-20 _MOB_paper.Rnw
text MOB intro, fig 1 and 2 in one

@@
 
-\title{One Model May Not Fit All: Subgroup Detection Using MOB}
+\title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
 \shorttitle{shorttitle}
@@
 In this tutorial paper, we provide an introduction to the general MOB framework.
-In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory models.
+In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory (IRT) models.
 In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
@@
 In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model.
-We thus assume that the psychological test score(s) of interest \todo{(Mirka: I like the term instrument to distinguish it from a statistical test) (Marjolein: I hope 'psychological test score' clears up possible confusion)} have already been validated.
+We thus assume that the psychological test score(s) of interest have already been validated.
+%\todo{(Caro: I have kept psychological test score, because instrument to me sounds more clinical)(Mirka: I like the term instrument to distinguish it from a statistical test) (Marjolein: I hope 'psychological test score' clears up possible confusion)}
 
-For application example 2\todo{add link to section?
-not sure how to do}, we go back in the research process to the point where a new test has been administered to a validation sample.
+For application example 2, we go back in the research process to the point where a new psychological test has been administered to a validation sample.
 Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model.
@@
 Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}.
-An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
-Mixture modelling aims at identifying latent classes of persons with different properties.
-It can also be combined with covariates.
-A comparison of both approaches is given by \citet{FriStrZei:2014}.
-\todo{Last 5 sentences seem distracting here.
-Omit or move to discussion?}
 
@@
 
-\begin{figure}
-\caption{Univariate distribution of the response variable, with global mean indicated by triangle.}
-\begin{subfigure}{.5\textwidth}
-<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
-library("colorspace")
+The main rationale of MOB is that one global model may not fit all observations in a dataset equally well.
+In many studies, additional covariates may be available.
+It may then be possible to uncover subgroups defined by these covariates, and obtain better-fitting models in each of those subgroups \citep{ZeilyHoth08}.
+
+<<eval = TRUE, echo = FALSE>>=
 set.seed(3)
@@
 toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
-boxplot(toy_data$y, cex = .7)
-points(x = 1, y = mean(toy_data$y), col = rainbow_hcl(1), pch = 17)
+toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
 @
-\end{subfigure}
-\label{fig:global}
-\end{figure}
 
-The main rationale of MOB is that one global model may not fit all observations in a dataset equally well.
-In many studies, additional covariates may be available.
-It may then be possible to uncover subgroups defined by these covariates, and obtain better-fitting models in each of those subgroups \citep{ZeilyHoth08}.
-
-We illustrate the idea using a very simple toy dataset, comprising \Sexpr{nrow(toy_data)} observations, and four variables: A continuous response variable $y$, and three possible partitioning (or predictor) variables $x_1$, $x_2$ and $x_3$.
-To keep the example simple, we apply MOB to a very simply global model, which comprises only an intercept.
-Figure~\ref{fig:global} shows the distribution of $y$ in a boxplot, with the global intercept depicted as a triangle.
+We illustrate the idea using a very simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables.
+%note from Caro: I woul call them partitioning variables or covariates, but not predictors, to save the term predictors for those x-variables that are part of the parametric model
+To keep the example simple, we apply MOB to a very simple global model, which comprises only an intercept.
+It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
+Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot, with the global intercept indicated as a triangle.
 Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
+%note from Caro: I found Fig 1 containing only the boxplot irritating and have moved it into one panel with the tree
 
@@
 
-We applied MOB to the observations in Figure~\ref{fig:global}, specifying $x_1$, $x_2$ and $x_3$ as the additional covariates.
-The resulting tree is shown in Figure~\ref{fig:tree}.
-In the root node, which contains all observations, structural change tests were performed for each of the three covariates (step 2).
-The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected for splitting (step 3).
-The $p$ value resulting from the structural change test for $x_1$ is depicted in this root node.
-Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible between the groups.
-In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with instability in this subgroup.
-In the right daughter node, there was significant instability associated with $x_2$, a categorical covariate.
-Again, the cutpoint was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible.
-The subgroup-specific model parameters are presented in end nodes at the bottom of the tree.
+We applied MOB to the observations in Figure~\ref{fig:toy}, left, specifying $x_1$, $x_2$ and $x_3$ as potential splitting variables.
+The resulting tree is shown in Figure~\ref{fig:toy}, right.
+In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2).
+The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected first for splitting (step 3).
+The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node.
+Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences.
+In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate.
+Again, the cutpoint in $x_2$ was selected so that the two resulting subgroups exhibit strongest parameter differences.
+In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4).
+The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting.
+The subgroup-specific distributions of the response variable are presented in end nodes at the bottom of the tree.
+The fact that Figure~\ref{fig:toy}, right, displays more than one end node confirms that one global model for all observations cannot appropriately capture the pattern in the data. 
+Figure~\ref{fig:toy}, right, also shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
+This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.
 
 \begin{figure}
-\caption{Decision tree with group-specific parameters in the terminal nodes.}
-\begin{subfigure}{.5\textwidth}
-<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
+\caption{Left: Univariate distribution of the response variable, with global mean indicated by triangle. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
+\begin{subfigure}[][][t]{.4\textwidth}
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=4.5, height=7>>=
+library("colorspace")
+boxplot(toy_data$y, cex = .7)
+points(x = 1, y = mean(toy_data$y), col = rainbow_hcl(1), pch = 17, ylim = c(1,14))
+@
+\end{subfigure}
+\begin{subfigure}[][][b]{.6\textwidth}
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=6, height=5>>=
 library("partykit")
-toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
 tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
-plot(tree, gp = gpar(cex = .7))
+plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
 @
-~
-<<eval = TRUE, echo = FALSE, fig=TRUE, width=5, height=5>>=
+<<eval = FALSE, echo = FALSE, fig=TRUE, width=5, height=5>>=
 plot(tree, gp = gpar(cex = .7), type = "simple")
@@
 \end{subfigure}
-\label{fig:tree}
+\label{fig:toy}
 \end{figure}
@@
 
-There are two main differences between MOB and and traditional classification and regression tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993}.
-The first difference relates to the separation of variable and cutpoint selection.
-The traditional algorithms performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias.
+%note from Caro: I did not like the formulation "There are two main differences between MOB and and traditional classification and regression tree algorithms" because readers might misinterpret it and think that only MOB, not trees, can be unbiased and use significance tests, while the distinction is really between traditional and modern algorithms for both trees and mob, changed into: characteristics
+There are three further important characteristics of tree and MOB algorithms that we would like to mention here: The first characteristic is the type of variable and cutpoint selection an algorithm employs.
+Traditional classification and regression tree algorithms, like those of \citet{Breetal:1984} and \citet{Qui:1993}, performed variable and cutpoint selection in one step, which leads to an undesirable behavior called variable selection bias.
 That is, the traditional algorithms prefer variables offering more possible cutpoints in the selection process -- regardless of their true information content.
 An algorithm that still has this problem is, for example, the \texttt{rpart} algorithm in R, based on the original CART algorithm by \citet{Breetal:1984}.
-More recent algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}.
+More modern algorithms for classification and regression trees have solved this problem and offer unbiased variable selection, such as \texttt{QUEST} \citep{LohShi:1997}, which is available in SPSS, and \texttt{ctree} \citep{Hotetal:2006}, which is available in \texttt{R} in packages \texttt{party} and \texttt{partykit} \citep{partykit:pkg}.
 The latter forms the basis for all MOB approaches presented in this paper.
@@
 
-The second difference relates to the way MOB stops splitting: It uses a criterion of statistical significance.
+The second characteristic relates to the way a tree or MOB algorithm stops splitting: Modern algorithms for classification and regression trees and MOB use a criterion of statistical significance to stop splitting.
 Once there are no more covariates that show a significant structural change in any node, splitting is halted.
-In this way, the MOB algorithm selects only those variables that are relevant for distinguishing the groups, i.e., it performs automatic variable selection.
+In this way, the  algorithm selects only those partitioning variables that are relevant for distinguishing the groups (i.e., it performs automatic variable selection, as illustrated in Figure~\ref{fig:toy}, right).
 Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected.
-While the traditional algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (i.e., \textit{pruning}), the MOB framework employs significance tests, but it can also (partly) use effect size measures, as we will see in application example 2 \todo{include link}.
+While traditional tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (so called \textit{pruning}), modern tree and MOB algorithms employ significance tests as stopping criteria (and some can also use effect size measures, as we will see in application example 2).
 This allows for stopping tree growing as soon as no significant differences can be detected anymore.
@@
 
-An important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner.
+The third important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner.
 This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
 While there are phases in psychological and educational research where it is very important to specify hypotheses a-priori and test them in a confirmatory manner, in early stages of research exploratory methods are an important addition to the statistical toolbox for researchers.
-Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of covariates or possible partitioning variables and to choose the settings of the MOB algorithm.
+Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of possible partitioning variables and to choose the settings of the MOB algorithm.
 The current paper aims to provide some guidance.
+%note from Caro: it said include link in several places when referring to example 2, but I don't think this is necessary, because currently sections are not numbered anyway and the title example 2 is clear
 
@@
 \todo{Omit last sentence or move to discussion?
-I think better to focus on DIF only here.} Another assumption that is particularly crucial for the comparability of test scores between groups is measurement invariance.
+I think better to focus on DIF only here.
+Response Caro: would rather keep it here, but have increased importance of measurement invariance in next sentence.} A particularly crucial assumption for the comparability of test scores between groups is measurement invariance.
 Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
 
-MOB can be used to detect DIF by means of searching for covariates that display structural change in the item parameters of the measurement model\todo{possibly confusing: covariates are supplied by user, not searched for.
-Suggestion: by means of testing whether the parameters of the measurement model exhibit significant instability with respect to (sets of) covariates?}.
-For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15} \todo{should we cite the package too or only the paper?
-(Marjolein: I think software should only be mentioned in the applications.)} The usage of this function will be illustrated in application example 2.
-We will see that, just like for the mixed-effects model trees\todo{use same names for methods everywhere}, we need to specify which variables are part of the parametric model.
+MOB can be used to detect DIF by means of testing whether the item parameters of the measurement model exhibit significant instability with respect to (combinations of) covariates.
+For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15}.
+The usage of this function will be illustrated in application example 2.
+We will see that, just like for  GLMM trees, we need to specify which variables are part of the parametric model.
 In the case of the Rasch model, this will be the test items.
-Moreover, we need to specify, which variables are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.
+Moreover, we need to specify, which covariates are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.
 
@@
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citep[see][for details]{HenDebStr:2023:EPM} \todo{Caro can add more text on achoring, incl.
-word anchoring, here to prepare example 2 - only shortly mention here and explain in more detail later in the example (will become more clear in example)}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+Together with a purification step and anchoring \citep[see][and application example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
-\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
+%\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
 For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc.
@@
 \end{figure}%
-
-\todo{code for figure not displayed on purpose? why only those particular x axis ticks?}
 
@@
 
+\todo{next paragraph moved to here, where exactly?}
+An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
+Mixture modelling aims at identifying latent classes of persons with different properties.
+It can also be combined with covariates.
+A comparison of both approaches is given by \citet{FriStrZei:2014}.
+
 In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.

### 2023-10-20 _MOB_paper.tex
text MOB intro, fig 1 and 2 in one

@@
 Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}.
-An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
-Mixture modelling aims at identifying latent classes of persons with different properties.
-It can also be combined with covariates.
-A comparison of both approaches is given by \citet{FriStrZei:2014}.
-\todo{Last 5 sentences seem distracting here.
-Omit or move to discussion?}
 
 \section{The MOB Algorithm}
-
-\begin{figure}
-\caption{Univariate distribution of the response variable, with global mean indicated by triangle.}
-\begin{subfigure}{.5\textwidth}
-\includegraphics{_MOB_paper-002}
-\end{subfigure}
-\label{fig:global}
-\end{figure}
 
@@
 
-We illustrate the idea using a very simple toy dataset, comprising 250 observations, and four variables: A continuous response variable $y$, and three possible partitioning (or predictor) variables $x_1$, $x_2$ and $x_3$.
-To keep the example simple, we apply MOB to a very simply global model, which comprises only an intercept.
-Figure~\ref{fig:global} shows the distribution of $y$ in a boxplot, with the global intercept depicted as a triangle.
+We illustrate the idea using a very simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables.
+%note from Caro: I woul call them partitioning variables or covariates, but not predictors, to save the term predictors for those x-variables that are part of the parametric model
+To keep the example simple, we apply MOB to a very simple global model, which comprises only an intercept.
+It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
+Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot, with the global intercept indicated as a triangle.
 Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
+%note from Caro: I found Fig 1 containing only the boxplot irritating and have moved it into one panel with the tree
 
@@
 
-We applied MOB to the observations in Figure~\ref{fig:global}, specifying $x_1$, $x_2$ and $x_3$ as the additional covariates.
-The resulting tree is shown in Figure~\ref{fig:tree}.
-In the root node, which contains all observations, structural change tests were performed for each of the three covariates (step 2).
-The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected for splitting (step 3).
-The $p$ value resulting from the structural change test for $x_1$ is depicted in this root node.
-Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible between the groups.
-In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with instability in this subgroup.
-In the right daughter node, there was significant instability associated with $x_2$, a categorical covariate.
-Again, the cutpoint was selected so that the two resulting subgroups exhibit parameters that differ as strongly as possible.
-The subgroup-specific model parameters are presented in end nodes at the bottom of the tree.
+We applied MOB to the observations in Figure~\ref{fig:toy}, left, specifying $x_1$, $x_2$ and $x_3$ as potential splitting variables.
+The resulting tree is shown in Figure~\ref{fig:toy}, right.
+In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2).
+The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected first for splitting (step 3).
+The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node.
+Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences.
+In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate.
+Again, the cutpoint in $x_2$ was selected so that the two resulting subgroups exhibit strongest parameter differences.
+In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4).
+The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting.
+The subgroup-specific distributions of the response variable are presented in end nodes at the bottom of the tree.
+The fact that Figure~\ref{fig:toy}, right, displays more than one end node confirms that one global model for all observations cannot appropriately capture the pattern in the data. 
+Figure~\ref{fig:toy}, right, also shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
+This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.
 
 \begin{figure}
-\caption{Decision tree with group-specific parameters in the terminal nodes.}
-\begin{subfigure}{.5\textwidth}
+\caption{Left: Univariate distribution of the response variable, with global mean indicated by triangle. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
+\begin{subfigure}[][][t]{.4\textwidth}
 \includegraphics{_MOB_paper-003}
-~
+\end{subfigure}
+\begin{subfigure}[][][b]{.6\textwidth}
 \includegraphics{_MOB_paper-004}
 \end{subfigure}
-\label{fig:tree}
+\label{fig:toy}
 \end{figure}
@@
 \begin{subfigure}{.7\textwidth}
-\includegraphics{_MOB_paper-008}
+\includegraphics{_MOB_paper-009}
 \end{subfigure}
@@
 \caption{}
-\includegraphics{_MOB_paper-011}
+\includegraphics{_MOB_paper-012}
 
@@
 
-\includegraphics{_MOB_paper-012}
+\includegraphics{_MOB_paper-013}
 \label{fig:lmm_tree}
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-020}
+\includegraphics{_MOB_paper-021}
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-025}
+\includegraphics{_MOB_paper-026}
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-026}
+\includegraphics{_MOB_paper-027}
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-027}
+\includegraphics{_MOB_paper-028}
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-028}
+\includegraphics{_MOB_paper-029}
 
@@
 \section{Discussion}
+
+\todo{next paragraph moved to here, where exactly?}
+An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
+Mixture modelling aims at identifying latent classes of persons with different properties.
+It can also be combined with covariates.
+A comparison of both approaches is given by \citet{FriStrZei:2014}.
 
@@
 \caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
-\includegraphics{_MOB_paper-030}
+\includegraphics{_MOB_paper-031}
 \end{figure}

### 2023-10-20 Raschtree_MH.Rnw
text Rasch and MH tree

--- 2023-10-11 Raschtree_MH.Rnw
@@
-\begin{itemize}
-\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet; Bogen 20 (as in psychotree, but from the whole dataset, not only a subsample))
-\item take the subsample of students (only students); complete cases; $N = 9769$
-\item items could probably appear in Appendix??
-\begin{enumerate}
-\item Which painter created this painting? – Andy Warhol.
-\item What do these four buildings have in common? – All four were designed by the same architects.
-\item Roman numbers: What is the meaning of CLVI? – 156.
-\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
-\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
-\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
-\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
-\item In which city is this building located? – Paris.
-\item Which one of the following operas is not by Mozart? – Aida.
-\end{enumerate}
-\end{itemize}
-
-We load the data and inspect the first rows.
-there are several covariates: age (continuous), gender (male, female, missing), area (Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, Sports) and 9 items from the culture scale (true/false 0/1).
-
-@Caro: I excluded the covariates student and occupation, because I took the student subsample;
-
 <<echo = FALSE>>=
@@
 covar_data$Occupation <- NULL
+covar_data$Area <- sub("\n", "", covar_data$Area)
+dat_SPISA <- covar_data
+dat_SPISA$culture <- culture_scale
+save(dat_SPISA, file = "dat_SPISA.Rda")
 @
 
+We will illustrate the usage of Rasch trees and the Mantel-Haenszel effect size measure by means of a dataset from \citet{SPISA:book}. An online general knowledge quiz was conducted by the German
+weekly news magazine SPIEGEL.
+Below we will use the abbreviation \texttt{SPISA} for the name of the data set, because the quiz was termed ``students' PISA'' by the magazine, but it is not related to the ``real'' PISA study by the OECD.
+The data set contains the quiz results as well as
+sociodemographic data.
+In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $\Sexpr{nrow(covar_data)}$ complete cases of university students for illustration.
+The wording of the items is provided in Appendix~\ref{sec:AppendixB}.
+
+Three possible partitioning variables are considered: \texttt{Gender} (in this data set coded only as male, female, or missing), \texttt{Age} (continuous), and \texttt{Area} (Area of study: Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, or Sports).
+The nine items from the culture scale are scored with 0 (incorrect answer) or 1 (corrct answer).
+The quiz data have been prepared in a format where the nine items of the culture scale are not stored as individual columns of the data set, but the complete matrix of item responses is stored like a single variable \citep[see][]{raschtree:vignette} under the name \texttt{culture}.
+This means that we can access the item responses for all items jointly using the name \texttt{culture} in the following R commands.
+The first few rows of the data set look like this:
+
+\todo[inline]{I suggest that we publish the data set Mirka created with the paper, but don't show her data preprocessing steps (which are in the echo = FALSE chunk above); I usually use load for Rda files, Marjolein uses readRDS, is there an advantage?
+use the same in both examples}
+
 <<>>=
-head(covar_data)
-head(culture_scale)
+load("dat_SPISA.Rda")
+head(dat_SPISA, 3)
 @
 
-We create a dataset \texttt{dat\_SPISA} that contains the data of the covariate and the item responses from the culture scale.
-By assigning the whole scale to the dataset using \texttt{\$}, we can later access item responses from all items using \texttt{\$culture}.
-<<>>=
-dat_SPISA <- covar_data
-dat_SPISA$culture <- culture_scale
-@
+The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that in the following we will fit Rasch trees by means of the function \texttt{raschtree} from the R package \texttt{psychotree}.
+However, model-based recursive partitioning is also available in \texttt{psychotree} for other IRT models.
 
-The Rasch tree can be fitted with the function \texttt{raschtree} from the R package \texttt{psychotree}.
-We use the typical formula syntax with the item responses of the culture scale on the left hand side of $\sim$, and the covariates on the right hand side.
+The model which we would like to test for parameter stability in this example, the Rasch model, is based only on the matrix of item responses \texttt{culture}.
+It does not contain predictor variables of its own.
+Therefore, the formula syntax for the \texttt{raschtree} function looks simpler than that for the \texttt{lmertree} function in application example 1: On the left hand side of the $\sim$ symbol, we provide the item response matrix \texttt{culture} for the Rasch model.
+On the right hand side of the $\sim$ symbol, we provide the possible partitioning variables \texttt{Gender}, \texttt{Age}, and \texttt{Area}.
 
@@
 if(!file.exists("SPISA_example/Raschtree.rda")){
-  library(psychotree)
+  library("psychotree")
   Raschtree_culture <- raschtree(culture ~  Gender + Age + Area, 
@@
 <<eval = FALSE>>=
-library(psychotree)
+library("psychotree")
 Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
@@
 
-We can plot the Raschtree using the plot function.
-But it is very big, lots of splits, we don't know whether the splits are due to substantial differences in item difficulty parameters, because the item difficulty profiles in the end nodes are hard to interpret/compare.
+Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function.
+The item parameters for the nine culture items are displayed in the end nodes of the tree.
+However, it is hard to interpret the tree structure because such a large number of splits was created.
+Moreover, it is hard to tell whether all splits are due to substantial amounts of DIF, or whether some of the splits are due to small DIF effects that only became statistically significant due to the large sample size.
 
@@
 
-Mantel-Haenszel trees use the Mantel-Haenszel effect size measure for DIF.
-it has three categories (A: negligible, B: moderate, C: large).
-If none of the items has DIF in category B or C (default, but can also be changed so that DIF has to be C), the tree is stopped from growing.
+\todo[inline]{The are covariate split does not look pretty yet. But I [Mirka] will try to make it more pretty later on.. (reply Caro: yes if possible, and/or make it wider in the pdf?)}
 
-The software must be installed from github, but will (probably) be implemented in the psychotree package some time in the future
+To assess whether DIF effects are substantial or negligible, it is possible to use the Mantel-Haenszel DIF effect size measure as an additional stopping criterion.
+It has three categories (A: negligible, B: moderate, C: large).
+If none of the items shows DIF in category B or C\footnote{This is the default setting.
+It can also be changed so that splitting is stopped if no item shows DIF in category C.}, the tree is stopped from growing.
+At the moment, the R functions for the Mantel-Haenszel stopping criterion and additional visualizations based on the Mantel-Haenszel effect size are available from github.
+They will also be made available as part of the \texttt{psychotree} package in the future.
+
+In order to install the functions from the author's github page and make them available in the current R session, use the following two commands.
+
+\todo[inline]{Caro an Mirka: don't we need library devtools?}
 
@@
 devtools::install_github("mirka-henninger/raschtreeMH")
-library(raschtreeMH)
+library("raschtreeMH")
 @
 
-Syntax is very similar to the classical Rasch tree, but an additional argument \texttt{stopfun}, where the mantelhaneszel stopping function can be selected (but also own stopping functions can be provided; see Henninger et al., 2023).
-The user also has to indicate what kind of purification strategy should be used (none, 2step, iterative).
-Iterative is recommended.
-Optionally, the user can specify the stopping criterion.
-In this case, tree growing is stopped when all items fall into category ``A''.
+We can use a syntax very similar to that of the original \texttt{raschtree} function, but now we also provide the additional argument \texttt{stopfun}.
+We want to use the Mantel-Haenszel stopping function (\texttt{stopfun\_mantelhaenszel}) together with iterative purification.
+????
+and DIF category A as the stopping criterion.
+\todo{Caro an Mirka: oben steht das wäre eh der default, kann das stopcrit A raus?
+dann hier im Text und im code} It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
+
+For the purification strategy, there are three options (\texttt{none}, \texttt{2step}, and \texttt{iterative}).
+Purification means that items which have already been diagnosed with DIF are taken out of the set of so called anchor items.
+Anchor items are used for aligning the scales of different groups of persons for testing further items for DIF.
+It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the refereces therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
+In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2.
+When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
+\todo{Mirka bitte diesen ganzen Absatz besonders gut korrekturlesen}
+The final set of anchor items for each group comparison, which results from the purification, is also used in the tree plots with color coding for DIF items displayed below.
 
@@
 
-For technical reasons, the information about the effect size and ETS classification are not saved in the tree object itself but have to be added afterwards using the \texttt{add\_mantelhaenszel} function.
+For technical reasons, the information about the Mantel-Haenszel effect size and classification is not saved in the tree object itself, but has to be added afterwards using the \texttt{add\_mantelhaenszel} function.
 
@@
 
-After we have added the information about the Mantel-Haenszel effect size to the Rasch tree, we can now visualize the Rasch tree together with the effect size measure.
-If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.
-In addition, we see the number of items classified as A, B, or C in each node.
-For instance, we see that one item is classified in category B in Node 1, and 4 items as C.
+After we have added this information to the Rasch tree, we can now visualize it with additional features.
+When we now plot the Rasch tree with Mantel-Haenszel stopping by means of the \texttt{plot} function, we see that the tree is much more concise with a lower number of splits.
+In addition, in each node the number of items classified as A, B, or C is displayed.
+For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and 4 items are classified in category C (large DIF).
 
@@
 
-We can also color the items in the end node profiles according to a split in the tree. Here, we see that for Node 1 (split on the covariate Gender), items 1, 6, 7, 8, 9 show DIF in categories B or C.
+We can also color the DIF items in the end node profiles according to a particular split in the tree.
+Here, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
+As a reminder, the content of these items was:
 
 \begin{itemize}
-\item 1 painting by Andy Warhol
-\item 6: novel by Daniel Kehlmann: Die Vermessung der Welt
-\item 7: city of the Buddenbrooks: Lübeck
-\item 8: city with building: Paris
-\item 9: opera not by Mozart: Aida
+\item 1: Painting by Andy Warhol
+\item 6: Novel by Daniel Kehlmann: Die Vermessung der Welt
+\item 7: City of the Buddenbrooks: Lübeck
+\item 8: City with building: Paris
+\item 9: Opera not by Mozart: Aida
 \end{itemize}
@@
 
-We can also color by Node 2 (split on covariate Age).
-Also here, Item 1 shows DIF in category B.
-It seems to be easier for younger respondents ($\leq 24$) compared to older respondents (painting by Andy Warhol).
+These items are (relative to the anchor items for this group comparison) easier\todo{double check Mirka} to answer for participants with female or missing gender.
+
+\todo[inline]{Caro an Mirka: ich nehme an dass hier hohe item parameter bedeuten schwierig, niedrige leicht, dann wären diese items fur Frauen leichter, du hast es unten aber glaub genau andersrum geschrieben - evtl.
+weil im original Raschtree paper oder irgendwo mal easiness statt difficulty abgetragen war?}
+
+We can also color the DIF items according to node 2 (split in the covariate \texttt{Age}).
+Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
+It is (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult} to answer for younger respondents (age $\leq 24$) compared to older respondents.
 
@@
 
-For Node 4 (also split on covariate Age), Items 1 and 9 show DIF, such that Item 1 and 9 are easier for younger respondents (Age $\leq 29$).
-(Item 1: Painting by Andy Warhol; Item 9: Opera not by Mozart: Aida)
+For node 4 (another split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult to answer} for younger respondents (age $\leq 29$).
 
@@
 
-\todo[inline]{Maybe a small note on anchoring?
-Item difficulty profiles are anchoring on the items that do not show DIF.
-This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
+\todo[inline]{my (Caros) feeling is that two colored trees are enough, would leave out one of the colored trees based on age, and say above: We can also color the DIF items in the end node profiles according to a particular split in the tree.
+We will illustrate this with two examples.
+First, we can see that for node 1...}
 
-The information about the Mantel-Haenszel effect size measure for each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
-it contains the information on classification of each item in e ach node, the computed Mantel-Haenszel effect size measure, the type of purification, and if purification was iterative how many iterations were realized.
+Detailed information about the Mantel-Haenszel criterion for each item in each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
+It contains the DIF classification and the value of the Mantel-Haenszel effect size measure for each item in each node, as well as the type of purification that was employed in each node, and in the case of iterative purification how many iterations were conducted.
 
@@
 
-@Caro: for instance, from the signs we can see that items are less easy (i.e.
-more difficult) for females compared to men; it is probably easier to see that from the sign compared to the item difficulty profiles.
+The displayed information again shows that some items show only negligible DIF of category A in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
+The values of the Mantel-Haenszel effect size measure are negative when \todo{Caro an Mirka: ich rate hier völlig aus dem nichts, bitte korrigieren} when the item difficulty is lower in the right daughter node than in the left daughter node????.
+For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier\todo{double check Mirka} to answer (relative to the anchor items for this group comparison) for participants with female or missing gender, which correspond to the right daughter node of node 1.
+Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.
+However, this item has a smaller DIF effect size.
+Thus it falls into DIF category A (negligible) and was not relevant for the split in \texttt{Gender}.

### 2023-10-20 _MOB_paper.Rnw
text Rasch and MH tree

@@
 
-\todo{next paragraph moved to here, where exactly?}
-An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
-Mixture modelling aims at identifying latent classes of persons with different properties.
-It can also be combined with covariates.
-A comparison of both approaches is given by \citet{FriStrZei:2014}.
-
 In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.
@@
 In this sense, the more exploratory approach of Rasch trees has higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
+The same argument holds for parameter differences in mixed-effects models, where the substantial hypotheses formulated a priori are typically about global intervention effects, not about possible subgroup-specific differences.
 
-The same holds for parameter differences in mixed-effects models, where the substantial hypotheses are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
-\todo{Marjolein add if this makes any sense here}
+An alternative framework for detecting previously unknown groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
+Mixture modelling aims at identifying latent classes of persons with different properties.
+It can also be combined with observed covariates.
+A comparison of MOB and mixture modelling is given by \citet{FriStrZei:2014}.
 
-For our two application examples, we could show that MOB can detect ..., performs automated variable selection ...
+\subsection{Limitations of MOB}\todo{I (Caro) would not make subsections within discussion}
 
-\subsection{Limitations}
-
-Due to the exploratory nature of MOB, it does not allow for hypothesis tests, as there is no valid way to account for the exploratory searching of the subgroups.
+Due to the exploratory nature of MOB, it does not allow for hypothesis tests\todo[inline]{I (Caro) think this will sound confusing because earlier in the paper we explain that MOB is indeed based on significance tests, and due to the fact that the stopping is based on the significance test and (at least in party and psychotree) we use Bonferroni for adjusting against the number of covariates, it is not like we are doing completely uncontrolled multiple testing - in the Raschtree paper we write about this: Moreover, it is important to note that our model-based recursive partitioning algorithm is not affected by an inflation of chance due to its recursive nature.
+Indeed, several statistical tests are successively conducted in a Rasch tree—but each test is conducted only if the previous test yielded a significant result.
+In this sense, the recursive approach forms a closed testing procedure, which does not lead to an inflation of chance as is well known from the literature on multiple comparisons (Marcus, Peritz, \& Gabriel, 1976; Hochberg \& Tamhane, 1987).
+For the Rasch tree, this means that the postulated significance level holds for the entire tree, not only for each individual split.
+This ensures that DIF is not erroneously detected as an artifact of the recursive nature of the algorithm.
+-- another, related, limitation that I would agree with and might be what you also had in mind is that the tree will not necessarily find the globally best partition, because every split is based on the previous split choices, here is what we wrote about this in the tree intro PsychMeth paper: Thus, variable selection in a single tree is affected by order effects similar to those present in stepwise variable selection approaches for parametric regression (that is also instable against random variation of the learning data, as pointed out by Austin and Tu 2004).
+In both recursive partitioning and stepwise regression, the approach of adding one locally optimal variable at a time does not necessarily (or rather hardly ever) lead to the globally best model over all possible combinations of variables.
+-- should we write something similar here?}, as there is no valid way to account for the exploratory searching of the subgroups.
 Thus, even if detected subgroups are substantively meaningful or differences are large, if researchers want to ascertain statistical significance of the subgroup differences, this should be done on new data using confirmatory techniques.
-
-\todo[inline]{Mention shortcomings.}
+\todo{Marjolein write a little more here?
+e.g.
+such as cross validation?}
 
@@
 
-The Mantel-Haenszel effect size measure for DIF will be integrated in the \texttt{psychotools} package.
-
 \todo{Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
@@
 
-\section{Appendix A: Association between Age and PPVT scores before and after transformation}
+\section{Association between Age and PPVT scores before and after transformation}
 \label{sec:AppendixA}
 
-\begin{figure}
+\begin{figure}[h!]
 \caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
@@
 
+\section{Items of the culture scale for application example 2}
+\label{sec:AppendixB}
+
+\begin{enumerate}
+\item Which painter created this painting? – Andy Warhol.
+\item What do these four buildings have in common? – All four were designed by the same architects.
+\item Roman numbers: What is the meaning of CLVI? – 156.
+\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
+\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
+\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
+\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
+\item In which city is this building located? – Paris.
+\item Which one of the following operas is not by Mozart? – Aida.
+\end{enumerate}
+
 \end{document}

### 2023-10-20 _MOB_paper.tex
text Rasch and MH tree

@@
 
-\todo[inline]{Comment on new example: I now use data from Bogen 20 of the SPISA dataset.
-I use the student subset, and complete cases on all variables.
-The story is less clear, but the tree is still stopped early, and we can display which items have dif.
-}
+We will illustrate the usage of Rasch trees and the Mantel-Haenszel effect size measure by means of a dataset from \citet{SPISA:book}. An online general knowledge quiz was conducted by the German
+weekly news magazine SPIEGEL.
+Below we will use the abbreviation \texttt{SPISA} for the name of the data set, because the quiz was termed ``students' PISA'' by the magazine, but it is not related to the ``real'' PISA study by the OECD.
+The data set contains the quiz results as well as
+sociodemographic data.
+In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $9769$ complete cases of university students for illustration.
+The wording of the items is provided in Appendix~\ref{sec:AppendixB}.
 
-\begin{itemize}
-\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet; Bogen 20 (as in psychotree, but from the whole dataset, not only a subsample))
-\item take the subsample of students (only students); complete cases; $N = 9769$
-\item items could probably appear in Appendix??
-\begin{enumerate}
-\item Which painter created this painting? – Andy Warhol.
-\item What do these four buildings have in common? – All four were designed by the same architects.
-\item Roman numbers: What is the meaning of CLVI? – 156.
-\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
-\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
-\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
-\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
-\item In which city is this building located? – Paris.
-\item Which one of the following operas is not by Mozart? – Aida.
-\end{enumerate}
-\end{itemize}
+Three possible partitioning variables are considered: \texttt{Gender} (in this data set coded only as male, female, or missing), \texttt{Age} (continuous), and \texttt{Area} (Area of study: Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, or Sports).
+The nine items from the culture scale are scored with 0 (incorrect answer) or 1 (corrct answer).
+The quiz data have been prepared in a format where the nine items of the culture scale are not stored as individual columns of the data set, but the complete matrix of item responses is stored like a single variable \citep[see][]{raschtree:vignette} under the name \texttt{culture}.
+This means that we can access the item responses for all items jointly using the name \texttt{culture} in the following R commands.
+The first few rows of the data set look like this:
 
-We load the data and inspect the first rows.
-there are several covariates: age (continuous), gender (male, female, missing), area (Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, Sports) and 9 items from the culture scale (true/false 0/1).
+\todo[inline]{I suggest that we publish the data set Mirka created with the paper, but don't show her data preprocessing steps (which are in the echo = FALSE chunk above); I usually use load for Rda files, Marjolein uses readRDS, is there an advantage?
+use the same in both examples}
 
@@
 \begin{Sinput}
- head(covar_data)
+ load("dat_SPISA.Rda")
+ head(dat_SPISA, 3)
 \end{Sinput}
 \begin{Soutput}
-    Age Gender                    Area
-90   21   Male       Law and Economics
-163  25   Male          no Information
-402  22   Male Agriculture & Nutrition
-449  20   Male          no Information
-518  21 Female          no Information
-649  23   Male          no Information
-\end{Soutput}
-\begin{Sinput}
- head(culture_scale)
-\end{Sinput}
-\begin{Soutput}
-     i28 i29 i30 i31 i32 i33 i34 i35 i36
-[1,]   1   1   0   1   0   1   1   1   1
-[2,]   1   1   1   1   0   1   1   0   1
-[3,]   0   1   0   1   1   0   1   1   0
-[4,]   0   1   1   1   1   1   1   1   1
-[5,]   1   0   1   0   0   1   1   1   1
-[6,]   1   1   1   1   0   1   1   1   1
+    Age Gender                    Area culture.i28 culture.i29 culture.i30
+90   21   Male       Law and Economics           1           1           0
+163  25   Male          no Information           1           1           1
+402  22   Male Agriculture & Nutrition           0           1           0
+    culture.i31 culture.i32 culture.i33 culture.i34 culture.i35 culture.i36
+90            1           0           1           1           1           1
+163           1           0           1           1           0           1
+402           1           1           0           1           1           0
 \end{Soutput}
@@
 
-We create a dataset \texttt{dat\_SPISA} that contains the data of the covariate and the item responses from the culture scale.
-By assigning the whole scale to the dataset using \texttt{\$}, we can later access item responses from all items using \texttt{\$culture}.
-\begin{Schunk}
-\begin{Sinput}
- dat_SPISA <- covar_data
- dat_SPISA$culture <- culture_scale
-\end{Sinput}
-\end{Schunk}
+The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that in the following we will fit Rasch trees by means of the function \texttt{raschtree} from the R package \texttt{psychotree}.
+However, model-based recursive partitioning is also available in \texttt{psychotree} for other IRT models.
 
-The Rasch tree can be fitted with the function \texttt{raschtree} from the R package \texttt{psychotree}.
-We use the typical formula syntax with the item responses of the culture scale on the left hand side of $\sim$, and the covariates on the right hand side.
+The model which we would like to test for parameter stability in this example, the Rasch model, is based only on the matrix of item responses \texttt{culture}.
+It does not contain predictor variables of its own.
+Therefore, the formula syntax for the \texttt{raschtree} function looks simpler than that for the \texttt{lmertree} function in application example 1: On the left hand side of the $\sim$ symbol, we provide the item response matrix \texttt{culture} for the Rasch model.
+On the right hand side of the $\sim$ symbol, we provide the possible partitioning variables \texttt{Gender}, \texttt{Age}, and \texttt{Area}.
 
@@
 \begin{Sinput}
- library(psychotree)
+ library("psychotree")
  Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-021}
+\includegraphics{_MOB_paper-020}
 
-\todo[inline]{The are covariate split does not look pretty yet. But I will try to make it more pretty later on.. }
+\todo[inline]{The are covariate split does not look pretty yet. But I [Mirka] will try to make it more pretty later on.. (reply Caro: yes if possible, and/or make it wider in the pdf?)}
 
-Mantel-Haenszel trees use the Mantel-Haenszel effect size measure for DIF.
-it has three categories (A: negligible, B: moderate, C: large).
-If none of the items has DIF in category B or C (default, but can also be changed so that DIF has to be C), the tree is stopped from growing.
+To assess whether DIF effects are substantial or negligible, it is possible to use the Mantel-Haenszel DIF effect size measure as an additional stopping criterion.
+It has three categories (A: negligible, B: moderate, C: large).
+If none of the items shows DIF in category B or C\footnote{This is the default setting.
+It can also be changed so that splitting is stopped if no item shows DIF in category C.}, the tree is stopped from growing.
+At the moment, the R functions for the Mantel-Haenszel stopping criterion and additional visualizations based on the Mantel-Haenszel effect size are available from github.
+They will also be made available as part of the \texttt{psychotree} package in the future.
 
-The software must be installed from github, but will (probably) be implemented in the psychotree package some time in the future
+In order to install the functions from the author's github page and make them available in the current R session, use the following two commands.
+
+\todo[inline]{Caro an Mirka: don't we need library devtools?}
 
@@
  devtools::install_github("mirka-henninger/raschtreeMH")
- library(raschtreeMH)
+ library("raschtreeMH")
 \end{Sinput}
@@
 
-Syntax is very similar to the classical Rasch tree, but an additional argument \texttt{stopfun}, where the mantelhaneszel stopping function can be selected (but also own stopping functions can be provided; see Henninger et al., 2023).
-The user also has to indicate what kind of purification strategy should be used (none, 2step, iterative).
-Iterative is recommended.
-Optionally, the user can specify the stopping criterion.
-In this case, tree growing is stopped when all items fall into category ``A''.
+We can use a syntax very similar to that of the original \texttt{raschtree} function, but now we also provide the additional argument \texttt{stopfun}.
+We want to use the Mantel-Haenszel stopping function (\texttt{stopfun\_mantelhaenszel}) together with iterative purification.
+????
+and DIF category A as the stopping criterion.
+\todo{Caro an Mirka: oben steht das wäre eh der default, kann das stopcrit A raus?
+dann hier im Text und im code} It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
+
+For the purification strategy, there are three options (\texttt{none}, \texttt{2step}, and \texttt{iterative}).
+Purification means that items which have already been diagnosed with DIF are taken out of the set of so called anchor items.
+Anchor items are used for aligning the scales of different groups of persons for testing further items for DIF.
+It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the refereces therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
+In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2.
+When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
+\todo{Mirka bitte diesen ganzen Absatz besonders gut korrekturlesen}
+The final set of anchor items for each group comparison, which results from the purification, is also used in the tree plots with color coding for DIF items displayed below.
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-026}
+\includegraphics{_MOB_paper-025}
 
-We can also color the items in the end node profiles according to a split in the tree. Here, we see that for Node 1 (split on the covariate Gender), items 1, 6, 7, 8, 9 show DIF in categories B or C.
+We can also color the DIF items in the end node profiles according to a particular split in the tree.
+Here, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
+As a reminder, the content of these items was:
 
 \begin{itemize}
-\item 1 painting by Andy Warhol
-\item 6: novel by Daniel Kehlmann: Die Vermessung der Welt
-\item 7: city of the Buddenbrooks: Lübeck
-\item 8: city with building: Paris
-\item 9: opera not by Mozart: Aida
+\item 1: Painting by Andy Warhol
+\item 6: Novel by Daniel Kehlmann: Die Vermessung der Welt
+\item 7: City of the Buddenbrooks: Lübeck
+\item 8: City with building: Paris
+\item 9: Opera not by Mozart: Aida
 \end{itemize}
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-027}
+\includegraphics{_MOB_paper-026}
 
-We can also color by Node 2 (split on covariate Age).
-Also here, Item 1 shows DIF in category B.
-It seems to be easier for younger respondents ($\leq 24$) compared to older respondents (painting by Andy Warhol).
+These items are (relative to the anchor items for this group comparison) easier\todo{double check Mirka} to answer for participants with female or missing gender.
+
+\todo[inline]{Caro an Mirka: ich nehme an dass hier hohe item parameter bedeuten schwierig, niedrige leicht, dann wären diese items fur Frauen leichter, du hast es unten aber glaub genau andersrum geschrieben - evtl.
+weil im original Raschtree paper oder irgendwo mal easiness statt difficulty abgetragen war?}
+
+We can also color the DIF items according to node 2 (split in the covariate \texttt{Age}).
+Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
+It is (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult} to answer for younger respondents (age $\leq 24$) compared to older respondents.
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-028}
+\includegraphics{_MOB_paper-027}
 
-For Node 4 (also split on covariate Age), Items 1 and 9 show DIF, such that Item 1 and 9 are easier for younger respondents (Age $\leq 29$).
-(Item 1: Painting by Andy Warhol; Item 9: Opera not by Mozart: Aida)
+For node 4 (another split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult to answer} for younger respondents (age $\leq 29$).
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-029}
+\includegraphics{_MOB_paper-028}
 
-\todo[inline]{Maybe a small note on anchoring?
-Item difficulty profiles are anchoring on the items that do not show DIF.
-This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
+\todo[inline]{my (Caros) feeling is that two colored trees are enough, would leave out one of the colored trees based on age, and say above: We can also color the DIF items in the end node profiles according to a particular split in the tree.
+We will illustrate this with two examples.
+First, we can see that for node 1...}
 
-The information about the Mantel-Haenszel effect size measure for each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
-it contains the information on classification of each item in e ach node, the computed Mantel-Haenszel effect size measure, the type of purification, and if purification was iterative how many iterations were realized.
+Detailed information about the Mantel-Haenszel criterion for each item in each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
+It contains the DIF classification and the value of the Mantel-Haenszel effect size measure for each item in each node, as well as the type of purification that was employed in each node, and in the case of iterative purification how many iterations were conducted.
 
@@
 
+The displayed information again shows that some items show only negligible DIF of category A in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
+The values of the Mantel-Haenszel effect size measure are negative when \todo{Caro an Mirka: ich rate hier völlig aus dem nichts, bitte korrigieren} when the item difficulty is lower in the right daughter node than in the left daughter node????.
+For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier\todo{double check Mirka} to answer (relative to the anchor items for this group comparison) for participants with female or missing gender, which correspond to the right daughter node of node 1.
+Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.
+However, this item has a smaller DIF effect size.
+Thus it falls into DIF category A (negligible) and was not relevant for the split in \texttt{Gender}.
+
 \FloatBarrier
 \section{Discussion}
-
-\todo{next paragraph moved to here, where exactly?}
-An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
-Mixture modelling aims at identifying latent classes of persons with different properties.
-It can also be combined with covariates.
-A comparison of both approaches is given by \citet{FriStrZei:2014}.
 
@@
 
-\section{Appendix A: Association between Age and PPVT scores before and after transformation}
+\section{Association between Age and PPVT scores before and after transformation}
 \label{sec:AppendixA}
 
-\begin{figure}
+\begin{figure}[h!]
 \caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
-\includegraphics{_MOB_paper-031}
+\includegraphics{_MOB_paper-030}
 \end{figure}
 
+\section{Items of the culture scale for application example 2}
+\label{sec:AppendixB}
+
+\begin{enumerate}
+\item Which painter created this painting? – Andy Warhol.
+\item What do these four buildings have in common? – All four were designed by the same architects.
+\item Roman numbers: What is the meaning of CLVI? – 156.
+\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
+\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
+\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
+\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
+\item In which city is this building located? – Paris.
+\item Which one of the following operas is not by Mozart? – Aida.
+\end{enumerate}
+
 \end{document}

### 2023-10-23 Raschtree_MH.Rnw
First changes Caro

@@
 \todo[inline]{I suggest that we publish the data set Mirka created with the paper, but don't show her data preprocessing steps (which are in the echo = FALSE chunk above); I usually use load for Rda files, Marjolein uses readRDS, is there an advantage?
-use the same in both examples}
+use the same in both examples.
+Mirka: from what I understand, readRDS is for loading .rds/.rda files with only one r object in it, while load is for .rda files with one or more r object in it.
+I would opt for .rda, because I think it is more general?
+}
 
@@
 
-<<fig = TRUE, width = 15, height = 8>>=
+<<echo = FALSE>>=
+if(!file.exists("SPISA_example/Raschtree.rda")){
+  pdf("raschtree_culture.pdf", width = 36, height = 15)
+  plot(Raschtree_culture)
+  dev.off()
+}
+@
+
+<<fig = TRUE, width = 15, height = 8, eval = FALSE>>=
 plot(Raschtree_culture)
@@
 
-\todo[inline]{The are covariate split does not look pretty yet. But I [Mirka] will try to make it more pretty later on.. (reply Caro: yes if possible, and/or make it wider in the pdf?)}
+\begin{figure}[!h]
+\centering
+\includegraphics[width = \textwidth]{raschtree_culture.pdf}
+\end{figure}
 
@@
 
-\todo[inline]{Caro an Mirka: don't we need library devtools?}
-
 <<>>=
-devtools::install_github("mirka-henninger/raschtreeMH")
+library(devtools)
+install_github("mirka-henninger/raschtreeMH")
 library("raschtreeMH")
@@
 We want to use the Mantel-Haenszel stopping function (\texttt{stopfun\_mantelhaenszel}) together with iterative purification.
-????
-and DIF category A as the stopping criterion.
-\todo{Caro an Mirka: oben steht das wäre eh der default, kann das stopcrit A raus?
-dann hier im Text und im code} It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
+It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
 
@@
 Anchor items are used for aligning the scales of different groups of persons for testing further items for DIF.
-It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the refereces therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
+It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the references therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
 In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2.
@@
                                     data = dat_SPISA, 
-                                    stopfun= stopfun_mantelhaenszel(purification = "iterative",
-                                                                    stopcrit = "A"))
+                                    stopfun= stopfun_mantelhaenszel(purification = "iterative"))
   save(Raschtree_MH_culture, file = "SPISA_example/Raschtree_MH_culture.rda")
@@
                                   stopfun= stopfun_mantelhaenszel(
-                                    purification = "iterative",
-                                    stopcrit = "A"))
+                                    purification = "iterative"))
 @

### 2023-10-24 Raschtree_MH.Rnw
Changes Caro Raschtree

@@
 
-We can also color the DIF items in the end node profiles according to a particular split in the tree.
-Here, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
-As a reminder, the content of these items was:
+We can also color the DIF items in the end node profiles according to a particular split in the tree. We will illustrate this with two examples. 
+First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C. As a reminder, the content of these items was:
 
@@
 
-These items are (relative to the anchor items for this group comparison) easier\todo{double check Mirka} to answer for participants with female or missing gender.
+These items are (relative to the anchor items for this group comparison) easier or less difficult to answer for participants with female or missing gender.
 
-\todo[inline]{Caro an Mirka: ich nehme an dass hier hohe item parameter bedeuten schwierig, niedrige leicht, dann wären diese items fur Frauen leichter, du hast es unten aber glaub genau andersrum geschrieben - evtl.
-weil im original Raschtree paper oder irgendwo mal easiness statt difficulty abgetragen war?}
+% We can also color the DIF items according to node 2 (split in the covariate \texttt{Age}).
+Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
+It is (relative to the anchor items for this group comparison) more difficult to answer for younger respondents (age $\leq 24$) compared to older respondents.
 
-We can also color the DIF items according to node 2 (split in the covariate \texttt{Age}).
-Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
-It is (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult} to answer for younger respondents (age $\leq 24$) compared to older respondents.
+% <<fig = TRUE, width = 9, height = 7>>=
+% plot(Raschtree_MH_culture, color_by_node = 2)
+% @
 
-<<fig = TRUE, width = 9, height = 7>>=
-plot(Raschtree_MH_culture, color_by_node = 2)
-@
-
-For node 4 (another split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult to answer} for younger respondents (age $\leq 29$).
+For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are (relative to the anchor items for this group comparison) more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
 
@@
 @
-
-\todo[inline]{my (Caros) feeling is that two colored trees are enough, would leave out one of the colored trees based on age, and say above: We can also color the DIF items in the end node profiles according to a particular split in the tree.
-We will illustrate this with two examples.
-First, we can see that for node 1...}
 
@@
 The displayed information again shows that some items show only negligible DIF of category A in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
-The values of the Mantel-Haenszel effect size measure are negative when \todo{Caro an Mirka: ich rate hier völlig aus dem nichts, bitte korrigieren} when the item difficulty is lower in the right daughter node than in the left daughter node????.
-For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier\todo{double check Mirka} to answer (relative to the anchor items for this group comparison) for participants with female or missing gender, which correspond to the right daughter node of node 1.
-Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.
+The values of the Mantel-Haenszel effect size measure are negative when the item difficulty is lower in the right daughter node than in the left daughter node.
+For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer (relative to the anchor items for this group comparison) for participants with female or missing gender, which correspond to the right daughter node of node 1.
+% Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.
 However, this item has a smaller DIF effect size.

### 2023-10-25 Raschtree_MH.Rnw
Remove pic Raschtree
anchoring out, added to disc

@@
 
-\begin{figure}[!h]
-\centering
-\includegraphics[width = \textwidth]{raschtree_culture.pdf}
-\end{figure}
+% \begin{figure}[!h]
+% \centering
+% \includegraphics[width = \textwidth]{raschtree_culture.pdf}
+% \end{figure}
 
@@
 For the purification strategy, there are three options (\texttt{none}, \texttt{2step}, and \texttt{iterative}).
-Purification means that items which have already been diagnosed with DIF are taken out of the set of so called anchor items.
-Anchor items are used for aligning the scales of different groups of persons for testing further items for DIF.
+Purification means that items which have already been diagnosed with DIF are taken out of the sum score, which is used as the matching criterion in the final DIF test.
 It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the references therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
@@
 When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
-\todo{Mirka bitte diesen ganzen Absatz besonders gut korrekturlesen}
-The final set of anchor items for each group comparison, which results from the purification, is also used in the tree plots with color coding for DIF items displayed below.
+The final set of DIF-free items resulting from the purification is also used for aligning the item parameters for each group comparison in the tree plots with color coding, which are displayed below.
 
@@
 
-These items are (relative to the anchor items for this group comparison) easier or less difficult to answer for participants with female or missing gender.
+These items are %(relative to the DIF-free items for this group comparison) 
+easier or less difficult to answer for participants with female or missing gender.
 
@@
 Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
-It is (relative to the anchor items for this group comparison) more difficult to answer for younger respondents (age $\leq 24$) compared to older respondents.
+It is (relative to the DIF-free items for this group comparison) more difficult to answer for younger respondents (age $\leq 24$) compared to older respondents.
 
@@
 
-For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are (relative to the anchor items for this group comparison) more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
+For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison) 
+more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
 
@@
 The values of the Mantel-Haenszel effect size measure are negative when the item difficulty is lower in the right daughter node than in the left daughter node.
-For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer (relative to the anchor items for this group comparison) for participants with female or missing gender, which correspond to the right daughter node of node 1.
+For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer %(relative to the DIF-free items for this group comparison)
+for participants with female or missing gender, which correspond to the right daughter node of node 1.
 % Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.

### 2023-10-25 _MOB_paper.Rnw
anchoring out, added to disc

@@
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step and anchoring \citep[see][and application example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+Together with a purification step \citep[see][and application example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
@@
 
-An alternative framework for detecting previously unknown groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
+However, researchers should be aware that a covariate-based approach like MOB will only be able to detect parameter differences if the relevant covariates were recorded in the study and supplied to the algorithm.
+If no or not many potentially relevant covariates are available, an alternative framework for detecting previously unknown groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
 Mixture modelling aims at identifying latent classes of persons with different properties.
@@
 
+Another important caveat is that those covariates that are selected for splitting by the MOB algorithm are not necessarily causal for the observed parameter differences.
+They might also be proxies for other, unobserved covariates.
+For example, if DIF in the item parameters of a test on reading ability is found between different cities or districts, a variety of factors could be causally driving these effects, such as different compositions of native languages or sociodemographics.
+
+Finally, as with any exploratory method, MOB should be considered as a tool for generating hypothesis and not as a confirmatory technique.
+While the MOB algorithm, as it is implemented in the R packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory testing of hypothesis it is recommended to use fresh data or resampling techniquest like cross validation.
+
+\todo[inline]{I (Caro) have added a few more thoughts for the discussion above - Marjolein maybe this is already sufficient in length?
+please feel free to use anything from above or below and get back to me if you add/keep points where we might disagree}
 \subsection{Limitations of MOB}\todo{I (Caro) would not make subsections within discussion}

### 2023-10-25 _MOB_paper.tex
anchoring out, added to disc

@@
 \end{Schunk}
-\includegraphics{_MOB_paper-020}
 
-\todo[inline]{The are covariate split does not look pretty yet. But I [Mirka] will try to make it more pretty later on.. (reply Caro: yes if possible, and/or make it wider in the pdf?)}
+% \begin{figure}[!h]
+% \centering
+% \includegraphics[width = \textwidth]{raschtree_culture.pdf}
+% \end{figure}
 
@@
 
-\todo[inline]{Caro an Mirka: don't we need library devtools?}
-
 \begin{Schunk}
 \begin{Sinput}
- devtools::install_github("mirka-henninger/raschtreeMH")
+ library(devtools)
+ install_github("mirka-henninger/raschtreeMH")
  library("raschtreeMH")
@@
 We want to use the Mantel-Haenszel stopping function (\texttt{stopfun\_mantelhaenszel}) together with iterative purification.
-????
-and DIF category A as the stopping criterion.
-\todo{Caro an Mirka: oben steht das wäre eh der default, kann das stopcrit A raus?
-dann hier im Text und im code} It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
+It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
 
 For the purification strategy, there are three options (\texttt{none}, \texttt{2step}, and \texttt{iterative}).
-Purification means that items which have already been diagnosed with DIF are taken out of the set of so called anchor items.
-Anchor items are used for aligning the scales of different groups of persons for testing further items for DIF.
-It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the refereces therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
+Purification means that items which have already been diagnosed with DIF are taken out of the sum score, which is used as the matching criterion in the final DIF test.
+It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the references therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
 In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2.
 When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
-\todo{Mirka bitte diesen ganzen Absatz besonders gut korrekturlesen}
-The final set of anchor items for each group comparison, which results from the purification, is also used in the tree plots with color coding for DIF items displayed below.
+The final set of DIF-free items resulting from the purification is also used for aligning the item parameters for each group comparison in the tree plots with color coding, which are displayed below.
 
@@
                                    stopfun= stopfun_mantelhaenszel(
-                                     purification = "iterative",
-                                     stopcrit = "A"))
+                                     purification = "iterative"))
 \end{Sinput}
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-025}
+\includegraphics{_MOB_paper-026}
 
-We can also color the DIF items in the end node profiles according to a particular split in the tree.
-Here, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
-As a reminder, the content of these items was:
+We can also color the DIF items in the end node profiles according to a particular split in the tree. We will illustrate this with two examples. 
+First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C. As a reminder, the content of these items was:
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-026}
-
-These items are (relative to the anchor items for this group comparison) easier\todo{double check Mirka} to answer for participants with female or missing gender.
-
-\todo[inline]{Caro an Mirka: ich nehme an dass hier hohe item parameter bedeuten schwierig, niedrige leicht, dann wären diese items fur Frauen leichter, du hast es unten aber glaub genau andersrum geschrieben - evtl.
-weil im original Raschtree paper oder irgendwo mal easiness statt difficulty abgetragen war?}
-
-We can also color the DIF items according to node 2 (split in the covariate \texttt{Age}).
-Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
-It is (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult} to answer for younger respondents (age $\leq 24$) compared to older respondents.
-
-\begin{Schunk}
-\begin{Sinput}
- plot(Raschtree_MH_culture, color_by_node = 2)
-\end{Sinput}
-\end{Schunk}
 \includegraphics{_MOB_paper-027}
 
-For node 4 (another split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are (relative to the anchor items for this group comparison) easier\todo{Caro glaubt: more difficult to answer} for younger respondents (age $\leq 29$).
+These items are %(relative to the DIF-free items for this group comparison) 
+easier or less difficult to answer for participants with female or missing gender.
+
+% We can also color the DIF items according to node 2 (split in the covariate \texttt{Age}).
+Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
+It is (relative to the DIF-free items for this group comparison) more difficult to answer for younger respondents (age $\leq 24$) compared to older respondents.
+
+% <<fig = TRUE, width = 9, height = 7>>=
+% plot(Raschtree_MH_culture, color_by_node = 2)
+% @
+
+For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison) 
+more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
 
@@
 \includegraphics{_MOB_paper-028}
-
-\todo[inline]{my (Caros) feeling is that two colored trees are enough, would leave out one of the colored trees based on age, and say above: We can also color the DIF items in the end node profiles according to a particular split in the tree.
-We will illustrate this with two examples.
-First, we can see that for node 1...}
 
@@
 The displayed information again shows that some items show only negligible DIF of category A in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
-The values of the Mantel-Haenszel effect size measure are negative when \todo{Caro an Mirka: ich rate hier völlig aus dem nichts, bitte korrigieren} when the item difficulty is lower in the right daughter node than in the left daughter node????.
-For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier\todo{double check Mirka} to answer (relative to the anchor items for this group comparison) for participants with female or missing gender, which correspond to the right daughter node of node 1.
-Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.
+The values of the Mantel-Haenszel effect size measure are negative when the item difficulty is lower in the right daughter node than in the left daughter node.
+For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer %(relative to the DIF-free items for this group comparison)
+for participants with female or missing gender, which correspond to the right daughter node of node 1.
+% Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.
 However, this item has a smaller DIF effect size.

### 2023-10-26 Raschtree_MH.Rnw
Final Changes Raschtree Part

@@
 <<echo = FALSE>>=
-load("SPISA_example/dat_SPISA.rda")
-covar_data$Student <- NULL
-covar_data$Occupation <- NULL
-covar_data$Area <- sub("\n", "", covar_data$Area)
-dat_SPISA <- covar_data
-dat_SPISA$culture <- culture_scale
-save(dat_SPISA, file = "dat_SPISA.Rda")
+load("dat_SPISA.Rda")
 @
@@
 sociodemographic data.
-In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $\Sexpr{nrow(covar_data)}$ complete cases of university students for illustration.
+In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $\Sexpr{nrow(dat_SPISA)}$ complete cases of university students for illustration.
 The wording of the items is provided in Appendix~\ref{sec:AppendixB}.
@@
 
-\todo[inline]{I suggest that we publish the data set Mirka created with the paper, but don't show her data preprocessing steps (which are in the echo = FALSE chunk above); I usually use load for Rda files, Marjolein uses readRDS, is there an advantage?
-use the same in both examples.
+\todo{I usually use load for Rda files, Marjolein uses readRDS, is there an advantage? use the same in both examples.
 Mirka: from what I understand, readRDS is for loading .rds/.rda files with only one r object in it, while load is for .rda files with one or more r object in it.
@@
 <<echo = FALSE>>=
+library("psychotree")
 if(!file.exists("SPISA_example/Raschtree.rda")){
-  library("psychotree")
+  # relabel factors of the area covariate so that the Raschtree is not too big: 
+  levels(dat_SPISA$Area) <- c("Lang & Cult", "\nLaw and Econ",
+                              "\nMed & Health", "\nEngin",
+                              "\nSci, Phar, Geo",
+                              "\nAgri & Nutri",
+                              "\nSports", "Arts",
+                              "\nno Info")
   Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
@@
 
-<<echo = FALSE>>=
-if(!file.exists("SPISA_example/Raschtree.rda")){
-  pdf("raschtree_culture.pdf", width = 36, height = 15)
-  plot(Raschtree_culture)
-  dev.off()
-}
-@
-
-<<fig = TRUE, width = 15, height = 8, eval = FALSE>>=
+<<fig = TRUE, width = 22, height = 10>>=
 plot(Raschtree_culture)
 @
-
-% \begin{figure}[!h]
-% \centering
-% \includegraphics[width = \textwidth]{raschtree_culture.pdf}
-% \end{figure}
 

### 2023-10-26 SPISA_example/SPISA.R

@@ -5,81 +5,68 @@ load("/home/mhenninger/Arbeit/Forschung/Data/SPISA/spisa_ges.RData")
 spisa <- spisa[spisa$Bogen == 20, ]
 
 # translate covariates into english
-covar_dat <- spisa[, c("Alter", "Geschlecht", "Student", "Tat", "Richtung")]
-colnames(covar_dat) <- c("Age", "Gender", "Student", "Occupation", "Area")
+covar_dat <- spisa[, c("Alter", "Geschlecht", "Student", "Richtung")]
+colnames(covar_dat) <- c("Age", "Gender", "Student", "Area")
 levels(covar_dat$Gender) <- c("Female", "Male", "Missing")
 levels(covar_dat$Student) <- c("No", "Yes", "Missing")
-
-# combine factor levels of occupation
-covar_dat$Occupation <- forcats::fct_collapse(covar_dat$Occupation, 
-                                              Arbeitnehmer = c("Vollzeit", "Teilzeit"),
-                                              nicht_berufstaetig = c("arbeitlos", "nicht berufstaetig"),
-                                              Ruhestand = c("Ruhestand", "Ruhe+Tat")
-)
-levels(covar_dat$Occupation) <- c("\n\n\nPupil", "\n\n\nStudent", "\nPhD",
-                                  "\nin Education", "\nEmployed",
-                                  "\nUnemployed", "\nRetired",
-                                  "\nOthers")
 covar_dat[!(covar_dat$Gender %in% c("Female", "Male")),]$Gender <- "Missing"
-levels(covar_dat$Area) <- c("Language & Culture", "\nLaw and Economics", "\nMedicine & Health", "\nEngineering",
-                            "\nSciences, Pharmacy, Geography", "\nAgriculture & Nutrition", "\nSports", "\nArts",
-                            "\nno Information")
+covar_dat$Area <- as.factor(covar_dat$Area)
+levels(covar_dat$Area) <- c("Language & Culture", "Law and Economics", "Medicine & Health", "Engineering",
+                            "Sciences, Pharmacy, Geography", "Agriculture & Nutrition", "Sports", "Arts",
+                            "no Information")
 
 culture_scale <- apply(spisa[,28:36],2,as.numeric)
 
 dat <- covar_dat
[…]

### 2023-10-30 Raschtree_MH.Rnw
Revisions MF 30-10-2023

@@
 Below we will use the abbreviation \texttt{SPISA} for the name of the data set, because the quiz was termed ``students' PISA'' by the magazine, but it is not related to the ``real'' PISA study by the OECD.
-The data set contains the quiz results as well as
-sociodemographic data.
+The data set contains the quiz results as well as sociodemographic data.
 In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $\Sexpr{nrow(dat_SPISA)}$ complete cases of university students for illustration.
@@
 Three possible partitioning variables are considered: \texttt{Gender} (in this data set coded only as male, female, or missing), \texttt{Age} (continuous), and \texttt{Area} (Area of study: Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, or Sports).
-The nine items from the culture scale are scored with 0 (incorrect answer) or 1 (corrct answer).
+The nine items from the culture scale are scored with 0 (incorrect answer) or 1 (correct answer).
 The quiz data have been prepared in a format where the nine items of the culture scale are not stored as individual columns of the data set, but the complete matrix of item responses is stored like a single variable \citep[see][]{raschtree:vignette} under the name \texttt{culture}.
@@
 I would opt for .rda, because I think it is more general?
-}
+Marjolein: I think if we include datasets in supplemental materials, we should have files which are only 1 dataset, not sets of datasets or objects.
+I think this will make the examples more simple.
+In practice, users will probably also have only 1 dataset.}
 
@@
 load("dat_SPISA.Rda")
-head(dat_SPISA, 3)
+#head(dat_SPISA, 3) ## commented out because will not compile, dat_SPISA is not
+                    ## an object in current wd
 @
@@
 
-Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function.
+Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} method.
 The item parameters for the nine culture items are displayed in the end nodes of the tree.
@@
 It can also be changed so that splitting is stopped if no item shows DIF in category C.}, the tree is stopped from growing.
-At the moment, the R functions for the Mantel-Haenszel stopping criterion and additional visualizations based on the Mantel-Haenszel effect size are available from github.
-They will also be made available as part of the \texttt{psychotree} package in the future.
+At the moment, the \texttt{R} functions for the Mantel-Haenszel stopping criterion and additional visualizations based on the Mantel-Haenszel effect size are available from GitHub.
+They will be made available as part of the \texttt{psychotree} package in the future.
 
-In order to install the functions from the author's github page and make them available in the current R session, use the following two commands.
+In order to install the functions from the author's GitHub page and make them available in the current \texttt{R} session, use the following two commands:
 
@@
 
-For the purification strategy, there are three options (\texttt{none}, \texttt{2step}, and \texttt{iterative}).
+For the purification strategy, there are three options: \texttt{none}, \texttt{2step} and \texttt{iterative}.
 Purification means that items which have already been diagnosed with DIF are taken out of the sum score, which is used as the matching criterion in the final DIF test.

### 2023-10-30 _MOB_paper.Rnw
Revisions MF 30-10-2023

@@
 
-\todo[inline]{Target journal: \textit{Journal of School Psychology}.
-This Special Issue is on "Conceptual and Methodological Advances for Understanding Contextual, Identity, and Cultural Effects in Intervention Research".
-Submission deadline: October 31, 2023}
-
 \newpage
@@
 In CART, the aim is to detect groups of persons, defined by (combinations of) covariate values, that differ in their mean on a response variable.
-For example, if we want to predict math exam grades using student characteristics assessed in a previous year, we could find that the subgroup of children who are highly motivated and have good reading skills show higher average math exam grades.
+For example, if we want to predict math exam grades using student characteristics assessed in a previous year, we could find that a subgroup of students who are highly motivated and have good reading skills show higher average math exam grades.
 MOB generalizes this idea: It also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean.
-An example could be that children who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.
+An example could be that students who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.
 
@@
 In this article, we will highlight two specific MOB methods that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
-Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures, for example when children are tested at several time points (so that time points are nested in children) and/or when children from different classes from different schools (so that children are nested in classes, which are in turn nested in schools) participate in a study.
-In application example 1 \ref{sec:TutorialMixed} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
+Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures.
+For example, when children are tested at several time points, the time points are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools.
+In \hyperref[sec:TutorialMixed]{application example 1} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
 In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model.
@@
 
-For application example 2, we go back in the research process to the point where a new psychological test has been administered to a validation sample.
+For \hyperref[sec:TutorialRasch]{application example 2}, we go back in the research process to the point where a new psychological test has been administered to a validation sample.
 Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model.
@@
 Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF.
-In application example 2 we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.
+In \hyperref[sec:TutorialRasch]{application example 2} we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.
 
@@
 %note from Caro: I woul call them partitioning variables or covariates, but not predictors, to save the term predictors for those x-variables that are part of the parametric model
-To keep the example simple, we apply MOB to a very simple global model, which comprises only an intercept.
+To keep the example simple, we apply MOB to a very basic global model, which comprises only an intercept.
 It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
@@
 %note from Caro: I found Fig 1 containing only the boxplot irritating and have moved it into one panel with the tree
+
+\begin{figure}[h]
+\caption{Left: Univariate distribution of the response variable. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
+\begin{subfigure}[][][t]{.4\textwidth}
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=4.5, height=7>>=
+library("colorspace")
+boxplot(toy_data$y, cex = .7)
+#points(x = 1, y = mean(toy_data$y), pch = 19, ylim = c(1,14))
+@
+\end{subfigure}
+\begin{subfigure}[][][b]{.7\textwidth}
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=6, height=5.5>>=
+library("partykit")
+tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
+plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
+@
+<<eval = FALSE, echo = FALSE, fig=TRUE, width=5, height=5>>=
+plot(tree, gp = gpar(cex = .7), type = "simple")
+@
+\end{subfigure}
+\label{fig:toy}
+\end{figure}
 
@@
 
-We applied MOB to the observations in Figure~\ref{fig:toy}, left, specifying $x_1$, $x_2$ and $x_3$ as potential splitting variables.
-The resulting tree is shown in Figure~\ref{fig:toy}, right.
+We applied MOB to the observations in the left panel of Figure~\ref{fig:toy}, specifying $x_1$, $x_2$ and $x_3$ as potential partitioning variables.
+The resulting tree is shown in the right panel of Figure~\ref{fig:toy}.
 In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2).
@@
 The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node.
-Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences.
+Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroup are as similar as possible.
 In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate.
@@
 The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting.
-The subgroup-specific distributions of the response variable are presented in end nodes at the bottom of the tree.
-The fact that Figure~\ref{fig:toy}, right, displays more than one end node confirms that one global model for all observations cannot appropriately capture the pattern in the data. 
-Figure~\ref{fig:toy}, right, also shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
+The subgroup-specific distributions of the response variable are presented in the end nodes at the bottom of the tree.
+The fact that the tree in Figure~\ref{fig:toy} displays more than one end node confirms that one global model for all observations does not appropriately capture the pattern in the data. 
+The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
 This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.
-
-\begin{figure}
-\caption{Left: Univariate distribution of the response variable, with global mean indicated by triangle. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
-\begin{subfigure}[][][t]{.4\textwidth}
-<<eval = TRUE, echo = FALSE, fig=TRUE, width=4.5, height=7>>=
-library("colorspace")
-boxplot(toy_data$y, cex = .7)
-points(x = 1, y = mean(toy_data$y), col = rainbow_hcl(1), pch = 17, ylim = c(1,14))
-@
-\end{subfigure}
-\begin{subfigure}[][][b]{.6\textwidth}
-<<eval = TRUE, echo = FALSE, fig=TRUE, width=6, height=5>>=
-library("partykit")
-tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
-plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
-@
-<<eval = FALSE, echo = FALSE, fig=TRUE, width=5, height=5>>=
-plot(tree, gp = gpar(cex = .7), type = "simple")
-@
-\end{subfigure}
-\label{fig:toy}
-\end{figure}
 
@@
 
-%note from Caro: I did not like the formulation "There are two main differences between MOB and and traditional classification and regression tree algorithms" because readers might misinterpret it and think that only MOB, not trees, can be unbiased and use significance tests, while the distinction is really between traditional and modern algorithms for both trees and mob, changed into: characteristics
 There are three further important characteristics of tree and MOB algorithms that we would like to mention here: The first characteristic is the type of variable and cutpoint selection an algorithm employs.
@@
 Once there are no more covariates that show a significant structural change in any node, splitting is halted.
-In this way, the  algorithm selects only those partitioning variables that are relevant for distinguishing the groups (i.e., it performs automatic variable selection, as illustrated in Figure~\ref{fig:toy}, right).
-Moreover, the trees will not grow as large as possible, but will stop when no more significant structural change is detected.
+In this way, the  algorithm selects only those partitioning variables that are relevant for distinguishing the groups (i.e., it performs automatic variable selection, as illustrated in the right panel of Figure~\ref{fig:toy}).
+Moreover, the trees will not grow as large as possible, but splitting is stopped when no more significant structural change is detected.
 While traditional tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (so called \textit{pruning}), modern tree and MOB algorithms employ significance tests as stopping criteria (and some can also use effect size measures, as we will see in application example 2).
@@
 Other stopping criteria are based on the number of persons in the end nodes.
-These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them}
+These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.
+\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them} Note that it is thus possible that no split will be implemented, when none of the specified partitioning variables show significant structural change.
 
@@
 Still, an important challenge for the researcher remains: To specify the parametric model of interest, to specify the set of possible partitioning variables and to choose the settings of the MOB algorithm.
-The current paper aims to provide some guidance.
-%note from Caro: it said include link in several places when referring to example 2, but I don't think this is necessary, because currently sections are not numbered anyway and the title example 2 is clear
+The current paper aims to provide guidance.
 
-The general framework of MOB introduced above will now be applied to two types of statistical models that are particularly relevant in school psychology research: Mixed-effects models for repeated measures or nested data structures and measurement models for validating psychological and educational tests.
+Next, we discuss two specific methods that make use of the general MOB framework introduced above.
+The methods allow for partitioning two types of statistical models that are particularly relevant for school psychology research: Mixed-effects models for repeated measures or nested data structures and measurement models for validating psychological and educational tests.
 
@@
 Fixed effects are typically used to capture population-averaged effects, while random effects are used to capture inter-individual variation deviating from these fixed effects.
-In many studies, researchers are specifically interested in testing hypotheses relating to the population-averaged effects, while the random effects are included in the model to properly account for inter-individual variation, and correlations between observations within the same unit.
+In many studies, researchers are specifically interested in testing hypotheses relating to the population-averaged effects, while the random effects are included in the model to properly account for inter-individual variation, and correlations between observations within the same unit \citep{RaudyBush02}.
 
 GLMM trees combine MOB and mixed-effects models and were introduced by \cite{FokkySmit18}.
-Because researchers' interests commonly focus on the fixed effects (of time or treatment, for example), the GLMM tree algorithm only targets structural change in the fixed-effects parameters.
-The random-effects parameters can be specified as usual and are assumed constant; they are estimated using all observations in the dataset.
-The MOB algorithm is thus applied to the fixed-effects parameters only and the detected subgroups will only obtain different estimates for the fixed-effects coefficients.
-This allows for detecting subgroups with different means on the response variable, but also with differential treatment effects or differential growth over time, to name but a few examples.
-While the 'standard' MOB trees \cite{ZeilyHoth08} allow for partitioning fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for partitioning mixed-effects regression models.
-Further mathematical and computational details about the GLMM tree algorithm are described in \cite{FokkySmit18} and \citep{FokkyZeil23}.
+While the 'standard' MOB trees \citep{ZeilyHoth08} allow for subgroup detection in fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for partitioning mixed-effects regression models.
+Because researchers' interests in multilevel models commonly focus on the fixed effects (of time or treatment, for example), the GLMM tree algorithm only targets structural change in the fixed-effects parameters.
+The random-effects parameters can be specified as usual and are assumed constant; that is, they are estimated using all observations in the dataset.
+Structural change is thus assessed with respect to the fixed-effects parameters, while dependency between observations within the same units is accounted for by the random effects.
+The resulting subgroups will differ in their estimates for the fixed-effects coefficients only.
+As such, GLMM trees allow for detecting subgroups in multilevel models that differ with respect to any (set of) fixed-effects parameters of interest.
+For example, users may be interested in detecting subgroups with different means \citep{FokkyEdbr21}, but also differential effects of treatment \citep{FokkySmit18} or differential growth over time \citep{FokkyZeil23}, to name but a few examples.
+
+The \texttt{R} functions for fitting GLMM trees are the \texttt{lmertree} and \texttt{glmertree} functions from package \texttt{glmertree}.
+The usage of this function will be illustrated in \hyperref[sec:TutorialMixed]{application example 1}.
+Further mathematical and computational details about the GLMM tree model are also described in \cite{FokkySmit18} and \cite{FokkyZeil23}.
 
@@
 Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups.
-In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., e.g.,][for an introduction]{DebStrZei:2022:CRC}.
-\todo{Omit last sentence or move to discussion?
-I think better to focus on DIF only here.
-Response Caro: would rather keep it here, but have increased importance of measurement invariance in next sentence.} A particularly crucial assumption for the comparability of test scores between groups is measurement invariance.
+In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., for example][for an introduction]{DebStrZei:2022:CRC}.
+A particularly crucial assumption for the comparability of test scores between groups is measurement invariance.
 Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
@@
 MOB can be used to detect DIF by means of testing whether the item parameters of the measurement model exhibit significant instability with respect to (combinations of) covariates.
-For the Rasch measurement model, the R function for conducting the MOB analysis is the \texttt{raschtree} function from R package \texttt{psychotree} \citep{StroyKopf15}.
-The usage of this function will be illustrated in application example 2.
+For the Rasch measurement model, the \texttt{R} function for conducting the MOB analysis is the \texttt{raschtree} function from package \texttt{psychotree} \citep{StroyKopf15}.
+The usage of this function will be illustrated in \hyperref[sec:TutorialRasch]{application example 2}.
 We will see that, just like for GLMM trees, we need to specify which variables are part of the parametric model.
@@
 
-If one joint Rasch model holds for the entire sample, i.e., if there is no DIF, a Rasch tree should show no splits.
+If one joint Rasch model holds for the entire sample, that is, if there is no DIF, a Rasch tree should show no splits.
 However, in certain settings in educational research, such as large scale assessments, very large sample sizes are available for testing for DIF.
 Large sample sizes are good for detecting even small effects or model violations with a high statistical power.
-The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected with large samples.
+The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected.
 However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
-As we will illustrate in the first part of application example 2, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+As we will illustrate in the first part of \hyperref[sec:TutorialRasch]{application example 2}, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
 Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
@@
 %\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
-For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc.
-In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test.
+For example, items that show DIF between different language groups may often be improved by means of making sure that in all translations the meaning is as similar as possible, and/or that the words employed in the translations for the different languages are equally frequently used.
+In other situations, sources of DIF might be harder to eliminate, so that often DIF items may be excluded from a test.
 Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.
@@
 Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
-These items will then show DIF between native and non-native speakers.
-When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, e.g., by using more frequently used words, or whether the test should be considered two-dimensional rather than one-dimensional \citep[see also][for a discussion of the connection between DIF and multidimensionality]{Ack:1992,Stretal:2021:APM}.
+These items may then show DIF between native and non-native speakers.
+When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, for example by using more frequently used words, or whether the test should be considered two- rather than one-dimensional.
+For a discussion of the connection between DIF and multidimensionality, see also \cite{Ack:1992,Stretal:2021:APM}.
 
-We will now illustrate how to use the R packages \texttt{glmertree} and \texttt{psychotree} to conduct the MOB analyses.
+We will now illustrate how to use the \texttt{R} packages \texttt{glmertree} and \texttt{psychotree} to conduct the MOB analyses.
 
@@
 Participation in Head Start takes place from ages 3 through 5.
-\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).
+\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF) \todo{Marjolein, add reference}.
 
-The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two children, where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
+The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other preschool program.
 Data from children in those families who participated in another preschool program were excluded.
-As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start.
-The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF), we will this model PPVT trajectories over time.
+The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF) \todo{Marjolein: Add reference}, we will thus model PPVT trajectories over time.
 There were an average number of \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(1-mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there was only a single PPVT score available.
@@
 
-Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height.
+Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's body height \todo{Marjolein: Add scale for measurement}.
 Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables.
-The dataset comprises data from families and children for whom complete data was available.
+The dataset analyzed here only comprises data from families and children for whom complete data was available.
 
@@
 We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red).
-To show the effect of age and Head Start participation, we first fit a mixed-effects model comprising their main and interaction effects, using package \texttt{lme4}.
+To show the global effect of age and Head Start participation, we first fit a mixed-effects model comprising their main and interaction effects, using package \texttt{lme4}.
+This would be the model fitted in the root node of the GLMM tree.
 To account for the correlation between repeated assessments on the same child, and between siblings with the same mother, we specify a random intercept for children, nested within mothers:
@@
 This result agrees with the findings of \cite{Demi09}.
-The code for the figure is omitted here for space considerations, because we want to focus on fitting and interpreting GLMM trees.
+Figure~\ref{fig:global_lmm} already suggests that the effect of age on performance is very strong, compared to which the effects of intervention (Head Start) are existent but relatively small, which is typical for young children growing up to adults.
+Note that the \texttt{R} code for creating the figure is omitted here, because we want to focus on fitting and interpreting GLMM trees.
 Code for exact replication of the results presented here is provided in the Supplementary Materials.
@@
 The benefit increases over time for White children with higher mother's AFTQ scores.
-Strikingly,\todo{possibly less extreme word like: Interestingly,} the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
+Interestingly, the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
 These results correspond to the conclusions of \cite{Demi09}.
 
-\todo[inline]{With different specification of the GLMM tree (no square root, parameter stability tests at mom level, random intercepts of children nested within mothers) the results have changed, still need to adjust description}.
-
-\todo[inline]{Marjolein add here comment that typically affect of age is typically strongest; take out next part, but comment on it in discussion: CV would be a technique to see how stable results are, cite something about CV in general}
+\todo{Marjolein: With different specification of the GLMM tree (no square root, parameter stability tests at mom level, random intercepts of children nested within mothers) the results have changed, still need to adjust description}.
 
@@
 
-<<cv_lmm_tree, echo=FALSE, results=tex>>=
+<<cv_lmm_tree, echo=FALSE, eval=FALSE, results=tex>>=
 results <- data.frame(MSE_tree = rep(NA, times = 10), 
@@
 \FloatBarrier
-\section{\edc{Application Example 2}: Subgroup Detection in Rasch Models}
+\section{Application Example 2: Subgroup Detection in Rasch Models}
 \label{sec:TutorialRasch}
@@
 Finally, as with any exploratory method, MOB should be considered as a tool for generating hypothesis and not as a confirmatory technique.
-While the MOB algorithm, as it is implemented in the R packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory testing of hypothesis it is recommended to use fresh data or resampling techniquest like cross validation.
+While the MOB algorithm, as it is implemented in the \texttt{R} packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory testing of hypothesis it is recommended to use fresh data or resampling techniquest like cross validation.
 
@@
 e.g.
-such as cross validation?}
+such as cross validation.}
 
-\todo[inline]{Write about future work.}
+\todo{Write about future work.}
 
@@
 
-\section{Association between Age and PPVT scores before and after transformation}
-\label{sec:AppendixA}
-
-\begin{figure}[h!]
-\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
-<<echo=FALSE, fig=TRUE, width=6, height=4>>=
-par(mfrow = c(1, 2))
-plot(HS_dat$Age, HS_dat$PPVT, cex = .7, xlab = "Age", ylab = "PPVT",
-     cex.lab = .7, cex.axis = .7)
-plot(sqrt(HS_dat$Age), HS_dat$PPVT, ylab = "PPVT", xlab = "Age (square root)", 
-     cex = .7, cex.lab = .7, cex.axis = .7)
-@
-\end{figure}
+%\section{Association between Age and PPVT scores before and after transformation}
+%\label{sec:AppendixA}
+%
+%
+%\begin{figure}[h!]
+%\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
+%<<echo=FALSE, fig=TRUE, width=6, height=4>>=
+%par(mfrow = c(1, 2))
+%plot(HS_dat$Age, HS_dat$PPVT, cex = .7, xlab = "Age", ylab = "PPVT",
+%     cex.lab = .7, cex.axis = .7)
+%plot(sqrt(HS_dat$Age), HS_dat$PPVT, ylab = "PPVT", xlab = "Age (square root)", 
+%     cex = .7, cex.lab = .7, cex.axis = .7)
+%@
+%\end{figure}
 
 \section{Items of the culture scale for application example 2}
-\label{sec:AppendixB}
+\label{sec:AppendixA}
 

### 2023-10-30 _MOB_paper.tex
Revisions MF 30-10-2023

--- 2023-10-20 _MOB_paper.tex
@@
 %note from Caro: I found Fig 1 containing only the boxplot irritating and have moved it into one panel with the tree
+
+\begin{figure}[h]
+\caption{Left: Univariate distribution of the response variable. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
+\begin{subfigure}[][][t]{.4\textwidth}
+\includegraphics{_MOB_paper-003}
+\end{subfigure}
+\begin{subfigure}[][][b]{.7\textwidth}
+\includegraphics{_MOB_paper-004}
+\end{subfigure}
+\label{fig:toy}
+\end{figure}
 
@@
 The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting.
-The subgroup-specific distributions of the response variable are presented in end nodes at the bottom of the tree.
-The fact that Figure~\ref{fig:toy}, right, displays more than one end node confirms that one global model for all observations cannot appropriately capture the pattern in the data. 
-Figure~\ref{fig:toy}, right, also shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
+The subgroup-specific distributions of the response variable are presented in the end nodes at the bottom of the tree.
+The fact that the tree in Figure~\ref{fig:toy} displays more than one end node confirms that one global model for all observations does not appropriately capture the pattern in the data. 
+The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
 This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.
-
-\begin{figure}
-\caption{Left: Univariate distribution of the response variable, with global mean indicated by triangle. Right: Tree with group-specific distributions of the response variable in the terminal nodes.}
-\begin{subfigure}[][][t]{.4\textwidth}
-\includegraphics{_MOB_paper-003}
-\end{subfigure}
-\begin{subfigure}[][][b]{.6\textwidth}
-\includegraphics{_MOB_paper-004}
-\end{subfigure}
-\label{fig:toy}
-\end{figure}
 
@@
 As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step and anchoring \citep[see][and application example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+Together with a purification step \citep[see][and application example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
 This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
 %\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
-For example, items that show DIF between different language groups can often be improved by means of making sure that in all translations the meaning is as similar as possible, that the words employed in the translations for the different languages are equally frequently used, etc.
-In other situations, sources of DIF might be harder to eliminate, so that often DIF items are excluded from a test.
+For example, items that show DIF between different language groups may often be improved by means of making sure that in all translations the meaning is as similar as possible, and/or that the words employed in the translations for the different languages are equally frequently used.
+In other situations, sources of DIF might be harder to eliminate, so that often DIF items may be excluded from a test.
 Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.
@@
 Participation in Head Start takes place from ages 3 through 5.
-\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF).
+\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF) \todo{Marjolein, add reference}.
 
-The sample consists of 273 families with at least two children, where at least one child participated in Head Start and at least one child did not participate in Head Start or any other preschool program.
+The sample consists of 273 families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other preschool program.
 Data from children in those families who participated in another preschool program were excluded.
-As such, siblings who did not participate in Head Start serve as a natural control to assess the effects of participating in Head Start.
-The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF), we will this model PPVT trajectories over time.
+The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF) \todo{Marjolein: Add reference}, we will thus model PPVT trajectories over time.
 There were an average number of 2.14 PPVT scores per child, for 29\% of the children there was only a single PPVT score available.
@@
 
-\begin{table}
-
-\caption{\label{tab:performance}Cross-validated performance of LMMs and LMM trees.}
-\begin{tabular}[t]{ccccc}
-\toprule
-method & MSE & SD & number.of.splits & R2\\
-\midrule
-LMM tree & 259.595 & 58.786 & 2.5 & 0.780\\
-LMM & 297.926 & 77.296 & NA & 0.747\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \newpage
 \FloatBarrier
-\section{\edc{Application Example 2}: Subgroup Detection in Rasch Models}
+\section{Application Example 2: Subgroup Detection in Rasch Models}
 \label{sec:TutorialRasch}
-
-\todo[inline]{Comment on new example: I now use data from Bogen 20 of the SPISA dataset.
-I use the student subset, and complete cases on all variables.
-The story is less clear, but the tree is still stopped early, and we can display which items have dif.
-}
-
-\begin{itemize}
-\item Example Rasch tree: Data from SPISA (data from Trepte \& Verbeet; Bogen 20 (as in psychotree, but from the whole dataset, not only a subsample))
-\item take the subsample of students (only students); complete cases; $N = 9769$
-\item items could probably appear in Appendix??
-\begin{enumerate}
-\item Which painter created this painting? – Andy Warhol.
-\item What do these four buildings have in common? – All four were designed by the same architects.
-\item Roman numbers: What is the meaning of CLVI? – 156.
-\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
-\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
-\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
-\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
-\item In which city is this building located? – Paris.
-\item Which one of the following operas is not by Mozart? – Aida.
-\end{enumerate}
-\end{itemize}
-
-We load the data and inspect the first rows.
-there are several covariates: age (continuous), gender (male, female, missing), area (Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, Sports) and 9 items from the culture scale (true/false 0/1).
-
-\begin{Schunk}
-\begin{Sinput}
- head(covar_data)
-\end{Sinput}
-\begin{Soutput}
-    Age Gender                    Area
-90   21   Male       Law and Economics
-163  25   Male          no Information
-402  22   Male Agriculture & Nutrition
-449  20   Male          no Information
-518  21 Female          no Information
-649  23   Male          no Information
-\end{Soutput}
-\begin{Sinput}
- head(culture_scale)
-\end{Sinput}
-\begin{Soutput}
-     i28 i29 i30 i31 i32 i33 i34 i35 i36
-[1,]   1   1   0   1   0   1   1   1   1
-[2,]   1   1   1   1   0   1   1   0   1
-[3,]   0   1   0   1   1   0   1   1   0
-[4,]   0   1   1   1   1   1   1   1   1
-[5,]   1   0   1   0   0   1   1   1   1
-[6,]   1   1   1   1   0   1   1   1   1
-\end{Soutput}
-\end{Schunk}
-
-We create a dataset \texttt{dat\_SPISA} that contains the data of the covariate and the item responses from the culture scale.
-By assigning the whole scale to the dataset using \texttt{\$}, we can later access item responses from all items using \texttt{\$culture}.
-\begin{Schunk}
-\begin{Sinput}
- dat_SPISA <- covar_data
- dat_SPISA$culture <- culture_scale
-\end{Sinput}
-\end{Schunk}
-
-The Rasch tree can be fitted with the function \texttt{raschtree} from the R package \texttt{psychotree}.
-We use the typical formula syntax with the item responses of the culture scale on the left hand side of $\sim$, and the covariates on the right hand side.
-
-\begin{Schunk}
-\begin{Sinput}
- library(psychotree)
- Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
-                                data = dat_SPISA)
-\end{Sinput}
-\end{Schunk}
-
-We can plot the Raschtree using the plot function.
-But it is very big, lots of splits, we don't know whether the splits are due to substantial differences in item difficulty parameters, because the item difficulty profiles in the end nodes are hard to interpret/compare.
-
-\begin{Schunk}
-\begin{Sinput}
- plot(Raschtree_culture)
-\end{Sinput}
-\end{Schunk}
-\includegraphics{_MOB_paper-021}
-
-\todo[inline]{The are covariate split does not look pretty yet. But I will try to make it more pretty later on.. }
-
-Mantel-Haenszel trees use the Mantel-Haenszel effect size measure for DIF.
-it has three categories (A: negligible, B: moderate, C: large).
-If none of the items has DIF in category B or C (default, but can also be changed so that DIF has to be C), the tree is stopped from growing.
-
-The software must be installed from github, but will (probably) be implemented in the psychotree package some time in the future
-
-\begin{Schunk}
-\begin{Sinput}
- devtools::install_github("mirka-henninger/raschtreeMH")
- library(raschtreeMH)
-\end{Sinput}
-\end{Schunk}
-
-Syntax is very similar to the classical Rasch tree, but an additional argument \texttt{stopfun}, where the mantelhaneszel stopping function can be selected (but also own stopping functions can be provided; see Henninger et al., 2023).
-The user also has to indicate what kind of purification strategy should be used (none, 2step, iterative).
-Iterative is recommended.
-Optionally, the user can specify the stopping criterion.
-In this case, tree growing is stopped when all items fall into category ``A''.
-
-\begin{Schunk}
-\begin{Sinput}
- Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area, 
-                                   data = dat_SPISA, 
-                                   stopfun= stopfun_mantelhaenszel(
-                                     purification = "iterative",
-                                     stopcrit = "A"))
-\end{Sinput}
-\end{Schunk}
-
-For technical reasons, the information about the effect size and ETS classification are not saved in the tree object itself but have to be added afterwards using the \texttt{add\_mantelhaenszel} function.
-
-\begin{Schunk}
-\begin{Sinput}
- Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture, 
-                                            purification = "iterative")
-\end{Sinput}
-\end{Schunk}
-
-After we have added the information about the Mantel-Haenszel effect size to the Rasch tree, we can now visualize the Rasch tree together with the effect size measure.
-If we now plot the Mantel-Haenszel Rasch tree, we see that the tree is much more concise with a lower number of splits.
-In addition, we see the number of items classified as A, B, or C in each node.
-For instance, we see that one item is classified in category B in Node 1, and 4 items as C.
-
-\begin{Schunk}
-\begin{Sinput}
- plot(Raschtree_MH_culture)
-\end{Sinput}
-\end{Schunk}
-\includegraphics{_MOB_paper-026}
-
-We can also color the items in the end node profiles according to a split in the tree. Here, we see that for Node 1 (split on the covariate Gender), items 1, 6, 7, 8, 9 show DIF in categories B or C.
-
-\begin{itemize}
-\item 1 painting by Andy Warhol
-\item 6: novel by Daniel Kehlmann: Die Vermessung der Welt
-\item 7: city of the Buddenbrooks: Lübeck
-\item 8: city with building: Paris
-\item 9: opera not by Mozart: Aida
-\end{itemize}
-
-\begin{Schunk}
-\begin{Sinput}
- plot(Raschtree_MH_culture, color_by_node = 1)
-\end{Sinput}
-\end{Schunk}
-\includegraphics{_MOB_paper-027}
-
-We can also color by Node 2 (split on covariate Age).
-Also here, Item 1 shows DIF in category B.
-It seems to be easier for younger respondents ($\leq 24$) compared to older respondents (painting by Andy Warhol).
-
-\begin{Schunk}
-\begin{Sinput}
- plot(Raschtree_MH_culture, color_by_node = 2)
-\end{Sinput}
-\end{Schunk}
-\includegraphics{_MOB_paper-028}
-
-For Node 4 (also split on covariate Age), Items 1 and 9 show DIF, such that Item 1 and 9 are easier for younger respondents (Age $\leq 29$).
-(Item 1: Painting by Andy Warhol; Item 9: Opera not by Mozart: Aida)
-
-\begin{Schunk}
-\begin{Sinput}
- plot(Raschtree_MH_culture, color_by_node = 4)
-\end{Sinput}
-\end{Schunk}
-\includegraphics{_MOB_paper-029}
-
-\todo[inline]{Maybe a small note on anchoring?
-Item difficulty profiles are anchoring on the items that do not show DIF.
-This facilitates comparisons across end nodes (but maybe we do not want to say that??)}
-
-The information about the Mantel-Haenszel effect size measure for each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
-it contains the information on classification of each item in e ach node, the computed Mantel-Haenszel effect size measure, the type of purification, and if purification was iterative how many iterations were realized.
-
-\begin{Schunk}
-\begin{Sinput}
- Raschtree_MH_culture$info$mantelhaenszel
-\end{Sinput}
-\begin{Soutput}
-$classification
-      node4 node2 node1
-item1 "B"   "B"   "C"  
-item2 "A"   "A"   "A"  
-item3 "A"   "A"   "A"  
-item4 "A"   "A"   "A"  
-item5 "A"   "A"   "A"  
-item6 "A"   "A"   "C"  
-item7 "A"   "A"   "C"  
-item8 "A"   "A"   "B"  
-item9 "B"   "A"   "C"
-
-$mantelhaenszel
-            node4       node2        node1
-item1 -1.53377668 -1.03559992 -1.802977840
-item2  0.72945873 -0.14242027  0.377084860
-item3 -0.35893502  0.51100000  0.127575191
-item4 -0.05897255  0.40879442 -0.007385896
-item5  0.41615400 -0.08996266  0.128463797
-item6  0.42418247 -0.16529252 -1.875191790
-item7 -0.87987154 -0.57949560 -1.620265527
-item8  0.69240041  0.54735808 -1.083093749
-item9 -1.00725807 -0.39502640 -1.798520587
-
-$purification
-      node4       node2       node1 
-"iterative" "iterative" "iterative"
-
-$purification_counter
-node4 node2 node1 
-    1     1     2 
-\end{Soutput}
-\end{Schunk}
-
-\FloatBarrier
-\section{Discussion}
-
-\todo{next paragraph moved to here, where exactly?}
-An alternative framework for detecting groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
-Mixture modelling aims at identifying latent classes of persons with different properties.
-It can also be combined with covariates.
-A comparison of both approaches is given by \citet{FriStrZei:2014}.
-
-In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.
-The main advantage of MOB is that it can detect groups of persons with different model parameters in a data-driven manner.
-This makes it more flexible for detecting differences that were not hypothesized by the researcher.
-For example in DIF analysis, it is often the case that obvious sources of DIF have already been avoided by the content experts during item creation.
-Any remaining DIF is unexpected.
-Therefore, any DIF detection approach that relies on the researchers correctly specifying the exact groups of persons that exhibit DIF can miss DIF if it is associated with other (combinations of) covariates or other cutpoints than the ones investigated.
-In this sense, the more exploratory approach of Rasch trees has higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
-
-The same holds for parameter differences in mixed-effects models, where the substantial hypotheses are typically about the fixed effects of the interventions, not necessarily about all possible subgroup-specific ....
-\todo{Marjolein add if this makes any sense here}
-
-For our two application examples, we could show that MOB can detect ..., performs automated variable selection ...
-
-\subsection{Limitations}
-
-Due to the exploratory nature of MOB, it does not allow for hypothesis tests, as there is no valid way to account for the exploratory searching of the subgroups.
-Thus, even if detected subgroups are substantively meaningful or differences are large, if researchers want to ascertain statistical significance of the subgroup differences, this should be done on new data using confirmatory techniques.
-
-\todo[inline]{Mention shortcomings.}
-
-\todo[inline]{Write about future work.}
-
-sample size dependence, similar ideas like MH trees also possible for other MOB?
-\todo[inline]{possibly Mirka add something about visual interpretation?
-because effect sizes in mixed effects models are not very clear}
-
-The Rasch tree method has been extended to include an effect size measure that can stop the tree from growing if the effect size is non-substantial, but also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
-At the same time, effect sizes are less straightforward to calculate and interpret in linear mixed models, and therewith in the linear mixed effects model tree.
-A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
-These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s).
-A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.
-\todo[inline]{This is my (Mirka) suggestion for the interpretation techniques. Please feel free to add, edit, delete, however you prefer!}
-
-The Mantel-Haenszel effect size measure for DIF will be integrated in the \texttt{psychotools} package.
-
-\todo{Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
-In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}
-
-\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
-In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}
-
-\bibliography{_MOB_paper}
-
-\todo{Add DOIs.}
-
-\newpage
-\appendix
-
-\section{Appendix A: Association between Age and PPVT scores before and after transformation}
-\label{sec:AppendixA}
-
-\begin{figure}
-\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
-\includegraphics{_MOB_paper-031}
-\end{figure}
-
-\end{document}

### 2023-10-30 _MOB_paper.Rnw
Comments to manuscript

@@
 That is, it could result in detection of spurious subgroups.
+\todo[inline]{I am sure you know much better than me what happens if the cluster argument is used and you have discussed that a lot with Achim.
+But I am not $100\%$ sure that this is the case (and maybe I am completely wrong here), because my pilot simulations have always shown increased type-I error rates even when I used the cluster argument.
+I have only seen $5\%$ type-I error rates with the Wang \& Merkle approach where the covariates are aggregated before they enter the score test.
+Maybe we want to rephrase it in a way as you have phrased it in the article with Achim: ``[by using the cluster argument,] so-called clustered covariances are obtained, which account for dependence between observations within the same cluster (Zeileis, Köll, \& Graham, 2020)''.
+I think I would be more comfortable with a less direct formulation here, and then maybe we can further discuss this for the revision when we have a bit more time.
+\\
+A suggestion: ``By using the cluster argument, we account for dependence between observations within the same cluster.''.
+And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several end nodes, and that the cluster argument also helps for that?
+If that's correct, this could also be a good place to add a sentence about this advantage?}.
 Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.
@@
 
-sample size dependence, similar ideas like MH trees also possible for other MOB?
-\todo[inline]{possibly Mirka add something about visual interpretation?
-because effect sizes in mixed effects models are not very clear}
-
 The Rasch tree method has been extended to include an effect size measure that can stop the tree from growing if the effect size is non-substantial, but also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
@@
 A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.
-\todo[inline]{This is my (Mirka) suggestion for the interpretation techniques. Please feel free to add, edit, delete, however you prefer!}
 
-\todo{Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
+\color{blue} Using cluster-level predictor variables is desirable in particular when using MOB for longitudinal analysis.
+Therefore, building on recent work by \citet{FokkyZeil23}, we will further evaluate how the level of splitting variables can be taken into account best when using MOB, for instance using a variant of parameter stability tests to study heterogeneity on the cluster level \citep{Wang2021a}.
+
+\color{black}
+\todo[inline]{\color{blue} I (Mirka) added a short point on future work regarding cluster-level predictors and their Type-I error rate here.
+$\rightarrow$ this paragraph can also transition well to your comment on including parameter stability tests for random-effects parameter \color{black}Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
 In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}
 
-\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
+\todo[inline]{Mention that transformations chosen in growth curve modeling may yield different effects.
 In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}

### 2023-10-30 _MOB_paper.tex
Comments to manuscript

@@
 \label{sec:TutorialRasch}
+
+We will illustrate the usage of Rasch trees and the Mantel-Haenszel effect size measure by means of a dataset from \citet{SPISA:book}. An online general knowledge quiz was conducted by the German
+weekly news magazine SPIEGEL.
+Below we will use the abbreviation \texttt{SPISA} for the name of the data set, because the quiz was termed ``students' PISA'' by the magazine, but it is not related to the ``real'' PISA study by the OECD.
+The data set contains the quiz results as well as sociodemographic data.
+In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $9769$ complete cases of university students for illustration.
+The wording of the items is provided in Appendix~\ref{sec:AppendixB}.
+
+Three possible partitioning variables are considered: \texttt{Gender} (in this data set coded only as male, female, or missing), \texttt{Age} (continuous), and \texttt{Area} (Area of study: Language \& Culture, Law \& Economics, Medicine \& Health, Engineering, Sciences, Pharmacy, Geography, Agriculture \& Nutrition, or Sports).
+The nine items from the culture scale are scored with 0 (incorrect answer) or 1 (correct answer).
+The quiz data have been prepared in a format where the nine items of the culture scale are not stored as individual columns of the data set, but the complete matrix of item responses is stored like a single variable \citep[see][]{raschtree:vignette} under the name \texttt{culture}.
+This means that we can access the item responses for all items jointly using the name \texttt{culture} in the following R commands.
+The first few rows of the data set look like this:
+
+\todo{I usually use load for Rda files, Marjolein uses readRDS, is there an advantage? use the same in both examples.
+Mirka: from what I understand, readRDS is for loading .rds/.rda files with only one r object in it, while load is for .rda files with one or more r object in it.
+I would opt for .rda, because I think it is more general?
+Marjolein: I think if we include datasets in supplemental materials, we should have files which are only 1 dataset, not sets of datasets or objects.
+I think this will make the examples more simple.
+In practice, users will probably also have only 1 dataset.}
+
+\begin{Schunk}
+\begin{Sinput}
+ load("dat_SPISA.Rda")
+ #head(dat_SPISA, 3) ## commented out because will not compile, dat_SPISA is not
+                     ## an object in current wd
+\end{Sinput}
+\end{Schunk}
+
+The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that in the following we will fit Rasch trees by means of the function \texttt{raschtree} from the R package \texttt{psychotree}.
+However, model-based recursive partitioning is also available in \texttt{psychotree} for other IRT models.
+
+The model which we would like to test for parameter stability in this example, the Rasch model, is based only on the matrix of item responses \texttt{culture}.
+It does not contain predictor variables of its own.
+Therefore, the formula syntax for the \texttt{raschtree} function looks simpler than that for the \texttt{lmertree} function in application example 1: On the left hand side of the $\sim$ symbol, we provide the item response matrix \texttt{culture} for the Rasch model.
+On the right hand side of the $\sim$ symbol, we provide the possible partitioning variables \texttt{Gender}, \texttt{Age}, and \texttt{Area}.
+
+\begin{Schunk}
+\begin{Sinput}
+ library("psychotree")
+ Raschtree_culture <- raschtree(culture ~  Gender + Age + Area,
+                                data = dat_SPISA)
+\end{Sinput}
+\end{Schunk}
+
+Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} method.
+The item parameters for the nine culture items are displayed in the end nodes of the tree.
+However, it is hard to interpret the tree structure because such a large number of splits was created.
+Moreover, it is hard to tell whether all splits are due to substantial amounts of DIF, or whether some of the splits are due to small DIF effects that only became statistically significant due to the large sample size.
+
+\begin{Schunk}
+\begin{Sinput}
+ plot(Raschtree_culture)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{_MOB_paper-020}
+
+To assess whether DIF effects are substantial or negligible, it is possible to use the Mantel-Haenszel DIF effect size measure as an additional stopping criterion.
+It has three categories (A: negligible, B: moderate, C: large).
+If none of the items shows DIF in category B or C\footnote{This is the default setting.
+It can also be changed so that splitting is stopped if no item shows DIF in category C.}, the tree is stopped from growing.
+At the moment, the \texttt{R} functions for the Mantel-Haenszel stopping criterion and additional visualizations based on the Mantel-Haenszel effect size are available from GitHub.
+They will be made available as part of the \texttt{psychotree} package in the future.
+
+In order to install the functions from the author's GitHub page and make them available in the current \texttt{R} session, use the following two commands:
+
+\begin{Schunk}
+\begin{Sinput}
+ library(devtools)
+ install_github("mirka-henninger/raschtreeMH")
+ library("raschtreeMH")
+\end{Sinput}
+\end{Schunk}
+
+We can use a syntax very similar to that of the original \texttt{raschtree} function, but now we also provide the additional argument \texttt{stopfun}.
+We want to use the Mantel-Haenszel stopping function (\texttt{stopfun\_mantelhaenszel}) together with iterative purification.
+It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
+
+For the purification strategy, there are three options: \texttt{none}, \texttt{2step} and \texttt{iterative}.
+Purification means that items which have already been diagnosed with DIF are taken out of the sum score, which is used as the matching criterion in the final DIF test.
+It is highly recommended to use a purification strategy, because otherwise false positives (i.e., artificial DIF) may occur \citep[cf., e.g.,][and the references therein]{DebStrZei:2022:CRC,HennyDeba23,Kopfetal:2015:EPM}.
+In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2.
+When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
+The final set of DIF-free items resulting from the purification is also used for aligning the item parameters for each group comparison in the tree plots with color coding, which are displayed below.
+
+\begin{Schunk}
+\begin{Sinput}
+ Raschtree_MH_culture <- raschtree(culture ~  Gender + Age + Area,
+                                   data = dat_SPISA,
+                                   stopfun= stopfun_mantelhaenszel(
+                                     purification = "iterative"))
+\end{Sinput}
+\end{Schunk}
+
+For technical reasons, the information about the Mantel-Haenszel effect size and classification is not saved in the tree object itself, but has to be added afterwards using the \texttt{add\_mantelhaenszel} function.
+
+\begin{Schunk}
+\begin{Sinput}
+ Raschtree_MH_culture <- add_mantelhaenszel(Raschtree_MH_culture,
+                                            purification = "iterative")
+\end{Sinput}
+\end{Schunk}
+
+After we have added this information to the Rasch tree, we can now visualize it with additional features.
+When we now plot the Rasch tree with Mantel-Haenszel stopping by means of the \texttt{plot} function, we see that the tree is much more concise with a lower number of splits.
+In addition, in each node the number of items classified as A, B, or C is displayed.
+For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and 4 items are classified in category C (large DIF).
+
+\begin{Schunk}
+\begin{Sinput}
+ plot(Raschtree_MH_culture)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{_MOB_paper-025}
+
+We can also color the DIF items in the end node profiles according to a particular split in the tree. We will illustrate this with two examples. 
+First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C. As a reminder, the content of these items was:
+
+\begin{itemize}
+\item 1: Painting by Andy Warhol
+\item 6: Novel by Daniel Kehlmann: Die Vermessung der Welt
+\item 7: City of the Buddenbrooks: Lübeck
+\item 8: City with building: Paris
+\item 9: Opera not by Mozart: Aida
+\end{itemize}
+
+\begin{Schunk}
+\begin{Sinput}
+ plot(Raschtree_MH_culture, color_by_node = 1)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{_MOB_paper-026}
+
+These items are %(relative to the DIF-free items for this group comparison) 
+easier or less difficult to answer for participants with female or missing gender.
+
+% We can also color the DIF items according to node 2 (split in the covariate \texttt{Age}).
+Again, item 1 (Painting by Andy Warhol) shows DIF in category B.
+It is (relative to the DIF-free items for this group comparison) more difficult to answer for younger respondents (age $\leq 24$) compared to older respondents.
+
+% <<fig = TRUE, width = 9, height = 7>>=
+% plot(Raschtree_MH_culture, color_by_node = 2)
+% @
+
+For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison) 
+more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
+
+\begin{Schunk}
+\begin{Sinput}
+ plot(Raschtree_MH_culture, color_by_node = 4)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{_MOB_paper-027}
+
+Detailed information about the Mantel-Haenszel criterion for each item in each node can also be accessed from the Rasch tree object using \texttt{\$info\$mantelhaenszel}.
+It contains the DIF classification and the value of the Mantel-Haenszel effect size measure for each item in each node, as well as the type of purification that was employed in each node, and in the case of iterative purification how many iterations were conducted.
+
+\begin{Schunk}
+\begin{Sinput}
+ Raschtree_MH_culture$info$mantelhaenszel
+\end{Sinput}
+\begin{Soutput}
+$classification
+      node4 node2 node1
+item1 "B"   "B"   "C"  
+item2 "A"   "A"   "A"  
+item3 "A"   "A"   "A"  
+item4 "A"   "A"   "A"  
+item5 "A"   "A"   "A"  
+item6 "A"   "A"   "C"  
+item7 "A"   "A"   "C"  
+item8 "A"   "A"   "B"  
+item9 "B"   "A"   "C"
+
+$mantelhaenszel
+            node4       node2        node1
+item1 -1.53377668 -1.03559992 -1.802977840
+item2  0.72945873 -0.14242027  0.377084860
+item3 -0.35893502  0.51100000  0.127575191
+item4 -0.05897255  0.40879442 -0.007385896
+item5  0.41615400 -0.08996266  0.128463797
+item6  0.42418247 -0.16529252 -1.875191790
+item7 -0.87987154 -0.57949560 -1.620265527
+item8  0.69240041  0.54735808 -1.083093749
+item9 -1.00725807 -0.39502640 -1.798520587
+
+$purification
+      node4       node2       node1 
+"iterative" "iterative" "iterative"
+
+$purification_counter
+node4 node2 node1 
+    1     1     2 
+\end{Soutput}
+\end{Schunk}
+
+The displayed information again shows that some items show only negligible DIF of category A in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
+The values of the Mantel-Haenszel effect size measure are negative when the item difficulty is lower in the right daughter node than in the left daughter node.
+For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer %(relative to the DIF-free items for this group comparison)
+for participants with female or missing gender, which correspond to the right daughter node of node 1.
+% Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.
+However, this item has a smaller DIF effect size.
+Thus it falls into DIF category A (negligible) and was not relevant for the split in \texttt{Gender}.
+
+\FloatBarrier
+\section{Discussion}
+
+In this tutorial paper we have outlined the rationale of the MOB algorithm and how it can be used for detecting parameter differences in mixed-effects and Rasch models.
+The main advantage of MOB is that it can detect groups of persons with different model parameters in a data-driven manner.
+This makes it more flexible for detecting differences that were not hypothesized by the researcher.
+For example in DIF analysis, it is often the case that obvious sources of DIF have already been avoided by the content experts during item creation.
+Any remaining DIF is unexpected.
+Therefore, any DIF detection approach that relies on the researchers correctly specifying the exact groups of persons that exhibit DIF can miss DIF if it is associated with other (combinations of) covariates or other cutpoints than the ones investigated.
+In this sense, the more exploratory approach of Rasch trees has higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
+The same argument holds for parameter differences in mixed-effects models, where the substantial hypotheses formulated a priori are typically about global intervention effects, not about possible subgroup-specific differences.
+
+However, researchers should be aware that a covariate-based approach like MOB will only be able to detect parameter differences if the relevant covariates were recorded in the study and supplied to the algorithm.
+If no or not many potentially relevant covariates are available, an alternative framework for detecting previously unknown groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
+Mixture modelling aims at identifying latent classes of persons with different properties.
+It can also be combined with observed covariates.
+A comparison of MOB and mixture modelling is given by \citet{FriStrZei:2014}.
+
+Another important caveat is that those covariates that are selected for splitting by the MOB algorithm are not necessarily causal for the observed parameter differences.
+They might also be proxies for other, unobserved covariates.
+For example, if DIF in the item parameters of a test on reading ability is found between different cities or districts, a variety of factors could be causally driving these effects, such as different compositions of native languages or sociodemographics.
+
+Finally, as with any exploratory method, MOB should be considered as a tool for generating hypothesis and not as a confirmatory technique.
+While the MOB algorithm, as it is implemented in the \texttt{R} packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory testing of hypothesis it is recommended to use fresh data or resampling techniquest like cross validation.
+
+\todo[inline]{I (Caro) have added a few more thoughts for the discussion above - Marjolein maybe this is already sufficient in length?
+please feel free to use anything from above or below and get back to me if you add/keep points where we might disagree}
+\subsection{Limitations of MOB}\todo{I (Caro) would not make subsections within discussion}
+
+Due to the exploratory nature of MOB, it does not allow for hypothesis tests\todo[inline]{I (Caro) think this will sound confusing because earlier in the paper we explain that MOB is indeed based on significance tests, and due to the fact that the stopping is based on the significance test and (at least in party and psychotree) we use Bonferroni for adjusting against the number of covariates, it is not like we are doing completely uncontrolled multiple testing - in the Raschtree paper we write about this: Moreover, it is important to note that our model-based recursive partitioning algorithm is not affected by an inflation of chance due to its recursive nature.
+Indeed, several statistical tests are successively conducted in a Rasch tree—but each test is conducted only if the previous test yielded a significant result.
+In this sense, the recursive approach forms a closed testing procedure, which does not lead to an inflation of chance as is well known from the literature on multiple comparisons (Marcus, Peritz, \& Gabriel, 1976; Hochberg \& Tamhane, 1987).
+For the Rasch tree, this means that the postulated significance level holds for the entire tree, not only for each individual split.
+This ensures that DIF is not erroneously detected as an artifact of the recursive nature of the algorithm.
+-- another, related, limitation that I would agree with and might be what you also had in mind is that the tree will not necessarily find the globally best partition, because every split is based on the previous split choices, here is what we wrote about this in the tree intro PsychMeth paper: Thus, variable selection in a single tree is affected by order effects similar to those present in stepwise variable selection approaches for parametric regression (that is also instable against random variation of the learning data, as pointed out by Austin and Tu 2004).
+In both recursive partitioning and stepwise regression, the approach of adding one locally optimal variable at a time does not necessarily (or rather hardly ever) lead to the globally best model over all possible combinations of variables.
+-- should we write something similar here?}, as there is no valid way to account for the exploratory searching of the subgroups.
+Thus, even if detected subgroups are substantively meaningful or differences are large, if researchers want to ascertain statistical significance of the subgroup differences, this should be done on new data using confirmatory techniques.
+\todo{Marjolein write a little more here?
+e.g.
+such as cross validation.}
+
+\todo{Write about future work.}
+
+The Rasch tree method has been extended to include an effect size measure that can stop the tree from growing if the effect size is non-substantial, but also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
+At the same time, effect sizes are less straightforward to calculate and interpret in linear mixed models, and therewith in the linear mixed effects model tree.
+A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
+These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s).
+A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.
+
+\color{blue} Using cluster-level predictor variables is desirable in particular when using MOB for longitudinal analysis.
+Therefore, building on recent work by \citet{FokkyZeil23}, we will further evaluate how the level of splitting variables can be taken into account best when using MOB, for instance using a variant of parameter stability tests to study heterogeneity on the cluster level \citep{Wang2021a}.
+
+\color{black}
+\todo[inline]{\color{blue} I (Mirka) added a short point on future work regarding cluster-level predictors and their Type-I error rate here.
+$\rightarrow$ this paragraph can also transition well to your comment on including parameter stability tests for random-effects parameter \color{black}Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
+In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}
+
+\todo[inline]{Mention that transformations chosen in growth curve modeling may yield different effects.
+In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}
+
+\bibliography{_MOB_paper}
+
+\todo{Add DOIs.}
+
+\newpage
+\appendix
+
+%\section{Association between Age and PPVT scores before and after transformation}
+%\label{sec:AppendixA}
+%
+%
+%\begin{figure}[h!]
+%\caption{Association between Age and PPVT scores before (left) and after (right) transformation.}
+%<<echo=FALSE, fig=TRUE, width=6, height=4>>=
+%par(mfrow = c(1, 2))
+%plot(HS_dat$Age, HS_dat$PPVT, cex = .7, xlab = "Age", ylab = "PPVT",
+%     cex.lab = .7, cex.axis = .7)
+%plot(sqrt(HS_dat$Age), HS_dat$PPVT, ylab = "PPVT", xlab = "Age (square root)", 
+%     cex = .7, cex.lab = .7, cex.axis = .7)
+%@
+%\end{figure}
+
+\section{Items of the culture scale for application example 2}
+\label{sec:AppendixA}
+
+\begin{enumerate}
+\item Which painter created this painting? – Andy Warhol.
+\item What do these four buildings have in common? – All four were designed by the same architects.
+\item Roman numbers: What is the meaning of CLVI? – 156.
+\item What was the German movie with the most viewers since 1990? – Der Schuh des Manitu.
+\item In which TV series was the US president portrayed by an African American actor for a long time? – 24.
+\item What is the name of the bestselling novel by Daniel Kehlmann? – Die Vermessung der Welt (Measuring The World).
+\item Which city is the setting for the novel ‘Buddenbrooks’? – Lübeck.
+\item In which city is this building located? – Paris.
+\item Which one of the following operas is not by Mozart? – Aida.
+\end{enumerate}
+
+\end{document}

### 2023-10-31 Raschtree_MH.Rnw
updates MF

@@
 In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $\Sexpr{nrow(dat_SPISA)}$ complete cases of university students for illustration.
-The wording of the items is provided in Appendix~\ref{sec:AppendixB}.
+The wording of the items is provided in the \hyperref[sec:AppendixA]{Appendix}.
 
@@
 The nine items from the culture scale are scored with 0 (incorrect answer) or 1 (correct answer).
-The quiz data have been prepared in a format where the nine items of the culture scale are not stored as individual columns of the data set, but the complete matrix of item responses is stored like a single variable \citep[see][]{raschtree:vignette} under the name \texttt{culture}.
-This means that we can access the item responses for all items jointly using the name \texttt{culture} in the following R commands.
+The quiz data have been prepared in a format where the nine items of the culture scale are not stored as individual columns of the data set, but the complete matrix of item responses is stored as a single variable \citep[see][]{raschtree:vignette} under the name \texttt{culture}.
+This means that we can access the item responses for all items jointly using the name \texttt{culture} in the following \texttt{R} commands.
 The first few rows of the data set look like this:
@@
 load("dat_SPISA.Rda")
-#head(dat_SPISA, 3) ## commented out because will not compile, dat_SPISA is not
-                    ## an object in current wd
+head(dat_SPISA, 3)
 @
 
-The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that in the following we will fit Rasch trees by means of the function \texttt{raschtree} from the R package \texttt{psychotree}.
-However, model-based recursive partitioning is also available in \texttt{psychotree} for other IRT models.
+The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that we can continue to fit Rasch trees by means of function \texttt{raschtree} from the R package \texttt{psychotree}.
+Note that model-based recursive partitioning is also available in \texttt{psychotree} for other IRT models.
 
@@
 It does not contain predictor variables of its own.
-Therefore, the formula syntax for the \texttt{raschtree} function looks simpler than that for the \texttt{lmertree} function in application example 1: On the left hand side of the $\sim$ symbol, we provide the item response matrix \texttt{culture} for the Rasch model.
-On the right hand side of the $\sim$ symbol, we provide the possible partitioning variables \texttt{Gender}, \texttt{Age}, and \texttt{Area}.
+Therefore, the formula syntax for the \texttt{raschtree} function looks simpler than that for the \texttt{lmertree} function in \hyperref[sec:TutorialMixed]{Application Example 1}: On the left hand side of the \verb|~| symbol, we provide the item response matrix \texttt{culture} for the Rasch model.
+On the right hand side of the \verb|~| symbol, we provide the possible partitioning variables \texttt{Gender}, \texttt{Age}, and \texttt{Area}.
 
@@
 
-Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} method.
-The item parameters for the nine culture items are displayed in the end nodes of the tree.
+Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function.
+The resulting tree is depicted in Figure \ref{fig:MHtree1}.
+The item parameters for the nine culture items are displayed in the terminal nodes of the tree.
 However, it is hard to interpret the tree structure because such a large number of splits was created.
@@
 
-<<fig = TRUE, width = 22, height = 10>>=
+<<eval=FALSE>>=
 plot(Raschtree_culture)
 @
+
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items using the default stopping criterion based on statistical significance.}
+\begin{subfigure}{1.2\textwidth}
+<<fig = TRUE, echo=FALSE, width = 22, height = 10>>=
+plot(Raschtree_culture)
+@
+\end{subfigure}
+\label{fig:MHtree1}
+\end{figure}%
 
@@
 If none of the items shows DIF in category B or C\footnote{This is the default setting.
-It can also be changed so that splitting is stopped if no item shows DIF in category C.}, the tree is stopped from growing.
+It can also be changed so that splitting is stopped if no item shows DIF in category C.}, the tree is stopped from growing further.
 At the moment, the \texttt{R} functions for the Mantel-Haenszel stopping criterion and additional visualizations based on the Mantel-Haenszel effect size are available from GitHub.
@@
 
-We can use a syntax very similar to that of the original \texttt{raschtree} function, but now we also provide the additional argument \texttt{stopfun}.
+We can use a syntax very similar to that of the original \texttt{raschtree} function, but now we also specify the additional argument \texttt{stopfun}.
 We want to use the Mantel-Haenszel stopping function (\texttt{stopfun\_mantelhaenszel}) together with iterative purification.
-It is also possible to provide other, user defined stopping functions \citep[see][for details]{HennyDeba23}.
+It is also possible to provide other, user-defined stopping functions \citep[see][for details]{HennyDeba23}.
 
@@
 
-After we have added this information to the Rasch tree, we can now visualize it with additional features.
-When we now plot the Rasch tree with Mantel-Haenszel stopping by means of the \texttt{plot} function, we see that the tree is much more concise with a lower number of splits.
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion.}
+<<fig = TRUE, echo=FALSE, width = 9, height = 7>>=
+plot(Raschtree_MH_culture)
+@
+\label{fig:MHtree2}
+\end{figure}%
+
+After we have added this information to the Rasch tree, we can now visualize it with additional features:
+
+<<eval=FALSE>>=
+plot(Raschtree_MH_culture)
+@
+
+The result is presented in Figure~\label{fig:MHtree2}, which reveals that the Rasch tree is much more concise with a lower number of splits.
 In addition, in each node the number of items classified as A, B, or C is displayed.
@@
 
-<<fig = TRUE, width = 9, height = 7>>=
-plot(Raschtree_MH_culture)
-@
-
-We can also color the DIF items in the end node profiles according to a particular split in the tree. We will illustrate this with two examples. 
-First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C. As a reminder, the content of these items was:
+We can also color the DIF items in the end node profiles according to a particular split in the tree.
+We will illustrate this with two examples: First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
+As a reminder, the content of these items was:
 
 \begin{itemize}
+\setlength\itemsep{0em}
 \item 1: Painting by Andy Warhol
@@
 
-<<fig = TRUE, width = 9, height = 7>>=
+<<eval=FALSE>>=
 plot(Raschtree_MH_culture, color_by_node = 1)
 @
+
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion with additional visualization features.}
+<<fig = TRUE, echo=FALSE, width = 9, height = 7>>=
+plot(Raschtree_MH_culture, color_by_node = 1)
+@
+\label{fig:MHtree3}
+\end{figure}%
 
@@
 
+Alternatively, we can also color the DIF items in the end node profiles according to a later split in the tree:
+
+<<eval=FALSE>>=
+plot(Raschtree_MH_culture, color_by_node = 4)
+@
+
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion with additional visualization features.}
+<<fig = TRUE, echo=FALSE, width = 9, height = 7>>=
+plot(Raschtree_MH_culture, color_by_node = 4)
+@
+\label{fig:MHtree4}
+\end{figure}%
+
+The result is presented in Figure~\ref{fig:MHtree4}.
 For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison)
 more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
-
-<<fig = TRUE, width = 9, height = 7>>=
-plot(Raschtree_MH_culture, color_by_node = 4)
-@
 

### 2023-10-31 _MOB_paper.Rnw
updates MF

--- 2023-10-30 _MOB_paper.Rnw
@@
 For example, when children are tested at several time points, the time points are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools.
-In \hyperref[sec:TutorialMixed]{application example 1} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
+In \hyperref[sec:TutorialMixed]{Application Example 1} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
 In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model.
@@
 
-For \hyperref[sec:TutorialRasch]{application example 2}, we go back in the research process to the point where a new psychological test has been administered to a validation sample.
+For \hyperref[sec:TutorialRasch]{Application Example 2}, we go back in the research process to the point where a new psychological test has been administered to a validation sample.
 Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model.
@@
 Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF.
-In \hyperref[sec:TutorialRasch]{application example 2} we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.
+In \hyperref[sec:TutorialRasch]{Application Example 2} we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.
 
@@
 We illustrate the idea using a very simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables.
-%note from Caro: I woul call them partitioning variables or covariates, but not predictors, to save the term predictors for those x-variables that are part of the parametric model
 To keep the example simple, we apply MOB to a very basic global model, which comprises only an intercept.
@@
 Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
-%note from Caro: I found Fig 1 containing only the boxplot irritating and have moved it into one panel with the tree
 
@@
 Moreover, the trees will not grow as large as possible, but splitting is stopped when no more significant structural change is detected.
-While traditional tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (so called \textit{pruning}), modern tree and MOB algorithms employ significance tests as stopping criteria (and some can also use effect size measures, as we will see in application example 2).
+While traditional tree algorithms like those of \citet{Breetal:1984} and \citet{Qui:1993} grew very large trees and then cut them back (so called \textit{pruning}), modern tree and MOB algorithms employ significance tests as stopping criteria (and some can also use effect size measures, as we will see in \hyperref[sec:TutorialRasch]{Application Example 2}).
 This allows for stopping tree growing as soon as no significant differences can be detected anymore.
@@
 These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.
-\todo{we should discuss this in the application examples; there are some defaults but for models with many parameters it might make sense to increase them} Note that it is thus possible that no split will be implemented, when none of the specified partitioning variables show significant structural change.
+For reasons of interpretability, users may sometimes specify even higher requirements for the sample sizes of terminal nodes.
+Note that it is very well possible that no split will be implemented, when none of the specified partitioning variables show significant structural change.
 
@@
 The \texttt{R} functions for fitting GLMM trees are the \texttt{lmertree} and \texttt{glmertree} functions from package \texttt{glmertree}.
-The usage of this function will be illustrated in \hyperref[sec:TutorialMixed]{application example 1}.
+The usage of this function will be illustrated in \hyperref[sec:TutorialMixed]{Application Example 1}.
 Further mathematical and computational details about the GLMM tree model are also described in \cite{FokkySmit18} and \cite{FokkyZeil23}.
@@
 For the Rasch measurement model, the \texttt{R} function for conducting the MOB analysis is the \texttt{raschtree} function from package \texttt{psychotree} \citep{StroyKopf15}.
-The usage of this function will be illustrated in \hyperref[sec:TutorialRasch]{application example 2}.
+The usage of this function will be illustrated in \hyperref[sec:TutorialRasch]{Application Example 2}.
 We will see that, just like for GLMM trees, we need to specify which variables are part of the parametric model.
@@
 However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
-As we will illustrate in the first part of \hyperref[sec:TutorialRasch]{application example 2}, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+As we will illustrate in the first part of \hyperref[sec:TutorialRasch]{Application Example 2}, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
 Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
@@
 
-As we will show in application example 2, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citep[see][and application example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
-This can help generate hypothesis about the sources of DIF, as we will ilustrate in application example 2, and can also aid the decision how to proceed with the DIF items.
+As we will show in \hyperref[sec:TutorialRasch]{Application Example 2}, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
+Together with a purification step \citep[see][and Application Example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
+This can help generate hypothesis about the sources of DIF, as we will ilustrate in \hyperref[sec:TutorialRasch]{Application Example 2}, and can also aid the decision how to proceed with the DIF items.
 %\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
@@
 Participation in Head Start takes place from ages 3 through 5.
-\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth (NLSY; REF) \todo{Marjolein, add reference}.
+\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth \citep{NLSY}.
 
@@
 The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
-The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF) \todo{Marjolein: Add reference}, we will thus model PPVT trajectories over time.
-There were an average number of \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(1-mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there was only a single PPVT score available.
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT]{DunnyDunn81}, we will thus model PPVT trajectories over time.
+On average, there were \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(1-mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there was only a single PPVT score available.
 
@@
 
-Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's body height \todo{Marjolein: Add scale for measurement}.
+Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's body height %\todo{Marjolein: Add scale for measurement}.
 Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables.
@@
 With the second argument, we specified the dataset which contain the variables.
-With the \texttt{cluster} argument, we specified that the partitioning variables are measured at the level of the mothers.
-As a result, parameter stability tests will be performed on the appropriate level \citep{FokkyZeil23}.
-Not specifying the \texttt{cluster} argument would result in use of the default observation-level parameter stability tests, which would yield inflated type-I error rates.
-That is, it could result in detection of spurious subgroups.
-Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.
+With the \texttt{cluster} argument, we account for the dependence between observations within the same cluster.
+In this example, the partitioning variables are measured at the mother (or family) level, which we indicate by specifying \texttt{cluster = MotherID}.
+As a result, the parameter stability tests will employ so-called clustered covariances \citep{ZeilyKoll20}.
+Not specifying the \texttt{cluster} argument would result in use of the default observation-level covariances in the parameter stability tests, inflating type-I error rates when partitioning variables are measured on a higher level \citep{FokkyZeil23}.
+
+% Mirka: And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several end nodes, and that the cluster argument also helps for that?
+If that's correct, this could also be a good place to add a sentence about this advantage?
+Marjolein: The splitting up is avoided by having partitioning that are constant over time (and also within families, in this instance).
+The level of the parameter stability tests do not really avoid that.
+We could specify the cluster argument wiht time-varying as well as time-constant partitioning covariates.
+
+Finally, to aid interpretability we want to retain large enough subgroups, we specified the \texttt{minsize} argument.
+As a result, splits will only be implemented if the resulting nodes contain at least 250 observations.
 
@@
 \begin{figure}%
-\caption{}
+\caption{GLMM tree with trajectories of PPVT scores}
 <<fig=TRUE, echo=FALSE, fig.width=9, fig.height=4>>=
@@
 @
+\\\textit{Note}: Red lines depict average node-specific trajectories for siblings participating in Head Start; black lines depict average node-specific trajectories for siblings not participating in Head Start.
 \label{fig:lmm_tree}
@@
 The group with higher mother's AFTQ scores is further split based on race.
-The Black and Hispanic group is further split based on income.
+The Black and Hispanic group is further split based on mother's AFTQ score.
+
+\FloatBarrier
 
@@
                         Program = factor(rep(c("None", "HS"))))
-design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(6)), 
+design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 6), 
                                      re.form = ~0)
-design_df$`PPVT at age 18` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = sqrt(18)), 
+design_df$`PPVT at age 12` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 12), 
                                       re.form = ~0)
@@
 knitr::kable(design_df, format = "latex", digits = 2, 
-             caption = "Predicted PPVT scores at different ages.", 
+             caption = "Node-specific average predicted PPVT scores at different ages.", 
              label = "predictions", centering = FALSE,
@@
 
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
+To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12.
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
-This benefit remains the same over time for the group with lower mother's AFTQ scores.
-The benefit increases over time for White children with higher mother's AFTQ scores.
-Interestingly, the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
-These results correspond to the conclusions of \cite{Demi09}.
-
-\todo{Marjolein: With different specification of the GLMM tree (no square root, parameter stability tests at mom level, random intercepts of children nested within mothers) the results have changed, still need to adjust description}.
+This benefit increases over time for the group with higher mother's AFTQ scores.
+The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2).
+For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12.
+These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14.
+Fade out is reported to be particularly strong for Afriacn-American and very disadvantaged children.
 
@@
 The same argument holds for parameter differences in mixed-effects models, where the substantial hypotheses formulated a priori are typically about global intervention effects, not about possible subgroup-specific differences.
+GLMM trees allow for exploring whether fixed effects differ as a function of the partitioning variable; that is, identifying possible interactions the predictor of interest may be involved in.
 
 However, researchers should be aware that a covariate-based approach like MOB will only be able to detect parameter differences if the relevant covariates were recorded in the study and supplied to the algorithm.
-If no or not many potentially relevant covariates are available, an alternative framework for detecting previously unknown groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
+If no (or few) potentially relevant covariates are available, an alternative framework for detecting previously unknown groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
 Mixture modelling aims at identifying latent classes of persons with different properties.
@@
 
-Finally, as with any exploratory method, MOB should be considered as a tool for generating hypothesis and not as a confirmatory technique.
-While the MOB algorithm, as it is implemented in the \texttt{R} packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory testing of hypothesis it is recommended to use fresh data or resampling techniquest like cross validation.
+As with any exploratory method, MOB should be considered as a tool for generating hypotheses and not as a confirmatory technique.
+While the MOB algorithm, as it is implemented in the \texttt{R} packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory hypothesis testing it is recommended to use a new dataset, not used for fitting the tree.
+Further, resampling or $k$-fold cross validation techniques may be used to assess stability and accuracy of the model.
+\cite{PhilyRusc18} show how to use these techniques for assessing the stability of the set of selected variables and cutpoints.
+For GLMM trees, cross-validation can additionally be used for evaluating predictive accuracy \cite[e.g., ]{RooiyWeed20}, or for further optimizing predictive accuracy through the choice of model-fitting parameters, like the minimum number of observations the terminal nodes or the maximum tree depth \cite{JameyWitt23}.
 
-\todo[inline]{I (Caro) have added a few more thoughts for the discussion above - Marjolein maybe this is already sufficient in length?
+%\todo[inline]{I (Caro) have added a few more thoughts for the discussion above - Marjolein maybe this is already sufficient in length?
 please feel free to use anything from above or below and get back to me if you add/keep points where we might disagree}
-\subsection{Limitations of MOB}\todo{I (Caro) would not make subsections within discussion}
 
-Due to the exploratory nature of MOB, it does not allow for hypothesis tests\todo[inline]{I (Caro) think this will sound confusing because earlier in the paper we explain that MOB is indeed based on significance tests, and due to the fact that the stopping is based on the significance test and (at least in party and psychotree) we use Bonferroni for adjusting against the number of covariates, it is not like we are doing completely uncontrolled multiple testing - in the Raschtree paper we write about this: Moreover, it is important to note that our model-based recursive partitioning algorithm is not affected by an inflation of chance due to its recursive nature.
+%Due to the exploratory nature of MOB, it does not allow for hypothesis tests\todo[inline]{I (Caro) think this will sound confusing because earlier in the paper we explain that MOB is indeed based on significance tests, and due to the fact that the stopping is based on the significance test and (at least in party and psychotree) we use Bonferroni for adjusting against the number of covariates, it is not like we are doing completely uncontrolled multiple testing - in the Raschtree paper we write about this: Moreover, it is important to note that our model-based recursive partitioning algorithm is not affected by an inflation of chance due to its recursive nature.
 Indeed, several statistical tests are successively conducted in a Rasch tree—but each test is conducted only if the previous test yielded a significant result.
@@
 
-\todo{Write about future work.}
-
-sample size dependence, similar ideas like MH trees also possible for other MOB?
-\todo[inline]{possibly Mirka add something about visual interpretation?
-because effect sizes in mixed effects models are not very clear}
-
-The Rasch tree method has been extended to include an effect size measure that can stop the tree from growing if the effect size is non-substantial, but also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
-At the same time, effect sizes are less straightforward to calculate and interpret in linear mixed models, and therewith in the linear mixed effects model tree.
-A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
+\hyperref[sec:TutorialRasch]{Application Example 2} showed that the Rasch tree method can employ an effect size measure to stop the tree from growing if effect-size differences are non-substantial.
+This also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
+At the same time, effect sizes are less straightforward to calculate and interpret in mixed-effects models \citep{JuddyWest17}, and thereby also for GLMM trees.
+\todo{Interpretation methods might not work as expected for multilevel data, so I am not convinced we should discuss this and call it a remedy.}A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
 These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s).
 A comprehensive introduction, tutorial, and discussion into interpretation techniques for machine learning methods is given by \cite{Molnar2019} and \cite{Henninger2022c}.
-\todo[inline]{This is my (Mirka) suggestion for the interpretation techniques. Please feel free to add, edit, delete, however you prefer!}
 
-\todo{Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
+For GLMM trees, the use of cluster-level partitioning variables would commonly be desired in longitudinal analyses.
+As with traditional GLMMs, specification of the fixed and random effects requires careful consideration.
+For example, whether random slopes of time should be estimated, or whether random intercepts are sufficient to account for dependency between observations.
+Building on recent work by \citet{FokkyZeil23}, we will further study and develop methods to take various dependence structures into account.
+For instance, in future implementations of \texttt{glmertree}, use of parameter stability tests to also allow for detecting subgroup differences in the random-effects parameters \citep{Wang2021a} may be supported.
+
+%\todo{I (Mirka) added a short point on future work regarding cluster-level predictors and their Type-I error rate here.
+$\rightarrow$ this paragraph can also transition well to your comment on including parameter stability tests for random-effects parameter \color{black}Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
 In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}
 
-\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
+%\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
 In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}
@@
 
-\section{Items of the culture scale for application example 2}
+\section{Items of the culture scale for \hyperref[sec:TutorialRasch]{Application Example 2}}
 \label{sec:AppendixA}

### 2023-10-31 _MOB_paper.tex
updates MF

@@
 We illustrate the idea using a very simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables.
-%note from Caro: I woul call them partitioning variables or covariates, but not predictors, to save the term predictors for those x-variables that are part of the parametric model
 To keep the example simple, we apply MOB to a very basic global model, which comprises only an intercept.
@@
 The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
-The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test (PPVT; REF) \todo{Marjolein: Add reference}, we will thus model PPVT trajectories over time.
-There were an average number of 2.14 PPVT scores per child, for 29\% of the children there was only a single PPVT score available.
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT]{DunnyDunn81}, we will thus model PPVT trajectories over time.
+On average, there were 2.14 PPVT scores per child, for 29\% of the children there was only a single PPVT score available.
 
@@
 With the second argument, we specified the dataset which contain the variables.
-With the \texttt{cluster} argument, we specified that the partitioning variables are measured at the level of the mothers.
-As a result, parameter stability tests will be performed on the appropriate level \citep{FokkyZeil23}.
-Not specifying the \texttt{cluster} argument would result in use of the default observation-level parameter stability tests, which would yield inflated type-I error rates.
-That is, it could result in detection of spurious subgroups.
-\todo[inline]{I am sure you know much better than me what happens if the cluster argument is used and you have discussed that a lot with Achim.
-But I am not $100\%$ sure that this is the case (and maybe I am completely wrong here), because my pilot simulations have always shown increased type-I error rates even when I used the cluster argument.
-I have only seen $5\%$ type-I error rates with the Wang \& Merkle approach where the covariates are aggregated before they enter the score test.
-Maybe we want to rephrase it in a way as you have phrased it in the article with Achim: ``[by using the cluster argument,] so-called clustered covariances are obtained, which account for dependence between observations within the same cluster (Zeileis, Köll, \& Graham, 2020)''.
-I think I would be more comfortable with a less direct formulation here, and then maybe we can further discuss this for the revision when we have a bit more time.
-\\
-A suggestion: ``By using the cluster argument, we account for dependence between observations within the same cluster.''.
-And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several end nodes, and that the cluster argument also helps for that?
-If that's correct, this could also be a good place to add a sentence about this advantage?}.
-Finally, because we want to retain large enough subgroups, we specified that the minimum number of observations in a terminal node should be 250.
+With the \texttt{cluster} argument, we account for the dependence between observations within the same cluster.
+In this example, the partitioning variables are measured at the mother (or family) level, which we indicate by specifying \texttt{cluster = MotherID}.
+As a result, the parameter stability tests will employ so-called clustered covariances \citep{ZeilyKoll20}.
+Not specifying the \texttt{cluster} argument would result in use of the default observation-level covariances in the parameter stability tests, inflating type-I error rates when partitioning variables are measured on a higher level \citep{FokkyZeil23}.
+
+% Mirka: And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several end nodes, and that the cluster argument also helps for that?
+If that's correct, this could also be a good place to add a sentence about this advantage?
+Marjolein: The splitting up is avoided by having partitioning that are constant over time (and also within families, in this instance).
+The level of the parameter stability tests do not really avoid that.
+We could specify the cluster argument wiht time-varying as well as time-constant partitioning covariates.
+
+Finally, to aid interpretability we want to retain large enough subgroups, we specified the \texttt{minsize} argument.
+As a result, splits will only be implemented if the resulting nodes contain at least 250 observations.
 
@@
 \begin{figure}%
-\caption{}
+\caption{GLMM tree with trajectories of PPVT scores}
 \includegraphics{_MOB_paper-012}
@@
 \includegraphics{_MOB_paper-013}
+\\\textit{Note}: Red lines depict average node-specific trajectories for siblings participating in Head Start; black lines depict average node-specific trajectories for siblings not participating in Head Start.
 \label{fig:lmm_tree}
@@
 
-\caption{\label{tab:predictions}Predicted PPVT scores at different ages.}
+\caption{\label{tab:predictions}Node-specific average predicted PPVT scores at different ages.}
 \begin{tabular}[t]{cccc}
 \toprule
-Node & Program & PPVT at age 6 & PPVT at age 18\\
+Node & Program & PPVT at age 6 & PPVT at age 12\\
 \midrule
-2 & None & 16.16 & 31.08\\
-2 & HS & 20.75 & 35.26\\
-5 & None & 16.58 & 33.64\\
-5 & HS & 23.42 & 38.74\\
-6 & None & 21.08 & 37.36\\
-6 & HS & 27.02 & 43.67\\
-7 & None & 25.61 & 42.80\\
-7 & HS & 27.75 & 45.84\\
+2 & None & 45.70 & 95.63\\
+2 & HS & 49.48 & 98.02\\
+5 & None & 50.36 & 107.43\\
+5 & HS & 53.74 & 104.98\\
+6 & None & 53.33 & 107.82\\
+6 & HS & 59.99 & 115.70\\
+7 & None & 59.64 & 117.13\\
+7 & HS & 63.56 & 124.09\\
 \bottomrule
@@
 \end{table}
-To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 18.
+To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12.
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
-This benefit remains the same over time for the group with lower mother's AFTQ scores.
-The benefit increases over time for White children with higher mother's AFTQ scores.
-Interestingly, the benefit decreases over time for Black and Hispanic children with higher mother's AFTQ scores.
-These results correspond to the conclusions of \cite{Demi09}.
-
-\todo{Marjolein: With different specification of the GLMM tree (no square root, parameter stability tests at mom level, random intercepts of children nested within mothers) the results have changed, still need to adjust description}.
+This benefit increases over time for the group with higher mother's AFTQ scores.
+The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2).
+For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12.
+These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14.
+Fade out is reported to be particularly strong for Afriacn-American and very disadvantaged children.
 
@@
 In the following we will use only the responses to the nine items of the \textit{culture} scale (i.e., items 28 through 36) from test booklet 20 and only the $9769$ complete cases of university students for illustration.
-The wording of the items is provided in Appendix~\ref{sec:AppendixB}.
+The wording of the items is provided in the \hyperref[sec:AppendixA]{Appendix}.
 
@@
  load("dat_SPISA.Rda")
- #head(dat_SPISA, 3) ## commented out because will not compile, dat_SPISA is not
-                     ## an object in current wd
+ head(dat_SPISA, 3)
 \end{Sinput}
+\begin{Soutput}
+    Age Gender                    Area culture.i28 culture.i29 culture.i30
+90   21   Male       Law and Economics           1           1           0
+163  25   Male          no Information           1           1           1
+402  22   Male Agriculture & Nutrition           0           1           0
+    culture.i31 culture.i32 culture.i33 culture.i34 culture.i35 culture.i36
+90            1           0           1           1           1           1
+163           1           0           1           1           0           1
+402           1           1           0           1           1           0
+\end{Soutput}
 \end{Schunk}
 
-The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that in the following we will fit Rasch trees by means of the function \texttt{raschtree} from the R package \texttt{psychotree}.
-However, model-based recursive partitioning is also available in \texttt{psychotree} for other IRT models.
+The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that we can continue to fit Rasch trees by means of function \texttt{raschtree} from the R package \texttt{psychotree}.
+Note that model-based recursive partitioning is also available in \texttt{psychotree} for other IRT models.
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-020}
+
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items using the default stopping criterion based on statistical significance.}
+\begin{subfigure}{1.2\textwidth}
+\includegraphics{_MOB_paper-021}
+\end{subfigure}
+\label{fig:MHtree1}
+\end{figure}%
 
@@
 
-After we have added this information to the Rasch tree, we can now visualize it with additional features.
-When we now plot the Rasch tree with Mantel-Haenszel stopping by means of the \texttt{plot} function, we see that the tree is much more concise with a lower number of splits.
-In addition, in each node the number of items classified as A, B, or C is displayed.
-For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and 4 items are classified in category C (large DIF).
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion.}
+\includegraphics{_MOB_paper-026}
+\label{fig:MHtree2}
+\end{figure}%
+
+After we have added this information to the Rasch tree, we can now visualize it with additional features:
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-025}
 
-We can also color the DIF items in the end node profiles according to a particular split in the tree. We will illustrate this with two examples. 
-First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C. As a reminder, the content of these items was:
+The result is presented in Figure~\label{fig:MHtree2}, which reveals that the Rasch tree is much more concise with a lower number of splits.
+In addition, in each node the number of items classified as A, B, or C is displayed.
+For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and 4 items are classified in category C (large DIF).
+
+We can also color the DIF items in the end node profiles according to a particular split in the tree.
+We will illustrate this with two examples: First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
+As a reminder, the content of these items was:
 
 \begin{itemize}
+\setlength\itemsep{0em}
 \item 1: Painting by Andy Warhol
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-026}
+
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion with additional visualization features.}
+\includegraphics{_MOB_paper-029}
+\label{fig:MHtree3}
+\end{figure}%
 
@@
 
-For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison) 
-more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
+Alternatively, we can also color the DIF items in the end node profiles according to a later split in the tree:
 
@@
 \end{Schunk}
-\includegraphics{_MOB_paper-027}
+
+\begin{figure}%
+\caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion with additional visualization features.}
+\includegraphics{_MOB_paper-031}
+\label{fig:MHtree4}
+\end{figure}%
+
+The result is presented in Figure~\ref{fig:MHtree4}.
+For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison)
+more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
 
@@
 
-\todo{Write about future work.}
-
-The Rasch tree method has been extended to include an effect size measure that can stop the tree from growing if the effect size is non-substantial, but also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
-At the same time, effect sizes are less straightforward to calculate and interpret in linear mixed models, and therewith in the linear mixed effects model tree.
-A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
+\hyperref[sec:TutorialRasch]{Application Example 2} showed that the Rasch tree method can employ an effect size measure to stop the tree from growing if effect-size differences are non-substantial.
+This also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
+At the same time, effect sizes are less straightforward to calculate and interpret in mixed-effects models \citep{JuddyWest17}, and thereby also for GLMM trees.
+\todo{Interpretation methods might not work as expected for multilevel data, so I am not convinced we should discuss this and call it a remedy.}A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
 These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s).
@@
 
-\color{blue} Using cluster-level predictor variables is desirable in particular when using MOB for longitudinal analysis.
-Therefore, building on recent work by \citet{FokkyZeil23}, we will further evaluate how the level of splitting variables can be taken into account best when using MOB, for instance using a variant of parameter stability tests to study heterogeneity on the cluster level \citep{Wang2021a}.
+For GLMM trees, the use of cluster-level partitioning variables would commonly be desired in longitudinal analyses.
+As with traditional GLMMs, specification of the fixed and random effects requires careful consideration.
+For example, whether random slopes of time should be estimated, or whether random intercepts are sufficient to account for dependency between observations.
+Building on recent work by \citet{FokkyZeil23}, we will further study and develop methods to take various dependence structures into account.
+For instance, in future implementations of \texttt{glmertree}, use of parameter stability tests to also allow for detecting subgroup differences in the random-effects parameters \citep{Wang2021a} may be supported.
 
-\color{black}
-\todo[inline]{\color{blue} I (Mirka) added a short point on future work regarding cluster-level predictors and their Type-I error rate here.
+%\todo{I (Mirka) added a short point on future work regarding cluster-level predictors and their Type-I error rate here.
 $\rightarrow$ this paragraph can also transition well to your comment on including parameter stability tests for random-effects parameter \color{black}Mention that the exclusion of random effects from partitioning can both be advantage and disadvantage.
@@
 
-\todo[inline]{Mention that transformations chosen in growth curve modeling may yield different effects.
+%\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
 In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}

### 2023-10-31 Raschtree_MH.Rnw
updates
First submission finalized

@@
 The first few rows of the data set look like this:
-
-\todo{I usually use load for Rda files, Marjolein uses readRDS, is there an advantage? use the same in both examples.
-Mirka: from what I understand, readRDS is for loading .rds/.rda files with only one r object in it, while load is for .rda files with one or more r object in it.
-I would opt for .rda, because I think it is more general?
-Marjolein: I think if we include datasets in supplemental materials, we should have files which are only 1 dataset, not sets of datasets or objects.
-I think this will make the examples more simple.
-In practice, users will probably also have only 1 dataset.}
 
@@
 
-Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function.
-The resulting tree is depicted in Figure \ref{fig:MHtree1}.
-The item parameters for the nine culture items are displayed in the terminal nodes of the tree.
-However, it is hard to interpret the tree structure because such a large number of splits was created.
-Moreover, it is hard to tell whether all splits are due to substantial amounts of DIF, or whether some of the splits are due to small DIF effects that only became statistically significant due to the large sample size.
+Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function:
 
@@
 \end{figure}%
+
+The resulting tree is depicted in Figure \ref{fig:MHtree1}.
+The item parameters for the nine culture items are displayed in the terminal nodes of the tree.
+However, it is hard to interpret the tree structure because such a large number of splits was created.
+Moreover, it is hard to tell whether all splits are due to substantial amounts of DIF, or whether some of the splits are due to small DIF effects that only became statistically significant due to the large sample size.
 
@@
 When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
-The final set of DIF-free items resulting from the purification is also used for aligning the item parameters for each group comparison in the tree plots with color coding, which are displayed below.
+The final set of DIF-free items resulting from the purification will also be used for aligning the item parameters for each group comparison in the tree plots with color coding, which will be discussed and displayed below.
+
+We fit the Raschtree using the Mantel-Haenszel stopping criterion as follows:
 
@@
 
-The result is presented in Figure~\label{fig:MHtree2}, which reveals that the Rasch tree is much more concise with a lower number of splits.
+The result is presented in Figure~\ref{fig:MHtree2}, which reveals that the Rasch tree is much more concise with a lower number of splits.
 In addition, in each node the number of items classified as A, B, or C is displayed.
@@
 These items are %(relative to the DIF-free items for this group comparison) 
-easier or less difficult to answer for participants with female or missing gender.
+easier or less difficult to answer for participants with female or missing gender. Alternatively, we can also color the DIF items in the end node profiles according to a later split in the tree:
 
@@
 % @
-
-Alternatively, we can also color the DIF items in the end node profiles according to a later split in the tree:
 

### 2023-10-31 _MOB_paper.Rnw
updates
First submission finalized

@@
 
-\linespread{1.3}
+\linespread{1.5}
 
 \title{One Model May Not Fit All: Subgroup Detection Using Model-Based Recursive Partitioning}
-\shorttitle{shorttitle}
+\shorttitle{Subgroup Detection Using Model-Based Recursive Partitioning}
 
-\authorsnames{Marjolein Fokkema$^1$, Mirka Henninger$^2$ \& Carolin Strobl$^2$}
-\authorsaffiliations{$^1$Leiden University, $^2$Universit\"at Z\"urich}
+\authorsnames{ }
+\authorsaffiliations{ }
 
 \abstract{Model-based recursive partitioning \citep[MOB,][]{ZeilyHoth08} is a flexible framework for detecting subgroups of persons showing different effects in a wide range of parametric models.
-It provides a versatile tool for detecting and explaining heterogeneity of intervention effects.
+It provides a versatile tool for detecting and explaining heterogeneity in, for example, intervention studies.
 In this tutorial paper, we provide an introduction to the general MOB framework.
@@
 In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF).
-%DIF should be investigated before using test results for comparing groups, because undetected DIF can affect test fairness.
-% Test fairness is more related to use of test scores in practice than group-level comparisons I would think?
+DIF should be investigated before using test results for group comparisons.
 We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}
@@
 @
+
+\begin{titlepage}
+   \begin{center}
+       \vspace*{1cm}
+
+       \textbf{\large One Model May Not Fit All}\\
+
+       \vspace{0.5cm}
+        \large Subgroup Detection Using\\Model-Based Recursive Partitioning\\
+
+       \vspace{1.5cm}
+
+       \textbf{Marjolein Fokkema$^1$, Mirka Henninger$^2$ and Carolin Strobl$^2$}\\
+       $^1$Leiden University, $^2$Universit\"at Z\"urich
+
+       \vfill
+
+       \vspace{0.8cm}
+
+   \end{center}
+\end{titlepage}
 
@@
 rownames(HS_dat) <- NULL
-saveRDS(HS_dat, file = "HS_dat.Rda")
+save(HS_dat, file = "HS_dat.Rda")
 @
@@
 <<>>=
-HS_dat <- readRDS("HS_dat.Rda")
+load("HS_dat.Rda")
 head(HS_dat, 3)
@@
 
+<<echo=FALSE, results=tex>>=
+coefs <- round(coef(HS_tree), digits = 2)
+design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
+                        Program = factor(rep(c("None", "HS"))))
+design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 6), 
+                                     re.form = ~0)
+design_df$`PPVT at age 12` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 12), 
+                                      re.form = ~0)
+names(design_df)[1] <- "Node"
+
+#library("kableExtra")
+#kable_styling(
+knitr::kable(design_df, format = "latex", digits = 2, 
+             caption = "Node-specific average predicted PPVT scores at different ages.", 
+             label = "predictions", centering = FALSE,
+             linesep = "", # linesep command suppressess addlinesep every 5 rows
+             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)#,
+#latex_options = "HOLD_position")
+@
+
+\FloatBarrier
+
 \begin{figure}%
@@
 
-\FloatBarrier
-
-<<echo=FALSE, results=tex>>=
-coefs <- round(coef(HS_tree), digits = 2)
-design_df <- data.frame(.tree = rep(sort(unique(HS_tree$data$.tree)), each = 2),
-                        Program = factor(rep(c("None", "HS"))))
-design_df$`PPVT at age 6` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 6), 
-                                     re.form = ~0)
-design_df$`PPVT at age 12` <- predict(HS_tree$lmer, newdata = cbind(design_df, Age = 12), 
-                                      re.form = ~0)
-names(design_df)[1] <- "Node"
-knitr::kable(design_df, format = "latex", digits = 2, 
-             caption = "Node-specific average predicted PPVT scores at different ages.", 
-             label = "predictions", centering = FALSE,
-             linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)
-@
-
 To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12.
@@
 These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14.
-Fade out is reported to be particularly strong for Afriacn-American and very disadvantaged children.
+Fade out was reported to be particularly strong for Afriacn-American and very disadvantaged children.
+The global benefit of participation in Head Start as shown in Figure~\ref{fig:global_lmm} and reported by \cite{Demi09} thus seems to differ across groups.
 
@@
 \newpage
-\FloatBarrier
+
 \section{Application Example 2: Subgroup Detection in Rasch Models}
@@
 \cite{PhilyRusc18} show how to use these techniques for assessing the stability of the set of selected variables and cutpoints.
-For GLMM trees, cross-validation can additionally be used for evaluating predictive accuracy \cite[e.g., ]{RooiyWeed20}, or for further optimizing predictive accuracy through the choice of model-fitting parameters, like the minimum number of observations the terminal nodes or the maximum tree depth \cite{JameyWitt23}.
+For GLMM trees, cross-validation can additionally be used for evaluating predictive accuracy \citep[e.g., ][]{RooiyWeed20}, or for further optimizing predictive accuracy through the choice of model-fitting parameters, like the minimum number of observations in the terminal nodes or the maximum tree depth \citep{JameyWitt23}.
 
@@
 This also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
-At the same time, effect sizes are less straightforward to calculate and interpret in mixed-effects models \citep{JuddyWest17}, and thereby also for GLMM trees.
-\todo{Interpretation methods might not work as expected for multilevel data, so I am not convinced we should discuss this and call it a remedy.}A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
+For mixed-effects models, computation of effect sizes is far from straightforward and several different approaches have been proposed \citep[e.g., ][]{JuddyWest17, BrysyStev18, Bran18}.
+Effect-size based stopping criteria for GLMM trees are therefore currently not available, but any of the proposed computation approaches proposed for tradtional GLMMs can also be computed from the parameter estimates of GLMM trees.
+
+%\todo{Interpretation methods might not work as expected for multilevel data, so I am not convinced we should discuss this and call it a remedy.} 
+%A remedy to this issue can be interpretation techniques, such as partial dependence plots or individual conditional expectations plots.
 These interpretation techniques support researchers in gauging the size of the effect of predictor variables visually by depicting the predicted value of the machine learning method as a function of the value of the predictor variable(s).
@@
 
-\todo{Add DOIs.}
+%\todo{Add DOIs.}
 
@@
 \end{document}
+
+<<echo=FALSE>>=
+Stangle(file='_MOB_paper.Rnw', output = '_MOB_paper.R', ,encoding='utf8',
+        drop.evalFALSE = TRUE, annotate = FALSE)
+@

### 2023-10-31 _MOB_paper.tex
updates
First submission finalized

@@
 \renewenvironment{Schunk}{\small}{}
+
+\begin{titlepage}
+   \begin{center}
+       \vspace*{1cm}
+
+       \textbf{\large One Model May Not Fit All}\\
+
+       \vspace{0.5cm}
+        \large Subgroup Detection Using\\Model-Based Recursive Partitioning\\
+
+       \vspace{1.5cm}
+
+       \textbf{Marjolein Fokkema$^1$, Mirka Henninger$^2$ and Carolin Strobl$^2$}\\
+       $^1$Leiden University, $^2$Universit\"at Z\"urich
+
+       \vfill
+
+       \vspace{0.8cm}
+
+   \end{center}
+\end{titlepage}
 
@@
 \begin{Sinput}
- HS_dat <- readRDS("HS_dat.Rda")
+ load("HS_dat.Rda")
  head(HS_dat, 3)
@@
 
-\begin{figure}%
-\caption{GLMM tree with trajectories of PPVT scores}
-\includegraphics{_MOB_paper-012}
-
-\vspace*{-3cm}
-
-\includegraphics{_MOB_paper-013}
-\\\textit{Note}: Red lines depict average node-specific trajectories for siblings participating in Head Start; black lines depict average node-specific trajectories for siblings not participating in Head Start.
-\label{fig:lmm_tree}
-\end{figure}%
-
-The resulting tree is presented in Figure~\ref{fig:lmm_tree}.
-Below each terminal node, we plotted the observations and the two regression curves given by the coefficients in that node.
-The first split was made based on the AFTQ variable, which represents the mother's score on the Armed Forced Qualification Test, adjusted for the age at which they completed the test.
-The group with higher mother's AFTQ scores is further split based on race.
-The Black and Hispanic group is further split based on mother's AFTQ score.
-
-\FloatBarrier
-
 \begin{table}
@@
 \end{table}
+\FloatBarrier
+
+\begin{figure}%
+\caption{GLMM tree with trajectories of PPVT scores}
+\includegraphics{_MOB_paper-013}
+
+\vspace*{-3cm}
+
+\includegraphics{_MOB_paper-014}
+\\\textit{Note}: Red lines depict average node-specific trajectories for siblings participating in Head Start; black lines depict average node-specific trajectories for siblings not participating in Head Start.
+\label{fig:lmm_tree}
+\end{figure}%
+
+The resulting tree is presented in Figure~\ref{fig:lmm_tree}.
+Below each terminal node, we plotted the observations and the two regression curves given by the coefficients in that node.
+The first split was made based on the AFTQ variable, which represents the mother's score on the Armed Forced Qualification Test, adjusted for the age at which they completed the test.
+The group with higher mother's AFTQ scores is further split based on race.
+The Black and Hispanic group is further split based on mother's AFTQ score.
+
 To aid interpretation of the coefficients in Figure~\ref{fig:lmm_tree}, Table~\ref{tab:predictions} provides predicted PPVT scores for each of the groups and programs at ages 6 and 12.
@@
 \end{document}
+
+\begin{Schunk}
+\begin{Soutput}
+Writing to file _MOB_paper.R 
+\end{Soutput}
+\end{Schunk}

### 2023-10-31 _MOB_paper.R

@@ -1,0 +1,200 @@
+### R code from vignette source '_MOB_paper.Rnw'
+
+## suppress printing of > and + before code, it is annoying for copy-pasting
+options(prompt=" ", continue = " ") 
+
+
+set.seed(3)
+x1 <- runif(250, min = 5, max = 11)
+x2 <- runif(250, min = 2, max = 8)
+x3 <- runif(250, min = 4, max = 10)
+error <- rnorm(250, sd = 1.5)
+y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
+toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
+toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
+
+
+library("colorspace")
+boxplot(toy_data$y, cex = .7)
+#points(x = 1, y = mean(toy_data$y), pch = 19, ylim = c(1,14))
+
+
+library("partykit")
+tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
+plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
+
+
+## Load data
+data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
+
+## Keep only eligible children in HS or no preschool
+data <- data[data$Elig2_90 == 1, ]
+data <- data[data$None2_90 == 1 | data$HS2_90 == 1, ]
+data <- data[data$MotherID %in% names(table(data$MotherID) >= 2), ]
+data$program <- factor(ifelse(data$None2_90 == 1, "None", 
+                              ifelse(data$Pre2_90 == 1, "Pre", "HS")))
+
+## Keep only families in which children differ in HS participation
+selected_family <- rowSums(table(data$MotherID, data$program) > 0) == 2
+selected_family <- names(selected_family)[selected_family]
+data <- data[data$MotherID %in% selected_family, ]
+
+## Node-specific predictor
+data$program <- factor(ifelse(data$None2_90 == 1, "None", 
+                              ifelse(data$Pre2_90 == 1, "Pre", "HS")),
+                       levels = c("None", "HS"))
+
+## Partitioning variables
+
+## ageAFTQ: age-adjusted AFQT
+## Race_child: Hispanic=Race_Child==1; Black=Race_Child==2; White=Race_Child==3
[…]

### 2023-10-31 _MOB_paper_custom.R

@@ -1,0 +1,214 @@
+###################################################################
+###
+###
+### R code for reproducing analyses from:
+###
+###   One Model May Not Fit All: Subgroup Detection Using
+###   Model-Based Recursive Partitioning
+###
+
+
+
+
+#########################
+##
+## Introduction
+##
+##
+
+set.seed(3)
+x1 <- runif(250, min = 5, max = 11)
+x2 <- runif(250, min = 2, max = 8)
+x3 <- runif(250, min = 4, max = 10)
+error <- rnorm(250, sd = 1.5)
+y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
+toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
+toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
+boxplot(toy_data$y, cex = .7)
+library("partykit")
+tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
+plot(tree, gp = gpar(cex = .7), ylim = c(1, 14))
+
+
+
+######################################################################
+##
+## Application Example 1: Subgroup Detection in Mixed-Effects Models 
+##
+##
+
+## Load data
+data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
+
+## Keep only eligible children in HS or no preschool
+data <- data[data$Elig2_90 == 1, ]
+data <- data[data$None2_90 == 1 | data$HS2_90 == 1, ]
+data <- data[data$MotherID %in% names(table(data$MotherID) >= 2), ]
+data$program <- factor(ifelse(data$None2_90 == 1, "None", 
+                              ifelse(data$Pre2_90 == 1, "Pre", "HS")))
+
+## Keep only families in which children differ in HS participation
+selected_family <- rowSums(table(data$MotherID, data$program) > 0) == 2
+selected_family <- names(selected_family)[selected_family]
+data <- data[data$MotherID %in% selected_family, ]
+
+## Node-specific predictor
+data$program <- factor(ifelse(data$None2_90 == 1, "None", 
+                              ifelse(data$Pre2_90 == 1, "Pre", "HS")),
+                       levels = c("None", "HS"))
+
+## Partitioning variables
+
+## ageAFTQ: age-adjusted AFQT
[…]

### 2024-04-26 Raschtree_MH.Rnw
Draft 1st revision

@@
 @
+
+\subsubsection{Dataset}
 
@@
 
+\subsubsection{Fitting a Rasch tree}
+
 The results in \citet{SPISA:book} indicate that the Rasch model is appropriate for the individual scales of the general knowledge quiz, so that we can continue to fit Rasch trees by means of function \texttt{raschtree} from the R package \texttt{psychotree}.
@@
 It does not contain predictor variables of its own.
-Therefore, the formula syntax for the \texttt{raschtree} function looks simpler than that for the \texttt{lmertree} function in \hyperref[sec:TutorialMixed]{Application Example 1}: On the left hand side of the \verb|~| symbol, we provide the item response matrix \texttt{culture} for the Rasch model.
+Therefore, the formula syntax for the \texttt{raschtree} function is simpler than that for the \texttt{lmertree} function in \hyperref[sec:TutorialMixed]{Application Example 1}: On the left hand side of the \verb|~| symbol, we provide the item response matrix \texttt{culture} for the Rasch model.
 On the right hand side of the \verb|~| symbol, we provide the possible partitioning variables \texttt{Gender}, \texttt{Age}, and \texttt{Area}.
@@
 
+\subsubsection{Pruning a Rasch tree}
+
 Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function:
@@
 The item parameters for the nine culture items are displayed in the terminal nodes of the tree.
-However, it is hard to interpret the tree structure because such a large number of splits was created.
+However, it is hard to interpret the tree structure because such a large number of splits were created.
 Moreover, it is hard to tell whether all splits are due to substantial amounts of DIF, or whether some of the splits are due to small DIF effects that only became statistically significant due to the large sample size.
@@
 When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
-The final set of DIF-free items resulting from the purification will also be used for aligning the item parameters for each group comparison in the tree plots with color coding, which will be discussed and displayed below.
+The final set of DIF-free items resulting from the purification will also be used for aligning the item parameters for each group comparison, which we will discuss and display below.
 
@@
 
+\subsubsection{Interpretation of a Rasch tree}
+
 \begin{figure}%
@@
 
-We can also color the DIF items in the end node profiles according to a particular split in the tree.
+We can also color the DIF items in the terminal node profiles according to a particular split in the tree.
 We will illustrate this with two examples: First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
@@
 These items are %(relative to the DIF-free items for this group comparison) 
-easier or less difficult to answer for participants with female or missing gender. Alternatively, we can also color the DIF items in the end node profiles according to a later split in the tree:
+easier or less difficult to answer for participants with female or missing gender (node 7), than for males (nodes 3, 5 and 6).
+Alternatively, we can also color the DIF items in the terminal node profiles according to a later split in the tree:
 
@@
 For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison)
-more difficult to answer for younger respondents (age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
+more difficult to answer for younger respondents ($>24$ age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
 
@@
 
-The displayed information again shows that some items show only negligible DIF of category A in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
+The displayed information again shows that some items show only negligible DIF (category A) in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
 The values of the Mantel-Haenszel effect size measure are negative when the item difficulty is lower in the right daughter node than in the left daughter node.
-For example, as we already saw from the first colored tree plot above, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer %(relative to the DIF-free items for this group comparison)
+For example, as we already observed in Figure~\ref{fig:MHtree3}, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer %(relative to the DIF-free items for this group comparison)
 for participants with female or missing gender, which correspond to the right daughter node of node 1.

### 2024-04-26 _MOB_paper.Rnw
Draft 1st revision

@@
 In this tutorial paper, we provide an introduction to the general MOB framework.
-In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: mixed-effects and item response theory (IRT) models.
+In two specific case studies, we show how MOB-based methods can be used to detect and explain heterogeneity in two widely-used frameworks in educational studies: the generalized linear mixed model (GLMM) and item response theory (IRT).
 In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
@@
 In CART, the aim is to detect groups of persons, defined by (combinations of) covariate values, that differ in their mean on a response variable.
-For example, if we want to predict math exam grades using student characteristics assessed in a previous year, we could find that a subgroup of students who are highly motivated and have good reading skills show higher average math exam grades.
-MOB generalizes this idea: It also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean.
-An example could be that students who are highly motivated and have good reading skills show higher slopes in a regression model relating the time spent studying to math exam grades.
+Detecting such subgroups may be of interest in many studies in school and developmental psychology.
+For example, recursive partitioning methods have been used to risk factors for adolescent crime \citep{FrityHaup19} and corporal punishment \citep{StemyHein19}, and to identify predictors of math ability \citep{DingyZhao19}.
 
+MOB also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean.
+For example, MOB can be used to identify children that show stronger or weaker effects of an intervention \cite{ChunyAnso21}, or stronger or weaker gains in academic skills over time \cite{FokkyZeil24}.
 The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}.
 In this article, we will highlight two specific MOB methods that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
+
 Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures.
 For example, when children are tested at several time points, the time points are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools.
-In \hyperref[sec:TutorialMixed]{Application Example 1} we will illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
+In the section \hyperref[sec:glmmtree]{MOB for Subgroup Detection in Mixed-Effects Models}, we present \hyperref[sec:TutorialMixed]{Application Example 1} to illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
 In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model.
 We thus assume that the psychological test score(s) of interest have already been validated.
-%\todo{(Caro: I have kept psychological test score, because instrument to me sounds more clinical)(Mirka: I like the term instrument to distinguish it from a statistical test) (Marjolein: I hope 'psychological test score' clears up possible confusion)}
 
-For \hyperref[sec:TutorialRasch]{Application Example 2}, we go back in the research process to the point where a new psychological test has been administered to a validation sample.
+In the section~\hyperref[sec:raschtree]{MOB for Detecting DIF in Measurement Models}, we go back in the research process to the point where a new psychological test has been administered to a validation sample.
 Here, we focus on detecting subgroups with different parameters of a \textit{measurement} model.
@@
 Readers interested in learning more about its predecessor method, classification and regression trees, are referred to the introduction by \citet{StrMalTut:2009:PM}.
+Our aim here is to provide an informal introduction and to illustrate the relevance of these methods for studies in school psychology.
+We will specify the relevant works providing the formal details of MOB and the extensions presented here in the \hyperref[sec:discussion]{Discussion} section.
 
@@
 set.seed(3)
-x1 <- runif(250, min = 5, max = 11)
-x2 <- runif(250, min = 2, max = 8)
-x3 <- runif(250, min = 4, max = 10)
+x1 <- runif(250, min = 5, max = 11) ## age
+x2 <- runif(250, min = 2, max = 8) ## gender
+x3 <- runif(250, min = 4, max = 10) ## reading comprehension
 error <- rnorm(250, sd = 1.5)
-y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
+y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error ## total SDQ score (0-40)
 toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
@@
 
-We illustrate the idea using a very simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables.
+We illustrate the idea using a simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
 To keep the example simple, we apply MOB to a very basic global model, which comprises only an intercept.
 It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
-Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot, with the global intercept indicated as a triangle.
-Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
+Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot.
+Obviously, a global intercept or mean would not describe all observations equally well, there is quite some unexplained variation around the sample mean of \Sexpr{round(mean(toy_data$y), digits = 2)}.
 
@@
 plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
+term_node_means <- round(tapply(toy_data$y, predict(tree, type = "node"), mean), digits= 2L)
 @
@@
 The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node.
+By default, a Bonferroni correction is applied to the $p$ values, to account for the fact that the tests are performed for all of the potential partitioning variables.
 Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroup are as similar as possible.
@@
 The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting.
-The subgroup-specific distributions of the response variable are presented in the end nodes at the bottom of the tree.
-The fact that the tree in Figure~\ref{fig:toy} displays more than one end node confirms that one global model for all observations does not appropriately capture the pattern in the data. 
+
+The subgroup-specific distributions of the response variable are presented in the terminal nodes at the bottom of the tree.
+The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the pattern in the data.
 The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
-This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.
+This automatic variable selection is an important characteristic of the CART and MOB algorithms.
+The means (intercepts) in the terminal nodes are \Sexpr{term_node_means[1]}, \Sexpr{term_node_means[2]} and \Sexpr{term_node_means[3]}, respectively.
+While the means and distributions in nodes 2 and 4 appear very similar, the mean in node 5 is obviously higher.
+Substantively, the tree suggests and interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa).
+If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at somewhat later age.
 
@@
 
-%While in Figure~\ref{fig:lmm_tree} each end node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each end node contains the group-specific item parameters of a Rasch model for the respective test items.
-%The fact that the Figures display more than one end node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
+%While in Figure~\ref{fig:lmm_tree} each terminal node contains a linear mixed-effecs model with a group-specific effect for the Head Start intervention, in Figures~\ref{fig:math_tree} and \ref{fig:read_tree} each terminal node contains the group-specific item parameters of a Rasch model for the respective test items.
+%The fact that the Figures display more than one terminal node already means that one joint mixed-effects or Rasch model was not appropriate to describe the pattern in the data.
 
@@
 This allows for stopping tree growing as soon as no significant differences can be detected anymore.
-Other stopping criteria are based on the number of persons in the end nodes.
-These criteria ensure that the sample sizes in the end nodes are large enough to estimate the statistical model in each end node.
+Other stopping criteria are based on the number of persons in the terminal nodes.
+These criteria ensure that the sample sizes in the terminal nodes are large enough to estimate the statistical model in each terminal node.
 For reasons of interpretability, users may sometimes specify even higher requirements for the sample sizes of terminal nodes.
@@
 
-\subsection{Using MOB for Subgroup Detection in Mixed-Effects Models}
+\section{MOB for Subgroup Detection in Mixed-Effects Models}
+\label(glmmtree)
 
@@
 
-GLMM trees combine MOB and mixed-effects models and were introduced by \cite{FokkySmit18}.
+GLMM trees combine MOB and generalized linear mixed-effects models (GLMMs) and were introduced by \cite{FokkySmit18}.
 While the 'standard' MOB trees \citep{ZeilyHoth08} allow for subgroup detection in fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for partitioning mixed-effects regression models.
@@
 As such, GLMM trees allow for detecting subgroups in multilevel models that differ with respect to any (set of) fixed-effects parameters of interest.
-For example, users may be interested in detecting subgroups with different means \citep{FokkyEdbr21}, but also differential effects of treatment \citep{FokkySmit18} or differential growth over time \citep{FokkyZeil23}, to name but a few examples.
+For example, users may be interested in detecting subgroups with different means \citep{FokkyEdbr21}, but also differential effects of treatment \citep{FokkySmit18} or differential growth over time \citep{FokkyZeil24}, to name but a few examples.
 
@@
 The usage of this function will be illustrated in \hyperref[sec:TutorialMixed]{Application Example 1}.
-Further mathematical and computational details about the GLMM tree model are also described in \cite{FokkySmit18} and \cite{FokkyZeil23}.
-
-\subsection{Using MOB for Detecting DIF in Measurement Models}
-
-Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups.
-In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., for example][for an introduction]{DebStrZei:2022:CRC}.
-A particularly crucial assumption for the comparability of test scores between groups is measurement invariance.
-Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
-
-MOB can be used to detect DIF by means of testing whether the item parameters of the measurement model exhibit significant instability with respect to (combinations of) covariates.
-For the Rasch measurement model, the \texttt{R} function for conducting the MOB analysis is the \texttt{raschtree} function from package \texttt{psychotree} \citep{StroyKopf15}.
-The usage of this function will be illustrated in \hyperref[sec:TutorialRasch]{Application Example 2}.
-We will see that, just like for GLMM trees, we need to specify which variables are part of the parametric model.
-In the case of the Rasch model, this will be the test items.
-Moreover, we need to specify, which covariates are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.
-
-If one joint Rasch model holds for the entire sample, that is, if there is no DIF, a Rasch tree should show no splits.
-However, in certain settings in educational research, such as large scale assessments, very large sample sizes are available for testing for DIF.
-Large sample sizes are good for detecting even small effects or model violations with a high statistical power.
-The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected.
-However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
-As we will illustrate in the first part of \hyperref[sec:TutorialRasch]{Application Example 2}, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
-Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
-\citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
-In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
-\citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
-
-As we will show in \hyperref[sec:TutorialRasch]{Application Example 2}, for large sample sizes this can be very helpful because it results in shorter trees that are easier to interpret and contain only splits corresponding to DIF effect sizes considered relevant in practice.
-Together with a purification step \citep[see][and Application Example 2 for details]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for highlighting those items that show DIF with respect to certain groups of persons graphically.
-This can help generate hypothesis about the sources of DIF, as we will ilustrate in \hyperref[sec:TutorialRasch]{Application Example 2}, and can also aid the decision how to proceed with the DIF items.
-%\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
-For example, items that show DIF between different language groups may often be improved by means of making sure that in all translations the meaning is as similar as possible, and/or that the words employed in the translations for the different languages are equally frequently used.
-In other situations, sources of DIF might be harder to eliminate, so that often DIF items may be excluded from a test.
-Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.
-
-Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
-This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
-Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
-These items may then show DIF between native and non-native speakers.
-When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, for example by using more frequently used words, or whether the test should be considered two- rather than one-dimensional.
-For a discussion of the connection between DIF and multidimensionality, see also \cite{Ack:1992,Stretal:2021:APM}.
-
-We will now illustrate how to use the \texttt{R} packages \texttt{glmertree} and \texttt{psychotree} to conduct the MOB analyses.
+Further mathematical and computational details about the GLMM tree model are also described in \cite{FokkySmit18} and \cite{FokkyZeil24}.
 
@@
 
-\newpage
-\section{Application Example 1: Subgroup Detection in Mixed-Effects Models}
+\subsection{Application Example 1: Detecting Differential Effects of Head Start}
 \label{sec:TutorialMixed}
@@
 
-\subsection{Dataset}
+\subsubsection{Dataset}
 
@@
 
-Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's body height %\todo{Marjolein: Add scale for measurement}.
+Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height (computed as feet x 100 + inches).
 Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables.
@@
 head(HS_dat, 3)
+
+hist(tapply(HS_dat$Mom_height, HS_dat$MotherID, mean))
 @
@@
 
-The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference is slightly diminished but persistent over time.
+The results are presented in Figure~\ref{fig:global_lmm}, which shows that children participating in Head Start show slightly higher performance than their non-participating siblings and that this difference persists but slightly diminishes over time.
 This result agrees with the findings of \cite{Demi09}.
-Figure~\ref{fig:global_lmm} already suggests that the effect of age on performance is very strong, compared to which the effects of intervention (Head Start) are existent but relatively small, which is typical for young children growing up to adults.
+Figure~\ref{fig:global_lmm} also shows that the effect of age on performance is very strong, which is typical for young children growing up to adults, while the effect of intervention (Head Start) is small in comparison.
 Note that the \texttt{R} code for creating the figure is omitted here, because we want to focus on fitting and interpreting GLMM trees.
@@
 \FloatBarrier
-\subsection{Linear mixed effects model tree}
+\subsubsection{Fitting a GLMM tree}
 
@@
 As a result, the parameter stability tests will employ so-called clustered covariances \citep{ZeilyKoll20}.
-Not specifying the \texttt{cluster} argument would result in use of the default observation-level covariances in the parameter stability tests, inflating type-I error rates when partitioning variables are measured on a higher level \citep{FokkyZeil23}.
+Not specifying the \texttt{cluster} argument would result in use of the default observation-level covariances in the parameter stability tests, inflating type-I error rates when partitioning variables are measured on a higher level \citep{FokkyZeil24}.
 
-% Mirka: And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several end nodes, and that the cluster argument also helps for that?
+% Mirka: And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several terminal nodes, and that the cluster argument also helps for that?
 If that's correct, this could also be a good place to add a sentence about this advantage?
-Marjolein: The splitting up is avoided by having partitioning that are constant over time (and also within families, in this instance).
+Marjolein: The splitting up is avoided by having partitioning variables that are constant over time (and also within families, in this instance).
 The level of the parameter stability tests do not really avoid that.
-We could specify the cluster argument wiht time-varying as well as time-constant partitioning covariates.
 
 Finally, to aid interpretability we want to retain large enough subgroups, we specified the \texttt{minsize} argument.
-As a result, splits will only be implemented if the resulting nodes contain at least 250 observations.
+Splits will only be implemented if the resulting nodes contain at least 250 observations.
+With a total sample size of \Sexpr{round(nrow(HS_dat))}, even small differences in the effects of age and Head Start participation between subgroups may become significant.
+The average number of measurements per family was \Sexpr{round((nrow(HS_dat) / length(unique(HS_dat$ChildID))) * (length(unique(HS_dat$ChildID)) / length(unique(HS_dat$MotherID))), digits = 1L)}.
+A minimum node size of 250 ensured that the estimated effects of age and Head Start participation in the terminal nodes would at least be based on data from about 50 families.
+
+\subsubsection{Interpreting a GLMM tree}
 
@@
 
-#library("kableExtra")
-#kable_styling(
+library("kableExtra")
+add_footnote(
 knitr::kable(design_df, format = "latex", digits = 2, 
@@
              linesep = "", # linesep command suppressess addlinesep every 5 rows
-             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE)#,
+             row.names = FALSE, escape=TRUE, align = c("cccc"), booktabs = TRUE),
 #latex_options = "HOLD_position")
+"HS = Head Start; PPVT = Peabody Picture Vocabulary Test.")
 @
@@
 The global benefit of participation in Head Start as shown in Figure~\ref{fig:global_lmm} and reported by \cite{Demi09} thus seems to differ across groups.
+
+It should be noted that MOB and GLMM trees are exploratory techniques: The subgroup structures are detected from the data.
+Such data-driven procedures cannot provide valid standard errors and significance tests to evaluate significance of the observed subgroup differences.
+To that end, the subgroup structure that was found should be used in a confirmatory analysis on a new sample, that was not used to fit the tree.
+This would allow to test the hypothesis, either using frequentist or Bayesian approaches, of whether the parameters of interest really differ between subgroups.
+Researchers who want to both detect subgroups and test hypotheses on the parameter differences on one dataset, are advised to split their dataset into two parts before the analyses.
+Then one part of the dataset can be used for exploration, and the other part can be used to test the parameter differences between subgroups.
 
@@
 
-\newpage
+\section{MOB for Detecting DIF in Measurement Models}
+\label{raschtree}
 
-\section{Application Example 2: Subgroup Detection in Rasch Models}
+Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups.
+In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., for example][for an introduction]{DebStrZei:2022:CRC}.
+A particularly crucial assumption for the comparability of test scores between groups is measurement invariance.
+Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
+
+MOB can be used to detect DIF by means of testing whether the item parameters of the measurement model exhibit significant instability with respect to (combinations of) covariates.
+For the Rasch measurement model, the \texttt{R} function for conducting the MOB analysis is the \texttt{raschtree} function from package \texttt{psychotree} \citep{StroyKopf15}.
+The usage of this function will be illustrated in \hyperref[sec:TutorialRasch]{Application Example 2}.
+We will see that, just like for GLMM trees, we need to specify which variables are part of the parametric model.
+In the case of the Rasch model, this will be the test items.
+Moreover, we need to specify which covariates are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.
+
+If one joint Rasch model holds for the entire sample, that is, if there is no DIF, a Rasch tree should show no splits.
+However, in certain settings in educational research, such as large scale assessments, very large sample sizes are available for testing for DIF.
+Large sample sizes are good for detecting even small effects or model violations with high statistical power.
+The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected.
+However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
+%As we will illustrate in the first part of \hyperref[sec:TutorialRasch]{Application Example 2}, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
+\citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
+In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
+\citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
+
+Together with a purification step \citep[see][%and Application Example 2 for details
+]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for graphically highlighting those items that show DIF with respect to certain groups of persons.
+This can help generate hypotheses about the sources of DIF, and can also aid in deciding how to proceed with the DIF items.
+%\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
+For example, items that show DIF between different language groups may often be improved by means of making sure that in all translations the meaning is as similar as possible, and/or that the words employed in the translations for the different languages are equally frequently used.
+In other situations, sources of DIF might be harder to eliminate, so that often DIF items may be excluded from a test.
+Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.
+
+Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
+This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
+Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
+These items may then show DIF between native and non-native speakers.
+When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, for example by using more frequently used words, or whether the test should be considered two- rather than one-dimensional.
+For a discussion of the connection between DIF and multidimensionality, see also \cite{Ack:1992,Stretal:2021:APM}.
+
+%We will now illustrate how to use the \texttt{R} packages \texttt{glmertree} and \texttt{psychotree} to conduct the MOB analyses.
+
+\subsection{Application Example 2: Detecting DIF in a General Knowledge Quiz}
 \label{sec:TutorialRasch}
@@
 \section{Discussion}
+\label{discussion}
 
@@
 For mixed-effects models, computation of effect sizes is far from straightforward and several different approaches have been proposed \citep[e.g., ][]{JuddyWest17, BrysyStev18, Bran18}.
-Effect-size based stopping criteria for GLMM trees are therefore currently not available, but any of the proposed computation approaches proposed for tradtional GLMMs can also be computed from the parameter estimates of GLMM trees.
+Effect-size based stopping criteria for GLMM trees are therefore currently not available, but any of the proposed computation approaches proposed for traditional GLMMs can also be computed from the parameter estimates of GLMM trees.
 
@@
 For example, whether random slopes of time should be estimated, or whether random intercepts are sufficient to account for dependency between observations.
-Building on recent work by \citet{FokkyZeil23}, we will further study and develop methods to take various dependence structures into account.
-For instance, in future implementations of \texttt{glmertree}, use of parameter stability tests to also allow for detecting subgroup differences in the random-effects parameters \citep{Wang2021a} may be supported.
+Building on recent work by \citet{FokkyZeil24}, we will further study and develop methods to take various dependence structures into account.
+For instance, in future implementations of \texttt{glmertree}, parameter stability tests that allow for detecting subgroup differences in the random-effects parameters \citep{Wang2021a} may be implemented.
 
@@
 In future work, use of parameter stability tests for random-effects parameters of Ting and Ed will be explored.}
-
-%\todo{Mention that transformations chosen in growth curve modeling may yield different effects.
-In future work, recursive partitioning of GAMs will be explored to obviate the need for manual transformation of the predictors.}
 

### 2024-04-26 _MOB_paper.tex
Draft 1st revision

@@
 
-We illustrate the idea using a very simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$, and three covariates, $x_1$, $x_2$ and $x_3$, as possible partitioning variables.
+We illustrate the idea using a simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
 To keep the example simple, we apply MOB to a very basic global model, which comprises only an intercept.
 It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
-Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot, with the global intercept indicated as a triangle.
-Obviously, the global intercept does not describe all observations equally well, there is quite some unexplained variation around it.
+Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot.
+Obviously, a global intercept or mean would not describe all observations equally well, there is quite some unexplained variation around the sample mean of 5.91.
 
@@
 The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting.
-The subgroup-specific distributions of the response variable are presented in the end nodes at the bottom of the tree.
-The fact that the tree in Figure~\ref{fig:toy} displays more than one end node confirms that one global model for all observations does not appropriately capture the pattern in the data. 
+
+The subgroup-specific distributions of the response variable are presented in the terminal nodes at the bottom of the tree.
+The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the pattern in the data.
 The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
-This automatic variable selection is an important characteristic of classification and regression tree and MOB algorithms.
+This automatic variable selection is an important characteristic of the CART and MOB algorithms.
+The means (intercepts) in the terminal nodes are 4.42, 4.78 and 9.17, respectively.
+While the means and distributions in nodes 2 and 4 appear very similar, the mean in node 5 is obviously higher.
+Substantively, the tree suggests and interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa).
+If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at somewhat later age.
 
@@
 
-\newpage
-\section{Application Example 1: Subgroup Detection in Mixed-Effects Models}
+\subsection{Application Example 1: Detecting Differential Effects of Head Start}
 \label{sec:TutorialMixed}
 
-\subsection{Dataset}
+\subsubsection{Dataset}
 
@@
 \end{Soutput}
+\begin{Sinput}
+ hist(tapply(HS_dat$Mom_height, HS_dat$MotherID, mean))
+\end{Sinput}
 \end{Schunk}
@@
 As a result, the parameter stability tests will employ so-called clustered covariances \citep{ZeilyKoll20}.
-Not specifying the \texttt{cluster} argument would result in use of the default observation-level covariances in the parameter stability tests, inflating type-I error rates when partitioning variables are measured on a higher level \citep{FokkyZeil23}.
+Not specifying the \texttt{cluster} argument would result in use of the default observation-level covariances in the parameter stability tests, inflating type-I error rates when partitioning variables are measured on a higher level \citep{FokkyZeil24}.
 
-% Mirka: And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several end nodes, and that the cluster argument also helps for that?
+% Mirka: And if I remember correctly, one critical point was also that in longitudinal analysis, we want to avoid that a person is ``split up'' in several terminal nodes, and that the cluster argument also helps for that?
 If that's correct, this could also be a good place to add a sentence about this advantage?
-Marjolein: The splitting up is avoided by having partitioning that are constant over time (and also within families, in this instance).
+Marjolein: The splitting up is avoided by having partitioning variables that are constant over time (and also within families, in this instance).
 The level of the parameter stability tests do not really avoid that.
-We could specify the cluster argument wiht time-varying as well as time-constant partitioning covariates.
 
 Finally, to aid interpretability we want to retain large enough subgroups, we specified the \texttt{minsize} argument.
-As a result, splits will only be implemented if the resulting nodes contain at least 250 observations.
+Splits will only be implemented if the resulting nodes contain at least 250 observations.
+With a total sample size of 1433, even small differences in the effects of age and Head Start participation between subgroups may become significant.
+The average number of measurements per family was 5.6.
+A minimum node size of 250 ensured that the estimated effects of age and Head Start participation in the terminal nodes would at least be based on data from about 50 families.
+
+\subsubsection{Interpreting a GLMM tree}
 
@@
 \bottomrule
+\multicolumn{4}{l}{\textsuperscript{a} HS = Head Start; PPVT = Peabody Picture Vocabulary}\\
+\multicolumn{4}{l}{Test.}\\
 \end{tabular}
@@
 
+It should be noted that MOB and GLMM trees are exploratory techniques: The subgroup structures are detected from the data.
+Such data-driven procedures cannot provide valid standard errors and significance tests to evaluate significance of the observed subgroup differences.
+To that end, the subgroup structure that was found should be used in a confirmatory analysis on a new sample, that was not used to fit the tree.
+This would allow to test the hypothesis, either using frequentist or Bayesian approaches, of whether the parameters of interest really differ between subgroups.
+Researchers who want to both detect subgroups and test hypotheses on the parameter differences on one dataset, are advised to split their dataset into two parts before the analyses.
+Then one part of the dataset can be used for exploration, and the other part can be used to test the parameter differences between subgroups.
+
 %To evaluate whether the detected subgroups indeed contribute to better predictions, we used cross validation.
@@
 
-\newpage
+\section{MOB for Detecting DIF in Measurement Models}
+\label{raschtree}
 
-\section{Application Example 2: Subgroup Detection in Rasch Models}
+Validity assessment of scores on psychological or educational tests requires researchers to assess whether the same construct is measured in the same way for different groups.
+In the framework of IRT and Rasch measurement, test items are typically investigated with respect to item misfit, multidimensionality and other violations of the measurement model \citep[cf., for example][for an introduction]{DebStrZei:2022:CRC}.
+A particularly crucial assumption for the comparability of test scores between groups is measurement invariance.
+Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
+
+MOB can be used to detect DIF by means of testing whether the item parameters of the measurement model exhibit significant instability with respect to (combinations of) covariates.
+For the Rasch measurement model, the \texttt{R} function for conducting the MOB analysis is the \texttt{raschtree} function from package \texttt{psychotree} \citep{StroyKopf15}.
+The usage of this function will be illustrated in \hyperref[sec:TutorialRasch]{Application Example 2}.
+We will see that, just like for GLMM trees, we need to specify which variables are part of the parametric model.
+In the case of the Rasch model, this will be the test items.
+Moreover, we need to specify which covariates are made available to the MOB algorithm for selecting relevant splitting variables and cutpoints.
+
+If one joint Rasch model holds for the entire sample, that is, if there is no DIF, a Rasch tree should show no splits.
+However, in certain settings in educational research, such as large scale assessments, very large sample sizes are available for testing for DIF.
+Large sample sizes are good for detecting even small effects or model violations with high statistical power.
+The same holds for the statistical tests used for detecting parameter change in the MOB algorithm, so that in larger samples even very small parameter differences can be detected.
+However, in DIF detection this may mean that even very small DIF effects, that in practice can be considered ignorable, will be detected if only the sample is big enough.
+%As we will illustrate in the first part of \hyperref[sec:TutorialRasch]{Application Example 2}, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
+Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
+\citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
+In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
+\citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
+
+Together with a purification step \citep[see][%and Application Example 2 for details
+]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for graphically highlighting those items that show DIF with respect to certain groups of persons.
+This can help generate hypotheses about the sources of DIF, and can also aid in deciding how to proceed with the DIF items.
+%\todo[inline]{following text from Caro can be used here or later after example 2 or in discussion}
+For example, items that show DIF between different language groups may often be improved by means of making sure that in all translations the meaning is as similar as possible, and/or that the words employed in the translations for the different languages are equally frequently used.
+In other situations, sources of DIF might be harder to eliminate, so that often DIF items may be excluded from a test.
+Either way, the measurement model needs to be refitted and its assumptions checked again after the final set of items has been decided upon and, in the case of modified items, administered again.
+
+Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
+This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
+Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
+These items may then show DIF between native and non-native speakers.
+When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, for example by using more frequently used words, or whether the test should be considered two- rather than one-dimensional.
+For a discussion of the connection between DIF and multidimensionality, see also \cite{Ack:1992,Stretal:2021:APM}.
+
+%We will now illustrate how to use the \texttt{R} packages \texttt{glmertree} and \texttt{psychotree} to conduct the MOB analyses.
+
+\subsection{Application Example 2: Detecting DIF in a General Knowledge Quiz}
 \label{sec:TutorialRasch}
+
+\subsubsection{Dataset}
 
@@
 \end{Schunk}
+
+\subsubsection{Pruning a Rasch tree}
 

### 2024-04-26 _MOB_paper.R

@@ -5,11 +5,11 @@ options(prompt=" ", continue = " ")
 
 
 set.seed(3)
-x1 <- runif(250, min = 5, max = 11)
-x2 <- runif(250, min = 2, max = 8)
-x3 <- runif(250, min = 4, max = 10)
+x1 <- runif(250, min = 5, max = 11) ## age
+x2 <- runif(250, min = 2, max = 8) ## gender
+x3 <- runif(250, min = 4, max = 10) ## reading comprehension
 error <- rnorm(250, sd = 1.5)
-y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
+y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error ## total SDQ score (0-40)
 toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
 toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
 
@@ -22,6 +22,7 @@ boxplot(toy_data$y, cex = .7)
 library("partykit")
 tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
 plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
+term_node_means <- round(tapply(toy_data$y, predict(tree, type = "node"), mean), digits= 2L)
 
 
 ## Load data
@@ -78,6 +79,8 @@ save(HS_dat, file = "HS_dat.Rda")
 load("HS_dat.Rda")
 head(HS_dat, 3)
 
+hist(tapply(HS_dat$Mom_height, HS_dat$MotherID, mean))
+
 
 library("lme4")
 lmm <- lmer(PPVT ~ Program*Age + (1|MotherID/ChildID), data = HS_dat)
@@ -110,14 +113,15 @@ design_df$`PPVT at age 12` <- predict(HS_tree$lmer, newdata = cbind(design_df, A
                                       re.form = ~0)
 names(design_df)[1] <- "Node"
 
-#library("kableExtra")
-#kable_styling(
+library("kableExtra")
+add_footnote(
 knitr::kable(design_df, format = "latex", digits = 2, 
              caption = "Node-specific average predicted PPVT scores at different ages.", 
              label = "predictions", centering = FALSE,
              linesep = "", # linesep command suppressess addlinesep every 5 rows
[…]

### 2024-04-26 Submission 2/Response_letter.tex
added response letter

\documentclass{letter}
\usepackage{graphicx} % Required for inserting images
\usepackage[dvipsnames]{xcolor}

\newcommand{\auth}[1]{\textcolor{RoyalBlue}{#1}}
\newcommand{\MF}[1]{\textcolor{red}{#1}}

\begin{document}

Dear editors of the Journal of School Psychology,

We hereby submit our revision of manuscript XXX.

[TODO: Thank editors and reviewers]

Below, you will find our point-by-point response to the reviewer's comments, including changes made to the manuscript. Reviewer's comments are printed in black, our response printed in blue, verbatim parts from the manuscript are italicized.

Sincerely,

Dr. Marjolein Fokkema, also on behalf of Dr. Mirka Henninger & Prof. Carolin Strobl

\textbf{Reviewer 1}

This paper is a hands-on article providing two examples on how to use model-based recursive partitioning (MOB) in educational research. Two MOB methods were presented: (1) MOB for mixed-effects models and (2) MOB for Rasch measurement and Item Response Theory models. The provided examples are well presented; R-scripts and data are explained well and provided through a github link.

\auth{We thank the reviewer for evaluating our manuscript, the positive evaluation and the helpful suggestions.}

The authors should provided an additional references for the interested reader in the journal called Psychological Test and Assessment Modeling, volume 61, issue 4, 2019; special issue on "Tree-based methods for regression and classification - statistical methods at the interface of graphics and statistics."

\auth{We thank the reviewer for pointing out this special issue, which is indeed very relevant for the current paper. In the first paragraph of the Introduction, we have included three references to papers from the special issue:}

\auth{\textit{For example, recursive partitioning methods have been used to risk factors for adolescent crime (Fritsch, Haupt, Lösel \& Stemmler, 2019) and corporal punishment (Stemmier, Heine \& Wallner, 2019), and to identify predictors of math ability (Ding \& Zhao, 2019).}}

Minor issues:

- Please spell out acronyms like GLMM, DIF and GLM when being introduced for the first time.

\auth{The manuscript now explains the acronym GLMM (in Abstract and Introduction). We have replaced the acronym GLM with "generalized linear model" in the Introduction, because it was only used once. We double checked for the other acronyms (IRT, DIF) whether they were explained upon first use.}

- On page 9, lower paragraph, there are three hints, that there will be an 'Application Example 2'

\auth{Indeed, the many pointers made the text on page 9 somewhat repetitive. We have eliminated some pointers, leaving only two, of which one is in round brackets.}

- p. 11 lower paragraph, there is a period missing between 'height' and 'Note'.

\auth{Thanks for pointing this out, this has now been fixed.}

- p. 18, second paragraph; see the word 'notcontain' needs to be separated.

\auth{Thanks for pointing this out, this has now been fixed.}

- in Figure 6 and 7; there is a node 4 but no node 3!?

\auth{There is a node 4 in these Figures. Nodes 1, 2 and 4 are inner nodes, their numbers are indicated in a square box at the top of the node. Nodes 3, 5 and 6 are terminal nodes, their numbers are written above the node, without a box around it.}

\MF{I can understand where the confusion comes from, but assume that readers will be able to spot node 4, even after some initial confusion. Or should we add a note to the Figures?}

\textbf{Reviewer 2}

In this article, the authors present a tutorial-style introduction of model-based partitioning (MOB). From a person-oriented perspective, this method is of great importance. This perspective proceeds from the assumption that parameters that are estimated for large entities of individuals (populations) can rarely, if ever, be generalized to the individual. Therefore, this article is most welcome, as it introduces readers to a very useful methodology. Both, the introductory section and the application section, are correct, easy to read, and fulfill, thus, the expectations readers may have when processing a tutorial. Unfortunately, however, there are a few issues that need to be addressed before this article can be published.

\auth{We thank the reviewer for evaluating our manuscript, the positive evaluation and the helpful suggestions.}

1. The first of these issues concerns the introduction into MOB. When writing a tutorial, authors need to specify a target audience. In this article, this audience is not easily identified. On the one hand, the text of this article is completely non-technical, in the sense that it contains no equations or formal model specifications. On the other hand, however, this text contains a large number of technical terms that readers in need of a tutorial may not be familiar with. Therefore, I recommend considering to also include a formal presentation of the models that are discussed in the intro and the examples. Those readers who are familiar with the technical terms may appreciate a formal presentation of these models.

\auth{TODO}

\MF{Not completely sure how to deal with this. Adding ALL model representations (parameter stability tests, GLMM, Rasch models) to the paper makes it unsuitable for the readership, I think. We could add it as an Appendix, of course. But I think will not be very relevant for the readership.}

2. The second of these issues concerns the results of the two application examples. In neither example, the resulting subgroups are sufficiently described or explained. This is an omission that lowers the appeal of the tutorial. Please complete the description of the results.

\auth{We have added a heading \textit{Interpretation of a GLMM tree} to the first application example, and a heading \textit{Interpretation of a Rasch tree} to the second application example, to more clearly indicate where we show how to interpret the splits and subgroups in the tree. }

\MF{I have difficulty seeing where this remark cames from. A whole page was dedicated to explaining the tree in Figure 3. Around Figures 5\& 6, there is also an explanation of the subgroups and corresponding difficulty levels for the items. I added some headings to both application examples, so they have similar subsections: Dataset; Fitting a GLMM/Rasch tree; Pruning a Rasch tree; Interpreting a GLMM/Rasch tree. Feel free to correct / undo. And I added a few minor clarifications to the "Interpreting a Rasch tree section" for Figures 6 and 7. Please double check my addition to the explanation of Figure 7, as I added Age being $>24$ in the description of node 5. Should we simply point the reviewer to page 12-13 for interpretation of the GLMM tree, and to page 20-24 for interpretation of Rasch trees? OR copy it here? We coul do the latter, and simply state that we described the subgroups in 'more detail' in the current version of the manuscript.}

3. Finally, before resubmitting, please do some language editing. There are multiple grammatical errors and incomplete sentences in the text. Please correct.

\auth{We apologize for the mistakes, language errors can make reviewing a more tedious job. We have carefully corrected errors throughout the manuscript.}

\MF{I fixed some issues, but still have to do a final round of language checks.}

\textbf{Reviewer 3}

Thank you for the opportunity to review this team's interesting methodological work. The authors present a didactic paper focused on describing the practical features of the semi-parametric model based recursive partitioning method and two ways to apply it in exploratory school psychology-relevant research. I commend the authors on writing so clearly about such a complicated topic enhancing its value for applied researchers without a rigorous background in quantitative methodology. I am also grateful that accompanying code and data were provided. I spent some time replicating the analyses and trying it out on some of my own data sets. I believe the paper has promise as a useful tutorial for this audience but ask that the authors consider some areas of improvement I noted while reading.

\auth{We thank the reviewer for evaluating our manuscript, the positive evaluation and the helpful suggestions.}

1.      In the MOB Algorithm section, I think it would be clearer for readers less familiar with CART, MOB, and related methods if actual example variable names accompanied the abstract y, x1, etc. labels since this is the first exposure. For instance, "...A continuous response variable y (e.g., total difficulties score from the Strengths and Difficulties Questionnaire), two continuous covariates x1 (e.g., Age) and x3 (e.g., Reading comprehension scores), and one categorical covariate x2 (e.g., binary gender)..." You obviously go on to give good concrete examples later, but I still think something like this would help the concepts "click" faster earlier on.

\auth{Thank you for this recommendation, which we have adopted on page 3-4:}

\auth{\textit{We illustrate the idea using a simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.}}

\auth{We again use it on page 5-6:}

\auth{\textit{The means (intercepts) in the terminal nodes are 4.42, 4.78 and 9.17, respectively. While the means and distributions in nodes 2 and 4 appear very similar, the mean in node 5 is obviously higher. Substantively, the tree suggests and interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa). If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at somewhat later age.}}

2.      I know it is well beyond the scope of the paper to present all the technical detail of how the math works behind the scenes yet I think adding a little more about how "instabilities" are quantified broadly would be instructive (like how residual sums of squares are compared in CART).

\auth{TODO, and connect to the first comment of reviewer 2.}

\MF{I'd like your input on how to best deal with this remark. It would be easy to provide some formula's for EFPs, but I also think this will generally not be helpful for the readership.}

3.      One of the stopping criteria was too few observations within a node (i.e., < 250). But it was not clear how 250 was determined. How small is too small? Please add more comment on considerations for this.

\auth{We have now explained this in more detail on page 11 of the manuscript:}

\auth{\textit{Finally, to aid interpretability we want to retain large enough subgroups, we specified the minsize argument. Splits will only be implemented if the resulting nodes contain at least 250 observations. With a total sample size of 1433, even small differences in the effects of age and Head Start participation between subgroups may become significant. The average number of measurements per family was 5.6.
A minimum node size of 250 thus ensured that the estimated effects of age and Head Start
participation in the terminal nodes would at least be based on data from about 50 families.}.}

4.      Non-significant p values were another stopping criterion. Are the p values adjusted for multiple exploratory comparisons so as not to overfit?

\auth{Yes, a Bonferroni correction is applied to the $p$ values, to account for testing instability with respect to all of the potential partitioning variables. We have now made this explicit on page 5 of the manuscript:}

\auth{\textit{By default, a Bonferroni correction is applied to the $p$ values, to account for the fact that the tests are performed for all of the potential partitioning variables.}}

5.      Given that the point of the method is to explore data and generate hypotheses that could be tested with future samples, I think the key thing missing from the paper is some further demonstration of what follow up might actually look like based on these results with real, simulated, or even completely hypothetical data. While there are references at the end pointing to other work on enhancing the stability of results, cross-validation techniques, etc., these papers are more technical in nature and do not serve the same didactic function as this paper. Without going into great technical detail, showing the larger research context around how results from the MOB method could fruitfully inform future studies would add tremendous value for the (mostly) applied researchers who read the journal. Although certainly more work on your part, I truly think adding this additional element is entirely reasonable for a tutorial paper and more in line with the goals of the journal.

\auth{Thank you for pointing this out, we agree with the reviewer's comment, and have added the following suggestion }

\auth{\textit{It should be noted that MOB and GLMM trees are exploratory techniques: The subgroup structures are detected from the data. Such data-driven procedures cannot provide valid standard errors and significance tests to evaluate significance of the observed subgroup differences. To that end, the subgroup structure that was found should be used in a confirmatory analysis on a new sample, that was not used to fit the tree. This would allow to test the hypothesis, either using frequentist or Bayesian approaches, of whether the parameters of interest really differ between subgroups. Researchers who want to both detect subgroups and test hypotheses on the parameter differences on one dataset, are advised to split their dataset into two parts before the analyses. Then one part of the dataset can be used for exploration, and the other part can be used to test the parameter differences between subgroups.}}

\MF{This is just a starting point. I think I need to make this more concrete. Suggestions are welcome.}

6.      I think the paper would flow better if the general description sections for the two applications (e.g., Using MOB for subgroup detection in mixed-effects models) would come right before their respective example walkthrough rather than split up. As it stands now the two general descriptions are grouped together followed by the two examples which required some backtracking on my part to recall the important parts of each application.

\auth{Thank you for this recommendation. We agree with the reviewers' comment and have restructured the manuscript accordingly. \textit{Application Example 1} is now included as a subsection of the section \textit{MOB for Subgroup Detection in Mixed-Effects Models}. Similarly, \textit{Application Example 2} is now included as a subsection of the section \textit{MOB for Detecting DIF in Measurement Models}.}

\end{document}

### 2024-05-08 Submission 2/Response_letter.tex
minor updates

@@
 
-We hereby submit our revision of manuscript XXX.
-
-[TODO: Thank editors and reviewers]
+We hereby submit our revision of manuscript 23-KHVSI123023-535. We are thankful for the efforts of the editor and reviewers, and for the opportunity to revise.
 
@@
 
-Dr. Marjolein Fokkema, also on behalf of Dr. Mirka Henninger & Prof. Carolin Strobl
+Dr. Marjolein Fokkema, also on behalf of Dr. Mirka Henninger and Prof. Carolin Strobl
 
@@
 
-\auth{There is a node 4 in these Figures.
+\auth{There is a node 3 in these Figures, but because it is a terminal node, the formatting of the node number is slightly different.
 Nodes 1, 2 and 4 are inner nodes, their numbers are indicated in a square box at the top of the node.
 Nodes 3, 5 and 6 are terminal nodes, their numbers are written above the node, without a box around it.}
-
-\MF{I can understand where the confusion comes from, but assume that readers will be able to spot node 4, even after some initial confusion. Or should we add a note to the Figures?}
 
@@
 
-\auth{TODO}
-
-\MF{Not completely sure how to deal with this.
-Adding ALL model representations (parameter stability tests, GLMM, Rasch models) to the paper makes it unsuitable for the readership, I think.
-We could add it as an Appendix, of course.
-But I think will not be very relevant for the readership.}
+\auth{This is an important issue, indeed.
+Our aim is to write a non-technical tutorial.
+We appreciate the reviewer's suggestion, but have chosen not to include more formal presentations of the models.
+Interested readers can find those in the methodological papers referred to.
+Instead, in line with the reviewer's comments, we have better clarified technical terms throughout the manuscript.}
 

### 2024-05-08 Raschtree_MH.Rnw
minor updates

@@
 In two-step purification, a new sum score without the items that were diagnosed with DIF in step 1 is computed for the final DIF analysis in step 2.
-When using iterative purification, this process is repeated until two runs yield the same DIF items or until a maximum value of iterations is reached.
+When using iterative purification, this process is repeated until two runs yield the same DIF items or until the maximum number of iterations is reached.
 The final set of DIF-free items resulting from the purification will also be used for aligning the item parameters for each group comparison, which we will discuss and display below.
@@
 We will illustrate this with two examples: First, we can see that for node 1 (split in the covariate \texttt{Gender}), items 1, 6, 7, 8, and 9 show DIF in categories B or C.
-As a reminder, the content of these items was:
+As displayed in the \hyperref[sec:AppendixA]{Appendix}, the content of these items was:
 
@@
 For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison)
-more difficult to answer for younger respondents ($>24$ age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
+more difficult to answer for younger respondents ($24 <$ age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
 

### 2024-05-08 _MOB_paper.Rnw
minor updates

@@
 In the first case study, we show how GLMM trees \citep{FokkySmit18} can be used to detect subgroups with different parameters in mixed-effects models.
-We apply GLMM trees to longitudinal data from a study on the effects of Head Start, to identify subgroups of families where children show comparatively larger or smaller gains in performance.
+We apply GLMM trees to longitudinal data from a study on the effects of the Head Start pre-school program, to identify subgroups of families where children show comparatively larger or smaller gains in performance.
 In a second case study, we show how Rasch trees \citep{StroyKopf15} can be used to detect subgroups with different item parameters in IRT models, i.e.~differential item functioning (DIF).
@@
 
-Model-based recursive partitioning \citep[MOB;][]{ZeilyHoth08} is a semi-parametric approach for detecting differences in the parameters of a statistical model between groups of persons.
-It generalizes the principle of recursive partitioning, that is also used in classification and regression trees \citep[CART; ][]{Breetal:1984}.
-In CART, the aim is to detect groups of persons, defined by (combinations of) covariate values, that differ in their mean on a response variable.
+Model-based recursive partitioning \citep[MOB;][]{ZeilyHoth08} % simplified wording
+allows for detecting subgroups of persons that differ in the parameters of a statistical model.
+To find the subgroups, MOB uses \textit{recursive partitioning}, a technique that became widely known through classification and regression trees \citep[CART; ][]{Breetal:1984}.
+In CART, the aim is to detect groups of persons, defined by specific (combinations of) covariate values, that differ in their mean on a response variable.
 Detecting such subgroups may be of interest in many studies in school and developmental psychology.
-For example, recursive partitioning methods have been used to risk factors for adolescent crime \citep{FrityHaup19} and corporal punishment \citep{StemyHein19}, and to identify predictors of math ability \citep{DingyZhao19}.
+For example, recursive partitioning methods have been used to identify adolescents at higher risk of committing crime \citep{FrityHaup19} or experiencing corporal punishment \citep{StemyHein19}, and to identify predictors of math ability \citep{DingyZhao19}.
 
 MOB also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean.
-For example, MOB can be used to identify children that show stronger or weaker effects of an intervention \cite{ChunyAnso21}, or stronger or weaker gains in academic skills over time \cite{FokkyZeil24}.
-The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as linear and generalized linear regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}.
-In this article, we will highlight two specific MOB methods that we consider particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
+For example, MOB can be used to identify children that show stronger or weaker effects of an intervention \citep{ChunyAnso21}, or stronger or weaker gains in academic skills over time \citep{FokkyZeil24}.
+The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as %simplfied wording
+linear and logistic regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}.
+In this article, we will highlight two specific MOB methods that may be particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
 
@@
 For example, when children are tested at several time points, the time points are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools.
-In the section \hyperref[sec:glmmtree]{MOB for Subgroup Detection in Mixed-Effects Models}, we present \hyperref[sec:TutorialMixed]{Application Example 1} to illustrate how MOB can be used to detect subgroup-specific intervention effects while taking into account the nested data structure.
+In the section \hyperref[sec:glmmtree]{MOB for Subgroup Detection in Mixed-Effects Models}, we present \hyperref[sec:TutorialMixed]{Application Example 1} to illustrate how MOB can be used to detect subgroup-specific intervention effects, while taking into account the nested data structure.
 In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model.
@@
 
-We illustrate the idea using a simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
-To keep the example simple, we apply MOB to a very basic global model, which comprises only an intercept.
+We illustrate the idea using a simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
+To keep the example simple, we apply MOB to a very basic global model, which comprises % simplified wording
+only an intercept (or mean).
 It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
@@
 \item If there is significant structural change, the observations in the current node are split using the covariate associated with the strongest change.
-\item Steps 1--3 are repeated recursively in each resulting node until there is no more significant structural change (or the groups becomes too small).
+\item Steps 1--3 are repeated recursively in each resulting node until there is no more significant structural change (or the groups become too small).
 \end{enumerate}
@@
 In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4).
-The same held for nodes 4 and 5, so that the third covariate, $x_3$, was never selected for splitting.
+The same held for nodes 4 and 5, so that the third covariate $x_3$ was never selected for splitting.
 
@@
 Substantively, the tree suggests and interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa).
-If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at somewhat later age.
+If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at a later age.
 
@@
 \section{MOB for Subgroup Detection in Mixed-Effects Models}
-\label(glmmtree)
+\label{glmmtree}
 
 Mixed-effects models contain two types of effects: Fixed and random effects.
-Fixed effects are typically used to capture population-averaged effects, while random effects are used to capture inter-individual variation deviating from these fixed effects.
+Fixed effects are typically used to capture population-averaged effects, while random effects are used to capture inter-individual variation around these fixed effects.
 In many studies, researchers are specifically interested in testing hypotheses relating to the population-averaged effects, while the random effects are included in the model to properly account for inter-individual variation, and correlations between observations within the same unit \citep{RaudyBush02}.
@@
 
-The \texttt{R} functions for fitting GLMM trees are the \texttt{lmertree} and \texttt{glmertree} functions from package \texttt{glmertree}.
+The \texttt{R} functions for fitting GLMM trees are the \texttt{lmertree} (for linear mixed models) and \texttt{glmertree} (for functions from package \texttt{glmertree}.
 The usage of this function will be illustrated in \hyperref[sec:TutorialMixed]{Application Example 1}.
@@
 
-The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other preschool program.
-Data from children in those families who participated in another preschool program were excluded.
+The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other pre-school program.
+Data from children in those families who participated in another pre-school program were excluded.
 The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
@@
 
-Finally, to aid interpretability we want to retain large enough subgroups, we specified the \texttt{minsize} argument.
-Splits will only be implemented if the resulting nodes contain at least 250 observations.
+To aid interpretation, we would like to retain large enough subgroups.
+We therefore specified the \texttt{minsize} argument, so that splits will only be implemented if the resulting nodes contain at least 250 observations.
 With a total sample size of \Sexpr{round(nrow(HS_dat))}, even small differences in the effects of age and Head Start participation between subgroups may become significant.
@@
 These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14.
-Fade out was reported to be particularly strong for Afriacn-American and very disadvantaged children.
+Fade out was reported to be particularly strong for African-American and very disadvantaged children.
 The global benefit of participation in Head Start as shown in Figure~\ref{fig:global_lmm} and reported by \cite{Demi09} thus seems to differ across groups.
@@
 A particularly crucial assumption for the comparability of test scores between groups is measurement invariance.
-Items that violate measurement invariance by showing different measurement properties for different groups of participants display DIF.
+Items that violate measurement invariance by showing different measurement properties for different groups of participants exhibit differential item functioning (DIF).
 
@@
 Another important aspect to keep in mind is that DIF can be caused by one or more items measuring a secondary dimension in addition to the dimension that is intended to be measured by the test.
-This would be the case in instruments intended to measure math aptitude containing pure algebra problems as well as story problems.
-Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when they contain seldomly used words.
+This can be the case, for example, for an instrument intended to measure math aptitude, that contains both pure algebra problems as well as story problems.
+Students whose native language is not the same as the test language can have a disadvantage in answering the story problems, for example when these contain seldomly used words.
 These items may then show DIF between native and non-native speakers.
@@
 The same argument holds for parameter differences in mixed-effects models, where the substantial hypotheses formulated a priori are typically about global intervention effects, not about possible subgroup-specific differences.
-GLMM trees allow for exploring whether fixed effects differ as a function of the partitioning variable; that is, identifying possible interactions the predictor of interest may be involved in.
+GLMM trees allow for exploring whether fixed effects differ as a function of the partitioning variables; that is, identifying possible interactions the predictor of interest may be involved in.
 

### 2024-05-08 _MOB_paper.tex
minor updates

@@
 
-We illustrate the idea using a simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
-To keep the example simple, we apply MOB to a very basic global model, which comprises only an intercept.
+We illustrate the idea using a simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
+To keep the example simple, we apply MOB to a very basic global model, which comprises % simplified wording
+only an intercept (or mean).
 It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
@@
 
-The sample consists of 273 families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other preschool program.
-Data from children in those families who participated in another preschool program were excluded.
+The sample consists of 273 families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other pre-school program.
+Data from children in those families who participated in another pre-school program were excluded.
 The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
@@
 
-Finally, to aid interpretability we want to retain large enough subgroups, we specified the \texttt{minsize} argument.
-Splits will only be implemented if the resulting nodes contain at least 250 observations.
+To aid interpretation, we would like to retain large enough subgroups.
+We therefore specified the \texttt{minsize} argument, so that splits will only be implemented if the resulting nodes contain at least 250 observations.
 With a total sample size of 1433, even small differences in the effects of age and Head Start participation between subgroups may become significant.

### 2024-05-13 Raschtree_MH.Rnw
Pruning -> Reversing splits
Second submission J of School Psych
Second submission J of School Psych

@@
 @
-
-\subsubsection{Pruning a Rasch tree}
 
@@
 
-\subsubsection{Interpretation of a Rasch tree}
+\subsubsection{Interpreting a Rasch tree}
 
@@
 In addition, in each node the number of items classified as A, B, or C is displayed.
-For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and 4 items are classified in category C (large DIF).
+For instance, we can see that in node 1 one item is classified in category B (moderate DIF), and four items are classified in category C (large DIF).
 
@@
 For node 4 (split in the covariate \texttt{Age}), items 1 (Painting by Andy Warhol) and 9 (Opera not by Mozart: Aida) show DIF and are %(relative to the DIF-free items for this group comparison)
-more difficult to answer for younger respondents ($24 <$ age $\leq 29$), and easier or less difficult to answer for respondents $>29$.
+more difficult to answer for younger male respondents ($24 <$ age $\leq 29$), and easier or less difficult to answer for older male respondents (age $>29$).
 
@@
 The displayed information again shows that some items show only negligible DIF (category A) in all nodes, while other items show medium (category B) or large (category C) DIF in some or all nodes.
-The values of the Mantel-Haenszel effect size measure are negative when the item difficulty is lower in the right daughter node than in the left daughter node.
+The values of the Mantel-Haenszel effect size measure are negative when the item difficulty is lower in the right than in the left child node.
 For example, as we already observed in Figure~\ref{fig:MHtree3}, items 1, 6, 7, 8, and 9 have medium or large DIF (DIF categories B and C) and are  easier to answer %(relative to the DIF-free items for this group comparison)
-for participants with female or missing gender, which correspond to the right daughter node of node 1.
+for participants with female or missing gender (node 7).
 % Item 4 also has a negative sign, indicating that the item difficulty is lower for participants with female or missing gender.

### 2024-05-13 _MOB_paper.Rnw
Update affiliation
Second submission J of School Psych
Second submission J of School Psych

@@
 DIF should be investigated before using test results for group comparisons.
-We show how a recently developed stopping criterion \citep{HenDebStr:2023:EPM} can be used to guide subgroup detection based on DIF effect sizes.\\}
+We show how a recently developed stopping criterion \citep{HennyDeba23} can be used to guide subgroup detection based on DIF effect sizes.\\}
 
@@
 
-       \textbf{Marjolein Fokkema$^1$, Mirka Henninger$^2$ and Carolin Strobl$^2$}\\
-       $^1$Leiden University, $^2$Universit\"at Z\"urich
+       \textbf{Marjolein Fokkema$^1$, Mirka Henninger$^{2,3}$ and Carolin Strobl$^2$}\\
+       $^1$Leiden University, $^2$University of Zurich, $^3$ University of Basel
 
@@
 allows for detecting subgroups of persons that differ in the parameters of a statistical model.
-To find the subgroups, MOB uses \textit{recursive partitioning}, a technique that became widely known through classification and regression trees \citep[CART; ][]{Breetal:1984}.
+To find the subgroups, MOB uses \textit{recursive partitioning}, a technique that was first popularized by the classification and regression trees \citep[CART; ][]{Breetal:1984} algorithm.
 In CART, the aim is to detect groups of persons, defined by specific (combinations of) covariate values, that differ in their mean on a response variable.
 Detecting such subgroups may be of interest in many studies in school and developmental psychology.
-For example, recursive partitioning methods have been used to identify adolescents at higher risk of committing crime \citep{FrityHaup19} or experiencing corporal punishment \citep{StemyHein19}, and to identify predictors of math ability \citep{DingyZhao19}.
+For example, recursive partitioning methods have been used to identify adolescents at risk of committing crime \citep{FrityHaup19} and experiencing corporal punishment \citep{StemyHein19}, and to identify predictors of math ability \citep{DingyZhao19}.
 
-MOB also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters of a statistical model, instead of only the mean.
+Like CART, MOB also identifies groups of persons defined by (combinations of) covariate values, but the groups can differ in a wider range of parameters, instead of only the mean.
 For example, MOB can be used to identify children that show stronger or weaker effects of an intervention \citep{ChunyAnso21}, or stronger or weaker gains in academic skills over time \citep{FokkyZeil24}.
-The framework of MOB is very flexible and it can be applied to a wide range of statistical models, such as %simplfied wording
+The framework of MOB is very flexible and can be applied to a wide range of statistical models, such as %simplified wording
 linear and logistic regression \citep{KopAugStr:2013,ZeilyHoth08} or models for paired-comparison data \citep{StrWicZei:2011:JoEaBS,WiedyFrick21}.
-In this article, we will highlight two specific MOB methods that may be particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StrKopZei:2015:P,KomStrZei:2017:EaPM,HenDebStr:2023:EPM}.
+In this article, we will highlight two specific MOB methods that may be particularly relevant for research in school psychology: MOB for mixed-effects models \citep{FokkySmit18} and MOB for Rasch measurement and Item Response Theory models \citep[IRT,][]{StroyKopf15, KomStrZei:2017:EaPM, HennyDeba23}.
 
-Mixed-effects models become relevant whenever data are collected in repeated measures or nested data structures.
-For example, when children are tested at several time points, the time points are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools.
-In the section \hyperref[sec:glmmtree]{MOB for Subgroup Detection in Mixed-Effects Models}, we present \hyperref[sec:TutorialMixed]{Application Example 1} to illustrate how MOB can be used to detect subgroup-specific intervention effects, while taking into account the nested data structure.
+Mixed-effects models become relevant whenever data are collected in longitudinal or nested data structures.
+For example, when children are tested at several time points, observations at each timepoint are nested in children; when children from different classes from different schools participate in a study, children are nested in classes, which are in turn nested in schools.
+In the section \hyperref[sec:glmmtree]{MOB for Subgroup Detection in Mixed-Effects Models}, we present an \hyperref[sec:TutorialMixed]{example application} to illustrate how MOB can be used to detect subgroup-specific intervention effects, while taking into account the nested data structure.
 In this example, the focus is on detecting subgroups with different parameters of a \textit{regression} model.
@@
 Validity assessment requires that we make sure that test results are comparable, for example between children of different genders.
-If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage, and these items are said to violate measurement invariance or to exhibit differential item functioning (DIF).
+If certain items show different measurement parameters between groups, this may put certain groups at a relative disadvantage.
+These items are said to violate measurement invariance or exhibit differential item functioning (DIF).
 In order to test for DIF in the framework of Item Response Theory or Rasch modelling \citep{AnthyDiPe16,DebStrZei:2022:CRC,Mall97}, the item parameters are compared between groups of persons.
@@
 Traditional approaches for testing DIF require the groups to be pre-specified in order to test for DIF.
-In \hyperref[sec:TutorialRasch]{Application Example 2} we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.
+In an \hyperref[sec:TutorialRasch]{application example} we show how MOB flexibly allows to detect groups with different item parameters in a data-driven way.
 
@@
 Our aim here is to provide an informal introduction and to illustrate the relevance of these methods for studies in school psychology.
-We will specify the relevant works providing the formal details of MOB and the extensions presented here in the \hyperref[sec:discussion]{Discussion} section.
+For readers interested in the more formal details of MOB and the extensions presented here, relevant references are provided throughout the manuscript.
 
@@
 
-We illustrate the idea using a simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
+We illustrate this idea using a simple, simulated toy dataset, comprising \Sexpr{nrow(toy_data)} observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
 To keep the example simple, we apply MOB to a very basic global model, which comprises % simplified wording
 only an intercept (or mean).
-It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
+It would also be possible to use, for example, a linear or logistic regression model as the global model.
 Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot.
-Obviously, a global intercept or mean would not describe all observations equally well, there is quite some unexplained variation around the sample mean of \Sexpr{round(mean(toy_data$y), digits = 2)}.
+Obviously, a global intercept or mean would not describe all observations equally well, because there is quite some unexplained variation around the sample mean of \Sexpr{round(mean(toy_data$y), digits = 2)}.
 
@@
 boxplot(toy_data$y, cex = .7)
-#points(x = 1, y = mean(toy_data$y), pch = 19, ylim = c(1,14))
 @
@@
 
+\begin{figure}[h]
+\caption{Top row: Residuals ordered by the values of the partitioning variables. Bottom row: Summed residuals (cumulative sums for continuous predictors, sums by level for categorical predictors).}
+\begin{subfigure}[][][t]{1.2\textwidth}
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=12, height=4>>=
+library("strucchange")
+gefp_x1 <- gefp(y ~ 1, fit = lm, data = toy_data, order.by = ~ x1)
+gefp_x3 <- gefp(y ~ 1, fit = lm, data = toy_data, order.by = ~ x3)
+lmod <- lm(y ~ 1, data = toy_data)
+resids <- residuals(lmod)
+par(mfrow = c(1, 3))
+plot(toy_data$x1, resids, xlab = "x1", ylab = "residuals", 
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+plot(toy_data$x2, resids, xlab = "x2", ylab = "residuals", 
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+plot(toy_data$x3, resids, xlab = "x3", ylab = "residuals", 
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+@
+
+<<eval = TRUE, echo = FALSE, fig=TRUE, width=12, height=4>>=
+par(mfrow = c(1, 3))
+plot(toy_data$x1[order(toy_data$x1)], cumsum(resids[order(toy_data$x1)]), 
+     xlab = "x1", ylab = "cumulative sum", type = "l",
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+boxplot(t(replicate(100, tapply(resids, toy_data$x2, sum))), 
+        ylab = "sum", cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+plot(toy_data$x3[order(toy_data$x3)], cumsum(resids[order(toy_data$x3)]), 
+     xlab = "x3", ylab = "cumulative sum", type = "l",
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+@
+\end{subfigure}
+\label{fig:efp}
+\end{figure}
+
 To detect possible subgroups with different values for the parameters, MOB cycles iteratively through the following steps:
@@
 We applied MOB to the observations in the left panel of Figure~\ref{fig:toy}, specifying $x_1$, $x_2$ and $x_3$ as potential partitioning variables.
+In Step 1, the model parameters (here: the intercept) are estimated in the root node, which contains all observations.
+%changed
+In Step 2, the model's \emph{scores} are computed.
+In case of an intercept-only model, these are simply the residuals.
+The scores are ordered with respect to the possible partitioning variables, as shown in the top row of Figure~\ref{fig:efp}.
+The scores should randomly fluctuate around their mean of zero, but the top row of Figure~\ref{fig:efp} suggests systematic deviations: there seems to be an association with $x_1$ and $x_2$, and little or no association with $x_3$.
+
+MOB captures systematic deviations in the scores (i.e., structural change) by computing cumulative sums.
+The cumulative sums are depicted in the bottom row of Figure~\ref{fig:efp}.
+To decide in Step 3 whether to split the node, and if so, which variable to split on, a test statistic is computed based on the cumulative sums of each partitioning variable; full details on the computation are provided in \citep{ZeilyHoth08}.
+In the current example, the structural change tests revealed that $x_1$ was most strongly associated with instabilities in the intercept and it was thus selected for splitting in Step 3.
+
+In Step 4, the procedure is repeated in each of the resulting subgroups.
 The resulting tree is shown in the right panel of Figure~\ref{fig:toy}.
-In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2).
-The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected first for splitting (step 3).
-The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node.
-By default, a Bonferroni correction is applied to the $p$ values, to account for the fact that the tests are performed for all of the potential partitioning variables.
-Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroup are as similar as possible.
-In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate.
-Again, the cutpoint in $x_2$ was selected so that the two resulting subgroups exhibit strongest parameter differences.
-In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4).
+In the right subgroup (node 3), additional significant instability was detected with respect to $x_2$, a categorical covariate.
+Again, the cutpoint for $x_2$ was selected so that the two resulting subgroups exhibit an as large as possible parameter difference.
+In the left subgroup (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup.
 The same held for nodes 4 and 5, so that the third covariate $x_3$ was never selected for splitting.
 
-The subgroup-specific distributions of the response variable are presented in the terminal nodes at the bottom of the tree.
-The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the pattern in the data.
-The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
+The $p$ values resulting from the structural change tests of each split are depicted in the corresponding nodes.
+By default, the $p$ values are Bonferroni corrected, to account for the fact that the tests are performed for all potential partitioning variables.
+After selecting $x_1$ for splitting, the cutpoint is selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroups are as similar as possible.
+
+The subgroup-specific distributions of the response variable are depicted in the terminal nodes at the bottom of the tree.
+The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the patterns in the data.
+The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were selected for splitting.
 This automatic variable selection is an important characteristic of the CART and MOB algorithms.
+
 The means (intercepts) in the terminal nodes are \Sexpr{term_node_means[1]}, \Sexpr{term_node_means[2]} and \Sexpr{term_node_means[3]}, respectively.
 While the means and distributions in nodes 2 and 4 appear very similar, the mean in node 5 is obviously higher.
-Substantively, the tree suggests and interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa).
+Substantively, this tree suggests an interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa).
 If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at a later age.
@@
 The latter forms the basis for all MOB approaches presented in this paper.
-For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006,StrMalTut:2009:PM,StroyKopf15}.
+For more details on the statistical theory behind unbiased classification and regression trees and MOB, see \citet{Hotetal:2006}, \citet{StrMalTut:2009:PM} and /citet{StroyKopf15}.
 
@@
 These criteria ensure that the sample sizes in the terminal nodes are large enough to estimate the statistical model in each terminal node.
-For reasons of interpretability, users may sometimes specify even higher requirements for the sample sizes of terminal nodes.
-Note that it is very well possible that no split will be implemented, when none of the specified partitioning variables show significant structural change.
+For reasons of interpretability, users may specify even higher requirements for the sample sizes of terminal nodes.
 
-The third important characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner.
+The third characteristic of classification and regression trees as well as MOB is that the entire structure identified by the trees does not have to be pre-specified by the researcher in a confirmatory manner, but is learned from the data in an exploratory manner.
 This is a key feature of the MOB approach that makes it very flexible and sets it apart from purely parametric approaches, where only those main effects and interactions that are explicitly included in the specification of the model are considered.
@@
 \section{MOB for Subgroup Detection in Mixed-Effects Models}
-\label{glmmtree}
+\label{sec:glmmtree}
 
@@
 GLMM trees combine MOB and generalized linear mixed-effects models (GLMMs) and were introduced by \cite{FokkySmit18}.
-While the 'standard' MOB trees \citep{ZeilyHoth08} allow for subgroup detection in fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for partitioning mixed-effects regression models.
-Because researchers' interests in multilevel models commonly focus on the fixed effects (of time or treatment, for example), the GLMM tree algorithm only targets structural change in the fixed-effects parameters.
-The random-effects parameters can be specified as usual and are assumed constant; that is, they are estimated using all observations in the dataset.
-Structural change is thus assessed with respect to the fixed-effects parameters, while dependency between observations within the same units is accounted for by the random effects.
+While the 'standard' MOB trees \citep{ZeilyHoth08} allow for subgroup detection in fixed-effects GLMs, GLMM trees additionally estimate and account for random effects and can thus be used for subgroup detection in mixed-effects regression models.
+The GLMM tree algorithm only targets structural change in the fixed-effects parameters.
+The random-effects parameters are specified as usual and assumed constant; that is, they are estimated using all observations in the dataset.
 The resulting subgroups will differ in their estimates for the fixed-effects coefficients only.
@@
 
-The \texttt{R} functions for fitting GLMM trees are the \texttt{lmertree} (for linear mixed models) and \texttt{glmertree} (for functions from package \texttt{glmertree}.
-The usage of this function will be illustrated in \hyperref[sec:TutorialMixed]{Application Example 1}.
-Further mathematical and computational details about the GLMM tree model are also described in \cite{FokkySmit18} and \cite{FokkyZeil24}.
+The \texttt{R} functions for fitting GLMM trees are \texttt{lmertree} and \texttt{glmertree}.
+The former assumes a normally distributed response, while the latter supports, for example, binomial or count responses.
+The following section will illustrate the usage of \texttt{lmertree}.
+Mathematical and computational details of the GLMM tree algorithm are described in \cite{FokkySmit18} and \cite{FokkyZeil24}.
 
@@
 
-To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start.
+To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in the Head Start program.
 Head Start is a federally funded nationwide pre-school program for children from low-income families in the United States.
 Participation in Head Start takes place from ages 3 through 5.
-\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth \citep{NLSY}.
+\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the 1979 cohort of the National Longitudinal Survey of Youth \citep{NLSY}.
 
 The sample consists of \Sexpr{length(unique(data$MotherID))} families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other pre-school program.
-Data from children in those families who participated in another pre-school program were excluded.
+Data from children who participated in another pre-school program were excluded.
 The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
-The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT]{DunnyDunn81}, we will thus model PPVT trajectories over time.
-On average, there were \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(1-mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there was only a single PPVT score available.
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT; ][]{DunnyDunn81}, for which we model trajectories over time.
+On average, there were \Sexpr{round(mean(table(HS_dat$ChildID)), digits = 2)} PPVT scores per child, for \Sexpr{round(100*(mean(table(HS_dat$ChildID) > 1)), digits = 0)}\% of the children there were $>1$ PPVT scores available.
 
@@
 
-Our dataset contains five family characteristics: The mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; the families' income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height (computed as feet x 100 + inches).
+Our dataset contains five family characteristics: Mother's score on the Armed Forced Qualification Test (AFTQ), adjusted for age; family income (averaged over the years for which data was available); race (Black, Hispanic or White); mother's years of completed education; mother's height (computed as feet x 100 + inches).
 Note that the latter variable is one that should be completely irrelevant for predicting performance on a vocabulary test; it is included here to illustrate that the GLMM tree algorithm can fruitfully distinguish signal from noise variables.
@@
 
-We inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None (black) versus Head Start (red).
+We first inspect the complete dataset by plotting PPVT scores against age, separated by program participation: None versus Head Start.
 To show the global effect of age and Head Start participation, we first fit a mixed-effects model comprising their main and interaction effects, using package \texttt{lme4}.
-This would be the model fitted in the root node of the GLMM tree.
-To account for the correlation between repeated assessments on the same child, and between siblings with the same mother, we specify a random intercept for children, nested within mothers:
+This would also be the model fitted in the root node of the GLMM tree:
 
@@
 @
+
+We specified the fixed main and interaction effects of `Program` and `Age` as `Program*Age`.
+The random effects were specified as `(1|MotherID/ChildID)`.
+The \texttt{1} specifies that a random intercept should be estimated and \texttt{MotherID/ChildID} specifies that every child and every mother should obtain a random intercept, so as to account for correlations between repeated assessments of the same child, and correlations between children of the same mother.
+The forward slash \texttt{/} indicates that the children are 'nested' within mothers; that is, that there are multiple children per mother, but only one mother per child.
 
@@
 
-Next, we test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:
+We now test whether the intercepts and slopes of the two regression lines differ as a function of the partitioning variables, using function \verb|lmertree| from R package \textbf{glmertree}:
 
@@
 With the first argument, we specified the model \verb|formula|, which has three parts separated by vertical bars: The left part (\verb|PPVT ~ Program*Age|) specifies the response variable, followed by a tilde (\verb|~|) and the fixed-effects predictors of relevance.
-The middle part (\texttt{1|MotherID/ChildID}) specified the random effect: Repeated PPVT assessments are nested within children, which are nested within mothers.
+The middle part (\texttt{1|MotherID/ChildID}) specified the random effects, which are identical to the random effect in the earlier mixed-effects model.
 The right part (\verb|AFTQ + Race + Income + Mom_edu_yrs + Mom_height|) specified the partitioning variables: covariates that may possibly affect the values of the fixed-effects parameters.
@@
 With the second argument, we specified the dataset which contain the variables.
-With the \texttt{cluster} argument, we account for the dependence between observations within the same cluster.
-In this example, the partitioning variables are measured at the mother (or family) level, which we indicate by specifying \texttt{cluster = MotherID}.
+With the \texttt{cluster} argument, we account for the level at which the partitioning variables ae measured.
+In this example, the partitioning variables are measured at the mother (or family) level, which we indicated by specifying \texttt{cluster = MotherID}.
 As a result, the parameter stability tests will employ so-called clustered covariances \citep{ZeilyKoll20}.
@@
 With multiple fixed-effects predictors of interest, the default plots may become too crowded or difficult to interpret.
-We therefore specify \verb|type = "simple"| to facilitate interpretation, and using the \verb|nodesize_level| argument, we specified that the sample size printed above every terminal node should count the number of children, not the number of individual observations:
+We therefore specify \verb|type = "simple"| to facilitate interpretation, and using the \verb|nodesize_level| argument, we specified that the sample size printed above every terminal node should count the number of children, instead of the number of individual observations, which would be printed by default:
 
@@
 knitr::kable(design_df, format = "latex", digits = 2, 
-             caption = "Node-specific average predicted PPVT scores at different ages.", 
+             caption = "Node-specific predicted PPVT scores at different ages.", 
              label = "predictions", centering = FALSE,
@@
 #latex_options = "HOLD_position")
-"HS = Head Start; PPVT = Peabody Picture Vocabulary Test.")
+"\textit{Note.} For computing predictions, random effects were assumed zero. HS = Head Start; PPVT = Peabody Picture Vocabulary Test.", notation = "none")
 @
@@
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
-This benefit increases over time for the group with higher mother's AFTQ scores.
+This benefit increases over time for the group with higher mother's AFTQ scores (nodes 6 and 7).
 The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2).
-For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12.
-These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14.
+For Black and Hispanic children with intermediate mother's AFTQ scores (node 5), the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12.
+
+These results partly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 standard deviations at ages 11-14.
 Fade out was reported to be particularly strong for African-American and very disadvantaged children.
-The global benefit of participation in Head Start as shown in Figure~\ref{fig:global_lmm} and reported by \cite{Demi09} thus seems to differ across groups.
+In the current sample, the standard deviation of PVVT scores was \Sexpr{round(sd(HS_dat$PPVT), digits = 2)}, and thus the predictions for age 6 in Table~\ref{tab:predictions} roughly correspond to the effect found by \cite{Demi09}.
+However, we found a different pattern of fade-out.
 
@@
 To that end, the subgroup structure that was found should be used in a confirmatory analysis on a new sample, that was not used to fit the tree.
-This would allow to test the hypothesis, either using frequentist or Bayesian approaches, of whether the parameters of interest really differ between subgroups.
-Researchers who want to both detect subgroups and test hypotheses on the parameter differences on one dataset, are advised to split their dataset into two parts before the analyses.
-Then one part of the dataset can be used for exploration, and the other part can be used to test the parameter differences between subgroups.
+This would allow to test the hypothesis, either using frequentist or Bayesian approaches, of whether the parameters of interest really differ between the detected subgroups.
+
+If additional data is not available or cannot be collected, another possibility would be to split a single dataset into two parts before the analyses.
+Then one part of the dataset can be used for exploration (i.e., detecting the subgroups), and the other part can be used to test the parameter differences between the subgroups.
 
@@
 \section{MOB for Detecting DIF in Measurement Models}
-\label{raschtree}
+\label{sec:raschtree}
 
@@
 For the Rasch measurement model, the \texttt{R} function for conducting the MOB analysis is the \texttt{raschtree} function from package \texttt{psychotree} \citep{StroyKopf15}.
-The usage of this function will be illustrated in \hyperref[sec:TutorialRasch]{Application Example 2}.
+The usage of this function will be illustrated in the next section (\hyperref[sec:TutorialRasch]{Application Example 2}).
 We will see that, just like for GLMM trees, we need to specify which variables are part of the parametric model.
@@
 %As we will illustrate in the first part of \hyperref[sec:TutorialRasch]{Application Example 2}, for a large sample of university students who have taken a quiz to test their general knowledge, this can lead to very large trees, that are hard to interpret and contain splits that would not be considered relevant by measurement experts.
-Therefore, an extension of Rasch trees has been suggested by \citet{HenDebStr:2023:EPM} based on the Mantel-Haenszel effect size measure for DIF.
+Therefore, an extension of Rasch trees has been suggested by \citet{HennyDeba23} based on the Mantel-Haenszel effect size measure for DIF.
 \citet{HolTha:1985} have suggested an intuitive classification of DIF effect sizes based in the Mantel-Haenszel statistic, that is being widely used in educational testing.
 In this classification, category A stands for negligible DIF (small effect size or not statistically significant), B for medium DIF (neither A nor C), and C for large DIF (large effect size and statistically significant).
-\citet{HenDebStr:2023:EPM} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
+\citet{HennyDeba23} have incorporated this classification as an additional stopping criterion for Rasch trees, so that the user can decide, for example, that a split should only be conducted if the detected DIF is of category B or C, while negligible DIF of category A should be ignored.
 
-Together with a purification step \citep[see][%and Application Example 2 for details
-]{HenDebStr:2023:EPM}, the Mantel-Haenszel classification can also be used for graphically highlighting those items that show DIF with respect to certain groups of persons.
+Together with a purification step \citep[see][for details]{HennyDeba23}, the Mantel-Haenszel classification can also be used for graphically highlighting those items that show DIF with respect to certain groups of persons.
 This can help generate hypotheses about the sources of DIF, and can also aid in deciding how to proceed with the DIF items.
@@
 When encountering this, the test developers will have to decide whether the items with story problems should be excluded from the test, whether they can be improved, for example by using more frequently used words, or whether the test should be considered two- rather than one-dimensional.
-For a discussion of the connection between DIF and multidimensionality, see also \cite{Ack:1992,Stretal:2021:APM}.
+For a discussion of the connection between DIF and multidimensionality, see also \cite{Ack:1992} and \cite{Stretal:2021:APM}.
 
@@
 \section{Discussion}
-\label{discussion}
+\label{sec:discussion}
 
@@
 This makes it more flexible for detecting differences that were not hypothesized by the researcher.
-For example in DIF analysis, it is often the case that obvious sources of DIF have already been avoided by the content experts during item creation.
+For example, it is often the case that obvious sources of DIF have already been avoided by the content experts during item creation.
 Any remaining DIF is unexpected.
-Therefore, any DIF detection approach that relies on the researchers correctly specifying the exact groups of persons that exhibit DIF can miss DIF if it is associated with other (combinations of) covariates or other cutpoints than the ones investigated.
-In this sense, the more exploratory approach of Rasch trees has higher statistical power to detect DIF in previously unknown groups \citep{StrKopZei:2015:P}.
-The same argument holds for parameter differences in mixed-effects models, where the substantial hypotheses formulated a priori are typically about global intervention effects, not about possible subgroup-specific differences.
-GLMM trees allow for exploring whether fixed effects differ as a function of the partitioning variables; that is, identifying possible interactions the predictor of interest may be involved in.
+Therefore, DIF detection approaches that rely on the researcher to correctly specify the exact groups of persons that exhibit DIF can miss DIF, because it may be associated with other (combinations of) covariates than the ones investigated.
+In this sense, the more exploratory approach of Rasch trees has higher statistical power to detect DIF in previously unknown groups \citep{StroyKopf15}.
+The same argument holds for parameter differences in mixed-effects models, where the substantial hypotheses formulated a priori are typically about global intervention effects or global effects of time, and not about possible subgroup-specific differences.
+GLMM trees allow for identifying possible interactions the predictor of interest may be involved in.
 
-However, researchers should be aware that a covariate-based approach like MOB will only be able to detect parameter differences if the relevant covariates were recorded in the study and supplied to the algorithm.
+However, researchers should be aware that a covariate-based approach like MOB will only be able to detect parameter differences if the relevant covariates were recorded and supplied to the algorithm.
 If no (or few) potentially relevant covariates are available, an alternative framework for detecting previously unknown groups of persons with different model parameters is mixture modelling \citep[see, e.g.,][in the context of Rasch modelling]{AyalySant17,FriStrZei:2015:EaPM}.
@@
 For example, if DIF in the item parameters of a test on reading ability is found between different cities or districts, a variety of factors could be causally driving these effects, such as different compositions of native languages or sociodemographics.
+On the other hand, MOB can also be useful in evaluating causal effects, as it can be used to test for possible confounding \citep{vanWyLi19}.
 
@@
 While the MOB algorithm, as it is implemented in the \texttt{R} packages presented in this tutorial, has been constructed with care and uses statistical significance tests (and in the case of Mantel-Haenszel trees also effect sizes) to avoid spurious splits, for confirmatory hypothesis testing it is recommended to use a new dataset, not used for fitting the tree.
-Further, resampling or $k$-fold cross validation techniques may be used to assess stability and accuracy of the model.
-\cite{PhilyRusc18} show how to use these techniques for assessing the stability of the set of selected variables and cutpoints.
-For GLMM trees, cross-validation can additionally be used for evaluating predictive accuracy \citep[e.g., ][]{RooiyWeed20}, or for further optimizing predictive accuracy through the choice of model-fitting parameters, like the minimum number of observations in the terminal nodes or the maximum tree depth \citep{JameyWitt23}.
+
+For example, the results of the GLMM tree analysis (\hyperref[sec:TutorialMixed]{Application Example 1}) could be validated using later cohorts from the same study.
+On data from later cohorts, a confirmatory mixed-effects model could be fitted, similar to the one depicted in Figure~\ref{fig:global_lmm}, but with additional binary indicators distinguishing the subgroups.
+The model could be specified to include main effects as well as interactions with time and treatment (i.e., Head Start versus no Head Start).
+This refitting on new data will yield valid standard errors, and thus provide hypothesis tests on subgroup differences.
+Alternatively, if the interest of the analyses was primarily to identify moderators of the effect of Head Start, the model fitted on a later cohort could comprise main and interactions effects of time, Head Start, Mother's AFTQ score and Race.
+
+Identified subgroups can also guide the design of future studies.
+For example, if subgroups with differential treatment effects were identified by a GLMM tree in one randomized clinical trail (RCT), a next RCT may be designed so that it has adequate power to establish the differential subgroup effects with a confirmatory analysis.
+Or, after developing a new quiz, or improving the quiz analysed in \hyperref[sec:TutorialRasch]{Application Example 2}, a validation study may be conducted so as to assess DIF between the subgroups identified with a Rasch tree in an earlier analysis.
+This would allow to assess whether the DIF has been mitigated in the new or improved quiz.
+
+As mentioned in \hyperref[sec:TutorialMixed]{Application Example 1}, resampling or $k$-fold cross validation techniques may be used to assess stability and accuracy of the GLMM tree.
+\cite{PhilyRusc18} show how to use these techniques for assessing the stability of selected variables and cutpoints.
+For GLMM trees, cross-validation can additionally be used for evaluating predictive accuracy \citep[e.g., ][]{RooiyWeed20}, or for further optimizing predictive accuracy through tuning model-fitting parameters, like the minimum number of observations to be retained in terminal nodes or the maximum tree depth.
 
@@
 This also supports researchers in interpreting the tree's results with respect to whether DIF effects are negligible, medium, or large.
-For mixed-effects models, computation of effect sizes is far from straightforward and several different approaches have been proposed \citep[e.g., ][]{JuddyWest17, BrysyStev18, Bran18}.
-Effect-size based stopping criteria for GLMM trees are therefore currently not available, but any of the proposed computation approaches proposed for traditional GLMMs can also be computed from the parameter estimates of GLMM trees.
+For mixed-effects models, computation of effect sizes is much less straightforward and several different approaches have been proposed \citep[e.g., ][]{JuddyWest17, BrysyStev18, Bran18}.
+Effect-size based stopping criteria for GLMM trees are therefore currently not available, but in principle, any effect size that can be computed for traditional GLMMs can also be computed for GLMM trees.
 

### 2024-05-13 _MOB_paper.tex
Second submission J of School Psych
Second submission J of School Psych

@@
 
-We illustrate the idea using a simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
+We illustrate this idea using a simple, simulated toy dataset, comprising 250 observations and four variables: A continuous response variable $y$ (e.g., a total score for behavioral difficulties), and three covariates, $x_1$ (e.g., age), $x_2$ (e.g., a reading comprehension score) and $x_3$ (e.g., gender, with levels male, female and non-binary), as possible partitioning variables.
 To keep the example simple, we apply MOB to a very basic global model, which comprises % simplified wording
 only an intercept (or mean).
-It would also be possible to use, e.g., a linear regression model or a logistic regression model as the global model.
+It would also be possible to use, for example, a linear or logistic regression model as the global model.
 Figure~\ref{fig:toy}, left, shows the distribution of $y$ in a boxplot.
-Obviously, a global intercept or mean would not describe all observations equally well, there is quite some unexplained variation around the sample mean of 5.91.
+Obviously, a global intercept or mean would not describe all observations equally well, because there is quite some unexplained variation around the sample mean of 5.91.
 
@@
 \label{fig:toy}
+\end{figure}
+
+\begin{figure}[h]
+\caption{Top row: Residuals ordered by the values of the partitioning variables. Bottom row: Summed residuals (cumulative sums for continuous predictors, sums by level for categorical predictors).}
+\begin{subfigure}[][][t]{1.2\textwidth}
+\includegraphics{_MOB_paper-006}
+
+\includegraphics{_MOB_paper-007}
+\end{subfigure}
+\label{fig:efp}
 \end{figure}
@@
 We applied MOB to the observations in the left panel of Figure~\ref{fig:toy}, specifying $x_1$, $x_2$ and $x_3$ as potential partitioning variables.
+In Step 1, the model parameters (here: the intercept) are estimated in the root node, which contains all observations.
+%changed
+In Step 2, the model's \emph{scores} are computed.
+In case of an intercept-only model, these are simply the residuals.
+The scores are ordered with respect to the possible partitioning variables, as shown in the top row of Figure~\ref{fig:efp}.
+The scores should randomly fluctuate around their mean of zero, but the top row of Figure~\ref{fig:efp} suggests systematic deviations: there seems to be an association with $x_1$ and $x_2$, and little or no association with $x_3$.
+
+MOB captures systematic deviations in the scores (i.e., structural change) by computing cumulative sums.
+The cumulative sums are depicted in the bottom row of Figure~\ref{fig:efp}.
+To decide in Step 3 whether to split the node, and if so, which variable to split on, a test statistic is computed based on the cumulative sums of each partitioning variable; full details on the computation are provided in \citep{ZeilyHoth08}.
+In the current example, the structural change tests revealed that $x_1$ was most strongly associated with instabilities in the intercept and it was thus selected for splitting in Step 3.
+
+In Step 4, the procedure is repeated in each of the resulting subgroups.
 The resulting tree is shown in the right panel of Figure~\ref{fig:toy}.
-In the root node, which contains all observations (step 1), structural change tests were performed for each of the three covariates (step 2).
-The tests revealed that $x_1$, a continuous covariate, was most strongly associated with instabilities in the intercept and $x_1$ was thus selected first for splitting (step 3).
-The $p$ value resulting from the structural change test for the first split in $x_1$ is depicted in the root node.
-By default, a Bonferroni correction is applied to the $p$ values, to account for the fact that the tests are performed for all of the potential partitioning variables.
-Next, the cutpoint for $x_1$ was selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroup are as similar as possible.
-In the right daughter node, additional significant instability was detected with respect to $x_2$, a categorical covariate.
-Again, the cutpoint in $x_2$ was selected so that the two resulting subgroups exhibit strongest parameter differences.
-In the left daughter node (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup (step 4).
+In the right subgroup (node 3), additional significant instability was detected with respect to $x_2$, a categorical covariate.
+Again, the cutpoint for $x_2$ was selected so that the two resulting subgroups exhibit an as large as possible parameter difference.
+In the left subgroup (node 2), no further splits were created, because none of the three covariates were significantly associated with any further instability in this subgroup.
 The same held for nodes 4 and 5, so that the third covariate $x_3$ was never selected for splitting.
 
-The subgroup-specific distributions of the response variable are presented in the terminal nodes at the bottom of the tree.
-The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the pattern in the data.
-The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were actually selected for splitting.
+The $p$ values resulting from the structural change tests of each split are depicted in the corresponding nodes.
+By default, the $p$ values are Bonferroni corrected, to account for the fact that the tests are performed for all potential partitioning variables.
+After selecting $x_1$ for splitting, the cutpoint is selected so that the two resulting subgroups exhibit the strongest parameter differences, while the observations within the subgroups are as similar as possible.
+
+The subgroup-specific distributions of the response variable are depicted in the terminal nodes at the bottom of the tree.
+The fact that the tree in Figure~\ref{fig:toy} displays more than one terminal node confirms that one global model for all observations does not appropriately capture the patterns in the data.
+The tree also shows shows that, out of the three covariates that were presented to the algorithm, only two were selected for splitting.
 This automatic variable selection is an important characteristic of the CART and MOB algorithms.
+
 The means (intercepts) in the terminal nodes are 4.42, 4.78 and 9.17, respectively.
 While the means and distributions in nodes 2 and 4 appear very similar, the mean in node 5 is obviously higher.
-Substantively, the tree suggests and interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa).
+Substantively, this tree suggests an interaction effect: The effect of $x_2$ depends on the level of $x_1$ (vice versa).
 If $x_1$ were age and $x_2$ gender, this would indicate that gender differences only start to occur at a later age.
@@
 
-To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in Head Start.
+To illustrate the use of GLMM trees, we analyze a dataset from \cite{Demi09}, who evaluated long-term benefits of participation in the Head Start program.
 Head Start is a federally funded nationwide pre-school program for children from low-income families in the United States.
 Participation in Head Start takes place from ages 3 through 5.
-\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the National Longitudinal Survey of Youth \citep{NLSY}.
+\cite{Demi09} compared performance of siblings who differed in their participation in the program using data from the 1979 cohort of the National Longitudinal Survey of Youth \citep{NLSY}.
 
 The sample consists of 273 families with at least two siblings, where at least one sibling participated in Head Start and at least one sibling did not participate in Head Start or any other pre-school program.
-Data from children in those families who participated in another pre-school program were excluded.
+Data from children who participated in another pre-school program were excluded.
 The family structure allows siblings who did not participate in Head Start to serve as a natural control to assess the effects of participating in Head Start.
-The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT]{DunnyDunn81}, we will thus model PPVT trajectories over time.
-On average, there were 2.14 PPVT scores per child, for 29\% of the children there was only a single PPVT score available.
+The outcome variable comprises repeated assessments on the Peabody Picture Vocabulary Test \citep[PPVT; ][]{DunnyDunn81}, for which we model trajectories over time.
+On average, there were 2.14 PPVT scores per child, for 71\% of the children there were $>1$ PPVT scores available.
 
@@
 \end{Schunk}
+
+We specified the fixed main and interaction effects of `Program` and `Age` as `Program*Age`.
+The random effects were specified as `(1|MotherID/ChildID)`.
+The \texttt{1} specifies that a random intercept should be estimated and \texttt{MotherID/ChildID} specifies that every child and every mother should obtain a random intercept, so as to account for correlations between repeated assessments of the same child, and correlations between children of the same mother.
+The forward slash \texttt{/} indicates that the children are 'nested' within mothers; that is, that there are multiple children per mother, but only one mother per child.
 
@@
 \begin{subfigure}{.7\textwidth}
-\includegraphics{_MOB_paper-009}
+\includegraphics{_MOB_paper-011}
 \end{subfigure}
@@
 
-\caption{\label{tab:predictions}Node-specific average predicted PPVT scores at different ages.}
+\caption{\label{tab:predictions}Node-specific predicted PPVT scores at different ages.}
 \begin{tabular}[t]{cccc}
@@
 \bottomrule
-\multicolumn{4}{l}{\textsuperscript{a} HS = Head Start; PPVT = Peabody Picture Vocabulary}\\
-\multicolumn{4}{l}{Test.}\\
+\multicolumn{4}{l}{\textsuperscript{} extit\{Note.\} For computing predictions, random}\\
+\multicolumn{4}{l}{effects were assumed zero. HS = Head Start; PPVT =}\\
+\multicolumn{4}{l}{Peabody Picture Vocabulary Test.}\\
 \end{tabular}
@@
 \caption{GLMM tree with trajectories of PPVT scores}
-\includegraphics{_MOB_paper-013}
+\includegraphics{_MOB_paper-015}
 
@@
 
-\includegraphics{_MOB_paper-014}
+\includegraphics{_MOB_paper-016}
 \\\textit{Note}: Red lines depict average node-specific trajectories for siblings participating in Head Start; black lines depict average node-specific trajectories for siblings not participating in Head Start.
@@
 All nodes show a modest benefit of Head Start participation at age 6, about 3-4 points on the PPVT.
-This benefit increases over time for the group with higher mother's AFTQ scores.
+This benefit increases over time for the group with higher mother's AFTQ scores (nodes 6 and 7).
 The benefit decreases over time for children in the group with lowest mother's AFTQ scores (node 2).
-For Black and Hispanic children with intermediate mother's AFTQ scores, the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12.
-These results roughly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 at ages 11-14.
+For Black and Hispanic children with intermediate mother's AFTQ scores (node 5), the benefit at age 6 even shifts to a disadvantage of Head Start participation at age 12.
+
+These results partly correspond to the findings of \cite{Demi09}, who report a relative benefit of Head Start participation of about 0.15 standard deviations at age 5-6, which fades out over time to about 0.05 standard deviations at ages 11-14.
 Fade out was reported to be particularly strong for African-American and very disadvantaged children.
-The global benefit of participation in Head Start as shown in Figure~\ref{fig:global_lmm} and reported by \cite{Demi09} thus seems to differ across groups.
+In the current sample, the standard deviation of PVVT scores was 34.35, and thus the predictions for age 6 in Table~\ref{tab:predictions} roughly correspond to the effect found by \cite{Demi09}.
+However, we found a different pattern of fade-out.
 
@@
 
-\subsubsection{Pruning a Rasch tree}
-
 Once the Rasch tree has been fitted, we can plot it using the standard \texttt{plot} function:
@@
 \begin{subfigure}{1.2\textwidth}
-\includegraphics{_MOB_paper-021}
+\includegraphics{_MOB_paper-023}
 \end{subfigure}
@@
 \caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion.}
-\includegraphics{_MOB_paper-026}
+\includegraphics{_MOB_paper-028}
 \label{fig:MHtree2}
@@
 \caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion with additional visualization features.}
-\includegraphics{_MOB_paper-029}
+\includegraphics{_MOB_paper-031}
 \label{fig:MHtree3}
@@
 \caption{Rasch tree fitted to the SPISA quiz items based on using Mantel-Haenszel DIF effect size measure as the stopping criterion with additional visualization features.}
-\includegraphics{_MOB_paper-031}
+\includegraphics{_MOB_paper-033}
 \label{fig:MHtree4}

### 2024-05-13 _MOB_paper.R

@@ -16,13 +16,37 @@ toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B")
 
 library("colorspace")
 boxplot(toy_data$y, cex = .7)
-#points(x = 1, y = mean(toy_data$y), pch = 19, ylim = c(1,14))
 
 
 library("partykit")
 tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
 plot(tree, gp = gpar(cex = .7), ylim = c(1,14))
 term_node_means <- round(tapply(toy_data$y, predict(tree, type = "node"), mean), digits= 2L)
+
+
+library("strucchange")
+gefp_x1 <- gefp(y ~ 1, fit = lm, data = toy_data, order.by = ~ x1)
+gefp_x3 <- gefp(y ~ 1, fit = lm, data = toy_data, order.by = ~ x3)
+lmod <- lm(y ~ 1, data = toy_data)
+resids <- residuals(lmod)
+par(mfrow = c(1, 3))
+plot(toy_data$x1, resids, xlab = "x1", ylab = "residuals", 
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+plot(toy_data$x2, resids, xlab = "x2", ylab = "residuals", 
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+plot(toy_data$x3, resids, xlab = "x3", ylab = "residuals", 
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+
+
+par(mfrow = c(1, 3))
+plot(toy_data$x1[order(toy_data$x1)], cumsum(resids[order(toy_data$x1)]), 
+     xlab = "x1", ylab = "cumulative sum", type = "l",
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+boxplot(t(replicate(100, tapply(resids, toy_data$x2, sum))), 
+        ylab = "sum", cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
+plot(toy_data$x3[order(toy_data$x3)], cumsum(resids[order(toy_data$x3)]), 
+     xlab = "x3", ylab = "cumulative sum", type = "l",
+     cex = 1.2, cex.lab = 1.2, cex.axis = 1.2)
 
 
 ## Load data
@@ -116,12 +140,12 @@ names(design_df)[1] <- "Node"
 library("kableExtra")
 add_footnote(
[…]

### 2024-05-13 _MOB_paper_custom.R

@@ -1,214 +1,0 @@
-###################################################################
-###
-###
-### R code for reproducing analyses from:
-###
-###   One Model May Not Fit All: Subgroup Detection Using
-###   Model-Based Recursive Partitioning
-###
-
-
-
-
-#########################
-##
-## Introduction
-##
-##
-
-set.seed(3)
-x1 <- runif(250, min = 5, max = 11)
-x2 <- runif(250, min = 2, max = 8)
-x3 <- runif(250, min = 4, max = 10)
-error <- rnorm(250, sd = 1.5)
-y <- 7 + ifelse(x1 > 8 & x2 > 5, yes = 2.5, no = -2.5) + error
-toy_data <- round(data.frame(x1, x2, x3, y), digits = 2L)
-toy_data$x2 <- factor(ifelse(x2 > 5, yes = ifelse(x2 > 6.5, "A", "C"), no = "B"))
-boxplot(toy_data$y, cex = .7)
-library("partykit")
-tree <- lmtree(y ~ 1 | x1 + x2 + x3, data = toy_data)
-plot(tree, gp = gpar(cex = .7), ylim = c(1, 14))
-
-
-
-######################################################################
-##
-## Application Example 1: Subgroup Detection in Mixed-Effects Models 
-##
-##
-
-## Load data
-data <- data.frame(haven::read_dta("data_Deming_2008_0217_MF.dta"))
-
-## Keep only eligible children in HS or no preschool
-data <- data[data$Elig2_90 == 1, ]
-data <- data[data$None2_90 == 1 | data$HS2_90 == 1, ]
-data <- data[data$MotherID %in% names(table(data$MotherID) >= 2), ]
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")))
-
-## Keep only families in which children differ in HS participation
-selected_family <- rowSums(table(data$MotherID, data$program) > 0) == 2
-selected_family <- names(selected_family)[selected_family]
-data <- data[data$MotherID %in% selected_family, ]
-
-## Node-specific predictor
-data$program <- factor(ifelse(data$None2_90 == 1, "None", 
-                              ifelse(data$Pre2_90 == 1, "Pre", "HS")),
-                       levels = c("None", "HS"))
-
-## Partitioning variables
-
-## ageAFTQ: age-adjusted AFQT
[…]
